### 2024-07-03 README.qmd
Initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# kwlenv

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh///master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): Quarto source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.qmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2024-07-14 analysis/paper/paper.qmd
upload qmd

---
title: "An Environmental Perspective on European Colonization in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute:
        - AS
institute:
  - AS:
      name: Academia Sinica
      address: 
  - IHP: 
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Introduction

European expansion since the 15th century marked the beginning of an era characterized by exploration, colonization, and global trade. Due to purposes and situations, the mode, pace, and degree of colonial control varied greatly across regions [@Jordan2009]. Understanding the processes of different colonial encounters has become a significant issue in archaeology to comprehend the dynamic outcomes of European colonization worldwide. In Asia, European colonization was primarily driven by the pursuit of Asian luxury goods and local natural resources for commercial profit. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European colonization, including resistance, negotiation, incorporation, persistence, and resilience [@Torrence2003; @Silliman2005; @Mitchell2010].

While many researchers focus on European colonies, places on the periphery of colonial centers were also significantly influenced by European powers and their associated trading systems [@Acabado2017; @Trabert2017]. These peripheral regions often experienced distinct forms of interaction and impact due to their unique geographical, cultural, and economic contexts. Exploring these areas is essential for gaining a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan is a good example for such colonial setting. Although it was on the periphery of European fortresses in Northern Taiwan in the 17th century, it was intensively involved in the regional exchange network stimulated by Europeans supported by the finding of large amounts of foreign trade goods at the site [@citation]. Additionally, Northeastern Taiwan was involved in the deerskin exchange system when the VOC (Dutch East India Company) formalized deerskin exports in the mid-17th century [@Cheng2017]. The impacts of European economic demands on the local economy and environment in Northeastern Taiwan still require further investigation. Understanding these effects can provide valuable insights into the ways in which Indigenous communities adapted to and were influenced by European colonial activities.

In this paper, I will explore the Indigenous experience of European colonization by examining archaeological evidence from the Kiwulan site (1350-1950 CE), a large indigenous settlement in northeastern Taiwan [@citation]. Kiwulan experienced periods of European presence in Northern Taiwan in the 17th century, followed by Han Chinese migration in the 19th century. Archaeological evidence related to subsistence practices is particularly valuable for this investigation, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European disruptions and the associated changes in their environment. Key research questions are: What were the resource use patterns of Indigenous people during the time of European colonization? Can we identify shifts in these patterns in response to colonial pressures? The later period of Chinese migration will be also included to have a long-term view. To address these questions, I will examine the faunal and plant remains from Kiwulan, with a preliminary analysis on ceramic residues. By adopting an environmental and ecological perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

## the mode of colonization

## trade and exchange of Han Chinese

# colonial studies in Archaeology
In this research, I will explore whether and how the introduction of foreign materials and practices, new forms of production, and migrated people from Europe and Chinese contributed to the changes in Indigenous social, environmental, and ecological systems since the protohistoric period (Liu and Wang, 2017). I will focus on archeological evidence from different parts in northeastern Taiwan to compare Indigenous reactions in a pericolonial context (Acabodo, 2017). Yilan is an ideal location as a starting point because of the frequent foreign contact with Europeans and Chinese traders in the protohistoric and historic period. Also, the ecological variation across the landscape from the coastal area to mountains provides multiscalar variability for exploring the effect of colonial contact on a regional scale. As a place located in the surrounding area of the European colonial base in northern Taiwan, Yilan enables a pericolonial perspective to understand the effect of European colonization. I will use historical ecological approaches to examine the interactions of people and environment in a colonial context, and their consequences on the development of the past and contemporary culture in a long-term view (Braje, 2017). This project will help us to understand social and ecological reactions of Indigenous communities to European colonization and cross-cultural contact that has not yet been fully understood.

I will use historical ecology as the conceptual framework to understand the human-environment interactions in northeastern Taiwan, which holds that human impacts on the ecological system are historically dependent (Balée, 1998; Braje, 2017; Fitzhugh et al., 2019). In a colonial context, historical ecological investigations can help us to understand the impacts of European presence and associated economic activities on the local environments and domestic economies (Oas and Hauser, 2018; Wallman et al., 2018). Located in northeastern Taiwan, Yilan is close to the Dutch colonial base in northern Taiwan with favorable soil conditions for rice growing and natural resources such as forests and deer. The frequent exchange activities with Europeans and Chinese merchants, and local interests in foreign goods could lead to further exploitation of deer and intensified rice growing as tributes for maintaining good relations with Europeans and exchanging more imported goods. 
My hypothesis is that if European colonial activities resulted in environmental and ecological changes in northeastern Taiwan due to deer hunting and rice growing, then we could observe changes in Indigenous subsistence and land use patterns. I will test my hypothesis by studying archaeological sediments, animal and plant remains, organic residues in locally made ceramics, and site locational preferences of Late Metal Age sites in Yilan to identify any changes after the colonial presence.

# Background
The Yilan Plain is the major coastal plain in northeastern Taiwan. Prehistory in the Yilan Plain begins from the Middle Neolithic period around 4000 B.P. and continues throughout the Metal Age period (2000-400 B.P.) (Liu 2011). The final Late Iron Age in Taiwan can be viewed as a protohistoric period due to the interaction of record-keeping Chinese and Europeans with local cultures. Most Late Metal Age sites in the Yilan Plain can be related to the local Indigenous people, the Kavalan, according to oral history and similarities between historical Kavalan material culture and archaeological artifacts (Chen 2004). 
European colonization of Taiwan began in the 1620s because of the competition between the Spanish and the Dutch for political and economic control of the Pacific (Andrade, 2007). Yilan is about 70 kilometers away from the first European northern base with mountains in between. Indigenous people in Yilan were not in direct contact with colonial powers until 1632, when they were attacked by the Spanish. Later in 1647, the Dutch attacked the Indigenous people and forced them to accept colonial rules and economic demands by paying annual tribute (Andrade, 2007). The Dutch also built a trading outpost in Trobiawan. According to the Dutch census, Indigenous people in northeastern Taiwan occupied over 40 settlements with approximately 10,000 people in the 17th century (Nakamura 1938), and those settlements were also described by the Chinese in later periods, demonstrating a continuous Indigenous habitation until the 19th century (Figure 1).
Due to the fear of European attacks, most local communities in Yilan maintained good relationships with the Europeans by paying tributes and Indigenous leaders could obtain foreign materials as gifts according to the degree of allegiance (Kang 2012). In addition to historical records, archaeological evidence also shows complex inter- and intra- exchange networks based on the findings of foreign goods, such as stonewares commonly found in European outposts in Southeast Asia, Chinese porcelains, and glass and stone beads (Chen 2005; Chen 2007; Liu 2011; National Museum of Taiwan History 2005, Wang and Liu 2007). This suggests frequent trade activities since the protohistoric period in the 14th century and later intensified through the global trade network stimulated by the Europeans and associated colonial activities. Moreover, foreign materials such as ceramics, beads, and textile were viewed as prestige goods in local communities before European arrival (Li and Wu, 2006).

Northeastern Taiwan is an especially interesting example of European and Indigenous interaction because the impact of European colonialism appears to have been much less pronounced compared to other places in this region. I define northeastern Taiwan as an area at the periphery of colonial center where the European power was indirect with a few sporadic, mostly non-violent direct contacts. This type of context has also been called pericolonialism (Acabado, 2017; Kang, 2016, pp. 31-78; Trabert, 2017). My PhD study at Kiwulan (14-19th A.D.), the biggest Indigenous settlement during the 17th century in Yilan, indicates that European contact and foreign goods resulted in an increased social differentiation in the local society through analyses of trade items, pottery production, and burial goods (Wang and Marwick, 2020a, 2020b, 2021). The results demonstrate that there was an indirect effect of European colonization in this region that could lead to changes in the organizational mode of a society that have not been identified before.
In addition to social changes, how did the interactions with Europeans and Chinese lead to changes in the local ecological system need further examination. For example, during the Dutch colonial period, deer hide was especially demanded in the Dutch colonization period as payment for tributes due to its high value for export to Japan, which led to intensified deer hunting in historical Taiwan (Andrade, 2007; Koo, 2011). While the commercial activities and a substantial decrease in Taiwan’s deer population were well documented, much less is known about how this historical consequence can be situated in a local ecological framework. Did the participation in these trade activities result in a change in Indigenous subsistence behaviors or the human-environment interactions? What was the response of local Indigenous people to the decrease of deer population? Those questions are important because they can help us to trace back the Indigenous cultures and their resilience under colonial interference since the protohistoric Taiwan.

# Methods
## Geoarchaeological and geochemical analysis for sediments
To explore changes in environments, I will examine sediments from archaeological sites in northeastern Taiwan, Yilan. I will mainly focus on archaeological sites recorded as Indigenous settlements that can be traced back to the protohistoric and historic period when they experienced the European and Chinese presence, such as Kiwulan I, Kiwulan II, and Yilan Agricultural Vocational High School site. Kiwulan and Yilan Agricultural Vocational High School are two major Late Metal Age sites with extensive excavation in recent years. Kiwulan (600-100 B.P.) is located at the northern part of Yilan plain (Chen, 2007), while Yilan Agricultural Vocational High School site (400-200 B.P.) is situated on the central part of Yilan (Li, 2014). Archaeological evidence from both sites includes a large amount of pottery sherds, imported stoneware and porcelains, wooden artifacts, stone and metal tools, pipes, thousands of glass beads, agate beads and metal ornaments, post holes with in-situ wooden posts, and burials. Based on their locations and the historical records, those two sites can be traced back to the 17th century when they were named Banouran and Beili according to the Dutch census (Nakamura 1938). Those archaeological materials are mainly stored in the Cultural Affairs Bureau in Yilan County, and it is feasible to analyze those materials and collect data based on previous collaboration during my PhD. 
For archaeological sediments, I will use geoarchaeological methods to measure the physical and chemical properties of sediments, including particle size, magnetic susceptibility, pH values, electrical conductivity, and organic matter and carbonates proportions to understand soil conditions and Indigenous landuse. I will also use organic chemical analysis to detect lipids in sediments to reconstruct past environments, fire and land use patterns associated with the colonial presence and long-term cultural interaction. Geoarchaeological and geochemical methods are useful and insightful methods especially when faunal and botanical evidence were poorly preserved, such as in a condition of acidic environments.

## Faunal and botanical analysis, and organic residue analysis
To explore any changes in the subsistence strategies, I will examine faunal and botanical evidence directly if they were preserved in good conditions. The historical records point out frequent exchange activities between Europeans, Chinese traders, and different Indigenous people, such as Kavalan and Basay during this period (Chen, 2008). According to historical records, Indigenous people in Yilan exchanged natural resources, such as deer hide and rice, for porcelains, metal items, and beads from Europeans, Chinese traders, and other Indigenous people, such as Basay (Borao, 2001). Deer hide was especially demanded in the Dutch colonization period as payment for tributes due to its high value as a commodity in East Asia. 
To understand change in subsistence behaviors or the human-environment interactions during and after the colonial period, I will examine two types of evidence from those sites. First is zooarchaeological and botanical data related to economic activities in a colonial context, especially animal bones such as deer and plant remains such as rice grains. Second, I will build on my previous pilot studies of isotope geochemical analysis for samples from Kiwulan, and greatly expand the sample sizes to include human remains (e.g., teeth), animal bones, and ceramic fabrics. This will enable me to measure dietary or subsistence changes related to environments, and stability before and after the European presence, and Chinese presence. 
To accomplish those analyses, I will draw on my existing collaborations with the Paleoclimatology and Organic Geochemistry lab in the University of Washington and the Microbial Metabolic Diversity lab in the Academia Sinica for organic residue analysis and DNA analysis, especially for pottery sherds. By analyzing the composition of the organic residues absorbed and trapped in potsherds, we can know what categories of food are cooked in pots, such as animal fats, plants, and aquatic resources that provides direct evidence to study ancient diet (Evershed 2008; Craig 2011; Cramp 2011). Similarly, microbes from food residues in pots offer a way to detect ancient diet because different types of microbes may be associated with different types of food (Bokulich et al. 2016). I will combine those advanced techniques to understand subsistence strategies in northeastern Taiwan.

# Results

<!-- Here's some example analysis code: -->

```{r}
#| label: get-data
#| eval: false
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r}
#| label: fig-demo-plot
#| fig-cap: "A plot of random numbers"
plot(rnorm(10))
```

@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion
Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period. This project will explore the environmental effect of colonial presence on local ecology on a regional scale using archaeological evidence that has not yet been fully understood. Archaeological materials can offer more direct evidence, allowing better interpretation for the environments where Indigenous live in that can also reduce biases from historical records. Through exploring the long-term dynamics of human-environmental relationship in colonial settings, we can better understand changes in local environments and Indigenous resilience. In addition to the Europeans, the interaction between Chinese traders and Indigenous groups in Taiwan since the Neolithic period, and an influx of Han Chinese immigrants in the 18th century, has contributed to the significant local cultural and environmental transformation. Those historical events are closely related to the Taiwan Indigenous people today, which is crucial to understanding the contemporary Indigenous cultures and socio-ecological landscape. 
The results of this project can advance our knowledge of the reaction of Indigenous people in Taiwan when facing dominant foreign power, and generate new anthropological insights about the adaptation and choices of Indigenous people. In the long term, this research will also benefit local Indigenous people, Kavalan, in northeastern Taiwan and help them to trace back their past and revitalize their cultural heritage by the plans of public presentation in the future. In addition, this project will combine geoarchaeological, geochemical, and computational methods to analyze archaeological evidence. The multidisciplinary approach can not only advance the development and application of scientific techniques in Taiwan archaeology, but also provide an opportunity to form collaborative relationships with scholars with different backgrounds, which will help to develop the academic network and create interdisciplinary collaboration.

This research will contribute to a broader understanding of the diverse ways Indigenous communities adapted to and resisted colonial forces, enhancing our comprehension of the long-term impacts of European colonization on local societies.

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2024-07-15 analysis/paper/paper.qmd
work on the intro section

@@
 ---
-title: "An Environmental Perspective on European Colonization in Northeastern Taiwan"
+title: "Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan"
 author:
@@
 # Introduction
-
-European expansion since the 15th century marked the beginning of an era characterized by exploration, colonization, and global trade.
-Due to purposes and situations, the mode, pace, and degree of colonial control varied greatly across regions [@Jordan2009].
-Understanding the processes of different colonial encounters has become a significant issue in archaeology to comprehend the dynamic outcomes of European colonization worldwide.
+<!-- introduction about European expansion -->
+European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world.
+Due to specific spatial and temporal contexts, the forms of European colonization and power relations varied across regions [@Jorden2009].
+Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004].
 In Asia, European colonization was primarily driven by the pursuit of Asian luxury goods and local natural resources for commercial profit.
 The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002].
-However, Indigenous communities did not passively accept European colonization.
-Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European colonization, including resistance, negotiation, incorporation, persistence, and resilience [@Torrence2003; @Silliman2005; @Mitchell2010].
+However, Indigenous communities did not just passively accept European colonization.
+Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European colonization, such as resistance, negotiation, incorporation, persistence, or resilience [@Torrence2003; @Silliman2005; @Mitchell2010].
 
-While many researchers focus on European colonies, places on the periphery of colonial centers were also significantly influenced by European powers and their associated trading systems [@Acabado2017; @Trabert2017].
-These peripheral regions often experienced distinct forms of interaction and impact due to their unique geographical, cultural, and economic contexts.
-Exploring these areas is essential for gaining a broader understanding of the varied consequences of colonial expansion.
-Northeastern Taiwan is a good example for such colonial setting.
-Although it was on the periphery of European fortresses in Northern Taiwan in the 17th century, it was intensively involved in the regional exchange network stimulated by Europeans supported by the finding of large amounts of foreign trade goods at the site [@citation].
-Additionally, Northeastern Taiwan was involved in the deerskin exchange system when the VOC (Dutch East India Company) formalized deerskin exports in the mid-17th century [@Cheng2017].
-The impacts of European economic demands on the local economy and environment in Northeastern Taiwan still require further investigation.
-Understanding these effects can provide valuable insights into the ways in which Indigenous communities adapted to and were influenced by European colonial activities.
+<!-- NE Taiwan as a case study -->
+While many studies focus on the center of European colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017].
+Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion.
+Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan.
+Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control.
+However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north.
+In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export, involving many Indigenous communities and Han Chinese.
+Despite some discussions on the policies and operations of the deerskin trade, direct archaeological investigations into the reactions of local Indigenous people to these commercial activities are rare.
+Prior to the deerskin trade, deer were already a substantial part of local subsistence.
+Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.
 
-In this paper, I will explore the Indigenous experience of European colonization by examining archaeological evidence from the Kiwulan site (1350-1950 CE), a large indigenous settlement in northeastern Taiwan [@citation].
-Kiwulan experienced periods of European presence in Northern Taiwan in the 17th century, followed by Han Chinese migration in the 19th century.
-Archaeological evidence related to subsistence practices is particularly valuable for this investigation, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European disruptions and the associated changes in their environment.
-Key research questions are: What were the resource use patterns of Indigenous people during the time of European colonization?
-Can we identify shifts in these patterns in response to colonial pressures?
-The later period of Chinese migration will be also included to have a long-term view.
-To address these questions, I will examine the faunal and plant remains from Kiwulan, with a preliminary analysis on ceramic residues.
+<!-- research aim and questions -->
+In this paper, I will explore the interactions between people and their environment in a colonial context by examining archaeological evidence from the Kiwulan site (1350-1950 CE), a large Indigenous settlement in northeastern Taiwan [@citation].
+Kiwulan experienced periods of European presence in northern Taiwan in the 17th century, followed by a large influx of Han Chinese migration in the 19th century.
+The later period of Chinese migration will also be included to provide a long-term perspective.
+Archaeological evidence related to subsistence practices is particularly focused in this research, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment.
+My research questions are: What were the resource use patterns of Indigenous people before and after the European presence?
+Can we identify shifts in these patterns in response to colonial economic demands?
+To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues.
 By adopting an environmental and ecological perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.
 
-## the mode of colonization
+# Historical background
+As an island close to the Asian continent, Northeast Asia, and Island Southeast Asia, Taiwan has been a place with frequent cross-cultural interactions since the Neolithic period and involved in a complex global trading network in the 17th century due to the European expansion.
+Based on the commercial purposes, the Europeans, including the Spanish and the Dutch, built fortresses and trading posts at different parts of Taiwan.
 
@@
 # colonial studies in Archaeology
+
 In this research, I will explore whether and how the introduction of foreign materials and practices, new forms of production, and migrated people from Europe and Chinese contributed to the changes in Indigenous social, environmental, and ecological systems since the protohistoric period (Liu and Wang, 2017).

### 2024-07-17 analysis/paper/paper.qmd
work on the section of background

@@
 European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world.
-Due to specific spatial and temporal contexts, the forms of European colonization and power relations varied across regions [@Jorden2009].
+According to specific spatial and temporal contexts, the forms of European colonization and power relations varied across regions [@Jorden2009].
 Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004].
-In Asia, European colonization was primarily driven by the pursuit of Asian luxury goods and local natural resources for commercial profit.
+In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources for commercial profit.
 The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002].
 However, Indigenous communities did not just passively accept European colonization.
-Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European colonization, such as resistance, negotiation, incorporation, persistence, or resilience [@Torrence2003; @Silliman2005; @Mitchell2010].
+Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, or entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009].
 
 <!-- NE Taiwan as a case study -->
-While many studies focus on the center of European colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017].
+While many studies focus on prominent colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017].
 Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion.
@@
 In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export, involving many Indigenous communities and Han Chinese.
-Despite some discussions on the policies and operations of the deerskin trade, direct archaeological investigations into the reactions of local Indigenous people to these commercial activities are rare.
-Prior to the deerskin trade, deer were already a substantial part of local subsistence.
+Despite some discussions on the policies and operations of the deerskin trade, direct archaeological investigations into the experiences of local Indigenous people to these commercial activities are rare.
+The European impacts on local ecological conditions have been recorded in many cases in colonial settings [@Jorden2009].
+Prior to the deerskin trade, deer has been a substantial part of local subsistence.
 Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.
@@
 <!-- research aim and questions -->
-In this paper, I will explore the interactions between people and their environment in a colonial context by examining archaeological evidence from the Kiwulan site (1350-1950 CE), a large Indigenous settlement in northeastern Taiwan [@citation].
-Kiwulan experienced periods of European presence in northern Taiwan in the 17th century, followed by a large influx of Han Chinese migration in the 19th century.
-The later period of Chinese migration will also be included to provide a long-term perspective.
-Archaeological evidence related to subsistence practices is particularly focused in this research, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment.
-My research questions are: What were the resource use patterns of Indigenous people before and after the European presence?
-Can we identify shifts in these patterns in response to colonial economic demands?
-To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues.
-By adopting an environmental and ecological perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.
+In this paper, I will explore the interactions between people and their environment in a colonial context by examining resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), a largest settlement in northeastern Taiwan [@citation].
+Kiwulan experienced periods of European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century.
+I will primarily focus on subsistence practices, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment.
+My research questions are: What were the subsistence practices of Indigenous people before and after the European presence?
+Can we identify shifts in these strategies in response to colonial economic demands?
+To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues for investigating culinary practices.
+Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.
 
-# Historical background
-As an island close to the Asian continent, Northeast Asia, and Island Southeast Asia, Taiwan has been a place with frequent cross-cultural interactions since the Neolithic period and involved in a complex global trading network in the 17th century due to the European expansion.
-Based on the commercial purposes, the Europeans, including the Spanish and the Dutch, built fortresses and trading posts at different parts of Taiwan.
+# Colonial encounter in Northeastern Taiwan
+<!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
+In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to avoid the interference in the Asian trade by the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company), which established a colony in southern Taiwan at the same time.
+In 1642, the Dutch defeated the Spanish and took over their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in 1662 [@Andrade2007].
+The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and the collection resources they needed among different Indigenous groups.
+For example, the Basay people, active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@citation].
+The European domination was mostly established through the conquest over indigenous villages, and the VOC further developed tribute system to request deerskins or products from indigenous people [].
+However, for most indigenous villages distant from the European colonial base, European colonial control was relatively less, indirect, and even unsuccessful.
 
-## trade and exchange of Han Chinese
+<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->
+Kavalan, or Yilan in the present day, in northeastern Taiwan is an example of slight European presence for most of the time.
+Kavalan is about 70 kilometers away from the European northern base with mountains in between.
+Local indigenous groups were Kavalan people, who first experienced European powers in 1632, when they were attacked by the Spanish.
+Later in 1647, a group of Dutch soldiers attacked many Kavalan villages and asked those Kavalan villages to pay tributes [@citation].
+The Dutch also built a trading outpost in Trobiawan located near the coastal area at Yilan, but there might be not much officials at any given time.
+According to the Dutch census, there were more than 40 Kavalan villages with approximately 10,000 people in the 17th century in this region [@Nakamura1938], which were also mentioned by the Chinese and Japanese in later periods, demonstrating a continuous Indigenous habitation until the 19th century.
+It appears that the Kavalan area was in the situation where trading activities were frequent and active stimulated by the European presence in northern Taiwan, with occasional pressures from colonial powers.
+In general, the European control over the Kavalan must have been largely indirect with limited colonialism [@].
 
-# colonial studies in Archaeology
+Before the arrival of the Europeans in northern Taiwan,
 
-In this research, I will explore whether and how the introduction of foreign materials and practices, new forms of production, and migrated people from Europe and Chinese contributed to the changes in Indigenous social, environmental, and ecological systems since the protohistoric period (Liu and Wang, 2017).
-I will focus on archeological evidence from different parts in northeastern Taiwan to compare Indigenous reactions in a pericolonial context (Acabodo, 2017).
-Yilan is an ideal location as a starting point because of the frequent foreign contact with Europeans and Chinese traders in the protohistoric and historic period.
-Also, the ecological variation across the landscape from the coastal area to mountains provides multiscalar variability for exploring the effect of colonial contact on a regional scale.
-As a place located in the surrounding area of the European colonial base in northern Taiwan, Yilan enables a pericolonial perspective to understand the effect of European colonization.
-I will use historical ecological approaches to examine the interactions of people and environment in a colonial context, and their consequences on the development of the past and contemporary culture in a long-term view (Braje, 2017).
-This project will help us to understand social and ecological reactions of Indigenous communities to European colonization and cross-cultural contact that has not yet been fully understood.
-
-I will use historical ecology as the conceptual framework to understand the human-environment interactions in northeastern Taiwan, which holds that human impacts on the ecological system are historically dependent (Balée, 1998; Braje, 2017; Fitzhugh et al., 2019).
-In a colonial context, historical ecological investigations can help us to understand the impacts of European presence and associated economic activities on the local environments and domestic economies (Oas and Hauser, 2018; Wallman et al., 2018).
-Located in northeastern Taiwan, Yilan is close to the Dutch colonial base in northern Taiwan with favorable soil conditions for rice growing and natural resources such as forests and deer.
-The frequent exchange activities with Europeans and Chinese merchants, and local interests in foreign goods could lead to further exploitation of deer and intensified rice growing as tributes for maintaining good relations with Europeans and exchanging more imported goods.
-My hypothesis is that if European colonial activities resulted in environmental and ecological changes in northeastern Taiwan due to deer hunting and rice growing, then we could observe changes in Indigenous subsistence and land use patterns.
-I will test my hypothesis by studying archaeological sediments, animal and plant remains, organic residues in locally made ceramics, and site locational preferences of Late Metal Age sites in Yilan to identify any changes after the colonial presence.
-
-# Background
 The Yilan Plain is the major coastal plain in northeastern Taiwan.
@@
 Most Late Metal Age sites in the Yilan Plain can be related to the local Indigenous people, the Kavalan, according to oral history and similarities between historical Kavalan material culture and archaeological artifacts (Chen 2004).
-European colonization of Taiwan began in the 1620s because of the competition between the Spanish and the Dutch for political and economic control of the Pacific (Andrade, 2007).
+
  Yilan is about 70 kilometers away from the first European northern base with mountains in between.
@@
 
+## trade and exchange of Han Chinese
+
+In this research, I will explore whether and how the introduction of foreign materials and practices, new forms of production, and migrated people from Europe and Chinese contributed to the changes in Indigenous social, environmental, and ecological systems since the protohistoric period (Liu and Wang, 2017).
+I will focus on archeological evidence from different parts in northeastern Taiwan to compare Indigenous reactions in a pericolonial context (Acabodo, 2017).
+Yilan is an ideal location as a starting point because of the frequent foreign contact with Europeans and Chinese traders in the protohistoric and historic period.
+Also, the ecological variation across the landscape from the coastal area to mountains provides multiscalar variability for exploring the effect of colonial contact on a regional scale.
+As a place located in the surrounding area of the European colonial base in northern Taiwan, Yilan enables a pericolonial perspective to understand the effect of European colonization.
+I will use historical ecological approaches to examine the interactions of people and environment in a colonial context, and their consequences on the development of the past and contemporary culture in a long-term view (Braje, 2017).
+This project will help us to understand social and ecological reactions of Indigenous communities to European colonization and cross-cultural contact that has not yet been fully understood.
+
+I will use historical ecology as the conceptual framework to understand the human-environment interactions in northeastern Taiwan, which holds that human impacts on the ecological system are historically dependent (Balée, 1998; Braje, 2017; Fitzhugh et al., 2019).
+In a colonial context, historical ecological investigations can help us to understand the impacts of European presence and associated economic activities on the local environments and domestic economies (Oas and Hauser, 2018; Wallman et al., 2018).
+Located in northeastern Taiwan, Yilan is close to the Dutch colonial base in northern Taiwan with favorable soil conditions for rice growing and natural resources such as forests and deer.
+The frequent exchange activities with Europeans and Chinese merchants, and local interests in foreign goods could lead to further exploitation of deer and intensified rice growing as tributes for maintaining good relations with Europeans and exchanging more imported goods.
+My hypothesis is that if European colonial activities resulted in environmental and ecological changes in northeastern Taiwan due to deer hunting and rice growing, then we could observe changes in Indigenous subsistence and land use patterns.
+I will test my hypothesis by studying archaeological sediments, animal and plant remains, organic residues in locally made ceramics, and site locational preferences of Late Metal Age sites in Yilan to identify any changes after the colonial presence.
+
 # Methods
+The later period of Chinese migration will also be included to provide a long-term perspective
+
 ## Geoarchaeological and geochemical analysis for sediments

### 2024-07-19 analysis/paper/paper.qmd
work on the background section

@@
 European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world.
-According to specific spatial and temporal contexts, the forms of European colonization and power relations varied across regions [@Jorden2009].
+The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jorden2009].
 Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004].
@@
 However, Indigenous communities did not just passively accept European colonization.
-Recent discussions on colonial interactions also highlight the diverse strategies employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, or entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009].
+Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009; @Alexander1998].
 
@@
 However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north.
-In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export, involving many Indigenous communities and Han Chinese.
-Despite some discussions on the policies and operations of the deerskin trade, direct archaeological investigations into the experiences of local Indigenous people to these commercial activities are rare.
-The European impacts on local ecological conditions have been recorded in many cases in colonial settings [@Jorden2009].
-Prior to the deerskin trade, deer has been a substantial part of local subsistence.
+In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export in a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001].
+Despite some discussions on the policies and operations of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people to the commercial activities are rare.
+The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jorden2009].
 Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.
@@
 <!-- research aim and questions -->
-In this paper, I will explore the interactions between people and their environment in a colonial context by examining resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), a largest settlement in northeastern Taiwan [@citation].
+In this paper, I will explore the interactions between people and their environment in a colonial context by examining resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), a largest settlement in northeastern Taiwan [@Chen2007].
 Kiwulan experienced periods of European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century.
@@
 <!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
-In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to avoid the interference in the Asian trade by the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company), which established a colony in southern Taiwan at the same time.
-In 1642, the Dutch defeated the Spanish and took over their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in 1662 [@Andrade2007].
-The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and the collection resources they needed among different Indigenous groups.
-For example, the Basay people, active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@citation].
-The European domination was mostly established through the conquest over indigenous villages, and the VOC further developed tribute system to request deerskins or products from indigenous people [].
-However, for most indigenous villages distant from the European colonial base, European colonial control was relatively less, indirect, and even unsuccessful.
+In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in Asian market, which established a colony in southern Taiwan two years before the Spanish in the north.
+In 1642, the Dutch defeated the Spanish and occupied their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in Taiwan in 1662 [@Andrade2007].
+The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and resources collections among different Indigenous groups.
+For example, the Basay people, an indigenous group active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@Kang2016].
+The colonial domination was mostly established through the conquest over indigenous villages, but the colonial control varied depending on the location of indigenous settlements.
+For most indigenous villages distant from the European colonial base, the colonial control was relatively less, indirect, and even unsuccessful.
+<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->
 
-<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->
-Kavalan, or Yilan in the present day, in northeastern Taiwan is an example of slight European presence for most of the time.
-Kavalan is about 70 kilometers away from the European northern base with mountains in between.
-Local indigenous groups were Kavalan people, who first experienced European powers in 1632, when they were attacked by the Spanish.
-Later in 1647, a group of Dutch soldiers attacked many Kavalan villages and asked those Kavalan villages to pay tributes [@citation].
-The Dutch also built a trading outpost in Trobiawan located near the coastal area at Yilan, but there might be not much officials at any given time.
-According to the Dutch census, there were more than 40 Kavalan villages with approximately 10,000 people in the 17th century in this region [@Nakamura1938], which were also mentioned by the Chinese and Japanese in later periods, demonstrating a continuous Indigenous habitation until the 19th century.
+Kavalan, or the present day Yilan, in northeastern Taiwan is an example of slight European presence for most of the time.
+Kavalan is about 70 kilometers away from the European northern fortress with mountains in between.
+Local indigenous groups were Kavalan people.
+According to the Dutch census, there were more than 40 Kavalan villages with approximately 10,000 people in the 17th century [@Nakamura1938], which were also mentioned by the Chinese and Japanese in later periods, suggesting a continuous indigenous habitation until the 19th century.
+Kavalan people first experienced European powers in 1632, when they were attacked by Spanish traders [@citation].
+Later in 1647, a group of Dutch soldiers attacked several Kavalan villages and asked those defeated villages to pay tributes [@Kang2016].
+According to Dutch written record, the VOC had built a trading outpost in Trobiawan, located near the coastal area in Yilan, but there might be not much officials at any given time.
 It appears that the Kavalan area was in the situation where trading activities were frequent and active stimulated by the European presence in northern Taiwan, with occasional pressures from colonial powers.
-In general, the European control over the Kavalan must have been largely indirect with limited colonialism [@].
+The European control over the Kavalan must have been largely indirect and limited, which allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes requested by the VOC, such as deerskins or products [@Kang2016].
 
-Before the arrival of the Europeans in northern Taiwan,
+Before the arrival of the Europeans in northern Taiwan, there had been a regional exchange network involving Chinese merchants, Japanese traders, and indigenous groups in northern and northeastern Taiwan since the 14th century [@citation].
+The local indigenous groups traded
 
@@
 The results demonstrate that there was an indirect effect of European colonization in this region that could lead to changes in the organizational mode of a society that have not been identified before.
+
 In addition to social changes, how did the interactions with Europeans and Chinese lead to changes in the local ecological system need further examination.

### 2024-07-21 analysis/paper/paper.qmd
revised paragraphs

@@
 Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004].
-In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources for commercial profit.
+In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources based on commercial interest [@Prakash1997].
 The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002].
@@
 <!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
-In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in Asian market, which established a colony in southern Taiwan two years before the Spanish in the north.
-In 1642, the Dutch defeated the Spanish and occupied their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in Taiwan in 1662 [@Andrade2007].
-The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and resources collections among different Indigenous groups.
-For example, the Basay people, an indigenous group active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@Kang2016].
-The colonial domination was mostly established through the conquest over indigenous villages, but the colonial control varied depending on the location of indigenous settlements.
+In 1626, the Spanish arrived Keelung in northern Taiwan and built Fort San Salvador at Heping Dao and later Fort Santo Domingo in Tamsui to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
+Despite an attempt to build a colonial city, San Salvador remained in its early stages of development and functioned as a presidio (military outpost), hosting mainly Spanish officers, soldiers, and missionaries [@Borao2009, p.103-134].
+The area outside of Fort San Salvador was inhabited by local indigenous groups, the Taparris and Quimaurris, who had resided at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81].
+The Spanish colonization may not have involved effective colonialism or asymmetrical power relations.
+Instead, the Spanish had to maintain good relations with the local indigenous people, as they relied on them for food supply to a certain degree [@Borao2001, SIT:170-171].
+Also, their territorial control could have been limited to a small area around the fortress, with minimal indigenous involvement only in exchange of goods.
+
+In 1642, the Dutch defeated the Spanish and took over their fortresses, officially expanding their colonial power to northern Taiwan until the end of the colonization in 1662.
+Compared with the Spanish, the Dutch were more strategic in consolidating their colonial power by forming alliances with the natives, imposing a bureaucratic and tribute system, and conquering indigenous villages that were unwilling to comply [@Kang2016; @Andrade2007].
+Both Spanish and Dutch worked with local indigenous people for the development of the colony, especially for the Dutch.
+For example, the Basay people active in Keelung, was recorded to have assisted in coordinating or collecting tribute among different Indigenous groups for the VOC governors and they also worked with the Spanish governors in early time [@Kang2016].
+The colonial domination was mostly established through the conquest over indigenous villages and offering protection to the indigenous people, but the colonial control varied depending on the location of indigenous settlements.
 For most indigenous villages distant from the European colonial base, the colonial control was relatively less, indirect, and even unsuccessful.
+
 <!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->
+
+A church was built later that reveals their religious activities [@Berrocal2018].
 

### 2024-07-22 analysis/paper/paper.qmd
work on the historical background

--- 2024-07-19 analysis/paper/paper.qmd
@@
 Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion.
-Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan.
+Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan for a better access to Asian market.
 Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control.
@@
 
-# Colonial encounter in Northeastern Taiwan
+# European encounters in Northeastern Taiwan
 <!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
-In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in Asian market, which established a colony in southern Taiwan two years before the Spanish in the north.
-In 1642, the Dutch defeated the Spanish and occupied their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in Taiwan in 1662 [@Andrade2007].
-The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and resources collections among different Indigenous groups.
-For example, the Basay people, an indigenous group active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@Kang2016].
-The colonial domination was mostly established through the conquest over indigenous villages, but the colonial control varied depending on the location of indigenous settlements.
-For most indigenous villages distant from the European colonial base, the colonial control was relatively less, indirect, and even unsuccessful.
+In 1626, a fleet from the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and built Fort San Salvador at Heping Dao in Keelung, followed by Fort Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had establish a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
+Despite the intention to establish a colonial city, the colony of San Salvador remained in its early developmental stages, primarily consisting of a presidio (military outpost), a church, and a population of Spanish officers, soldiers, and missionaries [@Borao2009, pp.
+103-134; @Berrocal2018].
+The area neighboring Fort San Salvador was inhabited by the Taparris and Quimaurris, local indigenous groups who had resided at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81].
+During the Spanish colonization (1626-1642), the colonial processes may not have involved extensive colonialism or asymmetrical power dynamics, with territorial control likely limited to the areas of fortresses.
+Written records mention that the Spanish appeared to rely on the local indigenous people for exchanging local goods and food supplies [@Borao2001, SIT:170-171].
+This is further supported by archaeological evidence at San Salvador [@Berrocal2018], where only a few European materials were found, alongside more indigenous and Chinese items that might be used by the Europeans, indicating cultural entanglement.
+
 <!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->
+In 1642, the Dutch defeated the Spanish and took over their fortresses, expanding their colonial power to northern and eastern Taiwan until 1662.
+Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007].
+For example, the VOC governors used various colonial institutions to secure their domination, including landdag ceremonies, taxation, village leasehold systems, selection of representatives, and census investigations [@Kang2016; @Andrade1997; @Andrade2007].
+After the Dutch settled northern Taiwan, they expanded their powers further to eastern Taiwan.
+However, their colonial control varied largely across different regions.
+For villages distant from the European colonial base, direct control was often challenging and mostly unsuccessful.
+In these cases, Europeans relied on indigenous intermediaries for colonial activities in remote areas.
+This cooperative interactions was evident in northern Taiwan, where the Dutch worked closely with the Basay people, an active indigenous group in Keelung, who assisted in coordination and collection of tributes among different indigenous groups in northern and northeastern Taiwan and had also worked with Spanish governors earlier [@Kang2016].
 
-Kavalan, or the present day Yilan, in northeastern Taiwan is an example of slight European presence for most of the time.
+<!-- Kavalan as an example -->
+Kavalan, the present day Yilan, in northeastern Taiwan is an example of slight European presence and indirect colonization for most of the time.
 Kavalan is about 70 kilometers away from the European northern fortress with mountains in between.
@@
 According to the Dutch census, there were more than 40 Kavalan villages with approximately 10,000 people in the 17th century [@Nakamura1938], which were also mentioned by the Chinese and Japanese in later periods, suggesting a continuous indigenous habitation until the 19th century.
-Kavalan people first experienced European powers in 1632, when they were attacked by Spanish traders [@citation].
+Kavalan people first experienced European powers in 1632, when they were attacked by a Spanish fleed [@citation].
 Later in 1647, a group of Dutch soldiers attacked several Kavalan villages and asked those defeated villages to pay tributes [@Kang2016].
-According to Dutch written record, the VOC had built a trading outpost in Trobiawan, located near the coastal area in Yilan, but there might be not much officials at any given time.
-It appears that the Kavalan area was in the situation where trading activities were frequent and active stimulated by the European presence in northern Taiwan, with occasional pressures from colonial powers.
-The European control over the Kavalan must have been largely indirect and limited, which allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes requested by the VOC, such as deerskins or products [@Kang2016].
+According to Dutch written record, the VOC had built a trading outpost in Trobiawan, located near the coastal area in Yilan, but not much information about the number of officials or its operation.
+It appears that the Kavalan area was in the situation where trading activities were frequent and active stimulated by the European presence in northern Taiwan, with occasional pressures from European attacks.
+The European control over the Kavalan must have been largely indirect and limited, which allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes, such as deerskins or products, requested by the VOC, which was even aborted later due to their refusal to pay [@Kang2016].
 
+<!-- previous trade network -->
 Before the arrival of the Europeans in northern Taiwan, there had been a regional exchange network involving Chinese merchants, Japanese traders, and indigenous groups in northern and northeastern Taiwan since the 14th century [@citation].

### 2024-07-26 analysis/paper/paper.qmd
fix the render issue and edit

--- 2024-07-19 analysis/paper/paper.qmd
@@
       orcid: 0000-0001-9060-2125
-      institute:
-        - AS
-institute:
-  - AS:
-      name: Academia Sinica
-      address: 
-  - IHP: 
+      institute: Academia Sinica
 title-block-published: "Last updated"  
@@
 European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world.
-The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jorden2009].
+The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jordan2009].
 Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004].
-In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources for commercial profit.
+In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources based on commercial interest [@Prakash1997].
 The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002].
 However, Indigenous communities did not just passively accept European colonization.
-Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jorden2009; @Alexander1998].
+Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998].
 
@@
 Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion.
-Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan.
+Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan for a better access to Asian market.
 Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control.
 However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north.
-In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export in a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001].
+In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export in a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017].
 Despite some discussions on the policies and operations of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people to the commercial activities are rare.
@@
 
-# Colonial encounter in Northeastern Taiwan
-<!-- the structure of power relations, colonial control in particular spatial and temporal contexts-->
-In 1626, the Spanish colonized northern Taiwan and built fortress in Tamsui and Keelung to the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in Asian market, which established a colony in southern Taiwan two years before the Spanish in the north.
-In 1642, the Dutch defeated the Spanish and occupied their fortresses, officially expanding their colonial power to northern Taiwan until the end of their colonization in Taiwan in 1662 [@Andrade2007].
-The form of European colonization in northern Taiwan relied to a certain degree on local Indigenous people for the development of the colony and resources collections among different Indigenous groups.
-For example, the Basay people, an indigenous group active in Keelung, were recorded to have assisted in the colonial activities of the VOC administrators, such as coordinating between indigenous villages [@Kang2016].
-The colonial domination was mostly established through the conquest over indigenous villages, but the colonial control varied depending on the location of indigenous settlements.
-For most indigenous villages distant from the European colonial base, the colonial control was relatively less, indirect, and even unsuccessful.
-<!-- domination and the processes by which it was established and resisted, how and why colonial powers were able to exert control (however limited) over Native populations -->
+# European encounters in the northern region of Taiwan
+<!-- the Spanish colonization and the power relations in particular spatial and temporal contexts-->
+In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
+Despite the intention to establish a colonial city, San Salvador remained in its early developmental stages, primarily consisting of a fortress and a church, with a population of Spanish officers, soldiers, and missionaries [@Borao2009, pp.
+103-134; @Berrocal2018].
+The area neighboring San Salvador was inhabited by the Taparris and Quimaurris, local indigenous groups who had resided at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81].
+During Spanish colonization (1626-1642), the colonial processes may not have involved extensive colonialism or asymmetrical power dynamics, with territorial control likely limited to the fort.
+Written records mention that the Spanish appeared to rely on the local indigenous people for exchanging local goods and food supplies [@Borao2001, SIT:170-171].
+This is further supported by archaeological evidence at San Salvador [@Berrocal2018], where only a few European materials were found, alongside more indigenous and Chinese items that might have been used by the Europeans, indicating cultural entanglement.
 
-Kavalan, or the present day Yilan, in northeastern Taiwan is an example of slight European presence for most of the time.
-Kavalan is about 70 kilometers away from the European northern fortress with mountains in between.
-Local indigenous groups were Kavalan people.
-According to the Dutch census, there were more than 40 Kavalan villages with approximately 10,000 people in the 17th century [@Nakamura1938], which were also mentioned by the Chinese and Japanese in later periods, suggesting a continuous indigenous habitation until the 19th century.
-Kavalan people first experienced European powers in 1632, when they were attacked by Spanish traders [@citation].
-Later in 1647, a group of Dutch soldiers attacked several Kavalan villages and asked those defeated villages to pay tributes [@Kang2016].
-According to Dutch written record, the VOC had built a trading outpost in Trobiawan, located near the coastal area in Yilan, but there might be not much officials at any given time.
-It appears that the Kavalan area was in the situation where trading activities were frequent and active stimulated by the European presence in northern Taiwan, with occasional pressures from colonial powers.
-The European control over the Kavalan must have been largely indirect and limited, which allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes requested by the VOC, such as deerskins or products [@Kang2016].
+<!-- the Dutch domination and the processes by which it was established, how and why colonial powers were able to exert control -->
+In 1642, the Dutch defeated the Spanish and took over the San Salvador and the Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662.
+Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007].
+For example, the VOC governors used various colonial institutions to secure their authority, including annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007].
+Among the political institutions, the tribute system operated from 1642 to 1647 and village lease system introduced in 1644 had great impacts on local indigenous economy and environment.
+Based on tribute system, each household of allied villages was required to pay rice or high-quality deerskin annually.
+The amount of deerskins collected from indigenous people in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007].
+Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on indigenous lands during certain periods [@Heyns2001].
+From 1633 to 1661, the average of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017].
+Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy .
 
-Before the arrival of the Europeans in northern Taiwan, there had been a regional exchange network involving Chinese merchants, Japanese traders, and indigenous groups in northern and northeastern Taiwan since the 14th century [@citation].
-The local indigenous groups traded
+<!--the variation of colonial powers over native populations, indirect colonization at Kavalan-->
+Although Dutch power expanded across nearly all of Taiwan in the mid-17th century, their colonial control and domination varied significantly in different geographical regions.
+In areas distant from the European colonial bases, colonial control was often challenging and largely unsuccessful.
+Yilan in northeastern Taiwan is an example of minimal European presence and indirect colonization, with fewer imposed colonial rules for most of the time [@Kang2012].
+Yilan is an alluvial plain located about 70 kilometers from the European colony in Keelung, separated by mountains, and inhabited mainly by the local indigenous group, the Kavalan.
+Given its strategic location on the route to Torrobouan, where the Dutch believed gold mines existed, they were eager to bring this area under their control [@DZI2000; @Kang2016, p47].
+In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory, and demanded tribute from the indigenous people [@Kang2016].
+nstead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the indigenous Basay people from Keelung to coordinate and collect tributes in Yilan [@Kang2016].
+This indirect approach allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes, such as deerskins or products, requested by the VOC [@Kang2016].
 
-The Yilan Plain is the major coastal plain in northeastern Taiwan.
-Prehistory in the Yilan Plain begins from the Middle Neolithic period around 4000 B.P.
-and continues throughout the Metal Age period (2000-400 B.P.) (Liu 2011).
-The final Late Iron Age in Taiwan can be viewed as a protohistoric period due to the interaction of record-keeping Chinese and Europeans with local cultures.
-Most Late Metal Age sites in the Yilan Plain can be related to the local Indigenous people, the Kavalan, according to oral history and similarities between historical Kavalan material culture and archaeological artifacts (Chen 2004).
+<!-- pre-exiting trade network -->
+Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the early 15th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005].
+Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the local Basay people [@Chen2005; @Andrade2007].
+This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003].
+This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005].
+Yilan experienced frequent trading activities driven by the pursuit of foreign goods even before the Europeans arrival, and these activities continued during the European presence in northern Taiwan, likely with some occasional pressure from European powers.
+This unqiue historical and spatial contexts make Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp.
+31-78; @Trabert2017].
 
- Yilan is about 70 kilometers away from the first European northern base with mountains in between.
-Indigenous people in Yilan were not in direct contact with colonial powers until 1632, when they were attacked by the Spanish.
-Later in 1647, the Dutch attacked the Indigenous people and forced them to accept colonial rules and economic demands by paying annual tribute (Andrade, 2007).
-The Dutch also built a trading outpost in Trobiawan.
-According to the Dutch census, Indigenous people in northeastern Taiwan occupied over 40 settlements with approximately 10,000 people in the 17th century (Nakamura 1938), and those settlements were also described by the Chinese in later periods, demonstrating a continuous Indigenous habitation until the 19th century (Figure 1).
-Due to the fear of European attacks, most local communities in Yilan maintained good relationships with the Europeans by paying tributes and Indigenous leaders could obtain foreign materials as gifts according to the degree of allegiance (Kang 2012).
-In addition to historical records, archaeological evidence also shows complex inter- and intra- exchange networks based on the findings of foreign goods, such as stonewares commonly found in European outposts in Southeast Asia, Chinese porcelains, and glass and stone beads (Chen 2005; Chen 2007; Liu 2011; National Museum of Taiwan History 2005, Wang and Liu 2007).
-This suggests frequent trade activities since the protohistoric period in the 14th century and later intensified through the global trade network stimulated by the Europeans and associated colonial activities.
-Moreover, foreign materials such as ceramics, beads, and textile were viewed as prestige goods in local communities before European arrival (Li and Wu, 2006).
+# Kiwulan site and previous studies
+This study focuses on archaeological evidence from Kiwulan, a large multi-component site of with an upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
+Kiwulan was inhabited by Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture and archaeological artifacts [@Chen2004].
+According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938].
+In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].
 
-Northeastern Taiwan is an especially interesting example of European and Indigenous interaction because the impact of European colonialism appears to have been much less pronounced compared to other places in this region.
-I define northeastern Taiwan as an area at the periphery of colonial center where the European power was indirect with a few sporadic, mostly non-violent direct contacts.
-This type of context has also been called pericolonialism (Acabado, 2017; Kang, 2016, pp.
-31-78; Trabert, 2017).
-My PhD study at Kiwulan (14-19th A.D.), the biggest Indigenous settlement during the 17th century in Yilan, indicates that European contact and foreign goods resulted in an increased social differentiation in the local society through analyses of trade items, pottery production, and burial goods (Wang and Marwick, 2020a, 2020b, 2021).
-The results demonstrate that there was an indirect effect of European colonization in this region that could lead to changes in the organizational mode of a society that have not been identified before.
+The excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, post-holes, and burials.
+These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004].
+In addition to locally made pottery, porcelains, stonewares, glass beads, carnelian beads and metal artifacts were largely found in the residential area and burials.
+Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area.
 
-In addition to social changes, how did the interactions with Europeans and Chinese lead to changes in the local ecological system need further examination.
-For example, during the Dutch colonial period, deer hide was especially demanded in the Dutch colonization period as payment for tributes due to its high value for export to Japan, which led to intensified deer hunting in historical Taiwan (Andrade, 2007; Koo, 2011).
-While the commercial activities and a substantial decrease in Taiwan’s deer population were well documented, much less is known about how this historical consequence can be situated in a local ecological framework.
-Did the participation in these trade activities result in a change in Indigenous subsistence behaviors or the human-environment interactions?
-What was the response of local Indigenous people to the decrease of deer population?
-Those questions are important because they can help us to trace back the Indigenous cultures and their resilience under colonial interference since the protohistoric Taiwan.
+# Previous research of
 
-## trade and exchange of Han Chinese
-
-In this research, I will explore whether and how the introduction of foreign materials and practices, new forms of production, and migrated people from Europe and Chinese contributed to the changes in Indigenous social, environmental, and ecological systems since the protohistoric period (Liu and Wang, 2017).
-I will focus on archeological evidence from different parts in northeastern Taiwan to compare Indigenous reactions in a pericolonial context (Acabodo, 2017).
-Yilan is an ideal location as a starting point because of the frequent foreign contact with Europeans and Chinese traders in the protohistoric and historic period.
-Also, the ecological variation across the landscape from the coastal area to mountains provides multiscalar variability for exploring the effect of colonial contact on a regional scale.
-As a place located in the surrounding area of the European colonial base in northern Taiwan, Yilan enables a pericolonial perspective to understand the effect of European colonization.
-I will use historical ecological approaches to examine the interactions of people and environment in a colonial context, and their consequences on the development of the past and contemporary culture in a long-term view (Braje, 2017).
-This project will help us to understand social and ecological reactions of Indigenous communities to European colonization and cross-cultural contact that has not yet been fully understood.
-
-I will use historical ecology as the conceptual framework to understand the human-environment interactions in northeastern Taiwan, which holds that human impacts on the ecological system are historically dependent (Balée, 1998; Braje, 2017; Fitzhugh et al., 2019).
-In a colonial context, historical ecological investigations can help us to understand the impacts of European presence and associated economic activities on the local environments and domestic economies (Oas and Hauser, 2018; Wallman et al., 2018).
-Located in northeastern Taiwan, Yilan is close to the Dutch colonial base in northern Taiwan with favorable soil conditions for rice growing and natural resources such as forests and deer.
-The frequent exchange activities with Europeans and Chinese merchants, and local interests in foreign goods could lead to further exploitation of deer and intensified rice growing as tributes for maintaining good relations with Europeans and exchanging more imported goods.
-My hypothesis is that if European colonial activities resulted in environmental and ecological changes in northeastern Taiwan due to deer hunting and rice growing, then we could observe changes in Indigenous subsistence and land use patterns.
-I will test my hypothesis by studying archaeological sediments, animal and plant remains, organic residues in locally made ceramics, and site locational preferences of Late Metal Age sites in Yilan to identify any changes after the colonial presence.
-
-# Methods
-The later period of Chinese migration will also be included to provide a long-term perspective
-
-## Geoarchaeological and geochemical analysis for sediments
-To explore changes in environments, I will examine sediments from archaeological sites in northeastern Taiwan, Yilan.
-I will mainly focus on archaeological sites recorded as Indigenous settlements that can be traced back to the protohistoric and historic period when they experienced the European and Chinese presence, such as Kiwulan I, Kiwulan II, and Yilan Agricultural Vocational High School site.
-Kiwulan and Yilan Agricultural Vocational High School are two major Late Metal Age sites with extensive excavation in recent years.
-Kiwulan (600-100 B.P.) is located at the northern part of Yilan plain (Chen, 2007), while Yilan Agricultural Vocational High School site (400-200 B.P.) is situated on the central part of Yilan (Li, 2014).
-Archaeological evidence from both sites includes a large amount of pottery sherds, imported stoneware and porcelains, wooden artifacts, stone and metal tools, pipes, thousands of glass beads, agate beads and metal ornaments, post holes with in-situ wooden posts, and burials.
-Based on their locations and the historical records, those two sites can be traced back to the 17th century when they were named Banouran and Beili according to the Dutch census (Nakamura 1938).
-Those archaeological materials are mainly stored in the Cultural Affairs Bureau in Yilan County, and it is feasible to analyze those materials and collect data based on previous collaboration during my PhD.
-For archaeological sediments, I will use geoarchaeological methods to measure the physical and chemical properties of sediments, including particle size, magnetic susceptibility, pH values, electrical conductivity, and organic matter and carbonates proportions to understand soil conditions and Indigenous landuse.
-I will also use organic chemical analysis to detect lipids in sediments to reconstruct past environments, fire and land use patterns associated with the colonial presence and long-term cultural interaction.
-Geoarchaeological and geochemical methods are useful and insightful methods especially when faunal and botanical evidence were poorly preserved, such as in a condition of acidic environments.
-
-## Faunal and botanical analysis, and organic residue analysis
-To explore any changes in the subsistence strategies, I will examine faunal and botanical evidence directly if they were preserved in good conditions.
-The historical records point out frequent exchange activities between Europeans, Chinese traders, and different Indigenous people, such as Kavalan and Basay during this period (Chen, 2008).
-According to historical records, Indigenous people in Yilan exchanged natural resources, such as deer hide and rice, for porcelains, metal items, and beads from Europeans, Chinese traders, and other Indigenous people, such as Basay (Borao, 2001).
-Deer hide was especially demanded in the Dutch colonization period as payment for tributes due to its high value as a commodity in East Asia.
-To understand change in subsistence behaviors or the human-environment interactions during and after the colonial period, I will examine two types of evidence from those sites.
-First is zooarchaeological and botanical data related to economic activities in a colonial context, especially animal bones such as deer and plant remains such as rice grains.
-Second, I will build on my previous pilot studies of isotope geochemical analysis for samples from Kiwulan, and greatly expand the sample sizes to include human remains (e.g., teeth), animal bones, and ceramic fabrics.
-This will enable me to measure dietary or subsistence changes related to environments, and stability before and after the European presence, and Chinese presence.
-To accomplish those analyses, I will draw on my existing collaborations with the Paleoclimatology and Organic Geochemistry lab in the University of Washington and the Microbial Metabolic Diversity lab in the Academia Sinica for organic residue analysis and DNA analysis, especially for pottery sherds.
-By analyzing the composition of the organic residues absorbed and trapped in potsherds, we can know what categories of food are cooked in pots, such as animal fats, plants, and aquatic resources that provides direct evidence to study ancient diet (Evershed 2008; Craig 2011; Cramp 2011).
-Similarly, microbes from food residues in pots offer a way to detect ancient diet because different types of microbes may be associated with different types of food (Bokulich et al.
-2016).
-I will combine those advanced techniques to understand subsistence strategies in northeastern Taiwan.
+# Mateirals and methods
 

### 2024-07-30 analysis/paper/paper.qmd
minor edits

@@
 # Kiwulan site and previous studies
-This study focuses on archaeological evidence from Kiwulan, a large multi-component site of with an upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
-Kiwulan was inhabited by Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture and archaeological artifacts [@Chen2004].
+This study focuses on archaeological evidence from Kiwulan, a large multi-component site where an upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
+Kiwulan was inhabited by the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in the historical records and archaeological artifacts [@Chen2004].
 According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938].
@@
 
-The excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, post-holes, and burials.
+The salvage excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, burials, and post-holes with posts in situ.
 These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004].
-In addition to locally made pottery, porcelains, stonewares, glass beads, carnelian beads and metal artifacts were largely found in the residential area and burials.
-Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area.
+In addition to locally made pottery with geometric patterns, foreign goods such as porcelains, stonewares, glass beads, carnelian beads and metal artifacts were largely found in the residential area and burials.
+Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area, mostly with the Basay people active in northern Taiwan.
 

### 2024-08-01 analysis/paper/paper.qmd
work on site description

@@
 
-# Kiwulan site and previous studies
-This study focuses on archaeological evidence from Kiwulan, a large multi-component site where an upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
+# Archaeological context at Kiwulan
+This study analyzes archaeological evidence from Kiwulan, a large multi-component site in Yilan, northeastern Taiwan.
+The upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) is the main focus [@Chen2007].
 Kiwulan was inhabited by the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in the historical records and archaeological artifacts [@Chen2004].
@@
 
-The salvage excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, burials, and post-holes with posts in situ.
+The salvage excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, burials, and post-holes with wooden posts in situ.
 These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004].
@@
 Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area, mostly with the Basay people active in northern Taiwan.
+This reveals the continuous participation of Kiwulan in the regional trade network closely associated with northern Taiwan since 15th century [@].
+Previous studies also indicates a possibly increased social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].
 
-# Previous research of
+The salvage excavation at Kiwulan uncovered 262 units (4 x 4 m), revealing numerous artifacts and features, including middens, burials, and post-holes with wooden posts in situ.
+These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side, and burials located to the north of the residential area [@Chen2004].
+Along with locally made pottery featuring geometric patterns, a variety of foreign goods—such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts—were found in both the residential area and burials.
+As there is no evidence of local production of beads or metallurgy, these items are believed to have been acquired through a regional exchange network, primarily with the Basay people active in northern Taiwan.
+This indicates that Kiwulan continuously participated in a regional trade network closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@].
+The use of foreign goods appears to have some cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].
+
+A wide range of animal bones and plant remains were found at Kiwulan, reflecting the diversity of local diet.
+The animals include deer, pigs, cows, birds, and fish, and plants mainly consisted of rice, peach, pandanus, and cucurbits [@Chen2004].
+Rice farming seems to be common but limited to a household scale according to historical records [@].
+Shellfish was also largely found, including bivalves such as clams and oysters, and univalves commonly from rivers and brackish water environments.
+Tools related to subsistence activities were found at Kiwulan, including iron arrowheads, iron sickles, and wooden paddles.
+The iron sickles indicate the cultivation and harvesting of crops, while the arrowheads suggest hunting activities targeting local wildlife such as deer and birds.
+The wooden paddle may have been used for fishing or other water-related activities, indicating a reliance on aquatic resources.
+The subsistence practices at Kiwulan should be primarily based on small scale rice farming, accompanied with hunting and fishing.
+Together, these findings suggest local residences engaged in a diverse range of subsistence activities, balancing agriculture with the exploitation of wild resources to meet their dietary needs.
 

### 2024-08-02 analysis/paper/paper.qmd
text editting

--- 2024-07-26 analysis/paper/paper.qmd
@@
 <!-- introduction about European expansion -->
-European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world.
-The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jordan2009].
-Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004].
-In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources based on commercial interest [@Prakash1997].
-The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002].
-However, Indigenous communities did not just passively accept European colonization.
-Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998].
+European expansion since the 15th century was usually associated with long-distance trade, exploration, and colonization in many parts of the world, which often led to profound impacts on indigenous societies.
+Depending on particular geographic and historical circumstances, many regions experienced different forms of European colonialism and power relations [@Jordan2009].
+Understanding the dynamics of these colonial processes has become a significant issue in archaeology because it provides critical insights into the complex outcomes of European colonization [@Gosden2004].
+In Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items from Asia for commercial purposes [@Prakash1997].
+Although the new exchange systems or colonial institutions introduced by Europeans could change local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002], indigenous groups did not just accept European colonization passively.
+Recent discussions on colonial interactions also suggest the active role of indigenous groups in responding to the arrival of Europeans and presenting resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998].
 
 <!-- NE Taiwan as a case study -->
-While many studies focus on prominent colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017].
+While many studies focus on prominent colonies, places on the periphery of colonial centers were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017].
 Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion.
-Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan for a better access to Asian market.
-Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control.
-However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north.
-In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export in a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017].
-Despite some discussions on the policies and operations of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people to the commercial activities are rare.
+Yilan, located in northeastern Taiwan, is an example of a periphery region where European control was limited and indirect for most of the time but intensively involved in regional trade activities stimulated by the European presence.
+In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017].
+Despite some discussions on the operations and policies of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people with the commercial activities are rare.
 The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jorden2009].
@@
 <!-- research aim and questions -->
-In this paper, I will explore the interactions between people and their environment in a colonial context by examining resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), a largest settlement in northeastern Taiwan [@Chen2007].
-Kiwulan experienced periods of European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century.
-I will primarily focus on subsistence practices, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment.
-My research questions are: What were the subsistence practices of Indigenous people before and after the European presence?
-Can we identify shifts in these strategies in response to colonial economic demands?
-To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues for investigating culinary practices.
+In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest settlement in northeastern Taiwan [@Chen2007].
+Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century.
+What was the resource use of Indigenous people before and after the foreign presence?
+Can we identify shifts in these practices in response to colonial economic demands?
+To address these questions, I will focus on subsistence practices because the choices made in the practices can reveal the strategies Indigenous people used to respond to European exploitation with associated changes in their environment.
+I will primarily examine the zooarchaeological evidence from Kiwulan for subsistence practices, coupled with ceramic residues for culinary practices.
 Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.
@@
 In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
-Despite the intention to establish a colonial city, San Salvador remained in its early developmental stages, primarily consisting of a fortress and a church, with a population of Spanish officers, soldiers, and missionaries [@Borao2009, pp.
+Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp.
 103-134; @Berrocal2018].
-The area neighboring San Salvador was inhabited by the Taparris and Quimaurris, local indigenous groups who had resided at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81].
-During Spanish colonization (1626-1642), the colonial processes may not have involved extensive colonialism or asymmetrical power dynamics, with territorial control likely limited to the fort.
-Written records mention that the Spanish appeared to rely on the local indigenous people for exchanging local goods and food supplies [@Borao2001, SIT:170-171].
-This is further supported by archaeological evidence at San Salvador [@Berrocal2018], where only a few European materials were found, alongside more indigenous and Chinese items that might have been used by the Europeans, indicating cultural entanglement.
+The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81].
+The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort.
+Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171].
+This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement.
 
 <!-- the Dutch domination and the processes by which it was established, how and why colonial powers were able to exert control -->
-In 1642, the Dutch defeated the Spanish and took over the San Salvador and the Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662.
+In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662.
 Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007].
-For example, the VOC governors used various colonial institutions to secure their authority, including annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007].
-Among the political institutions, the tribute system operated from 1642 to 1647 and village lease system introduced in 1644 had great impacts on local indigenous economy and environment.
-Based on tribute system, each household of allied villages was required to pay rice or high-quality deerskin annually.
+For example, the VOC governors used various colonial institutions to secure their authority, including the annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007].
+Among the political institutions, the tribute system operated from 1642 to 1647 and the village lease system introduced in 1644 had great impacts on the local indigenous economy and environment.
+Based on the tribute system, each household in allied villages was required to pay rice or high-quality deerskin annually.
 The amount of deerskins collected from indigenous people in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007].
 Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on indigenous lands during certain periods [@Heyns2001].
-From 1633 to 1661, the average of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017].
-Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy .
+From 1633 to 1661, the average number of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017].
+Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy.
 
 <!--the variation of colonial powers over native populations, indirect colonization at Kavalan-->
-Although Dutch power expanded across nearly all of Taiwan in the mid-17th century, their colonial control and domination varied significantly in different geographical regions.
-In areas distant from the European colonial bases, colonial control was often challenging and largely unsuccessful.
+Although the Dutch power expanded across nearly all of Taiwan in the middle of the 17th century, the extent of their colonial dominance and control differed greatly among different regions.
+Colonial rules in regions far from the European forts were usually difficult and unsuccessful.
 Yilan in northeastern Taiwan is an example of minimal European presence and indirect colonization, with fewer imposed colonial rules for most of the time [@Kang2012].
-Yilan is an alluvial plain located about 70 kilometers from the European colony in Keelung, separated by mountains, and inhabited mainly by the local indigenous group, the Kavalan.
+Yilan is an alluvial plain, located about 70 kilometers from the European colony in Keelung, with mountains in between.
+The main inhabitants in Yilan were the local indigenous group, the Kavalan.
 Given its strategic location on the route to Torrobouan, where the Dutch believed gold mines existed, they were eager to bring this area under their control [@DZI2000; @Kang2016, p47].
-In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory, and demanded tribute from the indigenous people [@Kang2016].
-nstead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the indigenous Basay people from Keelung to coordinate and collect tributes in Yilan [@Kang2016].
-This indirect approach allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes, such as deerskins or products, requested by the VOC [@Kang2016].
+In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory and demanded tribute from the indigenous people [@Kang2016].
+Instead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the Basay people active in northern Taiwan to coordinate and collect tributes in Yilan [@Kang2016].
+This indirect approach allowed the maintenance of autonomy and flexibility for the Kavalan people, who were able to negotiate tribute requested by the VOC [@Kang2016].
 
@@
 Yilan experienced frequent trading activities driven by the pursuit of foreign goods even before the Europeans arrival, and these activities continued during the European presence in northern Taiwan, likely with some occasional pressure from European powers.
-This unqiue historical and spatial contexts make Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp.
+This unique historical and spatial context makes Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp.
 31-78; @Trabert2017].
 
-# Kiwulan site and previous studies
-This study focuses on archaeological evidence from Kiwulan, a large multi-component site of with an upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
-Kiwulan was inhabited by Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture and archaeological artifacts [@Chen2004].
+# Archaeological context at Kiwulan
+Kiwulan, located in the northern part of Yilan, is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
+As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed.
+A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ.
+It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004].
+The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004].
 According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938].
@@
 
-The excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, post-holes, and burials.
-These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004].
-In addition to locally made pottery, porcelains, stonewares, glass beads, carnelian beads and metal artifacts were largely found in the residential area and burials.
-Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area.
+The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan.
+In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials.
+Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through a regional exchange network, primarily with the Basay people active in northern Taiwan.
+This indicates that Kiwulan has continuously participated in a regional trade network closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@].
+Moreover, the use of foreign goods appears to have some cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].
 
-# Previous research of
+There were a wide range of animal bones and plant remains uncovered, reflecting the diversity of the local diet and the environment close to river and coastal area.
+The animals include deer, pigs, cows, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004].
+Rice farming seems to be common but limited to a household scale, according to historical records [@].
+Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments, which indicates a reliance on aquatic and marine resources.
+Also, those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities.
+According to those pieces of evidence, the subsistence practices at Kiwulan should be primarily reliant on rice farming, supplemented by hunting and fishing for wild resources.
 

### 2024-08-06 analysis/paper/paper.qmd
minor edits

@@
 
-# European encounters in the northern region of Taiwan
-<!-- the Spanish colonization and the power relations in particular spatial and temporal contexts-->
+# European encounters in the northern Taiwan
+<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
 In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
@@
 
-<!-- the Dutch domination and the processes by which it was established, how and why colonial powers were able to exert control -->
+<!-- the Dutch domination, how and why colonial powers were able to exert control -->
 In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662.
@@
 
-<!--the variation of colonial powers over native populations, indirect colonization at Kavalan-->
+<!--variation of colonial powers across regions and indirect colonization at Kavalan-->
 Although the Dutch power expanded across nearly all of Taiwan in the middle of the 17th century, the extent of their colonial dominance and control differed greatly among different regions.
@@
 
-<!-- pre-exiting trade network -->
+<!-- the pre-existing trade network in norhtern Taiwan-->
 Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the early 15th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005].
@@
 # Archaeological context at Kiwulan
+<!--introduction of Kiwulan and the population-->
 Kiwulan, located in the northern part of Yilan, is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
@@
 
-The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan.
+<!--archaeological materials-->
+The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005].
 In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials.
@@
 
+<!--archaeological remains-->
 There were a wide range of animal bones and plant remains uncovered, reflecting the diversity of the local diet and the environment close to river and coastal area.

### 2024-08-07 analysis/code/kwl-seed.R
upload code file for KWL seeds

@@ -1,0 +1,102 @@
+# read data
+kwl_seed <- readxl::read_excel(here::here("analysis","data","raw_data","KWL-seed-P-H.xlsx"), sheet = 1,
+                                col_types = "text")
+
+kwl_seed_tidy <-
+  kwl_seed %>%
+  filter(`文化層` == "上文化層") %>%
+  filter(!種類或部位 %in% c("待判J", "待判", "殘件")) %>%
+  mutate(area = str_extract(`坑內位置`, "^.{1}")) %>%
+  left_join(kwl_chro_tidy_6,
+            by = c("Pit" = "Pit", "Layer" = "layer", "area" = "area")) %>%
+  mutate(`重量(g)` = as.numeric(str_remove(`重量(g)`, "以下")))
+
+rice_weight <-
+  kwl_seed_tidy %>%
+  filter(種類或部位 == "稻穀") %>%
+  group_by(period) %>%
+  summarize(`重量(g)` = sum(`重量(g)`)) %>%
+  filter(!is.na(period))
+
+rice_count <-
+  kwl_seed_tidy %>%
+  filter(種類或部位 == "稻穀") %>%
+  filter(!`件數`== "碎") %>%
+  mutate(`件數`= as.numeric(`件數`)) %>%
+  group_by(period) %>%
+  summarize(`件數` = sum(`件數`)) %>%
+  filter(!is.na(period))
+
+rice_weight_count <-
+  rice_weight %>%
+  left_join(rice_count) %>%
+  pivot_longer(-period, names_to = "稻穀", values_to = "value")
+
+# rice plot
+rice_num_weight_plot <-
+  rice_weight_count %>%
+  mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
+  ggplot(aes(x= period, y= value, fill = `稻穀`)) +
+  geom_bar(stat="identity", position = "dodge2") +
+  labs(x= NULL, y= NULL) +
+  #scale_y_continuous(breaks = seq(0, 90, by = 10)) +
+  theme_minimal()
+ggsave(here::here("analysis", "figures", "talk-seed-rice.png"), h = 3, w =4, units = "in")
+
+# other seed
+seed_number_period <-
+  kwl_seed_tidy %>%
+  filter(!is.na(period)) %>%
+  filter(!種類或部位 == "稻穀") %>%
+  count(period) %>%
+  rename(piece = n)
+
+seed_type_period <-
+  kwl_seed_tidy %>%
[…]

### 2024-08-19 analysis/paper/paper.qmd
work on the material section

@@
 In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials.
-Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through a regional exchange network, primarily with the Basay people active in northern Taiwan.
-This indicates that Kiwulan has continuously participated in a regional trade network closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@].
-Moreover, the use of foreign goods appears to have some cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].
+Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through exchange, primarily with the Basay people active in northern Taiwan.
+This indicates that Kiwulan has continuously participated in regional trade networks closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@Wang2007].
+Moreover, the use of foreign goods appears to have cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].
 
@@
 Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments, which indicates a reliance on aquatic and marine resources.
-Also, those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities.
+Those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities.
 According to those pieces of evidence, the subsistence practices at Kiwulan should be primarily reliant on rice farming, supplemented by hunting and fishing for wild resources.
+The resource use during the periods of external impacts still need further examination to better understands the changes over time.
 
 # Mateirals and methods
+To explore changes in the subsistence strategies before and after the European arrival with associated long-distance trade network, I examined zooarchaeological and botanical evidence related to economic activities in a colonial context from Kiwulan.
+The faunal and botanical assemblages analyzed in this study are from 40 units located in the residential area excavated during the first season (2001-2006) at Kiwulan.
+The 40 excavation units were selected due to their relativity intact chronological contexts and spatially continuous area.
+Because deerskin was the main commodity in local regional exchange since the 14th century and further largely demanded during the Dutch colonization period, deer bones and their butchery patterns are especially focused.
 

### 2024-08-21 analysis/paper/paper.qmd
work on the methods

@@
 To address these questions, I will focus on subsistence practices because the choices made in the practices can reveal the strategies Indigenous people used to respond to European exploitation with associated changes in their environment.
-I will primarily examine the zooarchaeological evidence from Kiwulan for subsistence practices, coupled with ceramic residues for culinary practices.
+I will primarily examine the zooarchaeological assemblages from Kiwulan for subsistence practices, coupled with botanical evidence as a supplementary evidence.
+Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as intercultural contact or European colonization (DeFrance, 2009).
 Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.
 
-# European encounters in the northern Taiwan
+# European Encounters in the Northern Taiwan
 <!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
@@
 
-# Archaeological context at Kiwulan
+# Archaeological Context at Kiwulan
 <!--introduction of Kiwulan and the population-->
@@
 
-# Mateirals and methods
-To explore changes in the subsistence strategies before and after the European arrival with associated long-distance trade network, I examined zooarchaeological and botanical evidence related to economic activities in a colonial context from Kiwulan.
-The faunal and botanical assemblages analyzed in this study are from 40 units located in the residential area excavated during the first season (2001-2006) at Kiwulan.
-The 40 excavation units were selected due to their relativity intact chronological contexts and spatially continuous area.
-Because deerskin was the main commodity in local regional exchange since the 14th century and further largely demanded during the Dutch colonization period, deer bones and their butchery patterns are especially focused.
+# Mateirals and Methods
+To explore changes in resource use before and after the European arrival with associated long-distance trade, I analyzed zooarchaeological and botanical assemblages from 40 adjacent units at the residential area excavated during the first season (2001-2004) at Kiwulan [@Chen2007].
+These 40 units were selected based on their relatively intact archaeological contexts and spatial continuity, providing a consistent framework for understanding daily food practices throughout different time periods on a settlement scale.
+
+To construct a reliable chronology, the archaeological records were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic data, sediment color, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
+These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009].
+Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above still provides a feasible framework for analyzing archaeological records before and after the European influence.
+Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with Chinese migrants.
+
+In this study, only the faunal assemblages identified to at least the family level were considered for consistent comparison across time sequences.
+To obtain a general pattern of resource use, measures of ubiquity and taxonomic abundance were applied.
+Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010].
+In this analysis, ubiquity values are expressed as the percentage of units with a given taxon to the total number of units containing cultural layers.
+Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method for a broad overview of taxa through time [@Grayson2004; @Lyman2012].
+For botanical assemblages, their number was served as supplementary evidence for food practices due to potential biases that caused by differential preservation conditions.
+For example, carbonized seeds usually preserve better than uncarbonized ones, which could lead to a misleading interpretation.
+
+Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions and butchery patterns of deer bones were examined.
+The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence.
+The proportion of deer skeletal elements was quantified using the Number of Identified Specimens (NISP) and classified into three groups: meaty elements (e.g., upper limb), mid-meaty elements (e.g., body parts), and non-meaty elements (e.g., cranial parts and antlers) [@Madrigal2004].
+For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified.
+The proportion of deer bones with cut marks was calculated using the normed NISP for comparison between temporal sequences [@Grayson2004].
+In addition, cut marks associated with the skinning process were examined.
+The skinning process can be identified by the encircling marks left around the lower limb, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].
 

### 2024-08-23 analysis/paper/paper.qmd
some more eidts on methods

@@
 Despite some discussions on the operations and policies of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people with the commercial activities are rare.
-The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jorden2009].
+The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jordan2009].
 Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.
@@
 # Mateirals and Methods
-To explore changes in resource use before and after the European arrival with associated long-distance trade, I analyzed zooarchaeological and botanical assemblages from 40 adjacent units at the residential area excavated during the first season (2001-2004) at Kiwulan [@Chen2007].
-These 40 units were selected based on their relatively intact archaeological contexts and spatial continuity, providing a consistent framework for understanding daily food practices throughout different time periods on a settlement scale.
+To explore changes in resource use before and after the European arrival with associated long-distance trade, I examined zooarchaeological and botanical remains from 40 adjacent units at the major residential area excavated during the first season (2001-2004) at Kiwulan [@Chen2007].
+The remains considered in this study are from dwelling spaces and trash pits that reflect daily dietary practices, whereas those from burials that might relate to ritual activities are excluded.
+These 40 units were selected because of their relatively intact archaeological contexts and spatial continuity, providing a consistent framework for understanding daily food practices over time on a settlement scale.
+Moreover, the same excavation and recovery methods applied at the site allow comparison between assembalges from different temporal sequences.
 
-To construct a reliable chronology, the archaeological records were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic data, sediment color, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
+To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
 These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009].
-Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above still provides a feasible framework for analyzing archaeological records before and after the European influence.
-Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with Chinese migrants.
+Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above provides a robust framework for comparing archaeological records before and after the European influence.
+Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the arrival of Chinese migrants in Yilan.
 
-In this study, only the faunal assemblages identified to at least the family level were considered for consistent comparison across time sequences.
-To obtain a general pattern of resource use, measures of ubiquity and taxonomic abundance were applied.
+Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences.
+To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied.
 Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010].
-In this analysis, ubiquity values are expressed as the percentage of units with a given taxon to the total number of units containing cultural layers.
-Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method for a broad overview of taxa through time [@Grayson2004; @Lyman2012].
+Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units containing cultural layers.
+Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method [@Grayson2004; @Lyman2012].
+Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages.
+Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size, and therefore is used throughout this study [@Grayson1984; @Lyman2008].
 For botanical assemblages, their number was served as supplementary evidence for food practices due to potential biases that caused by differential preservation conditions.

### 2024-08-24 analysis/code/kwl-faunal-layers.R
upload code file

@@ -9,51 +9,72 @@ kwl_fauna_HM <- readxl::read_excel(here::here("analysis","data","raw_data","KWL_
 kwl_fauna_broken <- readxl::read_excel(here::here("analysis","data","raw_data","broken_fauna.xlsx"), sheet = 1,
                                    col_types = "text")
 
-sample_p <- readr::read_csv(here::here("analysis", "data", "raw_data",
-                                       "kwl-list-of-sampling-squares.csv"))
+sample_p <- readr::read_csv(here::here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv"))
 
-# run "kwl-pots.R" first, and then tidy data and join chronology
+kwl_chro_6 <- readxl::read_excel(here::here("analysis", "data", "raw_data", "KWL_chronology.xlsx"), sheet = 5)
+
+# tidy chronological data
+kwl_chro_tidy_6 <-
+  kwl_chro_6 %>%
+  mutate(area = str_sub(Pit, 5, 7), Pit = str_sub(Pit, 1, 4)) %>%
+  mutate(area = ifelse(area == "", "ABCD", area)) %>%
+  mutate(area = sapply(strsplit(area, ""), paste, collapse= ",")) %>%
+  separate_rows(area) %>%
+  pivot_longer(cols = starts_with("L"), names_to = "layer", values_to = "period") %>%
+  mutate(layer = str_sub(layer, 2, 3)) %>%
+  mutate(layer = ifelse(nchar(layer) == 1, paste0("0", layer), layer))
+
+# combine faunal data from cultural layers and temporal sequences
 fauna_sam <-
   kwl_fauna %>%
   mutate(Pit = paste(context, pit, sep = "")) %>%
   filter(Pit %in% sample_p$the_sq) %>%
   filter(componant == "上文化層" &!layer == "00") %>% #remove sc
   filter(!str_detect(`總報告類別`, "[[:punct:]]+|[0-9]+")) %>%
-  #mutate(area = ifelse(is.na(area), "A", area)) %>%
-  left_join(kwl_chro_tidy_6,
-            by = c("Pit" = "Pit", "layer" = "layer", "area" = "area")) %>%
[…]

### 2024-08-25 analysis/paper/paper.qmd
minor edit

@@
 # Mateirals and Methods
-To explore changes in resource use before and after the European arrival with associated long-distance trade, I examined zooarchaeological and botanical remains from 40 adjacent units at the major residential area excavated during the first season (2001-2004) at Kiwulan [@Chen2007].
+To explore changes in resource use before and after the European arrival with associated long-distance trade, I examined zooarchaeological and botanical remains from 40 adjacent units at the major residential area excavated during the first season (2001-2003) at Kiwulan [@Chen2007].
 The remains considered in this study are from dwelling spaces and trash pits that reflect daily dietary practices, whereas those from burials that might relate to ritual activities are excluded.
 These 40 units were selected because of their relatively intact archaeological contexts and spatial continuity, providing a consistent framework for understanding daily food practices over time on a settlement scale.
-Moreover, the same excavation and recovery methods applied at the site allow comparison between assembalges from different temporal sequences.
+Moreover, the same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences.
 
@@
 # Results
-
-<!-- Here's some example analysis code: -->
 

### 2024-08-27 analysis/code/kwl-faunal-layers.R
work on code

@@ -161,8 +161,8 @@ fauna_deer_portion <-
     `部位/名稱` %in% c("掌骨","掌骨或蹠骨","蹠骨","踝骨","腕骨","metatarsal","metacarpal","腕或踝骨","腕骨或踝骨",
                    "蹠骨(或掌骨)","掌或蹠骨","蹠骨或掌骨","跟骨","astragalus","astragulas","calcanous",
                    "第1趾骨","第2趾骨","第3趾骨","左第2趾骨","右第2趾骨","carpal附骨？","附骨","趾骨,附骨") ~ "non-meaty\n(lower limb)",
-    `部位/名稱` %in% c("下顎及齒","上或下顎","上下顎齒","下顎骨","顱骨","第二頸椎","枕骨",
-                   "前臼齒","碎齒","齒","臼齒","上顎臼齒") ~ "non-meaty\n(cranial parts)")) #角基部, 犄角, 鹿角
+    `部位/名稱` %in% c("下顎及齒","上或下顎","下顎骨","顱骨","第二頸椎","枕骨"
+                   ) ~ "non-meaty\n(cranial parts)")) #角基部, 犄角, 鹿角, "上下顎齒", "前臼齒","碎齒","齒","臼齒","上顎臼齒"
 
 weight_head <-
 fauna_deer_portion %>%

### 2025-02-05 analysis/code/kwl-faunal-layers.R
code update

@@ -24,52 +24,51 @@ kwl_chro_tidy_6 <-
   mutate(layer = str_sub(layer, 2, 3)) %>%
   mutate(layer = ifelse(nchar(layer) == 1, paste0("0", layer), layer))
 
-# combine faunal data from cultural layers and temporal sequences
+# combine faunal data of cultural layers and temporal sequences
 fauna_sam <-
   kwl_fauna %>%
   mutate(Pit = paste(context, pit, sep = "")) %>%
   filter(Pit %in% sample_p$the_sq) %>%
   filter(componant == "上文化層" &!layer == "00") %>% #remove sc
-  filter(!str_detect(`總報告類別`, "[[:punct:]]+|[0-9]+")) %>%
+  #filter(!str_detect(`總報告類別`, "[[:punct:]]+|[0-9]+")) %>%
   mutate(area = ifelse(is.na(area), "C", area)) %>% # assign an area to the blanks, which won't change the results
-  left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area")) %>%
-  select(-section)
+  left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area"))
 
-broken <- kwl_fauna %>% filter(`部位/名稱` == "碎骨")
-
-# combine faunal data from middens and temporal sequences
+# combine faunal data of middens and temporal sequences
 fauna_H_sam <-
   kwl_fauna_HM %>%
+  select(-`空間號`, -`層   位`) %>%
   mutate(Pit = paste("P", pit, sep = "")) %>%
   filter(Pit %in% sample_p$the_sq) %>%
   filter(componant == "上文化層") %>%
-  filter(!str_detect(`總報告類別`, "[[:punct:]]+|[0-9]+")) %>%
   left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area"))
 
-# tidy data
+# combine and tidy data from midden and cultural layers
 fauna_combined_context <-
   rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
   mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
   mutate(`種屬/屬類` = ifelse(is.na(`種屬/屬類`), "none", `種屬/屬類`)) %>%
   mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
[…]

### 2025-02-05 analysis/paper/paper.qmd
work on Materials and Methods

@@
 <!-- research aim and questions -->
-In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest settlement in northeastern Taiwan [@Chen2007].
+In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest Iron Age settlement in northeastern Taiwan [@Chen2007].
 Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century.
@@
 
-# Mateirals and Methods
-To explore changes in resource use before and after the European arrival with associated long-distance trade, I examined zooarchaeological and botanical remains from 40 adjacent units at the major residential area excavated during the first season (2001-2003) at Kiwulan [@Chen2007].
-The remains considered in this study are from dwelling spaces and trash pits that reflect daily dietary practices, whereas those from burials that might relate to ritual activities are excluded.
-These 40 units were selected because of their relatively intact archaeological contexts and spatial continuity, providing a consistent framework for understanding daily food practices over time on a settlement scale.
-Moreover, the same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences.
+# Materials and Methods
+To explore changes in dietary practices before and after the European arrival with associated long-distance trade, I examined faunal assemblages from 40 adjacent units at the residential area excavated during the first season (2001-2003) at Kiwulan [@Chen2007].
+The remains considered in this study are from dwelling spaces and trash pits that could reflect daily dietary practices, whereas those from burials that might relate to ritual activities are excluded.
+These 40 units were selected because of their location that were believed to be the center of the settlement with thick archaeological deposits and relatively intact archaeological contexts.
+Also, the spatial continuity can provides a more consistent framework for understanding food practices over time on a settlement scale.
+The same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences.
 
@@
 These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009].
-Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above provides a robust framework for comparing archaeological records before and after the European influence.
-Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the arrival of Chinese migrants in Yilan.
+Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could offer a relatively robust framework for comparing archaeological records before and after the European influence.
+Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.
 
 Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences.
+Deer antlers that were not attached to cranial bones were excluded since they might not directly related to dietary practices.
 To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied.
 Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010].
-Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units containing cultural layers.
+Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers.
 Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method [@Grayson2004; @Lyman2012].
@@
 Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size, and therefore is used throughout this study [@Grayson1984; @Lyman2008].
-For botanical assemblages, their number was served as supplementary evidence for food practices due to potential biases that caused by differential preservation conditions.
-For example, carbonized seeds usually preserve better than uncarbonized ones, which could lead to a misleading interpretation.
+For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004].
+The normalized NISP is calculated by dividing NISP by the average number of skeletal elements of that given taxa.
 
-Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions and butchery patterns of deer bones were examined.
+Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions, fragmentation, and butchery patterns of deer bones were examined.
 The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence.
-The proportion of deer skeletal elements was quantified using the Number of Identified Specimens (NISP) and classified into three groups: meaty elements (e.g., upper limb), mid-meaty elements (e.g., body parts), and non-meaty elements (e.g., cranial parts and antlers) [@Madrigal2004].
-For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified.
-The proportion of deer bones with cut marks was calculated using the normed NISP for comparison between temporal sequences [@Grayson2004].
-In addition, cut marks associated with the skinning process were examined.
-The skinning process can be identified by the encircling marks left around the lower limb, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].
+The proportion of deer skeletal elements was quantified using NISP and classified into groups based on the meat yield: meaty elements (e.g., upper limb), mid-meaty elements (e.g., body parts), and non-meaty elements (e.g., skull bones and lower limb) [@Madrigal2004].
+While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain.
+Thus, non-meaty elements were considered as secondary source of nutrition in the interpretation.
+Similarly, the degree of fragmentation and the weight of broken unidentified bones could also imply the extraction of marrow or grease.
+For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified for a comparison between different temporal sequences.
+In addition, modifications such as cut marks, fracturing, and burning were examined, especially the cut marks associated with the skinning process.
+Skinning processes can be identified by the encircling and straight marks left on the lower limb, such as metapodials, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].
 

### 2025-02-06 analysis/code/kwl-faunal-layers.R
work on code

@@ -322,7 +322,7 @@ res <- t.test(cut_percent ~ period, data = fauna_deer_NNISP)
 ################### Cutmarks related to skinning ###################
 # Skeletal elements related to skinning
 deerskin_portion <-
-  fauna_combined_context %>%
+  fauna_deer_only %>%
   mutate(skinning_cut = case_when(str_detect(.$`部位/名稱`, "掌骨")|
                                   str_detect(.$`部位/名稱`, "蹠骨") ~ "metapodials",
                                   str_detect(.$`部位/名稱`, "腕骨")|
@@ -343,6 +343,16 @@ deerskin_portion_NISP <-
   rename(NISP_skin_bone = n) %>%
   drop_na()
 
+# plot bones associated with hide removal processes
+deer_skinning_related <-
+  deerskin_portion_NISP %>%
+  ggplot(aes(x = period, y = NISP_skin_bone , fill = skinning_cut)) +
+  scale_color_discrete(breaks=c("CL1", "CL2", "CL3", "CL4", "CL5", "CL6")) +
+  geom_bar(stat = "identity", position = position_dodge2(preserve = "single")) +
+  labs(x = NULL, y = "NISP") +
+  scale_fill_viridis_d(labels=c('carpal/\ntarsal', 'metapodials', 'phalanges','neck/\ntail regions')) +
+  theme_minimal()
+
 # NISP with circular and straight cutmarks
 deerskin_portion_NISP_cut <-
   deerskin_portion %>%
@@ -351,16 +361,15 @@ deerskin_portion_NISP_cut <-
   count() %>%
   left_join(deerskin_portion_NISP) %>%
   drop_na() %>%
-  #mutate_all(funs(ifelse(is.na(.), 0, .))) %>%
-  mutate(proportion = round(n/NISP_skin_bone, 2))
+  mutate(proportion = n/NISP_skin_bone)
 
-# plot
+# plot the proportion of skinning bones with cutmarks
 deer_cut_skinning <-
   deerskin_portion_NISP_cut %>%
-  ggplot(aes(x = period, y = proportion, fill = skinning_cut)) +
+  ggplot(aes(x = period, y = proportion , fill = skinning_cut)) +
   scale_color_discrete(breaks=c("CL1", "CL2", "CL3", "CL4", "CL5", "CL6")) +
   geom_bar(stat = "identity", position = position_dodge2(preserve = "single")) +
[…]

### 2025-02-08 analysis/code/kwl-faunal-layers.R
udpate code

@@ -46,32 +46,54 @@ fauna_H_sam <-
 # combine and tidy data from midden and cultural layers
 fauna_combined_context <-
   rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
+  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
+  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
   mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
-  mutate(`種屬/屬類` = ifelse(is.na(`種屬/屬類`), "none", `種屬/屬類`)) %>%
+  group_by(taxa) %>%
+  mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
+  ungroup() %>%
   mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
-  mutate(category = case_when(str_detect(animal, "deer") ~ "deer",
-                              str_detect(animal, "fish") ~ "fish",
+  mutate(category = case_when(str_detect(animal, "e deer") ~ "deer", str_detect(animal, "fish") ~ "fish",
                               str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
                               animal == "aves" ~ "bird", TRUE ~ animal)) %>%
-  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|
-                           str_detect(taxa, "erv")|str_detect(taxa, "Sus")|
-                           str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
+  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
+                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
                            str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
-                           str_detect(category, "fish") ~ "fish",
-                           str_detect(category, "bird") ~ "bird", TRUE ~ "reptile")) %>%
[…]

### 2025-02-08 analysis/paper/paper.qmd
work on tables and add one paragraph

@@
 Similarly, the degree of fragmentation and the weight of broken unidentified bones could also imply the extraction of marrow or grease.
+The degree of fragmentation is calculated by dividing Minimal Number of Elements (MNE) by NISP.
 For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified for a comparison between different temporal sequences.
@@
 
-```{r}
-#| label: get-data
-#| eval: false
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
+```{r setup, include= FALSE}
+library(tidyverse)
+library(knitr)
 ```
@@
 ```{r}
-#| label: fig-demo-plot
-#| fig-cap: "A plot of random numbers"
-plot(rnorm(10))
+#| label: get-data
+#| echo: false
+#| cache.extra: tools::md5sum('KWL_faunal_P.xlsx')
+
+# read data
+kwl_fauna <- readxl::read_excel(here::here("analysis","data","raw_data","KWL_faunal_P.xlsx"), sheet = 1,
+                                col_types = "text")
+
+kwl_fauna_HM <- readxl::read_excel(here::here("analysis","data","raw_data","KWL_faunal_HM.xlsx"), sheet = 1,
+                                   col_types = "text")
+
+kwl_fauna_broken <- readxl::read_excel(here::here("analysis","data","raw_data","broken_fauna.xlsx"), sheet = 1,
+                                   col_types = "text")
+
+sample_p <- readr::read_csv(here::here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv"))
+
+kwl_chro_6 <- readxl::read_excel(here::here("analysis", "data", "raw_data", "KWL_chronology.xlsx"), sheet = 5)
+
+# tidy chronological data
+kwl_chro_tidy_6 <-
+  kwl_chro_6 %>%
+  mutate(area = str_sub(Pit, 5, 7), Pit = str_sub(Pit, 1, 4)) %>%
+  mutate(area = ifelse(area == "", "ABCD", area)) %>%
+  mutate(area = sapply(strsplit(area, ""), paste, collapse= ",")) %>%
+  separate_rows(area) %>%
+  pivot_longer(cols = starts_with("L"), names_to = "layer", values_to = "period") %>%
+  mutate(layer = str_sub(layer, 2, 3)) %>%
+  mutate(layer = ifelse(nchar(layer) == 1, paste0("0", layer), layer))
+
+# combine faunal data of cultural layers and temporal sequences
+fauna_sam <-
+  kwl_fauna %>%
+  mutate(Pit = paste(context, pit, sep = "")) %>%
+  filter(Pit %in% sample_p$the_sq) %>%
+  filter(componant == "上文化層" &!layer == "00") %>% #remove sc
+  #filter(!str_detect(`總報告類別`, "[[:punct:]]+|[0-9]+")) %>%
+  mutate(area = ifelse(is.na(area), "C", area)) %>% # assign an area to the blanks, which won't change the results
+  left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area"))
+
+# combine faunal data of middens and temporal sequences
+fauna_H_sam <-
+  kwl_fauna_HM %>%
+  select(-`空間號`, -`層   位`) %>%
+  mutate(Pit = paste("P", pit, sep = "")) %>%
+  filter(Pit %in% sample_p$the_sq) %>%
+  filter(componant == "上文化層") %>%
+  left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area"))
+
+fauna_combined_context <-
+  rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
+  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
+  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
+  mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
+  group_by(taxa) %>%
+  mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
+  ungroup() %>%
+  mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
+  mutate(category = case_when(str_detect(animal, "e deer") ~ "deer", str_detect(animal, "fish") ~ "fish",
+                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
+                              animal == "aves" ~ "bird", TRUE ~ animal)) %>%
+  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
+                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
+                           str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
+                           str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
+                           TRUE ~ "reptile")) %>%
+  mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
+  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
+  mutate(modify = case_when(str_detect(`人為痕跡`, "切")|str_detect(`人為痕跡`, "削")|
+                            str_detect(`人為痕跡`, "砍")~ "cutmarks",
+                            str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒")~ "burn", TRUE ~ "no")) %>%
+  filter(is.na(refitted)) # for NISP
 ```
 
-@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
+```{r}
+#| label: tbl-counts
+#| echo: false
+#| tbl-cap: "Taxa, Number of Identified Specimens (NISP), and weight of faunal remains from the sampled area at Kiwulan."
+#| tbl-colwidths: [40,25,15,20]
+#| tbl-cap-location: bottom
+# taxa counts
+fauna_combined_taxa <-
+  fauna_combined_context %>%
+  count(taxa, `Weight (g)`) %>%
+  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
+  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Unidentified", tolower(taxa), sep =" "),
+                          TRUE ~ taxa)) %>%
+  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~ 
+                                   "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",   
+                                   str_detect(taxa, "Ree") ~ "Muntjac", str_detect(taxa, "Sus") ~ "Boar", 
+                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~ 
+                                   "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~ 
+                                   "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
+                                   str_detect(taxa, "cervid") ~ "Deer family")) %>%
+  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
+                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
+                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6, 
+                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
+                           str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
+                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
+  group_by(group) %>%
+  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
+  ungroup () %>%
+  rename(Taxon = taxa, NISP = n) %>%
+  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
+  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>% 
+  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`))
+
+total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
+unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Uni")) %>% summarise(sum(NISP))
+deer_NISP <- fauna_combined_taxa %>% 
+  filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "Ree")) %>% summarise(sum(NISP))
+deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 1)*100
+
+library(kableExtra)
+knitr::kable(fauna_combined_taxa, align = "l") 
+```
+
+The NISP of faunal assemblages from Kiwulan analyzed in this study is `r total_NISP`, where deer family is the majority that accounts for about `r deer_NISP_per`% of the identified bones, followed by boar and cattle as the second and third source based on NISP (@tbl-counts).
+Within the bones from deer family, sika deer's bone is the most common one, with a few bones from muntjact and a specimen indicating a possible affiliation with sambar deer.
+Although sika deer was the major animal source at the site, variations could be observed between different cultural layers across 600 years.
+To examine and compare general patterns of animal use in terms of their economic values over time, deer family were separated into two categories, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle (NISP = 1) were combined into the cattle category.
+
+```{r}
+#| label: ubiquity
+#| echo: false
+#| tbl-cap: "Ubiquity of faunal categories over the 40 units"
+#| tbl-colwidths: [60,40]
+# calculate the number of units for each cultural layers
+kwl_unit_per_period <-
+  kwl_chro_6 %>%
+  mutate(all = pmap_chr(select(., -Pit, -Grid), ~toString(unique(na.omit(c(...)))))) %>%  # unite and remove duplicates
+  mutate(unit = str_extract(Pit, "[A-Z]+[0-9]{3}")) %>%
+  distinct(unit, .keep_all = TRUE) %>%
+  mutate(CL1 = ifelse(str_detect(all, "CL1"), unit, NA)) %>%
+  mutate(CL2 = ifelse(str_detect(all, "CL2"), unit, NA)) %>%
+  mutate(CL3 = ifelse(str_detect(all, "CL3"), unit, NA)) %>%
+  mutate(CL4 = ifelse(str_detect(all, "CL4"), unit, NA)) %>%
+  mutate(CL5 = ifelse(str_detect(all, "CL5"), unit, NA)) %>%
+  mutate(CL6 = ifelse(str_detect(all, "CL6"), unit, NA))
+
+# count the units by cultural layers for later join
+unit_by_period <-
+  data.frame(unit_count = colSums(!is.na(kwl_unit_per_period))) %>%
+  slice_tail(n = 6) %>%
+  tibble::rownames_to_column("period")
+
+# ubiquity of taxa from the 40 units by temporal sequences
+fauna_ubiquity <-
+  fauna_combined_context %>%
+  select(Pit, layer, category, `重量(g)`, period, taxa) %>%
+  count(period, Pit, category) %>%
+  count(period, category) %>%
+  drop_na(period, category) %>%
+  left_join(unit_by_period) %>% # join the unit counts for each period
+  mutate(ubiquity = paste(round(n/unit_count *100, 2), "%")) %>%
+  select(-n, -unit_count) %>%
+  pivot_wider(names_from = period, values_from = ubiquity) %>%
+  mutate(across(CL1:CL6, ~replace_na(.x, "-"))) %>% 
+  arrange(category)
+
+knitr::kable(fauna_ubiquity)
+```
+
+@ubiquity shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
 

### 2025-02-10 analysis/code/kwl-faunal-layers.R
make plots for relative abundance

@@ -44,7 +44,7 @@ fauna_H_sam <-
   left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area"))
 
 # combine and tidy data from midden and cultural layers
-fauna_combined_context <-
+fauna_combined_context_all <-
   rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
   filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
   filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
@@ -53,7 +53,8 @@ fauna_combined_context <-
   mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
   ungroup() %>%
   mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
-  mutate(category = case_when(str_detect(animal, "e deer") ~ "deer", str_detect(animal, "fish") ~ "fish",
+  mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
+                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "fish") ~ "fish",
                               str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
                               animal == "aves" ~ "bird", TRUE ~ animal)) %>%
   mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
@@ -70,7 +71,7 @@ fauna_combined_context <-
 
 # taxa counts
 fauna_combined_taxa <-
-  fauna_combined_context %>%
+  fauna_combined_context_all %>%
   count(taxa, `Weight (g)`) %>%
   mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
   mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Unidentified", tolower(taxa), sep =" "),
@@ -86,7 +87,6 @@ fauna_combined_taxa <-
                            str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6, str_detect(`Common name`, "fish") ~ 7,
                            str_detect(taxa, "Tes") ~ 8, str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
[…]

### 2025-02-10 analysis/paper/paper.qmd
work on the result section

--- 2025-02-05 analysis/paper/paper.qmd
@@
 
-```{r}
-#| label: get-data
-#| eval: false
-# Note the path that we need to use to access our data files when rendering this document
-my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
+```{r setup, include= FALSE}
+library(tidyverse)
+library(knitr)
+library(viridis)
+library(cowplot)
 ```
@@
 ```{r}
-#| label: fig-demo-plot
-#| fig-cap: "A plot of random numbers"
-plot(rnorm(10))
+#| label: get-data
+#| echo: false
+
+# read data
+kwl_fauna <- readxl::read_excel(here::here("analysis","data","raw_data","KWL_faunal_P.xlsx"), sheet = 1,
+                                col_types = "text")
+
+kwl_fauna_HM <- readxl::read_excel(here::here("analysis","data","raw_data","KWL_faunal_HM.xlsx"), sheet = 1,
+                                   col_types = "text")
+
+kwl_fauna_broken <- readxl::read_excel(here::here("analysis","data","raw_data","broken_fauna.xlsx"), sheet = 1,
+                                   col_types = "text")
+
+sample_p <- readr::read_csv(here::here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv"))
+
+kwl_chro_6 <- readxl::read_excel(here::here("analysis", "data", "raw_data", "KWL_chronology.xlsx"), sheet = 5)
+
+# tidy chronological data
+kwl_chro_tidy_6 <-
+  kwl_chro_6 %>%
+  mutate(area = str_sub(Pit, 5, 7), Pit = str_sub(Pit, 1, 4)) %>%
+  mutate(area = ifelse(area == "", "ABCD", area)) %>%
+  mutate(area = sapply(strsplit(area, ""), paste, collapse= ",")) %>%
+  separate_rows(area) %>%
+  pivot_longer(cols = starts_with("L"), names_to = "layer", values_to = "period") %>%
+  mutate(layer = str_sub(layer, 2, 3)) %>%
+  mutate(layer = ifelse(nchar(layer) == 1, paste0("0", layer), layer))
+
+# combine faunal data of cultural layers and temporal sequences
+fauna_sam <-
+  kwl_fauna %>%
+  mutate(Pit = paste(context, pit, sep = "")) %>%
+  filter(Pit %in% sample_p$the_sq) %>%
+  filter(componant == "上文化層" &!layer == "00") %>% #remove sc
+  #filter(!str_detect(`總報告類別`, "[[:punct:]]+|[0-9]+")) %>%
+  mutate(area = ifelse(is.na(area), "C", area)) %>% # assign an area to the blanks, which won't change the results
+  left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area"))
+
+# combine faunal data of middens and temporal sequences
+fauna_H_sam <-
+  kwl_fauna_HM %>%
+  select(-`空間號`, -`層   位`) %>%
+  mutate(Pit = paste("P", pit, sep = "")) %>%
+  filter(Pit %in% sample_p$the_sq) %>%
+  filter(componant == "上文化層") %>%
+  left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area"))
+
+fauna_combined_context_all <-
+  rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
+  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
+  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
+  mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
+  group_by(taxa) %>%
+  mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
+  ungroup() %>%
+  mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
+  mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
+                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "fish") ~ "fish",
+                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
+                              animal == "aves" ~ "bird", TRUE ~ animal)) %>%
+  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
+                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
+                           str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
+                           str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
+                           TRUE ~ "reptile")) %>%
+  mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
+  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%  mutate(modify = case_when(str_detect(`人為痕跡`, "切")|str_detect(`人為痕跡`, "削")|
+                            str_detect(`人為痕跡`, "砍")~ "cutmarks",
+                            str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒")~ "burn", TRUE ~ "no")) %>%filter(is.na(refitted)) # for NISP
+
+fauna_combined_context <- fauna_combined_context_all %>% filter(!taxa == "Rattus sp.")
 ```
 
-@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
+```{r}
+#| label: taxa-NISP-counts
+#| echo: false
+# taxa counts
+fauna_combined_taxa <-
+  fauna_combined_context_all %>%
+  count(taxa, `Weight (g)`) %>%
+  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
+  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Unidentified", tolower(taxa), sep =" "),
+                          TRUE ~ taxa)) %>%
+  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~ 
+                                   "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",   
+                                   str_detect(taxa, "Ree") ~ "Muntjac", str_detect(taxa, "Sus") ~ "Boar", 
+                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~ 
+                                   "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~ 
+                                   "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
+                                   str_detect(taxa, "cervid") ~ "Deer family")) %>%
+  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
+                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
+                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6, 
+                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
+                           str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
+                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
+  group_by(group) %>%
+  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
+  ungroup () %>%
+  rename(Taxon = taxa, NISP = n) %>%
+  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
+  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>% 
+  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`))
+
+total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
+unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Uni")) %>% summarise(sum(NISP))
+deer_NISP <- fauna_combined_taxa %>% 
+  filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "Ree")) %>% summarise(sum(NISP))
+deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 3)*100
+```
+
+The NISP of faunal assemblages from Kiwulan analyzed in this study is `r total_NISP`, where deer family is the majority that accounts for about `r deer_NISP_per`% of the identified bones, followed by boar and cattle as the second and third taxon (@tbl-counts).
+Within the assemblages from deer family, sika deer is the most common animal, with a few assemblages identified as muntjac and a specimen indicating a possible affiliation with sambar deer.
+Although sika deer was the major animal source at the site, variations in animal use could be observed between different cultural layers across 600 years.
+To highlight general patterns of dietary practices in terms of their economic values over time, deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle.
+Rats (Rattus sp.) were excluded from further analysis as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.
+
+```{r}
+#| label: tbl-counts
+#| echo: false
+#| tbl-cap: "Taxa, Number of Identified Specimens (NISP), and weight of faunal remains from the sampled area at Kiwulan."
+#| tbl-colwidths: [40,25,15,20]
+#| tbl-cap-location: bottom
+library(kableExtra)
+knitr::kable(fauna_combined_taxa, align = "l") 
+```
+
+```{r}
+#| label: mammal-counts
+#| echo: false
+mammal_count <-
+  fauna_combined_context %>%
+  filter(class == "mammal") %>%
+  count(period, category) %>% 
+  drop_na() %>% 
+  bind_rows(summarise(., across(where(is.numeric), sum), .by = period))
+
+mutate(category = ifelse(is.na(category), "total", category))
+
+CL4 <- mammal_count %>% filter(period == "CL4") 
+CL4_boar <- round(CL4$n[2]/CL4$n[4], 3)*100
+
+```
+
+For animal classes, mammals are the primary dietary source throughout all layers, with fish as the secondary source in the layer 4 and birds in the layer 3 and the layer 2 (@fig-NISP-proportion: A).
+Fish and birds both identified from the layer 5 to the layer 3; however, their proportions (less then 10%) are much lower compared to mammals.
+Reptile is represented by one piece of turtle shell found in the layer 4, which might not be a common food content at the site.
+Within those identifiable mammals, deer is the major source from the layer 6 to the layer 2 that account for at least 85 % per layer, while the layer 1 is predominated only by cattle (@fig-NISP-proportion: B).
+Wild boar was present from the layer 6 to layer 3, and has a higher proportion in the layer 4 (`r CL4_boar` %), while muntjac appeared in the layer 5, and continuously present to the layer 2.
+Given that relative abundances could be impacted by sample size effect across layers, the relationship between NISP and sample size for major taxa were examined by linear regression model.
+The results shows that
+
+```{r}
+#| label: fig-NISP-proportion
+#| echo: false
+#| fig-cap: "A: Relative abundance of vertebrate classes by cultural layers based on normed NISP.
+The data is derived from Table 1.
+B: Relative abundance of mammals by cultural layers based on NISP.
+CL6 to CL1 prepresent cultural layers from the earliest to the latest layer.
+The NISP for each layer is CL6: 46, CL5: 65, CL4: 156, CL3: 28, CL2: 12, and CL1: 25."
+
+knitr::include_graphics(here::here("analysis/figures/NISP_pro_by_layer.png"))
+```
+
+```{r}
+#| label: ubiquity
+#| echo: false
+#| tbl-cap: "Ubiquity of faunal categories over the 40 units"
+#| tbl-colwidths: [60,40]
+# calculate the number of units for each cultural layers
+kwl_unit_per_period <-
+  kwl_chro_6 %>%
+  mutate(all = pmap_chr(select(., -Pit, -Grid), ~toString(unique(na.omit(c(...)))))) %>%  # unite and remove duplicates
+  mutate(unit = str_extract(Pit, "[A-Z]+[0-9]{3}")) %>%
+  distinct(unit, .keep_all = TRUE) %>%
+  mutate(CL1 = ifelse(str_detect(all, "CL1"), unit, NA)) %>%
+  mutate(CL2 = ifelse(str_detect(all, "CL2"), unit, NA)) %>%
+  mutate(CL3 = ifelse(str_detect(all, "CL3"), unit, NA)) %>%
+  mutate(CL4 = ifelse(str_detect(all, "CL4"), unit, NA)) %>%
+  mutate(CL5 = ifelse(str_detect(all, "CL5"), unit, NA)) %>%
+  mutate(CL6 = ifelse(str_detect(all, "CL6"), unit, NA))
+
+# count the units by cultural layers for later join
+unit_by_period <-
+  data.frame(unit_count = colSums(!is.na(kwl_unit_per_period))) %>%
+  slice_tail(n = 6) %>%
+  tibble::rownames_to_column("period")
+
+# ubiquity of taxa from the 40 units by temporal sequences
+fauna_ubiquity <-
+  fauna_combined_context %>%
+  select(Pit, layer, category, `重量(g)`, period, taxa) %>%
+  count(period, Pit, category) %>%
+  count(period, category) %>%
+  drop_na(period, category) %>%
+  left_join(unit_by_period) %>% # join the unit counts for each period
+  mutate(ubiquity = paste(round(n/unit_count *100, 2), "%")) %>%
+  select(-n, -unit_count) %>%
+  pivot_wider(names_from = period, values_from = ubiquity) %>%
+  mutate(across(CL1:CL6, ~replace_na(.x, "-"))) %>% 
+  arrange(category)
+
+knitr::kable(fauna_ubiquity)
+```
+
+@ubiquity shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
 

### 2025-02-11 analysis/paper/paper.qmd
work on the result section

@@
 The NISP of faunal assemblages from Kiwulan analyzed in this study is `r total_NISP`, where deer family is the majority that accounts for about `r deer_NISP_per`% of the identified bones, followed by boar and cattle as the second and third taxon (@tbl-counts).
-Within the assemblages from deer family, sika deer is the most common animal, with a few assemblages identified as muntjac and a specimen indicating a possible affiliation with sambar deer.
+Within the assemblages from the deer family, sika deer is the most common animal, with a few assemblages identified as muntjac and a specimen indicating a possible affiliation with sambar deer.
 Although sika deer was the major animal source at the site, variations in animal use could be observed between different cultural layers across 600 years.
-To highlight general patterns of dietary practices in terms of their economic values over time, deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle.
+To highlight general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle.
 Rats (Rattus sp.) were excluded from further analysis as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.
@@
   drop_na() %>% 
-  bind_rows(summarise(., across(where(is.numeric), sum), .by = period))
-
+  bind_rows(summarise(., across(where(is.numeric), sum), .by = period)) %>% 
+  mutate(category = as.character(category)) %>% 
   mutate(category = ifelse(is.na(category), "total", category))
@@
 
+# taxonomic diversity
+category_count <-
+  fauna_combined_context %>%
+  count(period, category) %>% 
+  count(period, category) %>% 
+  drop_na() %>% 
+  bind_rows(summarise(., across(where(is.numeric), sum), .by = period)) %>% 
+  slice(24:29) %>% 
+  select(period, n) %>% 
+  rename(diversity = n)
+
+# wide format and join
+wider_mammal_count <-
+  mammal_count %>% 
+  pivot_wider(id_cols = period, names_from = category, values_from = n, values_fill = 0) %>% 
+  left_join(category_count)
+
+boar_reg <- summary(lm(boar ~ total, data = wider_mammal_count))
+boar_lm <- c(round(boar_reg$adj.r.squared, 2), sprintf('%.3f', boar_reg$coefficients[,4][2]))
+munt_reg <- summary(lm(muntjac ~ total, data = wider_mammal_count))
+munt_lm <- c(round(munt_reg$adj.r.squared, 2), sprintf('%.2f', munt_reg$coefficients[,4][2]))
+deer_reg <- summary(lm(deer ~ total, data = wider_mammal_count))
+div_reg <- summary(lm(diversity ~ total, data = wider_mammal_count))
+div_lm <- c(round(div_reg$adj.r.squared, 2), sprintf('%.2f', div_reg$coefficients[,4][2]))
+
 ```
 
-For animal classes, mammals are the primary dietary source throughout all layers, with fish as the secondary source in the layer 4 and birds in the layer 3 and the layer 2 (@fig-NISP-proportion: A).
-Fish and birds both identified from the layer 5 to the layer 3; however, their proportions (less then 10%) are much lower compared to mammals.
-Reptile is represented by one piece of turtle shell found in the layer 4, which might not be a common food content at the site.
-Within those identifiable mammals, deer is the major source from the layer 6 to the layer 2 that account for at least 85 % per layer, while the layer 1 is predominated only by cattle (@fig-NISP-proportion: B).
-Wild boar was present from the layer 6 to layer 3, and has a higher proportion in the layer 4 (`r CL4_boar` %), while muntjac appeared in the layer 5, and continuously present to the layer 2.
-Given that relative abundances could be impacted by sample size effect across layers, the relationship between NISP and sample size for major taxa were examined by linear regression model.
-The results shows that
+For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in layer 4 and birds in layer 3 and layer 2 (@fig-NISP-proportion: A).
+Fish and birds are both identified from layer 5 to layer 3; however, their proportions (less than 10%) are much lower compared to mammals.
+The reptile is represented by one piece of turtle shell found in layer 4, which might not be a common food content at the site.
+Within those identifiable mammals, deer is the major source from layer 6 to layer 2 that accounts for at least 85% per layer, while layer 1 is predominated only by cattle (@fig-NISP-proportion: B).
+Wild boar was present from layer 6 to layer 3, and had a higher proportion in layer 4 (`r CL4_boar` %), while muntjac appeared in layer 5, and was continuously present until layer 2.
+In general, more diverse animals were found from layer 5 to layer 3, especially in layer 4, where boar and fish have slightly higher proportions compared with other layers.
+Given that the taxonomic richness, or diversity of animals, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model.
+The result shows no correlation between the diversity of animals and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), which rules out the potential sample size bias.
 
@@
 ```{r}
-#| label: ubiquity
+#| label: tbl-ubiquity
 #| echo: false
 #| tbl-cap: "Ubiquity of faunal categories over the 40 units"
-#| tbl-colwidths: [60,40]
+#| tbl-cap-location: bottom
 # calculate the number of units for each cultural layers
@@
 
-@ubiquity shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
+@tbl-ubiquity shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
 

### 2025-02-12 analysis/paper/paper.qmd
add paragraphs

@@
 The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence.
-The proportion of deer skeletal elements was quantified using NISP and classified into groups based on the meat yield: meaty elements (e.g., upper limb), mid-meaty elements (e.g., body parts), and non-meaty elements (e.g., skull bones and lower limb) [@Madrigal2004].
+The proportion of deer skeletal elements was quantified using NISP and classified into groups based on the meat yield: meaty elements (e.g., upper forelimb and upper hindlimb ), mid-meaty elements (e.g., body parts), and less-meaty elements (e.g., skull bones and lower limb) [@Madrigal2004].
 While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain.
 Thus, non-meaty elements were considered as secondary source of nutrition in the interpretation.
-Similarly, the degree of fragmentation and the weight of broken unidentified bones could also imply the extraction of marrow or grease.
+Similarly, the degree of fragmentation could also imply the extraction of marrow or grease.
 The degree of fragmentation is calculated by dividing Minimal Number of Elements (MNE) by NISP.
@@
 
-For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in layer 4 and birds in layer 3 and layer 2 (@fig-NISP-proportion: A).
-Fish and birds are both identified from layer 5 to layer 3; however, their proportions (less than 10%) are much lower compared to mammals.
-The reptile is represented by one piece of turtle shell found in layer 4, which might not be a common food content at the site.
-Within those identifiable mammals, deer is the major source from layer 6 to layer 2 that accounts for at least 85% per layer, while layer 1 is predominated only by cattle (@fig-NISP-proportion: B).
-Wild boar was present from layer 6 to layer 3, and had a higher proportion in layer 4 (`r CL4_boar` %), while muntjac appeared in layer 5, and was continuously present until layer 2.
-In general, more diverse animals were found from layer 5 to layer 3, especially in layer 4, where boar and fish have slightly higher proportions compared with other layers.
-Given that the taxonomic richness, or diversity of animals, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model.
+For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in CL4 and birds in CL3 and CL2 (@fig-NISP-proportion: A).
+Fish and birds are both identified from CL5 to CL3; however, their proportions (less than 10%) are much lower compared to mammals.
+The reptile is represented by one piece of turtle shell found in CL4, which might not be a common food content at the site.
+Within those identifiable mammals, deer is the major source from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by cattle (@fig-NISP-proportion: B).
+Wild boar was present from CL6 to CL3, and had a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5, and was continuously present until CL2.
+In general, there is an increase in animal diversity in CL4 and CL5, especially in CL4, where boar and fish have slightly higher proportions compared with other layers.
+Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model.
 The result shows no correlation between the diversity of animals and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), which rules out the potential sample size bias.
@@
 #| echo: false
-#| tbl-cap: "Ubiquity of faunal categories over the 40 units"
+#| tbl-cap: "Ubiquity of faunal categories across the sampling area at Kiwulan."
 #| tbl-cap-location: bottom
@@
 
+CL4_ubi_deer <- fauna_ubiquity$CL4[1]
+CL4_ubi_boar <- fauna_ubiquity$CL4[2]
+CL4_ubi_munt <- fauna_ubiquity$CL4[3]
+
 knitr::kable(fauna_ubiquity)
@@
 
-@tbl-ubiquity shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`
+@tbl-ubiquity shows deer had been the most frequent animal from CL6 to CL2, with the highest ubiquity in CL4 and CL5 (`r CL4_ubi_deer`), but disappeared in CL1.
+The second and third frequent animal are boar and muntjac.
+Although both animals reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for boar, and `r CL4_ubi_munt` for muntjac), their presence patterns are slightly different.
+Boar was utilized as early as deer in CL6, declined in CL3, and disappeared in CL2, whereas muntjac showed a cultural lag, appearing one phase later than boar.
+Cattle had a low ubiquity in CL5 and high ubiquity in CL6.
+Small animals (bird, fish, and turtle) had much lower ubiquity than large animals and they had a slightly increase in ubiquity in CL3.
+The result of animal ubiquity correspond with the result of their quantity (NISP), suggesting that deer is the most common resource at the site, with an increase utilization of boar and muntjac in CL4.
 
 ```{r}
-#| label: demo-inline-code
+#| label: fig-deer-portion
 #| echo: false
-x <- round(pi, 2)
+#| fig-cap: "The frequency of skeletal portions using NISP of deer across cultural layers at Kiwulan"
+
+knitr::include_graphics(here::here("analysis/figures/deer_portion.png"))
 ```
 
-Here is an example of inline code `r x` in the middle of a sentence.
+Since deer remains was the major component of faunal assemblage at Kiwulan according to their high quantity and ubiquity, their body parts were examined to explore any changes in dietary practices and preferences.
 
 # Discussion
+
+played a consistent role in subsistence strategies.
+A taxon can have high NISP but low ubiquity if concentrated in a few contexts (e.g., feasting deposits).
+A taxon with high ubiquity but low NISP may have been used frequently but in small amounts.
+
+Other than quantity, the measure of presence and absence of a given taxon across the sampeld units could give insights into the frequency and distribution consistency of taxa at a settlement scale.
+
+High-ubiquity taxa (e.g., deer, pigs) likely played a consistent role in subsistence strategies.
+Low-ubiquity taxa (e.g., rare birds, marine species) may indicate opportunistic hunting, ritual use, or seasonal availability.
+Changes in ubiquity across layers may reflect environmental shifts or human adaptations.
+For example, a decline in large mammal ubiquity might suggest deforestation or overhunting, leading to a shift toward smaller prey.
+
+Issue of Taphonomic & Preservation Bias: Some taxa may be less ubiquitous due to poor bone preservation rather than actual scarcity in the past.
+Small and fragile bones (e.g., fish, rodents) might have low ubiquity due to taphonomic processes.
 

### 2025-02-13 analysis/code/kwl-faunal-layers.R
work on plots

@@ -114,7 +114,7 @@ verte_class_barplot <-
   mutate(class = factor(class, levels = c("mammal", "bird", "fish", "reptile"), ordered = TRUE)) %>%
   ggplot(aes(x = period, fill = class))+
   geom_bar(position = "fill", width = 0.7) + #position_dodge2(preserve = "single")
-  labs(y = "NISP", x = NULL) +
+  labs(y = "NISP", x = "cultural layer") +
   scale_y_continuous(labels = scales::percent) +
   theme_minimal() +
   theme(legend.title=element_blank())
@@ -134,19 +134,19 @@ verte_class_barplot_normed <-
   mutate(`%NNISP` = (NNISP/total_per_period)*100) %>%
   ggplot(aes(x = period, y = `%NNISP`))+
   geom_bar(stat = "identity", aes(fill = class), width = 0.7) +
-  labs(y = "%NNISP", x = NULL) +
+  labs(y = "%NNISP", x = "cultural layer") +
   theme_minimal()
 
 # barplot III: relative abundance of mammals by period
 library(viridis)
 verte_mammal_barplot <-
   fauna_combined_context %>%
-  filter(class == "mammal" &!category == "rat") %>%
+  filter(class == "mammal") %>%
   drop_na(period, category) %>% # remove category to get absolute abundance
   mutate(category = factor(category, levels = c("deer", "boar", "muntjac", "cattle"), ordered = TRUE)) %>%
   ggplot(aes(x = period, fill = category))+
   geom_bar(position = "fill", width = 0.7) +
-  labs(y = "%NISP", x = NULL) +
+  labs(y = "%NISP", x = "cultural layer") +
   scale_y_continuous(labels = scales::percent) +
   theme_minimal() +
   scale_fill_viridis_d(option="magma", begin = 0.2, end = 0.9) +  #plasma
@@ -156,9 +156,8 @@ library(cowplot)
 plot_grid(verte_class_barplot_normed, verte_mammal_barplot,
[…]

### 2025-02-14 analysis/paper/paper.qmd
work on the result section

@@
 
-Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions, fragmentation, and butchery patterns of deer bones were examined.
-The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence.
-The proportion of deer skeletal elements was quantified using NISP and classified into groups based on the meat yield: meaty elements (e.g., upper forelimb and upper hindlimb ), mid-meaty elements (e.g., body parts), and less-meaty elements (e.g., skull bones and lower limb) [@Madrigal2004].
-While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain.
-Thus, non-meaty elements were considered as secondary source of nutrition in the interpretation.
+Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequency of skeletal elements and surface modifications such as cutmarks of deer bones were examined to assess butchery and transport strategy.
+This can also shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network.
+Skeletal elements was quantified using NISP and classified into groups based on meat yield: meaty (e.g., upper forelimb and upper hindlimb ), mid-meaty (e.g., body parts), and less meaty portions (e.g., skull and lower limb) [@Madrigal2004; @Munro2004].
+Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation compared to othe parts.
+While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002].
+Thus, less meaty elements were considered as secondary source of nutrition in the interpretation.
 Similarly, the degree of fragmentation could also imply the extraction of marrow or grease.
@@
 div_lm <- c(round(div_reg$adj.r.squared, 2), sprintf('%.2f', div_reg$coefficients[,4][2]))
-
 ```
@@
 
+# get the bone mineral desity data
+deer_element_BMD <- readr::read_csv(here::here("analysis", "data", "derived_data", "deer_BMD.csv"))
+cor_sur_BMD <- cor.test(deer_element_BMD$NISP, deer_element_BMD$BMD)
+survivor_cor <- round(c(cor_sur_BMD$estimate, cor_sur_BMD$p.value),2)
+
 knitr::kable(fauna_ubiquity)
@@
 
+Since deer remains was the major component of faunal assemblage at Kiwulan according to their high quantity and ubiquity, the frequency of skeletal elements were examined to explore any changes in processing decisions of meat portions between different phases.
+Deer bones were grouped into five body portions based on their nutritional value [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), and lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus, and phalanges).
+The result (@fig-deer-portion: A) shows that the meaty portion, including upper forelimb and upper hindlimb, especially upper forelimb, is the most frequent part throughout the cultural layers, followed by the less-meaty portions, such as lowerlimb and foot.
+The consistency of higher numbers of upperlimb and lowerlimb suggests that those elements should be transported together (e.g., @Stiner1991), or marrows from upperlimb were commonly utilized.
+It should be noted that the category of axial such as ribs were generally low across different phases, which might be caused by highly fragmentation, butchery practices or dietary preferences.
+A significant change is observed from the frequency of head bones that shows a few in CL6, increased and reach the highest number in CL4, and then decreased since then.
+
+Considering varying frequencies of skeletal parts might be mediated by differential bone density, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation.
+The values of bone survivorship was collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%.
+The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between skeletal part survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones.
+
 ```{r}
@@
 #| echo: false
-#| fig-cap: "The frequency of skeletal portions using NISP of deer across cultural layers at Kiwulan"
+#| fig-cap: "A: The frequency of skeletal parts of deer using NISP across cultural layers at Kiwulan. B: Bone minerial density agaist %survivorship of deer remains."
 
-knitr::include_graphics(here::here("analysis/figures/deer_portion.png"))
+knitr::include_graphics(here::here("analysis/figures/deer_elements.png"))
 ```
-
-Since deer remains was the major component of faunal assemblage at Kiwulan according to their high quantity and ubiquity, their body parts were examined to explore any changes in dietary practices and preferences.
 

### 2025-02-14 analysis/paper/paper.qmd
fixed grammar errors

@@
 
-Since deer remains was the major component of faunal assemblage at Kiwulan according to their high quantity and ubiquity, the frequency of skeletal elements were examined to explore any changes in processing decisions of meat portions between different phases.
+Since deer remain was the major component of faunal assemblage at Kiwulan according to their high quantity and ubiquity, the frequency of skeletal elements was examined to explore any changes in processing decisions of meat portions between different phases.
 Deer bones were grouped into five body portions based on their nutritional value [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), and lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus, and phalanges).
-The result (@fig-deer-portion: A) shows that the meaty portion, including upper forelimb and upper hindlimb, especially upper forelimb, is the most frequent part throughout the cultural layers, followed by the less-meaty portions, such as lowerlimb and foot.
-The consistency of higher numbers of upperlimb and lowerlimb suggests that those elements should be transported together (e.g., @Stiner1991), or marrows from upperlimb were commonly utilized.
-It should be noted that the category of axial such as ribs were generally low across different phases, which might be caused by highly fragmentation, butchery practices or dietary preferences.
-A significant change is observed from the frequency of head bones that shows a few in CL6, increased and reach the highest number in CL4, and then decreased since then.
+The result (@fig-deer-portion: A) shows that the meaty portion, including upper forelimb and upper hindlimb, especially upper forelimb, is the most frequent part throughout the cultural layers, followed by the less-meaty portions, such as lower limb and foot.
+The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991), or marrows from upper limb were commonly utilized.
+It should be noted that the frequencies of axial elements, such as ribs, were generally low across different phases, likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
+A different pattern is observed in the frequency of head bones, which were a few in CL6, increased to the highest number in CL4, and then decreased thereafter.
 
 Considering varying frequencies of skeletal parts might be mediated by differential bone density, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation.
-The values of bone survivorship was collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%.
+The values of bone survivorship were collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%.
 The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between skeletal part survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones.

### 2025-02-17 analysis/code/kwl-faunal-layers.R
update code

@@ -64,9 +64,10 @@ fauna_combined_context_all <-
                            TRUE ~ "reptile")) %>%
   mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
   mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
-  mutate(modify = case_when(str_detect(`人為痕跡`, "切")|str_detect(`人為痕跡`, "削")|
-                            str_detect(`人為痕跡`, "砍")~ "cutmarks",
-                            str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒")~ "burn", TRUE ~ "no")) %>%
+  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"),"yes", "no"),
+         fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
+                            str_detect(`人為痕跡`, "折"), "yes", "no"),
+         burning =  ifelse(str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒"), "yes", "no")) %>%
   filter(is.na(refitted)) # for NISP
 
 # taxa counts
@@ -134,7 +135,7 @@ verte_class_barplot_normed <-
   mutate(`%NNISP` = (NNISP/total_per_period)*100) %>%
   ggplot(aes(x = period, y = `%NNISP`))+
   geom_bar(stat = "identity", aes(fill = class), width = 0.7) +
-  labs(y = "%NNISP", x = "cultural layer") +
+  labs(y = "%NNISP", x = "Cultural layer") +
   theme_minimal()
 
 # barplot III: relative abundance of mammals by period
@@ -146,7 +147,7 @@ verte_mammal_barplot <-
   mutate(category = factor(category, levels = c("deer", "boar", "muntjac", "cattle"), ordered = TRUE)) %>%
   ggplot(aes(x = period, fill = category))+
   geom_bar(position = "fill", width = 0.7) +
-  labs(y = "%NISP", x = "cultural layer") +
+  labs(y = "%NISP", x = "Cultural layer") +
   scale_y_continuous(labels = scales::percent) +
[…]

### 2025-02-21 analysis/code/kwl-faunal-layers.R
work on proportion plot for cut marks

@@ -106,7 +106,7 @@ fauna_combined_context <-
   fauna_combined_context_all %>%
   filter(!taxa == "Rattus sp.")
 
-################### Taxonomic abundance ###################
+################### NISP & Taxonomic abundance ###################
 # barplot I: NISP by period (NISP)
 NISP_barplot <-
   fauna_combined_context %>%
@@ -118,40 +118,7 @@ NISP_barplot <-
   theme_minimal() +
   theme(legend.title=element_blank())
 
-# barplot II: classes of vertebrates by period (normed NISP across classes)
-verte_class_barplot_normed <-
-  fauna_combined_context %>%
-  count(period, class) %>% #class
-  drop_na(period, class) %>%
-  mutate(NNISP = case_when(str_detect(class, "mammal") ~ n/205,
-                           str_detect(class, "bird") ~ n/120,
-                           str_detect(class, "fish") ~ n/150,
-                           str_detect(class, "reptile") ~ n/200)) %>%
-  mutate(class = factor(class, levels = c("mammal", "bird", "fish", "reptile"), ordered = TRUE)) %>%
-  group_by(period) %>%
-  summarise(total_per_period = sum(NNISP), across()) %>%
-  mutate(`%NNISP` = (NNISP/total_per_period)*100) %>%
-  ggplot(aes(x = period, y = `%NNISP`))+
-  geom_bar(stat = "identity", aes(fill = class)) +
-  labs(y = "%NNISP", x = NULL) +
-  theme_minimal()
-
-# barplot III: relative abundance of mammals by period
-library(viridis)
-verte_mammal_barplot <-
-  fauna_combined_context %>%
-  filter(class == "mammal") %>%
-  drop_na(period, category) %>% # remove category to get absolute abundance
-  mutate(category = factor(category, levels = c("deer", "boar", "muntjac", "cattle"), ordered = TRUE)) %>%
-  ggplot(aes(x = period, fill = category))+
-  geom_bar(position = "fill") +
-  labs(y = "%NISP", x = NULL) +
-  scale_y_continuous(labels = scales::percent) +
-  theme_minimal() +
[…]

### 2025-02-22 analysis/paper/paper.qmd
work on the result section

@@
 
-To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
+To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six cultural layers (CL6-CL1) spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
 These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009].
@@
 Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers.
-Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method [@Grayson2004; @Lyman2012].
+Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012].
 Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages.
-Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size, and therefore is used throughout this study [@Grayson1984; @Lyman2008].
+Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size [@Grayson1984; @Lyman2008].
 For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004].
-The normalized NISP is calculated by dividing NISP by the average number of skeletal elements of that given taxa.
+The normalized NISP is calculated by dividing NISP by the number of skeletal elements.
 
-Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequency of skeletal elements and surface modifications such as cutmarks of deer bones were examined to assess butchery and transport strategy.
-This can also shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network.
-Skeletal elements was quantified using NISP and classified into groups based on meat yield: meaty (e.g., upper forelimb and upper hindlimb ), mid-meaty (e.g., body parts), and less meaty portions (e.g., skull and lower limb) [@Madrigal2004; @Munro2004].
-Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation compared to othe parts.
+Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal elements and surface modifications of deer bones were examined to assess butchery and transport strategies.
+This can shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network.
+Skeletal elements were quantified using NISP and classified into groups based on meat yield: meaty (e.g., upper forelimb and upper hindlimb), mid-meaty (e.g., body parts), and less meaty portions (e.g., skull and lower limb) [@Madrigal2004; @Munro2004].
+Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation compared to other parts.
 While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002].
 Thus, less meaty elements were considered as secondary source of nutrition in the interpretation.
-Similarly, the degree of fragmentation could also imply the extraction of marrow or grease.
-The degree of fragmentation is calculated by dividing Minimal Number of Elements (MNE) by NISP.
-For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified for a comparison between different temporal sequences.
-In addition, modifications such as cut marks, fracturing, and burning were examined, especially the cut marks associated with the skinning process.
-Skinning processes can be identified by the encircling and straight marks left on the lower limb, such as metapodials, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].
+Similarly, the degree of fragmentation could also imply the extraction of marrow or grease, which is calculated by dividing the Minimal Number of Elements (MNE) by NISP.
+Bone surface modifications relative to human butchering and cooking activities, such as cut marks, fracturing, and burning damage, were recorded, as well as natural marks such as weathering and root etching for understanding taphonomic effects.
+Cut marks on major limb joints, such as shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were further quantified to explore any changes in butchering patterns between temporal sequences.
+Moreover, cut marks on lower limbs relative to skinning processes were examined and compared.
+Those marks are charactersitic of circular and straight incisions left by the series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].
 
@@
   mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
-  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%  mutate(modify = case_when(str_detect(`人為痕跡`, "切")|str_detect(`人為痕跡`, "削")|
-                            str_detect(`人為痕跡`, "砍")~ "cutmarks",
-                            str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒")~ "burn", TRUE ~ "no")) %>%filter(is.na(refitted)) # for NISP
+  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
+  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"),"yes", "no"),
+         fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
+                            str_detect(`人為痕跡`, "折"), "yes", "no"),
+         burning =  ifelse(str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒"), "yes", "no")) %>%
+  filter(is.na(refitted)) # for NISP
 
@@
   mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
-  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Unidentified", tolower(taxa), sep =" "),
+  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                           TRUE ~ taxa)) %>%
@@
 unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Uni")) %>% summarise(sum(NISP))
-deer_NISP <- fauna_combined_taxa %>% 
+deer_fami_NISP <- fauna_combined_taxa %>% 
   filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "Ree")) %>% summarise(sum(NISP))
-deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 3)*100
+deer_fami_NISP_per <- round(deer_fami_NISP/(total_NISP-unident), 3)*100
 ```
 
-The NISP of faunal assemblages from Kiwulan analyzed in this study is `r total_NISP`, where deer family is the majority that accounts for about `r deer_NISP_per`% of the identified bones, followed by boar and cattle as the second and third taxon (@tbl-counts).
-Within the assemblages from the deer family, sika deer is the most common animal, with a few assemblages identified as muntjac and a specimen indicating a possible affiliation with sambar deer.
-Although sika deer was the major animal source at the site, variations in animal use could be observed between different cultural layers across 600 years.
-To highlight general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle.
-Rats (Rattus sp.) were excluded from further analysis as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.
+The faunal assemblage analyszed in this paper from Kiwulan is composed of `r total_NISP` bone fragments based on NISP, which increased over time from CL6, peaked at CL4, and then declined dramatically after CL4 (@fig-NISP-proportion: A).
+A similar trend also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
+Among the funal remains, deer family is the majority that accounts for about `r deer_fami_NISP_per`% of the identified bones, followed by boar and cattle as the second and third taxon (@tbl-counts).
+Within the deer family, sika deer is prevalent, with a few bones identified as muntjac and a specimen indicating a possible affiliation with sambar deer.
+Although sika deer was the major animal source at the site, variations in their quantities and proportions could be observed between temperoal sequences across 600 years.
+To further compare general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle.
+Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.
 
@@
 
-For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in CL4 and birds in CL3 and CL2 (@fig-NISP-proportion: A).
+For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in CL4 and birds in CL3 and CL2 (@fig-NISP-proportion: C).
 Fish and birds are both identified from CL5 to CL3; however, their proportions (less than 10%) are much lower compared to mammals.
 The reptile is represented by one piece of turtle shell found in CL4, which might not be a common food content at the site.
-Within those identifiable mammals, deer is the major source from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by cattle (@fig-NISP-proportion: B).
+Within those identifiable mammals, deer is the major source from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by cattle (@fig-NISP-proportion: D).
 Wild boar was present from CL6 to CL3, and had a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5, and was continuously present until CL2.
-In general, there is an increase in animal diversity in CL4 and CL5, especially in CL4, where boar and fish have slightly higher proportions compared with other layers.
+In general, there is an increase in animal diversity in CL4 and CL5, especially in CL4, where boar and fish have slightly higher proportions compared to those in other layers.
 Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model.
@@
 #| echo: false
-#| fig-cap: "A: Relative abundance of vertebrate classes by cultural layers based on normed NISP.
-The data is derived from Table 1.
-B: Relative abundance of mammals by cultural layers based on NISP.
-CL6 to CL1 prepresent cultural layers from the earliest to the latest layer.
-The NISP for each layer is CL6: 46, CL5: 65, CL4: 156, CL3: 28, CL2: 12, and CL1: 25."
+#| fig-cap: "A: The number of identified specimen per cultural layers.
+The NISP data is derived from Table 1 .
+CL6 represents the earliest layer and CL1 is the youngest.
+B: The weight of the unidentified small bone fragments for each cultural layer.
+C: Relative abundance of vertebrate classes by cultural layers based on normed NISP.
+D: Relative abundance of identified mammals by cultural layers based on NISP."
 
@@
 
-Since deer remain was the major component of faunal assemblage at Kiwulan according to their high quantity and ubiquity, the frequency of skeletal elements was examined to explore any changes in processing decisions of meat portions between different phases.
+Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases.
 Deer bones were grouped into five body portions based on their nutritional value [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), and lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus, and phalanges).
-The result (@fig-deer-portion: A) shows that the meaty portion, including upper forelimb and upper hindlimb, especially upper forelimb, is the most frequent part throughout the cultural layers, followed by the less-meaty portions, such as lower limb and foot.
-The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991), or marrows from upper limb were commonly utilized.
-It should be noted that the frequencies of axial elements, such as ribs, were generally low across different phases, likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
-A different pattern is observed in the frequency of head bones, which were a few in CL6, increased to the highest number in CL4, and then decreased thereafter.
+The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part throughout the cultural layers, followed by upper hindlimb and the less-meaty portions, such as lower limb and foot.
+The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991).
+The frequencies of axial elements, such as ribs, were generally low at Kiwulan, likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
+A different pattern between layers is observed in the frequency of head bones, which were a few in CL6, increased to the highest number in CL4, and then decreased thereafter.
+In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.
 
-Considering varying frequencies of skeletal parts might be mediated by differential bone density, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation.
+Considering the impacted of differential bone density on preservation of skeletal elements, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation.
 The values of bone survivorship were collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%.
-The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between skeletal part survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones.
+The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones.
 
@@
 
+```{r}
+#| label: tbl-deer-modify
+#| echo: false
+#| tbl-cap: "Number and proportion of deer bones with surface modifications at Kiwulan."
+#| tbl-cap-location: bottom
+
+fauna_deer_only <-
+  fauna_combined_context %>%
+  filter(category %in% c("deer") & !is.na(period))
+
+deer_NISP <- fauna_deer_only %>% count(period) %>% rename(NISP = n)
+
+# bones with surface modifications
+burned_deer <- fauna_deer_only %>% filter(burning == "yes") %>% count(period) %>% rename(burned = n)
+cutmark_deer <- fauna_deer_only %>% filter(cutmarks == "yes") %>% count(period) %>% rename(cutmarks = n)
+frac_deer <- fauna_deer_only %>% filter(fractures  == "yes") %>%
+  filter(`部位/名稱` %in% c("脛骨","股骨","肱骨","橈骨","上肢骨","掌骨","掌骨或蹠骨","蹠骨")) %>%
+  count(period) %>% rename(fractures = n)
+list_modif <- list(burned_deer, cutmark_deer, frac_deer, deer_NISP)
+
+# join
+deer_modify <-
+  list_modif %>%
+  reduce(inner_join, by = 'period') %>%
+  mutate(cut_per = round(cutmarks/NISP, 3) * 100, burn_per = round(burned/NISP, 3) * 100,
+         frac_per = round(fractures/NISP, 3) * 100) %>%
+  mutate(burned = paste(burned, paste0("(", burn_per, " %)")),
+         cutmarks = paste(cutmarks, paste0("(", cut_per, " %)")),
+         fractures = paste(fractures, paste0("(", frac_per, " %)"))) %>%
+  rename(`cut marks` = cutmarks) %>%
+  select(-cut_per, -frac_per, -burn_per)
+
+knitr::kable(deer_modify)
+```
+
+Surface modifications, including cut marks, fractures, and burning damage, were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify).
+Burned bones, characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4.
+Cut marks were present on 50% of bones in CL2, 30% of bones in earlier period (CL6 to CL4), and at a lower proportion of 23% in CL3.
+Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching.
+@fig-deer-cutmark shows that at least 50 % of joints of each deer part in CL4 present cut marks, with a higher proportion of 75% at the hip.
+A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4.
+Not all joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult.
+Nevertheless, the overall high proprotion of joints with cut marks suggests intensive proceessing of deer throughout the site's occupation.
+Cut marks on lower limbs can be associeted with skinning and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).
+
+Bone breakage patterns, including spiral fractures, transverse fractures, longitudinal splits, percurssion scars, and chop marks [@Haynes1983; @Gifford2018; @Fisher1995], were observed esepically on long bones across different layers.
+These fractures indicate intentional marrow extraction at Kiwulan.
+CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5.
+This suggests more intensive marrow extraction in later periods.
+Most chop and cut marks were likly made by metal tools according to clean, smooth cut surfaces and finer marks on bones.
+The use of metal tools can be further supported by the finding of iron knives at Kiwulan [@Chen2007].
+In addition to anthropogenic modifications, natural alterations to bones resulting from post-deposition impacts, such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@tbl-deer-modify).
+
+```{r}
+#| label: fig-deer-cutmark
+#| echo: false
+#| fig-cap: "The proportion of cut marks on major limb joints of deer across cultural layers. The method for clasifying and quantifiying limb joints is based on Lyman (2005)."
+
+knitr::include_graphics(here::here("analysis/figures/deer_cutmarks.png"))
+```
+
 # Discussion
@@
 
+frequencies of deer bone elements are consistent with the proportion of joints. 
 Other than quantity, the measure of presence and absence of a given taxon across the sampeld units could give insights into the frequency and distribution consistency of taxa at a settlement scale.
@@
 Small and fragile bones (e.g., fish, rodents) might have low ubiquity due to taphonomic processes.
+
+These post-depositional impacts might have led to an underestimation of cut marks due to the diffculty of surface identification.
 

### 2025-02-24 analysis/code/kwl-faunal-layers.R
edit code and add references

@@ -64,9 +64,9 @@ fauna_combined_context_all <-
                            TRUE ~ "reptile")) %>%
   mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
   mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
-  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"),"yes", "no"),
+  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"), "yes", "no"),
          fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
-                            str_detect(`人為痕跡`, "折"), "yes", "no"),
+                            str_detect(`人為痕跡`, "折")|str_detect(`人為痕跡`, "敲"), "yes", "no"),
          burning =  ifelse(str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒"), "yes", "no")) %>%
   filter(is.na(refitted)) # for NISP
 
@@ -102,9 +102,7 @@ fauna_combined_MNI <-
   filter(!is.na(`部位/左右`)) %>%
   count(`部位/左右`, category, `部位/名稱`)
 
-fauna_combined_context <-
-  fauna_combined_context_all %>%
-  filter(!taxa == "Rattus sp.")
+fauna_combined_context <- fauna_combined_context_all %>% filter(!taxa == "Rattus sp.") # remove rat
 
 ################### NISP & Taxonomic abundance ###################
 # barplot I: NISP by period (NISP)
@@ -336,7 +334,7 @@ weight_head <-
 #ggsave(here::here("analysis", "figures", "talk-fauna-6.png"), h = 2, w =4, units = "in")
 
 ################### Cut marks ###################
-# count cut marks throughout all taxa
+# count cut marks across taxa
 fauna_total_cut <-
   fauna_combined_context %>%
   filter(cutmarks == "yes") %>%
@@ -344,7 +342,7 @@ fauna_total_cut <-
   count() %>%
   drop_na(period, taxa)
 
-# cutmarks on deer bones
+# joints of major limbs for deer
 fauna_deer_joints <- # based on Lyman (2005)
   fauna_deer_portion  %>%
   mutate(part = paste0(`部位/名稱`,`部位/位置`)) %>%
[…]

### 2025-02-25 analysis/paper/paper.qmd
add figure captions and sentences

@@
 
-Surface modifications, including cut marks, fractures, and burning damage, were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify).
-Burned bones, characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4.
+Surface modifications, including cut marks, burning damage, and fractures were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify).
 Cut marks were present on 50% of bones in CL2, 30% of bones in earlier period (CL6 to CL4), and at a lower proportion of 23% in CL3.
 Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching.
+Some joints displyed chop marks along with cut marks (@fig-deer-cutmark), of which the clean, smooth cut surfaces and sharp marks suggest they were likly made by metal tools [@Greenfield1999].
 @fig-deer-cutmark shows that at least 50 % of joints of each deer part in CL4 present cut marks, with a higher proportion of 75% at the hip.
 A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4.
-Not all joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult.
+Not all types of joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult.
 Nevertheless, the overall high proprotion of joints with cut marks suggests intensive proceessing of deer throughout the site's occupation.
-Cut marks on lower limbs can be associeted with skinning and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).
+In addition, cut marks on lower limbs can be associeted with skinning (@fig-deer-cutmark: e) and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).
 
-Bone breakage patterns, including spiral fractures, transverse fractures, longitudinal splits, percurssion scars, and chop marks [@Haynes1983; @Gifford2018; @Fisher1995], were observed esepically on long bones across different layers.
+Burned bones, characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4.
+Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed on long bones across different layers (@fig-deer-fracture).
 These fractures indicate intentional marrow extraction at Kiwulan.
@@
 This suggests more intensive marrow extraction in later periods.
-Most chop and cut marks were likly made by metal tools according to clean, smooth cut surfaces and finer marks on bones.
-The use of metal tools can be further supported by the finding of iron knives at Kiwulan [@Chen2007].
-In addition to anthropogenic modifications, natural alterations to bones resulting from post-deposition impacts, such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@tbl-deer-modify).
+Chewing marks were also observed on some bones that show a series of shallow and smotth features (@fig-deer-fracture: e).
+In addition to anthropogenic modifications, natural alterations to bones resulting from post-deposition impacts, such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-deer-fracture: c,d; @tbl-deer-modify).
 
@@
 knitr::include_graphics(here::here("analysis/figures/deer_cutmarks.png"))
+```
+
+```{r}
+#| label: fig-cutmark-img
+#| echo: false
+#| fig-cap: "Examples of sika deer bones with cut marks, chop surfaces, and burning damage.
+Cut marks are indicated with red squares, and chop surfaces are indicated with black arrows.
+Bone fragments are (a) proximal radius with cut marks and a chop surface, (b) distal humerus with cut marks, (c) acetabulum with chop surfaces, (d) charred astragulus, (e) metatarsal with cut marks."
+
+knitr::include_graphics(here::here("analysis/figures/deer_cutmark_img.png"))
+```
+
+```{r}
+#| label: fig-fracture-img
+#| echo: false
+#| fig-cap: "Examples of sika deer bones with fractures, natural modifications, and chewing marks.
+Bone fragments are (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing spiral fracture, (c) distal radius with root etching, (d) femur shaft showing transverse fracture on the bottom, a percurssion scar in the middle with longitudinal split, and weathering, (e) scapula with chewing marks."
+
+knitr::include_graphics(here::here("analysis/figures/deer_fracture_img.png"))
 ```

### 2025-02-25 analysis/paper/paper.qmd
made some edits

@@
 total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
-unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Uni")) %>% summarise(sum(NISP))
+unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "In")) %>% summarise(sum(NISP))
 deer_fami_NISP <- fauna_combined_taxa %>% 
@@
 The faunal assemblage analyszed in this paper from Kiwulan is composed of `r total_NISP` bone fragments based on NISP, which increased over time from CL6, peaked at CL4, and then declined dramatically after CL4 (@fig-NISP-proportion: A).
-A similar trend also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
-Among the funal remains, deer family is the majority that accounts for about `r deer_fami_NISP_per`% of the identified bones, followed by boar and cattle as the second and third taxon (@tbl-counts).
+A similar trend was also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
+Among the faunal remains, the deer family is the majority accounting for about `r deer_fami_NISP_per`% of the bones identified to class level, followed by boar and cattle as the second and third taxon (@tbl-counts).
 Within the deer family, sika deer is prevalent, with a few bones identified as muntjac and a specimen indicating a possible affiliation with sambar deer.
@@
 Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases.
-Deer bones were grouped into five body portions based on their nutritional value [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), and lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus, and phalanges).
-The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part throughout the cultural layers, followed by upper hindlimb and the less-meaty portions, such as lower limb and foot.
+Deer bones were grouped into six body portions based on their nutritional values [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges).
+The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot.
 The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991).
-The frequencies of axial elements, such as ribs, were generally low at Kiwulan, likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
+The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
 A different pattern between layers is observed in the frequency of head bones, which were a few in CL6, increased to the highest number in CL4, and then decreased thereafter.
@@
 Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching.
-Some joints displyed chop marks along with cut marks (@fig-deer-cutmark), of which the clean, smooth cut surfaces and sharp marks suggest they were likly made by metal tools [@Greenfield1999].
 @fig-deer-cutmark shows that at least 50 % of joints of each deer part in CL4 present cut marks, with a higher proportion of 75% at the hip.
@@
 Nevertheless, the overall high proprotion of joints with cut marks suggests intensive proceessing of deer throughout the site's occupation.
-In addition, cut marks on lower limbs can be associeted with skinning (@fig-deer-cutmark: e) and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).
+In addition, cut marks on lower limbs can be associeted with skinning (@fig-cutmark-img: e) and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).
+Some joints displayed clean, smooth cut surfaces from chopping (@fig-cutmark-img: a, c), along with narrow V-shaped cut marks observed on most bones, suggesting they were likely made with metal tools [@Greenfield1999].
 
-Burned bones, characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4.
-Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed on long bones across different layers (@fig-deer-fracture).
+Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4.
+Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed frequently on long bones across different layers (@fig-fracture-img).
 These fractures indicate intentional marrow extraction at Kiwulan.
 CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5.
-This suggests more intensive marrow extraction in later periods.
-Chewing marks were also observed on some bones that show a series of shallow and smotth features (@fig-deer-fracture: e).
-In addition to anthropogenic modifications, natural alterations to bones resulting from post-deposition impacts, such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-deer-fracture: c,d; @tbl-deer-modify).
+Chewing marks were observed on a few bones that show a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023].
+In addition, natural alterations to bones resulting from post-deposition impacts were also observed , such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify).
 
@@
 #| fig-cap: "Examples of sika deer bones with cut marks, chop surfaces, and burning damage.
-Cut marks are indicated with red squares, and chop surfaces are indicated with black arrows.
+Cut marks are indicated by red squares, and chop surfaces are indicated by black arrows.
 Bone fragments are (a) proximal radius with cut marks and a chop surface, (b) distal humerus with cut marks, (c) acetabulum with chop surfaces, (d) charred astragulus, (e) metatarsal with cut marks."
@@
 #| fig-cap: "Examples of sika deer bones with fractures, natural modifications, and chewing marks.
-Bone fragments are (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing spiral fracture, (c) distal radius with root etching, (d) femur shaft showing transverse fracture on the bottom, a percurssion scar in the middle with longitudinal split, and weathering, (e) scapula with chewing marks."
+Bone fragments are (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing spiral fracture, (c) distal radius showing tranverse fracture and root etching, (d) femur shaft showing longitudinal split, weathering, and a percurssion scar indicated by an arrow, (e) scapula with chewing marks."
 

### 2025-02-27 analysis/paper/paper.qmd
minor changes

@@
 Although the new exchange systems or colonial institutions introduced by Europeans could change local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002], indigenous groups did not just accept European colonization passively.
-Recent discussions on colonial interactions also suggest the active role of indigenous groups in responding to the arrival of Europeans and presenting resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998].
+Recent discussions on colonial interactions also suggest the active role of indigenous groups in responding to the arrival of Europeans and presenting resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998; @Noe2023].
 
@@
 Can we identify shifts in these practices in response to colonial economic demands?
-To address these questions, I will focus on subsistence practices because the choices made in the practices can reveal the strategies Indigenous people used to respond to European exploitation with associated changes in their environment.
-I will primarily examine the zooarchaeological assemblages from Kiwulan for subsistence practices, coupled with botanical evidence as a supplementary evidence.
-Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as intercultural contact or European colonization (DeFrance, 2009).
-Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.
+To address these questions, I will focus on subsistence practices because the choices made in daily practices can reveal the strategies Indigenous people used to respond to resource explortations by foreign presence.
+I will examine the zooarchaeological assemblages from the upper culture layer at Kiwulan, a large settlement dating to 14-19th centuries.
+Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as European colonization or intercultural contact [@Defrance2009].
+Through a zooarchaeological perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.
 
@@
 Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences.
-Deer antlers that were not attached to cranial bones were excluded since they might not directly related to dietary practices.
+Deer antlers were excluded since they were less likely related to dietary practices.
 To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied.
@@
 
-Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal elements and surface modifications of deer bones were examined to assess butchery and transport strategies.
+Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies.
 This can shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network.
-Skeletal elements were quantified using NISP and classified into groups based on meat yield: meaty (e.g., upper forelimb and upper hindlimb), mid-meaty (e.g., body parts), and less meaty portions (e.g., skull and lower limb) [@Madrigal2004; @Munro2004].
-Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation compared to other parts.
+Skeletal parts were quantified using NISP and classified into groups based on meat yield and portions: meaty (e.g., upperlimb), mid-meaty (e.g., vertebrae), and less meaty parts (e.g., skull and lower limb) [@Madrigal2004; @Munro2004].
+Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation and quantities compared to other parts.
 While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002].
@@
 The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
-A different pattern between layers is observed in the frequency of head bones, which were a few in CL6, increased to the highest number in CL4, and then decreased thereafter.
+A different pattern between layers is observed in the frequency of cranial bones, which were present in CL5, increased to the highest number in CL4, and then decreased thereafter.
+None of cranial bones were found in CL6 and CL2.
 In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.
@@
   count(period) %>% rename(fractures = n)
-list_modif <- list(burned_deer, cutmark_deer, frac_deer, deer_NISP)
+post_deposit <-fauna_deer_only %>% filter(!is.na(`保存狀態`)) %>% count(period) %>%
+  rename(`post-deposition` = n)
+
+list_modif <- list(burned_deer, cutmark_deer, frac_deer, post_deposit, deer_NISP)
 
@@
   mutate(cut_per = round(cutmarks/NISP, 3) * 100, burn_per = round(burned/NISP, 3) * 100,
-         frac_per = round(fractures/NISP, 3) * 100) %>%
+         frac_per = round(fractures/NISP, 3) * 100, post_per= round(`post-deposition`/NISP, 3) * 100) %>%
   mutate(burned = paste(burned, paste0("(", burn_per, " %)")),
          cutmarks = paste(cutmarks, paste0("(", cut_per, " %)")),
-         fractures = paste(fractures, paste0("(", frac_per, " %)"))) %>%
+         fractures = paste(fractures, paste0("(", frac_per, " %)")),
+         `post-deposition` = paste(`post-deposition`, paste0("(", post_per, " %)"))) %>%
   rename(`cut marks` = cutmarks) %>%
-  select(-cut_per, -frac_per, -burn_per)
+  select(-cut_per, -frac_per, -burn_per, -post_per)
 
@@
 CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5.
-Chewing marks were observed on a few bones that show a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023].
+Chewing marks were observed on a few bones that show a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023], suggesting carnivore gnawing.
 In addition, natural alterations to bones resulting from post-deposition impacts were also observed , such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify).
@@
 #| echo: false
-#| fig-cap: "Examples of sika deer bones with cut marks, chop surfaces, and burning damage.
+#| fig-cap: "Examples of deer bones with cut marks, chop surfaces, and burning damage.
 Cut marks are indicated by red squares, and chop surfaces are indicated by black arrows.
@@
 #| echo: false
-#| fig-cap: "Examples of sika deer bones with fractures, natural modifications, and chewing marks.
+#| fig-cap: "Examples of deer bones with fractures, natural modifications, and chewing marks.
 Bone fragments are (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing spiral fracture, (c) distal radius showing tranverse fracture and root etching, (d) femur shaft showing longitudinal split, weathering, and a percurssion scar indicated by an arrow, (e) scapula with chewing marks."
@@
 # Discussion
+<!--summary of key results-->
+The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption since CL6 until CL2, and then shifted to cattle in CL1.
+Despite deer had been a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals shows changes over time.
+Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences relative to foreign presences for the following discussion.
+The sequences include the pre-European period (CL6, CL5), the European period (CL4), post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1).
+It should be noted that the European period does not necessary means a solid European colonial control over the Kiwulan, but a broad impacts from European occupation and activities in northern Taiwan, including the involvement into the long-distance trade network, part of the tribute system, and experiences of occasional attacks in cases in Yilan.
+
+Faunal remains recovered from the layer associated with the European presence reveals the highest quantity, followed by the pre-European period.
+During the European period, in addition to the predominance of deer, an increased proportion of boar among faunal remains suggests a slightly greater reliance on boar as a secondary meat source compared to other time periods, along with a broader range of medium and small animals, such as muntjac, birds, fish, and reptiles.
+A similar pattern was observed in the ubiquity of animals, which shows the highest ubiquity of deer, boar and munjact remains during the European period.
+Interestingly, deer bones from the pre-European period, specifically CL5, presented the same ubiquity as in the European period but in only half the quantity.
+This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but with double the quantity of deer consumed during the European period.
+During the Chinese period, the low quantity of deer in CL2 and the predomination of cattle in CL1 suggest a sudden shift in dietary practices after the Chinese arrival, indicating a Chinese influence and the decline of indigeonus population in this region.
+
+The skeletal parts of deer demonstrate a different
+
+The cutmarks found on bone portions related to skinning correspond to the description of deerskin as a commodity during the maritime era (Chen, 2004; Cheng, 2017; Höllmann, 1991).
+Interestingly, a similar butchering pattern related to skinning suggests that hides were produced at the same rate during the European period, despite the high demand for deerskin.
+However, the dramatic decrease in deer bones during the Chinese period may indicate overhunting that persisted for centuries or a decline in the Indigenous population (Zhan, 1995).
+Moreover, the double increase in pig bones after the European arrival may suggest an increased reliance on pigs.
+The shape of the third molars of pigs indicates that most were domesticated.
+This implies that pigs played a more significant role in the diet after Europeans arrived.
+Eventually, cows replaced pigs during the Chinese period, which might be a Chinese influence.
+
+By contextralizing the sampled faunal remains from the major residential area at Kiwulan   into the time periods with foreign contact, we can observe the
+
+<!--population dynamics and pottery assembalges-->
+
+<!--environmental changes in general and taphonomic issues-->
+
+<!--resilence-->
 
@@
 
+The European maritime cultural exchange to some extent altered the material culture of the Indigenous peoples in northeastern Taiwan by adopting foreign goods or customs into their daily lives.
+However, a sign of indigenous resilience in the face of European influences and resource exploitation may be seen in the continuing tradition of pottery production and deer consumption with few changes, even though changes were observed through the uneven distribution of prestige items, such as glass beads and porcelains, across different households in the living area and burials (Wang and Marwick, 2020a; 2021).
+
+The continuous practices of ceramic making and deer consumption by the Indigenous peoples before and after the European arrival can be viewed as a way of negotiating their identities and maintaining their traditions.
+Everyday practices are essential in connecting history and social memories, which helps maintain tradition in a setting of frequent cultural exchange or asymmetrical relations (Silliman, 2009).
+Moreover, ceramics could be an item that conveys social meaning through its style, such as decorations, and be used to show group identity (Croucher and Wynne-Jones, 2006; Pikirayi, 2007).
+It is worth noting that the pottery found at Kiwulan, with its typical round body and incised geometric motifs, is also commonly found across many sites during this period in northeastern Taiwan (Taiwan History, 2005).
+Similar geometric motifs and vessel shapes in the region might suggest the communication of Indigenous identity within the context of ongoing cross-cultural interactions with outsiders, including Europeans and Chinese.
+Furthermore, despite the availability of foreign items, such as iron pans that could have potentially replaced local pottery for cooking purposes, local ceramics have been used continuously for over 6,000 years (Chen, 2007).
+This emphasizes the significant role of pottery-making in the lives of Indigenous communities.
+
+By situating the findings within a pericolonial context (Acabado, 2017), locally made pottery and faunal remains reveal cultural persistence and resilience among Indigenous peoples in northeast Taiwan.
+The interaction between Europeans and Indigenous peoples in the region appears to have been based on negotiation and bartering (Torrence, 2000).
+For instance, the Dutch reduced and eventually waived annual tribute in 1648 in response to the resistance to paying tribute by most Indigenous settlements in Yilan (Blussé and Everts, 2000, pp.
+534, 536, 558; Kang, 2012).
+The continuous consumption of deer and consistent skinning patterns before and after the arrival of Europeans indicate a more adaptable tribute payment system, whereby the Indigenous peoples did not need to hunt more deer to meet European demand.
+However, it is important to note that there may have been a gradual decline in deer numbers since the post-European period, which is supported by the increased number of pigs found in the same period.
+This could suggest cultural resilience through dietary changes to adapt to deer overhunting.
+A more comprehensive investigation across different Indigenous settlements in northeastern Taiwan could provide a clearer understanding of overall dietary patterns.
+
 # Conclusion
+
+Through contextralizing the faunal remains into time periods relaitve to different foreign presences, including the Europeans in the 17th century corresponding to CL4, and the Chinese immigrants since the late 19th century corresponding to CL1, the resutls reveals how indigenous people respond to the foreign influences on their economic aspect.
+
 Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period.

### 2025-02-27 analysis/code/kwl-faunal-layers.R

@@ -46,8 +46,8 @@ fauna_H_sam <-
 # combine and tidy data from midden and cultural layers
 fauna_combined_context_all <-
   rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
-  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
-  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
+  filter(!`部位/名稱` %in% c("角","犄角","角?", "角基部")) %>%  #not associated with diet
+  #filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
   mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
   group_by(taxa) %>%
   mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
@@ -108,10 +108,11 @@ fauna_combined_context <- fauna_combined_context_all %>% filter(!taxa == "Rattus
 # barplot I: NISP by period (NISP)
 NISP_barplot <-
   fauna_combined_context %>%
-  count (period) %>%
+  count (period, animal) %>%
   drop_na() %>%
-  ggplot(aes(x = period, y = n)) +
-  geom_col(width = 0.6) + #position_dodge2(preserve = "single")
+  ggplot(aes(x = period, y = n, fill = animal)) +
+  geom_bar(stat = "identity", #  geom_col(width = 0.6) +  #
+           position = position_dodge2(preserve = "single"), widtg = 0.6) +
   labs(y = "NISP", x = NULL) +
   theme_minimal() +
   theme(legend.title=element_blank())
@@ -267,8 +268,8 @@ fauna_deer_portion <-
     `部位/名稱` %in% c("肩胛骨","肱骨","尺骨","橈骨") ~ "upper forelimb", #upper limb, "上肢骨"
     `部位/名稱` %in% c("脛骨","髖骨","股骨") ~ "upper hindlimb",
     `部位/名稱` %in% c("掌骨","掌骨或蹠骨","蹠骨","跗骨","腕骨","跟骨","astragalus") ~ "lower limb",
[…]

### 2025-02-27 analysis/code/kwl-faunal-layers.R
code edits

@@ -46,8 +46,8 @@ fauna_H_sam <-
 # combine and tidy data from midden and cultural layers
 fauna_combined_context_all <-
   rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
-  filter(!`部位/名稱` %in% c("角","犄角","角?", "角基部")) %>%  #not associated with diet
-  #filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
+  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
+  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
   mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
   group_by(taxa) %>%
   mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
@@ -108,11 +108,11 @@ fauna_combined_context <- fauna_combined_context_all %>% filter(!taxa == "Rattus
 # barplot I: NISP by period (NISP)
 NISP_barplot <-
   fauna_combined_context %>%
-  count (period, animal) %>%
+  count (period) %>%
   drop_na() %>%
-  ggplot(aes(x = period, y = n, fill = animal)) +
-  geom_bar(stat = "identity", #  geom_col(width = 0.6) +  #
-           position = position_dodge2(preserve = "single"), widtg = 0.6) +
+  ggplot(aes(x = period, y = n)) +
+  geom_col(width = 0.6) + #geom_bar(stat = "identity", #  geom_col(width = 0.6) +  #
+           #position = position_dodge2(preserve = "single"), widtg = 0.6) +
   labs(y = "NISP", x = NULL) +
   theme_minimal() +
   theme(legend.title=element_blank())
@@ -268,7 +268,7 @@ fauna_deer_portion <-
     `部位/名稱` %in% c("肩胛骨","肱骨","尺骨","橈骨") ~ "upper forelimb", #upper limb, "上肢骨"
     `部位/名稱` %in% c("脛骨","髖骨","股骨") ~ "upper hindlimb",
     `部位/名稱` %in% c("掌骨","掌骨或蹠骨","蹠骨","跗骨","腕骨","跟骨","astragalus") ~ "lower limb",
[…]

### 2025-02-27 analysis/paper/paper.qmd
work on discussion section

@@
 Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences.
-Deer antlers were excluded since they were less likely related to dietary practices.
+Deer antlers without attaching to crania were excluded since they were less likely related to dietary practices.
 To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied.
@@
   mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
-  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"),"yes", "no"),
+  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"), "yes", "no"),
          fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
-                            str_detect(`人為痕跡`, "折"), "yes", "no"),
+                            str_detect(`人為痕跡`, "折")|str_detect(`人為痕跡`, "敲"), "yes", "no"),
          burning =  ifelse(str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒"), "yes", "no")) %>%
@@
 
+# calculate MNE
+deer_MNE <-
+  fauna_deer_only  %>%
+  filter(!is.na(`部位/左右`)&is.na(overlap)) %>% # remove overlapped fragments
+  count(period,`部位/左右`,`部位/名稱`,`部位/位置`) %>%
+  group_by(period) %>%
+  summarise(MNE = sum(n), across())
+
+deer_NISP <- fauna_deer_only %>% count(period) %>% rename(NISP = n)
+frag_index <- left_join(deer_MNE, deer_NISP) %>% mutate(index = MNE/NISP) %>% distinct(period, index) %>% 
+  mutate(index = round(index, 2))
+frag_CL3_CL4 <- c(frag_index$index[2], frag_index$index[3])
+
 knitr::kable(fauna_ubiquity)
@@
 The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
-A different pattern between layers is observed in the frequency of cranial bones, which were present in CL5, increased to the highest number in CL4, and then decreased thereafter.
-None of cranial bones were found in CL6 and CL2.
+A different pattern between layers is observed in the frequency of cranial bones, for which a few present in CL6, increased to the highest number in CL4, and then decreased thereafter.
 In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.
@@
 The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones.
+For the degree of bone fragmentation, CL3 shows higher fragmentation (index = `r frag_CL3_CL4[1]`), while bones in CL4 is less fragmented (index = `r frag_CL3_CL4[2]`).
+However, the fragmentation rates across all layers are not significantly differed.
 
@@
 #| tbl-cap-location: bottom
+#| tbl-colwidths: [10,20,20,20,20,10]
 
@@
 This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but with double the quantity of deer consumed during the European period.
-During the Chinese period, the low quantity of deer in CL2 and the predomination of cattle in CL1 suggest a sudden shift in dietary practices after the Chinese arrival, indicating a Chinese influence and the decline of indigeonus population in this region.
+During the Chinese period, the low quantity of deer in CL2 and the predominate of cattle in CL1 suggest a sudden shift in dietary practices after the Chinese arrival, indicating a Chinese influence and the decline of indigenous population in this region.
 
-The skeletal parts of deer demonstrate a different
+The skeletal part frequencies of deer demonstrate more complete carcasses processed on site during the European period and pre-European period compared to post-European period and Chinese period.
+The whole carcasses processing and a substantial rise in head parts (eg.
+crania, mandibles) in the European period align with its high quantity of deer, suggesting more deer were brought back intact to the settlement rather than selectively butchered in the field.
+On the other hand, the low frequency of head parts in the earliest layer and Chinese period might indicate deer could be processed at hunting location with mostly meaty (eg.
+upperlimb) and mid-meaty portions (eg.
+ribs) transported back.
+Interestingly, front quarters (eg.
+humeri, radius) are more frequently found compared to forequarters (eg.
+femur, tibia) among the meaty portions.
+In addition to culinary preferences or differential processing, a likely explanation for this is that femur and tibia were heavily fragmented due to marrow extraction for their high yields [@Madrigal2002], which could lead to underrepresented resulting from unidentifiable small fragments.
 

### 2025-03-01 analysis/code/kwl-faunal-layers.R
udpate code

@@ -54,19 +54,20 @@ fauna_combined_context_all <-
   ungroup() %>%
   mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
   mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
-                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "fish") ~ "fish",
+                                str_detect(animal, "e deer")~ "deer", str_detect(animal, "pig")|
+                                str_detect(animal, "boar") ~ "suid", str_detect(animal, "fish") ~ "fish",
                               str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
                               animal == "aves" ~ "bird", TRUE ~ animal)) %>%
   mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
-                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
-                           str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
+                             str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
+                             str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
                            str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
                            TRUE ~ "reptile")) %>%
   mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
-  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
+  mutate(category = fct_relevel(category, "deer","suid","muntjac","cattle","rat","bird","fish","turtle")) %>%
   mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"), "yes", "no"),
          fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
[…]

### 2025-03-01 analysis/paper/paper.qmd
work on discussion section

@@
   mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
-                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "fish") ~ "fish",
+                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "pig")|
+                                str_detect(animal, "boar") ~ "suid", str_detect(animal, "fish") ~ "fish",
                               str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
@@
   mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
-  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
+  mutate(category = fct_relevel(category, "deer","suid","muntjac","cattle","rat","bird","fish","turtle")) %>%
   mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"), "yes", "no"),
@@
                                    "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
-                                   str_detect(taxa, "Ree") ~ "Muntjac", str_detect(taxa, "Sus") ~ "Boar", 
+                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Wild boar",
                                    str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~ 
@@
                                    "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
-                                   str_detect(taxa, "cervid") ~ "Deer family")) %>%
+                                   str_detect(taxa, "cervid") ~ "Deer", taxa == "Sus scrofa domestica" ~ 
+                                   "Domestic pig", taxa == "Aves" ~ "Birds")) %>%
   mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
@@
 A similar trend was also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
-Among the faunal remains, the deer family is the majority accounting for about `r deer_fami_NISP_per`% of the bones identified to class level, followed by boar and cattle as the second and third taxon (@tbl-counts).
+Among the faunal remains, the deer family is the majority accounting for about `r deer_fami_NISP_per`% of the bones identified to class level, followed by suid and cattle as the second and third taxon (@tbl-counts).
 Within the deer family, sika deer is prevalent, with a few bones identified as muntjac and a specimen indicating a possible affiliation with sambar deer.
-Although sika deer was the major animal source at the site, variations in their quantities and proportions could be observed between temperoal sequences across 600 years.
+Although sika deer was the major animal source at the site, variations in their quantities and proportions could be observed between temporal sequences across 600 years.
 To further compare general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle.
+Similarly, wild boar and domestic pig (n = 2, from CL4 and CL6) were grouped into the category of suid.
 Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.
@@
 CL4 <- mammal_count %>% filter(period == "CL4") 
-CL4_boar <- round(CL4$n[2]/CL4$n[4], 3)*100
+CL4_boar <- round(CL4$n[3]/CL4$n[4], 3)*100
 
@@
 
-boar_reg <- summary(lm(boar ~ total, data = wider_mammal_count))
+boar_reg <- summary(lm(suid ~ total, data = wider_mammal_count))
 boar_lm <- c(round(boar_reg$adj.r.squared, 2), sprintf('%.3f', boar_reg$coefficients[,4][2]))
@@
 Within those identifiable mammals, deer is the major source from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by cattle (@fig-NISP-proportion: D).
-Wild boar was present from CL6 to CL3, and had a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5, and was continuously present until CL2.
-In general, there is an increase in animal diversity in CL4 and CL5, especially in CL4, where boar and fish have slightly higher proportions compared to those in other layers.
+Suid was present from CL6 to CL3, and had a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5, and was continuously present until CL2.
+In general, there is an increase in animal diversity in CL4 and CL5, especially in CL4, where suid and fish have slightly higher proportions compared to those in other layers.
 Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model.
@@
 
+# for deer
+fauna_deer_only <-
+  fauna_combined_context %>%
+  filter(category %in% c("deer") & !is.na(period))
+
+deer_NISP <- fauna_deer_only %>% count(period) %>% rename(NISP = n)
+
 # calculate MNE
@@
 @tbl-ubiquity shows deer had been the most frequent animal from CL6 to CL2, with the highest ubiquity in CL4 and CL5 (`r CL4_ubi_deer`), but disappeared in CL1.
-The second and third frequent animal are boar and muntjac.
-Although both animals reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for boar, and `r CL4_ubi_munt` for muntjac), their presence patterns are slightly different.
-Boar was utilized as early as deer in CL6, declined in CL3, and disappeared in CL2, whereas muntjac showed a cultural lag, appearing one phase later than boar.
+The second and third frequent animal are suid and muntjac.
+Although both animals reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for suid, and `r CL4_ubi_munt` for muntjac), their presence patterns are slightly different.
+Suid was utilized as early as deer in CL6, declined in CL3, and disappeared in CL2, whereas muntjac showed a cultural lag, appearing one phase later than suid.
 Cattle had a low ubiquity in CL5 and high ubiquity in CL6.
 Small animals (bird, fish, and turtle) had much lower ubiquity than large animals and they had a slightly increase in ubiquity in CL3.
-The result of animal ubiquity correspond with the result of their quantity (NISP), suggesting that deer is the most common resource at the site, with an increase utilization of boar and muntjac in CL4.
+The result of animal ubiquity correspond with the result of their quantity (NISP), suggesting that deer is the most common resource at the site, with an increase utilization of suid and muntjac in CL4.
 
@@
 For the degree of bone fragmentation, CL3 shows higher fragmentation (index = `r frag_CL3_CL4[1]`), while bones in CL4 is less fragmented (index = `r frag_CL3_CL4[2]`).
-However, the fragmentation rates across all layers are not significantly differed.
+However, the fragmentation rates are not significantly differed between layers.
 
@@
 #| tbl-colwidths: [10,20,20,20,20,10]
-
-fauna_deer_only <-
-  fauna_combined_context %>%
-  filter(category %in% c("deer") & !is.na(period))
-
-deer_NISP <- fauna_deer_only %>% count(period) %>% rename(NISP = n)
 
@@
 # Discussion
-<!--summary of key results-->
+<!--results and chronology -->
 The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption since CL6 until CL2, and then shifted to cattle in CL1.
-Despite deer had been a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals shows changes over time.
+Despite deer had been a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals shows changes that could reflect the indigenous reactions to foreign presence.
 Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences relative to foreign presences for the following discussion.
 The sequences include the pre-European period (CL6, CL5), the European period (CL4), post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1).
-It should be noted that the European period does not necessary means a solid European colonial control over the Kiwulan, but a broad impacts from European occupation and activities in northern Taiwan, including the involvement into the long-distance trade network, part of the tribute system, and experiences of occasional attacks in cases in Yilan.
+It should be noted that the European period does not necessary means a solid European colonial control over the Kiwulan, but a broad impacts from European occupation and economic activities on indigenous societies in Yilan, such as getting involvement into the long-distance trade network, being part of the tribute system, or experiences of occasional European attacks.
 
-Faunal remains recovered from the layer associated with the European presence reveals the highest quantity, followed by the pre-European period.
-During the European period, in addition to the predominance of deer, an increased proportion of boar among faunal remains suggests a slightly greater reliance on boar as a secondary meat source compared to other time periods, along with a broader range of medium and small animals, such as muntjac, birds, fish, and reptiles.
-A similar pattern was observed in the ubiquity of animals, which shows the highest ubiquity of deer, boar and munjact remains during the European period.
+<!--summary of overall observations-->
+Faunal remains recovered from the layer associated with the European presence reveal the highest quantity, followed by the pre-European period.
+During the European period, in addition to the predominance of deer, an increased proportion of suid among faunal remains suggests a slightly greater reliance on suid as a secondary meat source compared to other time periods, along with a broader range of medium and small animals, such as muntjac, birds, fish, and reptiles.
+A similar pattern was observed in the ubiquity of animals, which shows the highest ubiquity of deer, suid, and munjact remains during the European period.
 Interestingly, deer bones from the pre-European period, specifically CL5, presented the same ubiquity as in the European period but in only half the quantity.
-This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but with double the quantity of deer consumed during the European period.
-During the Chinese period, the low quantity of deer in CL2 and the predominate of cattle in CL1 suggest a sudden shift in dietary practices after the Chinese arrival, indicating a Chinese influence and the decline of indigenous population in this region.
+This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but consumption of deer was intensified during the European period.
+During the Chinese period, the low quantity of deer in CL2 and the predominate of cattle in CL1 suggest a sudden shift in dietary practices after the Chinese arrival, indicating a Chinese influence and the decline of Indigenous population in this region.
 
-The skeletal part frequencies of deer demonstrate more complete carcasses processed on site during the European period and pre-European period compared to post-European period and Chinese period.
-The whole carcasses processing and a substantial rise in head parts (eg.
-crania, mandibles) in the European period align with its high quantity of deer, suggesting more deer were brought back intact to the settlement rather than selectively butchered in the field.
-On the other hand, the low frequency of head parts in the earliest layer and Chinese period might indicate deer could be processed at hunting location with mostly meaty (eg.
-upperlimb) and mid-meaty portions (eg.
-ribs) transported back.
-Interestingly, front quarters (eg.
-humeri, radius) are more frequently found compared to forequarters (eg.
-femur, tibia) among the meaty portions.
-In addition to culinary preferences or differential processing, a likely explanation for this is that femur and tibia were heavily fragmented due to marrow extraction for their high yields [@Madrigal2002], which could lead to underrepresented resulting from unidentifiable small fragments.
+<!--key results of skeletal part frequencies and butchery patterns-->
+The skeletal part frequencies and butchery patterns of deer demonstrate more complete carcasses processed within the settlement during the European period and pre-European period compared to the post-European period and Chinese period.
+Although a more complete and evenly distributed deer parts with cut marks were found in both the European period and before European arrival, the substantial rise in head parts (e.g., crania and mandibles) and overall high quantity of bones in the European period suggests more deer were brought back intact to the settlement.
+The low frequency of head parts in the earliest layer and Chinese period might indicate deer could be mostly processed at the hunting location with only meaty (e.g., upper limb) and mid-meaty portions (e.g., ribs) selectively butchered and transported back.
+Interestingly, among the meaty portions throughout all periods, front quarters (e.g., humeri, radius) are more frequently found compared to forequarters (e.g., femur, tibia).
+One likely explanation is that femur and tibia might be heavily fragmented due to being broken for their high marrow yields [@Madrigal2002], leading to underestimation due to difficult identification.
+The behavior of marrow extraction can be supported by around one-third long bones being fractured and a moderate degree of fragmentation on average at Kiwulan.
+Another possibility is forequarters could be traded since deer legs as commodity were mentioned in European records [@Blusse2006, pp.220-224], but unfortunately, there was no detail about which portion of legs were normally traded.
 
-The cutmarks found on bone portions related to skinning correspond to the description of deerskin as a commodity during the maritime era (Chen, 2004; Cheng, 2017; Höllmann, 1991).
-Interestingly, a similar butchering pattern related to skinning suggests that hides were produced at the same rate during the European period, despite the high demand for deerskin.
-However, the dramatic decrease in deer bones during the Chinese period may indicate overhunting that persisted for centuries or a decline in the Indigenous population (Zhan, 1995).
-Moreover, the double increase in pig bones after the European arrival may suggest an increased reliance on pigs.
-The shape of the third molars of pigs indicates that most were domesticated.
+<!--skinning-->
+The cut marks found on the lower limbs of deer at Kiwulan should be associated with the traces left by skinning processes, which confirms that deer skins were removed and likely used for trading with external traders.
+A large quantity of foreign goods, such as ornaments, porcelains, and metal items, supports frequent trading activities between Kiwulan residents and outsiders [@Wang2020ornament; @Chen2007].
+Deer hides and venison, were important commodities in the East Asian trade network since the mid-15th century, connecting Han Chinese, Japanese, and Indigenous people, especially in western Taiwan [@Cheng2017; @Hollmann1991].
+This deer trade was later monopolized by the Europeans in the 17th century, and deer products were largely exported to Japan.
+The evidence of the removal of deer skins at Kiwulan suggests that not only western Taiwan but also northeastern Taiwan involved and impacted by the long-distance trade.
+Deer bones with skinning cuts did not show a substantial increase in the European period; however, the disproportionate increase in head parts hints at a likelihood of brain tanning when preparing hides.
+Brain tanning is a traditional method for softening and preserving hides commonly used in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016].
+The present-day Bunun group in the central highlands of Taiwan has revived hide tanning based on oral history, primarily using peanuts or pig lard, with a few examples using animal brains [@Li1999, pp.
+48; @Xu1994].
+Although materials for tanning used by the Kavalan people are unknown due to the loss of traditions, brain tanning is a possible method given the availability of animal brains from hunting and their high-fat content.
+
+<!--decline of deer, ceramics, and population dynamics of Kiwulan-->
+The dramatic decrease in deer bones during the Chinese period may indicate overhunting that persisted for centuries or a decline in the Indigenous population (Zhan, 1995).
+The double increase in suid after the European arrival may suggest an increased reliance on pigs.
 This implies that pigs played a more significant role in the diet after Europeans arrived.
 Eventually, cows replaced pigs during the Chinese period, which might be a Chinese influence.
-
-By contextralizing the sampled faunal remains from the major residential area at Kiwulan   into the time periods with foreign contact, we can observe the
-
-<!--population dynamics and pottery assembalges-->
 
@@
 <!--resilence-->
+By contextralizing the sampled faunal remains from the major residential area at Kiwulan   into the time periods with foreign contact, we can observe the
 

### 2025-03-03 analysis/code/kwl-faunal-layers.R
code edits

@@ -80,7 +80,7 @@ fauna_combined_taxa <-
                           TRUE ~ taxa)) %>%
   mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~
                                      "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
-                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Wild boar",
+                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Suid",
                                    str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~
                                      "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~
                                      "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
@@ -98,7 +98,8 @@ fauna_combined_taxa <-
   rename(Taxon = taxa, NISP = n) %>%
   select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
   bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>%
-  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`))
+  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`)) %>%
+  mutate(Taxon = ifelse(Taxon == "Sus scrofa", "Sus spp.", Taxon))
 
 # calculate MNI
 fauna_combined_MNI <-
@@ -112,11 +113,23 @@ fauna_combined_context <- fauna_combined_context_all %>% filter(!taxa == "Rattus
 # barplot I: NISP by period (NISP)
 NISP_barplot <-
   fauna_combined_context %>%
-  count (period) %>%
+  count(period) %>%
   drop_na() %>%
   ggplot(aes(x = period, y = n)) +
   geom_col(width = 0.6) + #geom_bar(stat = "identity", #  geom_col(width = 0.6) +  #
            #position = position_dodge2(preserve = "single"), widtg = 0.6) +
+  labs(y = "NISP", x = NULL) +
+  theme_minimal() +
+  theme(legend.title=element_blank())
+
+# barplot: across taxon
[…]

### 2025-03-03 analysis/paper/paper.qmd
work on discussion section

@@
 ---
-title: "Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan"
+title: "Exploring colonial impacts on indigenous hunting and animal use in northeastern Taiwan"
 author:
@@
 In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest Iron Age settlement in northeastern Taiwan [@Chen2007].
-Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century.
+Kiwulan encountered European colonial power in the 17th century, followed by a large influx of Han Chinese migration in the 19th century.
 What was the resource use of Indigenous people before and after the foreign presence?
-Can we identify shifts in these practices in response to colonial economic demands?
-To address these questions, I will focus on subsistence practices because the choices made in daily practices can reveal the strategies Indigenous people used to respond to resource explortations by foreign presence.
-I will examine the zooarchaeological assemblages from the upper culture layer at Kiwulan, a large settlement dating to 14-19th centuries.
-Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as European colonization or intercultural contact [@Defrance2009].
-Through a zooarchaeological perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.
+Can we identify subsistence strategies in response to impacts resulting from integration into trading network?
+To address these questions, I will analyze the zooarchaeological assemblages from Kiwulan, because they provide valuable insights into subsistence practices and accommodation within specific social contexts, such as European colonization or being part of the regional trade network [@Defrance2009; @Crabtree1990].
+Through a perspective from subsistence, this study aims to shed light on Indigenous persistence and resilience in the face of foreign presence.
 
@@
                                    "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
-                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Wild boar",
+                                   str_detect(taxa, "Ree") ~ "Muntjac", taxa == "Sus scrofa" ~ "Suid",
                                    str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~ 
@@
   bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>% 
-  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`))
+  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`)) %>% 
+  mutate(Taxon = ifelse(Taxon == "Sus scrofa", "Sus spp.", Taxon))
 
@@
 To further compare general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle.
-Similarly, wild boar and domestic pig (n = 2, from CL4 and CL6) were grouped into the category of suid.
+Similarly, domestic pig (n = 2, from CL4 and CL6) and wild boar were grouped into the category of suid.
 Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.
@@
 In addition, natural alterations to bones resulting from post-deposition impacts were also observed , such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify).
+Those post-depositional impacts might have led to an underestimation of cut marks due to the difficulty of surface identification.
 
@@
 <!--results and chronology -->
-The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption since CL6 until CL2, and then shifted to cattle in CL1.
-Despite deer had been a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals shows changes that could reflect the indigenous reactions to foreign presence.
-Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences relative to foreign presences for the following discussion.
-The sequences include the pre-European period (CL6, CL5), the European period (CL4), post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1).
-It should be noted that the European period does not necessary means a solid European colonial control over the Kiwulan, but a broad impacts from European occupation and economic activities on indigenous societies in Yilan, such as getting involvement into the long-distance trade network, being part of the tribute system, or experiences of occasional European attacks.
+The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption from CL6 to CL2, and then shifted to cattle in CL1.
+Despite deer was a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals show changes that could reflect the indigenous reactions to foreign presence.
+Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences in relation to foreign presences for further discussion.
+The temporal sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1).
+It should be noted that the European period does not necessarily mean a solid European colonial control over the Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial activities on indigenous societies in Yilan, such as getting involved in the long-distance trade network, being part of the tribute system, and experiencing occasional European attacks.
 
-<!--summary of overall observations-->
-Faunal remains recovered from the layer associated with the European presence reveal the highest quantity, followed by the pre-European period.
-During the European period, in addition to the predominance of deer, an increased proportion of suid among faunal remains suggests a slightly greater reliance on suid as a secondary meat source compared to other time periods, along with a broader range of medium and small animals, such as muntjac, birds, fish, and reptiles.
-A similar pattern was observed in the ubiquity of animals, which shows the highest ubiquity of deer, suid, and munjact remains during the European period.
-Interestingly, deer bones from the pre-European period, specifically CL5, presented the same ubiquity as in the European period but in only half the quantity.
-This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but consumption of deer was intensified during the European period.
-During the Chinese period, the low quantity of deer in CL2 and the predominate of cattle in CL1 suggest a sudden shift in dietary practices after the Chinese arrival, indicating a Chinese influence and the decline of Indigenous population in this region.
+<!--summary of overall observations across time periods -->
+Faunal remains recovered from the European period reveal the highest quantity, followed by the late pre-European period.
+During the European period, in addition to the predominance of deer, an increased proportion of suid suggests a slightly greater reliance on suid as a secondary meat source.
+A similar pattern was observed from a higher ubiquity of suids during the European period compared to other periods.
+Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but in only half the quantity.
+This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but deer utilization was intensified during the European period.
+The dramatic decrease in deer bones after the European period might be related to both the decline of the Kiwulan population and a decrease in the wild deer population.
+Sika deer populations in western Taiwan have significantly shrunk since the 18th century because of habitat loss and overhunting [@Chen2000animal].
+Despite this, deer remained a predominant resource for local consumption at Kiwulan during the post-European and early Chinese period.
+In the late Chinese period, the dietary preference shifts from deer to cattle, suggesting a Chinese influence and the relocation of the Kavalan people to Hualien due to population pressures from Chinese immigration [@Zhan1995].
 
-<!--key results of skeletal part frequencies and butchery patterns-->
-The skeletal part frequencies and butchery patterns of deer demonstrate more complete carcasses processed within the settlement during the European period and pre-European period compared to the post-European period and Chinese period.
-Although a more complete and evenly distributed deer parts with cut marks were found in both the European period and before European arrival, the substantial rise in head parts (e.g., crania and mandibles) and overall high quantity of bones in the European period suggests more deer were brought back intact to the settlement.
-The low frequency of head parts in the earliest layer and Chinese period might indicate deer could be mostly processed at the hunting location with only meaty (e.g., upper limb) and mid-meaty portions (e.g., ribs) selectively butchered and transported back.
-Interestingly, among the meaty portions throughout all periods, front quarters (e.g., humeri, radius) are more frequently found compared to forequarters (e.g., femur, tibia).
+<!--skeletal part frequencies and butchery patterns of deer-->
+Despite being the main resource throughout most periods, deer utilization shows different patterns over time.
+Dear body parts and butchery patterns demonstrate more complete carcasses processed within the settlement during the European and pre-European period compared to the post-European and Chinese period.
+Although more complete and evenly distributed joints with cut marks were found in both the European period and before the European arrival, the substantial rise in head parts (e.g., crania and mandibles) with the overall high quantity suggests more deer were brought back intact to the settlement.
+On the contrary, the low frequency of head parts in the early pre-European period and Chinese period might indicate deer could be mostly processed at the hunting location with mostly meaty (e.g., upper limb) and mid-meaty portions (e.g., ribs) selectively butchered and transported back.
+Interestingly, among the meaty portions, front quarters (e.g., humeri, radius) are more frequently found compared to forequarters (e.g., femur, tibia) throughout all periods.
 One likely explanation is that femur and tibia might be heavily fragmented due to being broken for their high marrow yields [@Madrigal2002], leading to underestimation due to difficult identification.
-The behavior of marrow extraction can be supported by around one-third long bones being fractured and a moderate degree of fragmentation on average at Kiwulan.
-Another possibility is forequarters could be traded since deer legs as commodity were mentioned in European records [@Blusse2006, pp.220-224], but unfortunately, there was no detail about which portion of legs were normally traded.
+Marrow extraction at Kiwulan can be supported by around one-third of long bones being fractured and a moderate degree of fragmentation on average.
+Another possibility is forequarters could be traded since deer legs and meats as commodities were commonly mentioned in European records [@Blusse2006, pp.220-224], but there is no mention of any specific legs commonly traded.
+Considering the generally high frequencies of meaty portions found at Kiwulan, deer might not just for consumption at the site, but also for trading.
 
-<!--skinning-->
-The cut marks found on the lower limbs of deer at Kiwulan should be associated with the traces left by skinning processes, which confirms that deer skins were removed and likely used for trading with external traders.
-A large quantity of foreign goods, such as ornaments, porcelains, and metal items, supports frequent trading activities between Kiwulan residents and outsiders [@Wang2020ornament; @Chen2007].
-Deer hides and venison, were important commodities in the East Asian trade network since the mid-15th century, connecting Han Chinese, Japanese, and Indigenous people, especially in western Taiwan [@Cheng2017; @Hollmann1991].
-This deer trade was later monopolized by the Europeans in the 17th century, and deer products were largely exported to Japan.
-The evidence of the removal of deer skins at Kiwulan suggests that not only western Taiwan but also northeastern Taiwan involved and impacted by the long-distance trade.
-Deer bones with skinning cuts did not show a substantial increase in the European period; however, the disproportionate increase in head parts hints at a likelihood of brain tanning when preparing hides.
+<!--deer hides-->
+Deer hides were important commodities since the mid-15th century, connecting Han Chinese, Japanese, and Indigenous people, and later monopolized by the Europeans in the 17th century, when deer products were largely exported to Japan, especially from western Taiwan [@Cheng2017; @Hollmann1991].
+The cut marks found on the lower limbs of deer at Kiwulan should be associated with the traces left by skinning, which confirms the removalof  deer skins and likely being traded with external traders.
+A large quantity of foreign goods, such as ornaments, porcelains, and metal items, supports frequent exchange between Kiwulan residents and outsiders [@Wang2020ornament; @Chen2007].
+The evidence of skinning and foreign goods at Kiwulan suggests that not only western Taiwan but also northeastern Taiwan was involved and impacted by the long-distance trade.
+Although deer bones with skinning cuts did not show a substantial increase in the European period, the disproportionate increase in head parts might hint at a likelihood of brain tanning when preparing hides.
 Brain tanning is a traditional method for softening and preserving hides commonly used in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016].
@@
 48; @Xu1994].
-Although materials for tanning used by the Kavalan people are unknown due to the loss of traditions, brain tanning is a possible method given the availability of animal brains from hunting and their high-fat content.
+Despite the materials for tanning used by the Kavalan people being unknown due to the loss of traditions, brain tanning is a practicable method given the availability of deer brains from hunting.
 
-<!--decline of deer, ceramics, and population dynamics of Kiwulan-->
-The dramatic decrease in deer bones during the Chinese period may indicate overhunting that persisted for centuries or a decline in the Indigenous population (Zhan, 1995).
-The double increase in suid after the European arrival may suggest an increased reliance on pigs.
-This implies that pigs played a more significant role in the diet after Europeans arrived.
-Eventually, cows replaced pigs during the Chinese period, which might be a Chinese influence.
+<!-- population dynamics and arguments-->
+The population dynamics at Kiwulan must be considered when discussing faunal changes over time.
+A larger pottery vessels during the European period suggests population growth [@Wang2020pot; @Hsieh2009], which may contribute to the increasing animal consumption overall.
+However, the impacts from foreign presence on faunal assemblages can be revealed through butchery patterns of deer parts, the relative proportions of animals, and taxon ubiquity.
+First, deer bones with skinning evidence were mostly found during the European period, followed by the late pre-European phase, corresponding with the rise of regional trade networks since the 15th century and intensified in the 17th century.
+Second, if the high frequency of deer head parts during the European period solely resulted from food scarcity due to population growth and increased consumption of brains for nutrition, a parallel rise in marrow extraction would be expected.
+However, fractured long bones did not increase during the European period or the previous phase; instead, they were 10-30 % lower than in other periods.
+Third, population growth should have led to a more proportional increase in both large and small animals.
+Although there was a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion compared to other animals.
+This suggests that intensified deer hunting for trade may have driven indigenous people to rely more on suids as an alternative prey, likely due to their comparable body size and meat return rates to deer [@Broughton2011prey].
+Suids as a secondary protein source after deer is also observed in prehistoric western Taiwan [@Chuang2021].
 
-<!--environmental changes in general and taphonomic issues-->
+<!--indigenous reaction and resilience-->
+Compelling the results of faunal analysis and reports from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but also a shift driven by European presence and associated intensified long-distance trade.
+It is clear to see deer is both the first priority of food source and commodities at Kiwulan since the late pre-European period and intensified during the European period.
+Based on the dietary preference, the increased use of suids during the European period could be a temporary subsistence strategy in response to the high demands of deer products.
+After the Europeans left, deer hide trade continued but on a smaller scale, with around 10,000 hides exported annually in the 18th century, which is 20 times less than during the European period [@].
+This change likely reduced the reliance of suids, as shown by their decreased proportion at Kiwulan.
+The slightly increase in muntjact during the post-European and early Chinese period might be associated with the decline of deer population, leading people turned to use muntjact hides for trade instead [].
+Nevertheless, deer continued to be consumed until the early Chinese period.
+Everyday practices play an essential role for connecting history and social memories, enabling Indigenous people maintain tradition in a setting of frequent cultural exchange or asymmetrical relations [@Silliman2009; e.g., @Noe2023].
+The persistence of deer consumption at Kiwulan reflect identity negotiation and tradition maintenance, while the increased utilization of suids likely responded to the high demand for deer products in large scale trade.
+This highlights how Kiwulan residents reacted to foreign presence and demonstrated cultural resilience.
 
-<!--resilence-->
-By contextralizing the sampled faunal remains from the major residential area at Kiwulan   into the time periods with foreign contact, we can observe the
+<!--taphonomic & preservation bias -->
 
-played a consistent role in subsistence strategies.
-A taxon can have high NISP but low ubiquity if concentrated in a few contexts (e.g., feasting deposits).
-A taxon with high ubiquity but low NISP may have been used frequently but in small amounts.
+# Conclusion
 
-frequencies of deer bone elements are consistent with the proportion of joints. 
-Other than quantity, the measure of presence and absence of a given taxon across the sampeld units could give insights into the frequency and distribution consistency of taxa at a settlement scale.
-
-High-ubiquity taxa (e.g., deer, pigs) likely played a consistent role in subsistence strategies.
-Low-ubiquity taxa (e.g., rare birds, marine species) may indicate opportunistic hunting, ritual use, or seasonal availability.
-Changes in ubiquity across layers may reflect environmental shifts or human adaptations.
-For example, a decline in large mammal ubiquity might suggest deforestation or overhunting, leading to a shift toward smaller prey.
-
-Issue of Taphonomic & Preservation Bias: Some taxa may be less ubiquitous due to poor bone preservation rather than actual scarcity in the past.
-Small and fragile bones (e.g., fish, rodents) might have low ubiquity due to taphonomic processes.
-
-These post-depositional impacts might have led to an underestimation of cut marks due to the diffculty of surface identification.
-
-The European maritime cultural exchange to some extent altered the material culture of the Indigenous peoples in northeastern Taiwan by adopting foreign goods or customs into their daily lives.
-However, a sign of indigenous resilience in the face of European influences and resource exploitation may be seen in the continuing tradition of pottery production and deer consumption with few changes, even though changes were observed through the uneven distribution of prestige items, such as glass beads and porcelains, across different households in the living area and burials (Wang and Marwick, 2020a; 2021).
-
-The continuous practices of ceramic making and deer consumption by the Indigenous peoples before and after the European arrival can be viewed as a way of negotiating their identities and maintaining their traditions.
-Everyday practices are essential in connecting history and social memories, which helps maintain tradition in a setting of frequent cultural exchange or asymmetrical relations (Silliman, 2009).
-Moreover, ceramics could be an item that conveys social meaning through its style, such as decorations, and be used to show group identity (Croucher and Wynne-Jones, 2006; Pikirayi, 2007).
-It is worth noting that the pottery found at Kiwulan, with its typical round body and incised geometric motifs, is also commonly found across many sites during this period in northeastern Taiwan (Taiwan History, 2005).
-Similar geometric motifs and vessel shapes in the region might suggest the communication of Indigenous identity within the context of ongoing cross-cultural interactions with outsiders, including Europeans and Chinese.
-Furthermore, despite the availability of foreign items, such as iron pans that could have potentially replaced local pottery for cooking purposes, local ceramics have been used continuously for over 6,000 years (Chen, 2007).
-This emphasizes the significant role of pottery-making in the lives of Indigenous communities.
+Through contextralizing the faunal remains into time periods relaitve to different foreign presences, including the Europeans in the 17th century corresponding to CL4, and the Chinese immigrants since the late 19th century corresponding to CL1, the resutls reveals how indigenous people respond to the foreign influences on their economic aspect.
 
@@
 
-# Conclusion
-
-Through contextralizing the faunal remains into time periods relaitve to different foreign presences, including the Europeans in the 17th century corresponding to CL4, and the Chinese immigrants since the late 19th century corresponding to CL1, the resutls reveals how indigenous people respond to the foreign influences on their economic aspect.
-
 Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period.
@@
 Those historical events are closely related to the Taiwan Indigenous people today, which is crucial to understanding the contemporary Indigenous cultures and socio-ecological landscape.
+
 The results of this project can advance our knowledge of the reaction of Indigenous people in Taiwan when facing dominant foreign power, and generate new anthropological insights about the adaptation and choices of Indigenous people.

### 2025-03-05 analysis/paper/paper.qmd
work on discussion, conclusion, and abstract sections

@@
 ---
-title: "Exploring colonial impacts on indigenous hunting and animal use in northeastern Taiwan"
+title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan, a late Iron site in Northeastern Taiwan"
 author:
@@
 abstract: |
-  Text of abstract
+  Colonial encounters between Europeans and Indigenous people often led to profound impacts on many Indigenous societies and cultures around the world.
+In northeastern Taiwan, local Indigenous groups experienced foreign presences through regional trade networks in East Asia, which later expanded into a larger-scale trade network after the arrival of Europeans in the 17th century.
+This research focuses on local subsistence through faunal remains to understand how Indigenous communities responded to the regional trade activities driven by the European and Han Chinese, particularly the high demand for natural resources, such as deer.
+By analyzing faunal remains from the upper cultural layer of Kiwulan (AD 1450-1950), a late Iron Age site in northeastern Taiwan, the local participation in deer hide trade and their adaptive strategies were discussed.
+The results suggest indigenous cultural persistence of deer consumption practiced by local people, along with increased reliance on suids during the European period as the demands for deer products intensified.
+This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European territories, a perspective rarely discussed.
+
 keywords: |
-  keyword 1; keyword 2; keyword 3
+  Zooarchaeology; Iron Age; Taiwan; sika deer; trade; European colonization
 highlights: |
@@
 The temporal sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1).
-It should be noted that the European period does not necessarily mean a solid European colonial control over the Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial activities on indigenous societies in Yilan, such as getting involved in the long-distance trade network, being part of the tribute system, and experiencing occasional European attacks.
+It should be noted that the European period does not necessarily mean a solid European colonial control over Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial economic activities on indigenous people in Yilan, such as getting involved in the long-distance trade network, tribute paying, and suffered from occasional European attacks.
 
 <!--summary of overall observations across time periods -->
-Faunal remains recovered from the European period reveal the highest quantity, followed by the late pre-European period.
-During the European period, in addition to the predominance of deer, an increased proportion of suid suggests a slightly greater reliance on suid as a secondary meat source.
-A similar pattern was observed from a higher ubiquity of suids during the European period compared to other periods.
+The sampled area at Kiwulan shows a large quantity of faunal remains in the European period, followed by the late pre-European period.
 Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but in only half the quantity.
 This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but deer utilization was intensified during the European period.
+During the European period, in addition to the predominance of deer, an increased relative proportion and ubiquity of suid suggests a greater reliance on suid than in other periods.
 The dramatic decrease in deer bones after the European period might be related to both the decline of the Kiwulan population and a decrease in the wild deer population.
 Sika deer populations in western Taiwan have significantly shrunk since the 18th century because of habitat loss and overhunting [@Chen2000animal].
+An increase in muntjac use may suggest they served as a substitute, replacing the role of suids in earlier periods.
 Despite this, deer remained a predominant resource for local consumption at Kiwulan during the post-European and early Chinese period.
@@
 <!--skeletal part frequencies and butchery patterns of deer-->
-Despite being the main resource throughout most periods, deer utilization shows different patterns over time.
-Dear body parts and butchery patterns demonstrate more complete carcasses processed within the settlement during the European and pre-European period compared to the post-European and Chinese period.
-Although more complete and evenly distributed joints with cut marks were found in both the European period and before the European arrival, the substantial rise in head parts (e.g., crania and mandibles) with the overall high quantity suggests more deer were brought back intact to the settlement.
-On the contrary, the low frequency of head parts in the early pre-European period and Chinese period might indicate deer could be mostly processed at the hunting location with mostly meaty (e.g., upper limb) and mid-meaty portions (e.g., ribs) selectively butchered and transported back.
-Interestingly, among the meaty portions, front quarters (e.g., humeri, radius) are more frequently found compared to forequarters (e.g., femur, tibia) throughout all periods.
-One likely explanation is that femur and tibia might be heavily fragmented due to being broken for their high marrow yields [@Madrigal2002], leading to underestimation due to difficult identification.
-Marrow extraction at Kiwulan can be supported by around one-third of long bones being fractured and a moderate degree of fragmentation on average.
-Another possibility is forequarters could be traded since deer legs and meats as commodities were commonly mentioned in European records [@Blusse2006, pp.220-224], but there is no mention of any specific legs commonly traded.
-Considering the generally high frequencies of meaty portions found at Kiwulan, deer might not just for consumption at the site, but also for trading.
+Although deer was the primary resource at Kiwulan, differences in skeletal completeness across periods suggests changes in butchery locations and cooking practices.
+There were more complete carcasses in the European and late pre-European period compared to the post-European and Chinese periods.
+Moreover, the substantial rise in fragmented head parts (cranium and mandible) during the European period indicates greater use of soft tissues, such as brains and tongues, or the transport of more intact deer to the settlement for processing.
+This aligns with the historical account in 1603, Record of Formosa (Dongfan ji) [@Chen1603], which describes deer tongue consumption and group hunting for deer in western Taiwan.
+Changes in cooking practices are also evident.
+The decline in burned bones over time with a lowest percentage in the European period suggests a shift from roasting to other cooking methods, such as boiling.
+The presence of small chopped bones that could fit in local ceramics at Kiwulan indicates boiling as a possible practice since the European period.
+The shift in cooking practices suggests that the increased need for deer may have led people to adopt boiling due to its additional benefits, such as maximizing residual meat use, facilitating grease rendering, and improving quality of lean meat [@Lupo1997].
+This could be further confirmed through organic residue analysis in the future.
+
+A consistent pattern was also identified throughout all periods: High frequencies of upper limb bones but low frequencies of vertebrae and ribs, suggesting selectively butchered of meaty portions.
+Moreover, it is interesting that front quarters (humerus and radius) are much more abundant than hindquarters (femur and tibia).
+One possible explanation is that femurs and tibiae were heavily fragmented for marrow extraction due to their high yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage.
+Marrow extraction can be supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation.
+Another likelihood is that some hindquarters might be traded, as European records mention deer legs and meats as commodities in addition to hides [@Blusse2006, pp.220-224], but without specifying which legs.
+Considering the overall high proportion of meaty portions at Kiwulan, deer were likely processed not only for local consumption but also for trade.
 
 <!--deer hides-->
-Deer hides were important commodities since the mid-15th century, connecting Han Chinese, Japanese, and Indigenous people, and later monopolized by the Europeans in the 17th century, when deer products were largely exported to Japan, especially from western Taiwan [@Cheng2017; @Hollmann1991].
-The cut marks found on the lower limbs of deer at Kiwulan should be associated with the traces left by skinning, which confirms the removalof  deer skins and likely being traded with external traders.
-A large quantity of foreign goods, such as ornaments, porcelains, and metal items, supports frequent exchange between Kiwulan residents and outsiders [@Wang2020ornament; @Chen2007].
-The evidence of skinning and foreign goods at Kiwulan suggests that not only western Taiwan but also northeastern Taiwan was involved and impacted by the long-distance trade.
-Although deer bones with skinning cuts did not show a substantial increase in the European period, the disproportionate increase in head parts might hint at a likelihood of brain tanning when preparing hides.
-Brain tanning is a traditional method for softening and preserving hides commonly used in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016].
+Deer hides trade had prevailed since 15th century in East Asia, and monopolized by the Europeans in the 17th century, when large hides were exported to Japan, especially from western Taiwan [@Cheng2017; @Hollmann1991; @Wang1981].
+The cut marks identified on deer lower limbs at Kiwulan confirm the removal of deer skins and likely being traded with external traders.
+Abundant foreign goods at Kiwulan, including ornaments, porcelains, and metal items, supports frequent exchange between residents and outsiders [@Wang2020ornament; @Chen2007].
+This indicates that not only western Taiwan but also northeastern Taiwan was intensively involved and impacted by the long-distance trade.
+Although lower limbs with skinning cuts did not show a substantial increase in the European period, the disproportionate increase in head parts might hint at a likelihood of brain tanning when preparing hides in addition to tissue consumption.
+Brain tanning is a traditional way for softening and preserving hides in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016].
 The present-day Bunun group in the central highlands of Taiwan has revived hide tanning based on oral history, primarily using peanuts or pig lard, with a few examples using animal brains [@Li1999, pp.
 48; @Xu1994].
-Despite the materials for tanning used by the Kavalan people being unknown due to the loss of traditions, brain tanning is a practicable method given the availability of deer brains from hunting.
+Despite the materials for tanning used by the Kavalan people being unknown due to the loss of traditions, brain tanning is a practicable method given the availability of deer heads from hunting.
 
 <!-- population dynamics and arguments-->
-The population dynamics at Kiwulan must be considered when discussing faunal changes over time.
-A larger pottery vessels during the European period suggests population growth [@Wang2020pot; @Hsieh2009], which may contribute to the increasing animal consumption overall.
-However, the impacts from foreign presence on faunal assemblages can be revealed through butchery patterns of deer parts, the relative proportions of animals, and taxon ubiquity.
-First, deer bones with skinning evidence were mostly found during the European period, followed by the late pre-European phase, corresponding with the rise of regional trade networks since the 15th century and intensified in the 17th century.
-Second, if the high frequency of deer head parts during the European period solely resulted from food scarcity due to population growth and increased consumption of brains for nutrition, a parallel rise in marrow extraction would be expected.
-However, fractured long bones did not increase during the European period or the previous phase; instead, they were 10-30 % lower than in other periods.
+The population dynamics at Kiwulan must be carefully considered when interpreting the faunal results, because high population during the European period inferred by pottery could contribute to the increasing animal use overall [@Wang2020pot; @Hsieh2009].
+However, the impacts from foreign presence can be still identified through butchery patterns of deer parts, the relative proportions of animals, and taxon ubiquity.
+First, deer bones with skinning evidence were mostly found during the European period, followed by the late pre-European phase, corresponding with the rise of regional trade networks since the 15th century and intensified trade in the 17th century.
+Second, if the high frequency of deer parts during the European period, especially heads relative to brains or tongues consumption, solely resulted from food stress due to population growth, a parallel rise in marrow extraction would be expected.
+In fact, long bones with fracture patterns were 10-30 % lower during the European period  and the previous phase than in other periods.
 Third, population growth should have led to a more proportional increase in both large and small animals.
-Although there was a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion compared to other animals.
-This suggests that intensified deer hunting for trade may have driven indigenous people to rely more on suids as an alternative prey, likely due to their comparable body size and meat return rates to deer [@Broughton2011prey].
-Suids as a secondary protein source after deer is also observed in prehistoric western Taiwan [@Chuang2021].
+Although there was a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion.
+This indicates that trade of deer products may have driven indigenous people to rely more on suids as an alternative prey, likely due to their comparable body size and meat return rates to deer [@Broughton2011prey].
+Suids as a secondary protein source after deer are also observed in prehistoric western Taiwan [@Chuang2021].
+This hints the role of suids as a supplement when deer was not sufficient.
 
 <!--indigenous reaction and resilience-->
-Compelling the results of faunal analysis and reports from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but also a shift driven by European presence and associated intensified long-distance trade.
-It is clear to see deer is both the first priority of food source and commodities at Kiwulan since the late pre-European period and intensified during the European period.
-Based on the dietary preference, the increased use of suids during the European period could be a temporary subsistence strategy in response to the high demands of deer products.
-After the Europeans left, deer hide trade continued but on a smaller scale, with around 10,000 hides exported annually in the 18th century, which is 20 times less than during the European period [@].
-This change likely reduced the reliance of suids, as shown by their decreased proportion at Kiwulan.
-The slightly increase in muntjact during the post-European and early Chinese period might be associated with the decline of deer population, leading people turned to use muntjact hides for trade instead [].
-Nevertheless, deer continued to be consumed until the early Chinese period.
-Everyday practices play an essential role for connecting history and social memories, enabling Indigenous people maintain tradition in a setting of frequent cultural exchange or asymmetrical relations [@Silliman2009; e.g., @Noe2023].
-The persistence of deer consumption at Kiwulan reflect identity negotiation and tradition maintenance, while the increased utilization of suids likely responded to the high demand for deer products in large scale trade.
+Compelling the zooarchaeological evidence and descriptions from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but also reflecting intensified hunting driven by deer trade.
+The results suggest deer is the first priority of food source and commodities at Kiwulan since the late pre-European period and their economic benefits were intensified during the European period.
+It should be noted that the role of the Dutch is both colonist who demanded tribute from Yilan, and another trader who were incorporated into the pre-existing inter-insular trade network involving Han Chinese merchants and other indigenous groups in Northern Taiwan [e.g., @Torrence2000].
+The prestige values embedded in foreign goods in indigenous culture might lead Kiwulan aggrandizers actively participated in the trade rather than resisted [@Wang2020ornament].
+After the Dutch left, deer hide trade continued but back to a small scale, with around 10,000 hides exported annually, which is 10 times less than during the European period [@Nakamura1997].
+A small deer trade market and low population at Kiwulan might lead to a reduced reliance of suids, as shown by a low proportion of suids in the post-European period and eventually disappeared during the Chinese period.
+
+<!--Final remarks -->
+Everyday practices play an essential role for maintaining Indigenous traditions within a colonial setting or frequent cultural exchange [@Silliman2009; e.g., @Noe2023].
+The persistence of deer consumption with similar butchery patterns at Kiwulan reflect a maintained tradition, showing community identities.
+
+while the increased utilization of suids likely responded to the high demand for deer products in large scale trade.
+The small sample size after separating them into different periods might limits a further interpretation; however, changes in different phases were still provide evident patterns reflecting changes in animal use.
 This highlights how Kiwulan residents reacted to foreign presence and demonstrated cultural resilience.
 
-<!--taphonomic & preservation bias -->
+Deer continued to be consumed until the early Chinese period.
 
 # Conclusion
-
-Through contextralizing the faunal remains into time periods relaitve to different foreign presences, including the Europeans in the 17th century corresponding to CL4, and the Chinese immigrants since the late 19th century corresponding to CL1, the resutls reveals how indigenous people respond to the foreign influences on their economic aspect.
-
-By situating the findings within a pericolonial context (Acabado, 2017), locally made pottery and faunal remains reveal cultural persistence and resilience among Indigenous peoples in northeast Taiwan.
-The interaction between Europeans and Indigenous peoples in the region appears to have been based on negotiation and bartering (Torrence, 2000).
-For instance, the Dutch reduced and eventually waived annual tribute in 1648 in response to the resistance to paying tribute by most Indigenous settlements in Yilan (Blussé and Everts, 2000, pp.
-534, 536, 558; Kang, 2012).
-The continuous consumption of deer and consistent skinning patterns before and after the arrival of Europeans indicate a more adaptable tribute payment system, whereby the Indigenous peoples did not need to hunt more deer to meet European demand.
-However, it is important to note that there may have been a gradual decline in deer numbers since the post-European period, which is supported by the increased number of pigs found in the same period.
-This could suggest cultural resilience through dietary changes to adapt to deer overhunting.
-A more comprehensive investigation across different Indigenous settlements in northeastern Taiwan could provide a clearer understanding of overall dietary patterns.
-
-Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period.
-This project will explore the environmental effect of colonial presence on local ecology on a regional scale using archaeological evidence that has not yet been fully understood.
-Archaeological materials can offer more direct evidence, allowing better interpretation for the environments where Indigenous live in that can also reduce biases from historical records.
-Through exploring the long-term dynamics of human-environmental relationship in colonial settings, we can better understand changes in local environments and Indigenous resilience.
-In addition to the Europeans, the interaction between Chinese traders and Indigenous groups in Taiwan since the Neolithic period, and an influx of Han Chinese immigrants in the 18th century, has contributed to the significant local cultural and environmental transformation.
-Those historical events are closely related to the Taiwan Indigenous people today, which is crucial to understanding the contemporary Indigenous cultures and socio-ecological landscape.
-
-The results of this project can advance our knowledge of the reaction of Indigenous people in Taiwan when facing dominant foreign power, and generate new anthropological insights about the adaptation and choices of Indigenous people.
-In the long term, this research will also benefit local Indigenous people, Kavalan, in northeastern Taiwan and help them to trace back their past and revitalize their cultural heritage by the plans of public presentation in the future.
-In addition, this project will combine geoarchaeological, geochemical, and computational methods to analyze archaeological evidence.
-The multidisciplinary approach can not only advance the development and application of scientific techniques in Taiwan archaeology, but also provide an opportunity to form collaborative relationships with scholars with different backgrounds, which will help to develop the academic network and create interdisciplinary collaboration.
-
-This research will contribute to a broader understanding of the diverse ways Indigenous communities adapted to and resisted colonial forces, enhancing our comprehension of the long-term impacts of European colonization on local societies.
+By contextualizing zooarchaeological evidence into time periods relative to foreign presence, including the Europeans in the 17th century, and Chinese immigrants from the late 19th century, this study reveals the participation of indigenous groups in regional trade work.
+The frequencies, butchering patterns, and human modifications on deer bones confirms the active deer hides trade since the early 15th century and intensified during the European period.
+The results demonstrates even indigenous people in peripheral areas with relatively weak colonial control were also profoundly impacted than previously thought.
+In pericolonial setting, the continued prioritization of deer consumption at Kiwulan suggests a cultural persistence of food practices, while the strategic reliance on suids during the European period demonstrates adaptation to high demands for deer products, and later incorporating more muntjac in response to overhunting of sika deer.
+The long-term dynamics of animal use in northeastern Taiwan can enhance our understanding of Indigenous resilience in face of foreign impacts and environmental change.
+In addition, the role of indigenous people might be more active in the exchange network considering the large amount of foreign goods that were incorporated into local cultures and used as prestige items.
+A more comprehensive investigation across Indigenous settlements in northeastern Taiwan and including other dietary evidence, such as botanical remains and organic residue analysis in the future could provide a deeper insights into the regional dietary patterns and human-environment dynamics.
 

### 2025-03-06 analysis/paper/paper.qmd
edit the introduction section and add references

@@
 ---
-title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan, a late Iron site in Northeastern Taiwan"
+title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
 author:
@@
   Colonial encounters between Europeans and Indigenous people often led to profound impacts on many Indigenous societies and cultures around the world.
-In northeastern Taiwan, local Indigenous groups experienced foreign presences through regional trade networks in East Asia, which later expanded into a larger-scale trade network after the arrival of Europeans in the 17th century.
-This research focuses on local subsistence through faunal remains to understand how Indigenous communities responded to the regional trade activities driven by the European and Han Chinese, particularly the high demand for natural resources, such as deer.
+In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contacts through due to regional trade networks in East Asia, which later expanded into a larger-scale trade after the arrival of Europeans in the 17th century.
+This research focuses on local subsistence through faunal remains to understand how Indigenous communities responded to the regional trade activities driven by the Europeans and the Han Chinese, particularly the high demand for natural resources, such as deer.
 By analyzing faunal remains from the upper cultural layer of Kiwulan (AD 1450-1950), a late Iron Age site in northeastern Taiwan, the local participation in deer hide trade and their adaptive strategies were discussed.
-The results suggest indigenous cultural persistence of deer consumption practiced by local people, along with increased reliance on suids during the European period as the demands for deer products intensified.
-This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European territories, a perspective rarely discussed.
+The results suggest indigenous cultural persistence of deer consumption, along with increased reliance on suids during the European period as the demands for deer products intensified.
+This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely discussed.
 
@@
 highlights: |
-  These are the highlights. 
+  Indigenous community in northeastern Taiwan engaged in regional deer trade were identified from faunal remains.
+  The high freqiencies of deer bones at Kiwulan reveals continously cultural persistence in deer consumption.
+  Changes in deer butchery patterns and cooking practices indicates an intensified hunting during the European colonization period.
+  Increased reliance on suids during the European period suggests adaptive strategies to sustain local subsistence.
 ---
@@
 Understanding the dynamics of these colonial processes has become a significant issue in archaeology because it provides critical insights into the complex outcomes of European colonization [@Gosden2004].
-In Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items from Asia for commercial purposes [@Prakash1997].
-Although the new exchange systems or colonial institutions introduced by Europeans could change local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002], indigenous groups did not just accept European colonization passively.
-Recent discussions on colonial interactions also suggest the active role of indigenous groups in responding to the arrival of Europeans and presenting resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998; @Noe2023].
+In East Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items for commercial purposes [@Prakash1997].
+The new exchange systems and colonial institutions introduced by Europeans often affected local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002].
+However, many recent studies also suggest that indigenous groups did not just accept European colonization passively but actively responded to the arrival of Europeans and colonialism through resistance, negotiation, and cultural persistence and resilience, showing entangled colonial experiences [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998; @Noe2023].
 
 <!-- NE Taiwan as a case study -->
-While many studies focus on prominent colonies, places on the periphery of colonial centers were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017].
+While many studies focus on prominent colonies, places on the periphery of colonial bases were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017].
 Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion.
-Yilan, located in northeastern Taiwan, is an example of a periphery region where European control was limited and indirect for most of the time but intensively involved in regional trade activities stimulated by the European presence.
-In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017].
-Despite some discussions on the operations and policies of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people with the commercial activities are rare.
-The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jordan2009].
-Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.
+Yilan, located in northeastern Taiwan, is an example of a remote region where European control was limited and indirect for most of the time but was intensively involved in regional trade activities with foreign groups.
+In the mid-17th century, the VOC (Dutch East India Company) monopolized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017].
+Despite many discussions on the operations and policies of the deerskin trade through Dutch and Chinese written records [@Lin2014deer], direct investigations on the colonial experiences of local indigenous people through archaeological evidence are rare.
+The interaction between Indigenous people and their environments in the colonial setting of economic demands requires more Indigenous perspectives from archaeological studies.
 
 <!-- research aim and questions -->
-In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest Iron Age settlement in northeastern Taiwan [@Chen2007].
-Kiwulan encountered European colonial power in the 17th century, followed by a large influx of Han Chinese migration in the 19th century.
-What was the resource use of Indigenous people before and after the foreign presence?
-Can we identify subsistence strategies in response to impacts resulting from integration into trading network?
-To address these questions, I will analyze the zooarchaeological assemblages from Kiwulan, because they provide valuable insights into subsistence practices and accommodation within specific social contexts, such as European colonization or being part of the regional trade network [@Defrance2009; @Crabtree1990].
-Through a perspective from subsistence, this study aims to shed light on Indigenous persistence and resilience in the face of foreign presence.
+In this paper, I will investigate the Indigenous subsistence in a colonial context through the case study of the Kiwulan site (AD 1350-1950), one of the largest Iron Age settlements in northeastern Taiwan [@Chen2007].
+Kiwulan encountered European colonial powers, including the Spanish and Dutch in the 17th century, followed by a large influx of Han Chinese migration in the 19th century, making it a suitable example to explore the reactions of Indigenous groups to the foreign presence over time.
+My research questions are: What was the resource use of Indigenous people before and after those foreign presences, especially during the Dutch period with large-scale trade?
+Can we identify the evidence of getting involved in the deer trade from faunal remains?
+If so, were there subsistence strategies in response to the impacts of economic demands by the Europeans and Chinese?
+To address these questions, I will analyze the faunal remains with a focus on vertebrate species from Kiwulan because zooarchaeology is one useful approach to studying local consumption and exchange systems [@Defrance2009; @Crabtree1990].
+Through zooarchaeological evidence, this study aims to stress the active role of Indigenous people at the periphery of colonial bases in regional trade networks.
 
@@
 <!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
-In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
+In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
 Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp.
@@
 <!--introduction of Kiwulan and the population-->
-Kiwulan, located in the northern part of Yilan, is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
+Kiwulan, located in the northern part of Yilan (@fig-KWL-map), is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
 As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed.
@@
 The resource use during the periods of external impacts still need further examination to better understands the changes over time.
+
+```{r}
+#| label: fig-KWL-map
+#| echo: false
+#| fig-cap: "The geographic location of Kiwulan and European fortress in northern Taiwan (adapted from Wang and Marwick, 2020). Map data from naturalearthdata.com."
+
+knitr::include_graphics(here::here("analysis/figures/KWL_map.png"))
+```
 
@@
 The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot.
-The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together (e.g., @Stiner1991).
+The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together [e.g., @Stiner1991].
 The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
@@
 It should be noted that the European period does not necessarily mean a solid European colonial control over Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial economic activities on indigenous people in Yilan, such as getting involved in the long-distance trade network, tribute paying, and suffered from occasional European attacks.
+The small sample size after separating them into different periods might limits a comprehensive interpretation.
+However, comparing the results in a continous sequences still provide evident patterns reflecting changes in animal use.
 
@@
 
-A consistent pattern was also identified throughout all periods: High frequencies of upper limb bones but low frequencies of vertebrae and ribs, suggesting selectively butchered of meaty portions.
-Moreover, it is interesting that front quarters (humerus and radius) are much more abundant than hindquarters (femur and tibia).
+A consistent pattern in deer part frequencies was also identified throughout all periods.
+There were high frequencies of upper limb bones but low frequencies of vertebrae and ribs, suggesting selectively butchered and preferences of meaty portions.
+Moreover, among those meaty portions, front quarters (humerus and radius) are always more abundant than hindquarters (femur and tibia).
 One possible explanation is that femurs and tibiae were heavily fragmented for marrow extraction due to their high yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage.
-Marrow extraction can be supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation.
-Another likelihood is that some hindquarters might be traded, as European records mention deer legs and meats as commodities in addition to hides [@Blusse2006, pp.220-224], but without specifying which legs.
+Marrow extraction can be supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation in general.
+Another likelihood is that some hindquarters might be used for trading, as European records mention deer legs and meats as commodities in addition to hides [@Blusse2006, pp.220-224], but without specifying which legs.
 Considering the overall high proportion of meaty portions at Kiwulan, deer were likely processed not only for local consumption but also for trade.
@@
 <!-- population dynamics and arguments-->
-The population dynamics at Kiwulan must be carefully considered when interpreting the faunal results, because high population during the European period inferred by pottery could contribute to the increasing animal use overall [@Wang2020pot; @Hsieh2009].
-However, the impacts from foreign presence can be still identified through butchery patterns of deer parts, the relative proportions of animals, and taxon ubiquity.
-First, deer bones with skinning evidence were mostly found during the European period, followed by the late pre-European phase, corresponding with the rise of regional trade networks since the 15th century and intensified trade in the 17th century.
-Second, if the high frequency of deer parts during the European period, especially heads relative to brains or tongues consumption, solely resulted from food stress due to population growth, a parallel rise in marrow extraction would be expected.
-In fact, long bones with fracture patterns were 10-30 % lower during the European period  and the previous phase than in other periods.
-Third, population growth should have led to a more proportional increase in both large and small animals.
-Although there was a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion.
-This indicates that trade of deer products may have driven indigenous people to rely more on suids as an alternative prey, likely due to their comparable body size and meat return rates to deer [@Broughton2011prey].
-Suids as a secondary protein source after deer are also observed in prehistoric western Taiwan [@Chuang2021].
+The high population during the European period inferred by pottery could contribute to the increasing animal use overall [@Wang2020pot; @Hsieh2009].
+However, the impacts from foreign presence can be still identified through deer butchery patterns, relative proportions of animals, and taxon ubiquity.
+First, lower limbs with skinning cuts were mostly in the European period, followed by the late pre-European phase, corresponding with the rise of regional trade since the 15th century that were later intensified in the 17th century.
+Second, if the high frequencies of deer parts, especially heads for brains or tongues consumption, during the European period solely resulted from food stress due to population growth, a parallel rise in marrow extraction would be expected.
+However, long bones with fracture patterns were 10-30 % lower during the European period  and the previous phase than in other periods.
+Furthrmore, population growth should have led to a more proportional increase in both large and small animals.
+Despite a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion.
+This indicates that deer products trade may have driven indigenous people to rely more on suids as an alternative prey, likely because of their comparable body size and meat return rates to deer [@Broughton2011prey].
+Suids as a secondary resource after deer in indigenous cultures were observed in western Taiwan [@Chuang2021].
 This hints the role of suids as a supplement when deer was not sufficient.
+After the Dutch left, deer hide trade continued but in a small scale with around 10,000 hides exported annually, which is much less than 100,000 hides in the European period [@Nakamura1997; @Tsao2011].
+The small trade market and low population at Kiwulan might lead to a reduced reliance of suids, as shown by a low proportion of suids in the post-European period and eventually disappeared during the Chinese period.
 
 <!--indigenous reaction and resilience-->
-Compelling the zooarchaeological evidence and descriptions from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but also reflecting intensified hunting driven by deer trade.
-The results suggest deer is the first priority of food source and commodities at Kiwulan since the late pre-European period and their economic benefits were intensified during the European period.
-It should be noted that the role of the Dutch is both colonist who demanded tribute from Yilan, and another trader who were incorporated into the pre-existing inter-insular trade network involving Han Chinese merchants and other indigenous groups in Northern Taiwan [e.g., @Torrence2000].
-The prestige values embedded in foreign goods in indigenous culture might lead Kiwulan aggrandizers actively participated in the trade rather than resisted [@Wang2020ornament].
-After the Dutch left, deer hide trade continued but back to a small scale, with around 10,000 hides exported annually, which is 10 times less than during the European period [@Nakamura1997].
-A small deer trade market and low population at Kiwulan might lead to a reduced reliance of suids, as shown by a low proportion of suids in the post-European period and eventually disappeared during the Chinese period.
-
-<!--Final remarks -->
-Everyday practices play an essential role for maintaining Indigenous traditions within a colonial setting or frequent cultural exchange [@Silliman2009; e.g., @Noe2023].
-The persistence of deer consumption with similar butchery patterns at Kiwulan reflect a maintained tradition, showing community identities.
-
-while the increased utilization of suids likely responded to the high demand for deer products in large scale trade.
-The small sample size after separating them into different periods might limits a further interpretation; however, changes in different phases were still provide evident patterns reflecting changes in animal use.
-This highlights how Kiwulan residents reacted to foreign presence and demonstrated cultural resilience.
-
-Deer continued to be consumed until the early Chinese period.
+Compelling the zooarchaeological evidence and descriptions from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but reflect intensified hunting driven by deer trade.
+Deer is the first priority of food source and commodities at Kiwulan since the late pre-European period and their economic benefits were intensified during the European period.
+It should be recognized that the role of the Dutch is both colonist who demanded tribute from Yilan, and another trader who were incorporated into the pre-existing inter-insular trade involving Han Chinese merchants and other indigenous groups in Northern Taiwan [e.g., @Torrence2000].
+Local indigenous people might take a more active participation in the trade rather than resisted due to the benefits from exchanging foreign necessities (e.g.
+metal knives and pan) and goods (e.g.
+ornaments and porcelains) that were highly valued in local cultures [@Wang2020ornament].
+Despite an indirect and weak Dutch colonial control in Yilan, the stress from colonial powers can not be totally neglected given the local experiences of the Dutch attacks occasionally.
+Everyday practices play an essential role for maintaining Indigenous traditions within a colonial setting or frequent cultural exchange [@Silliman2009; @Noe2023].
+The persistence of deer consumption with similar butchery patterns over time at Kiwulan reflect a maintained tradition even after other animals were introduced such as cattle, showing community identities.
 
@@
 In pericolonial setting, the continued prioritization of deer consumption at Kiwulan suggests a cultural persistence of food practices, while the strategic reliance on suids during the European period demonstrates adaptation to high demands for deer products, and later incorporating more muntjac in response to overhunting of sika deer.
+The This highlights how Kiwulan residents reacted to foreign presence and demonstrated cultural resilience.
 The long-term dynamics of animal use in northeastern Taiwan can enhance our understanding of Indigenous resilience in face of foreign impacts and environmental change.
@@
 
-<!-- The following line inserts a page break  -->
+I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the zooarchaeological remains used in this study and the archaeological team for the early documentation of faunal remains that provides a foundation for this study.
+Special thanks to Chien-Hsiang Lin at the Biodiversity Research Center, Academia Sinica, for his guidance and assistance in identifying fish bones.
+I also appreciate Jui-Yang Tsou for his valuable comments and assistance during the identification process.
+I would like to express my gratitude to Yin-Ru Chiang and my lab colleagues at the Biodiversity Research Center, Academia Sinica, for providing research space to finalize this work.
+I am also grateful to the reviewers of this manuscript.
 

### 2025-03-07 analysis/code/kwl-faunal-layers.R
made a minor change to animal category

@@ -56,7 +56,7 @@ fauna_combined_context_all <-
   mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
                                 str_detect(animal, "e deer")~ "deer", str_detect(animal, "pig")|
                                 str_detect(animal, "boar") ~ "suid", str_detect(animal, "fish") ~ "fish",
-                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
+                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "bovine",
                               animal == "aves" ~ "bird", TRUE ~ animal)) %>%
   mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
                              str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
@@ -64,7 +64,7 @@ fauna_combined_context_all <-
                            str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
                            TRUE ~ "reptile")) %>%
   mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
-  mutate(category = fct_relevel(category, "deer","suid","muntjac","cattle","rat","bird","fish","turtle")) %>%
+  mutate(category = fct_relevel(category, "deer","suid","muntjac","bovine","rat","bird","fish","turtle")) %>%
   mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"), "yes", "no"),
          fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
                               str_detect(`人為痕跡`, "折")|str_detect(`人為痕跡`, "敲"), "yes", "no"),
@@ -178,7 +178,7 @@ verte_mammal_barplot <-
   fauna_combined_context %>%
   filter(class == "mammal") %>%
   drop_na(period, category) %>% # remove category to get absolute abundance
-  mutate(category = factor(category, levels = c("deer", "suid", "muntjac", "cattle"), ordered = TRUE)) %>%
+  mutate(category = factor(category, levels = c("deer", "suid", "muntjac", "bovine"), ordered = TRUE)) %>%
[…]

### 2025-03-07 analysis/paper/paper.qmd
work on discussion section and add references

@@
 abstract: |
-  Colonial encounters between Europeans and Indigenous people often led to profound impacts on many Indigenous societies and cultures around the world.
-In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contacts through due to regional trade networks in East Asia, which later expanded into a larger-scale trade after the arrival of Europeans in the 17th century.
-This research focuses on local subsistence through faunal remains to understand how Indigenous communities responded to the regional trade activities driven by the Europeans and the Han Chinese, particularly the high demand for natural resources, such as deer.
-By analyzing faunal remains from the upper cultural layer of Kiwulan (AD 1450-1950), a late Iron Age site in northeastern Taiwan, the local participation in deer hide trade and their adaptive strategies were discussed.
-The results suggest indigenous cultural persistence of deer consumption, along with increased reliance on suids during the European period as the demands for deer products intensified.
-This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely discussed.
+  Colonial encounters between Europeans and Indigenous peoples often had profound impacts on Indigenous societies and cultures in many parts of the world.
+In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contact through regional trade networks in East Asia, which expanded further with the arrival of Europeans in the 17th century.
+This study focuses on local subsistence through faunal remains to explore how Indigenous communities responded to trade activities driven by Europeans and Han Chinese, particularly the high demand for natural resources such as deer.
+By analyzing faunal remains from the upper component of Kiwulan (AD 1350-1850), a late Iron Age site in northeastern Taiwan, this research dicusses the Indigenous participation in the deerskin trade and their adaptive strategies.
+The results suggest cultural persistence in deer consumption, along with increased reliance on suids during the European period as demand for deer products intensified.
+This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely explored.
 
@@
 <!-- introduction about European expansion -->
-European expansion since the 15th century was usually associated with long-distance trade, exploration, and colonization in many parts of the world, which often led to profound impacts on indigenous societies.
-Depending on particular geographic and historical circumstances, many regions experienced different forms of European colonialism and power relations [@Jordan2009].
-Understanding the dynamics of these colonial processes has become a significant issue in archaeology because it provides critical insights into the complex outcomes of European colonization [@Gosden2004].
-In East Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items for commercial purposes [@Prakash1997].
-The new exchange systems and colonial institutions introduced by Europeans often affected local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002].
-However, many recent studies also suggest that indigenous groups did not just accept European colonization passively but actively responded to the arrival of Europeans and colonialism through resistance, negotiation, and cultural persistence and resilience, showing entangled colonial experiences [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998; @Noe2023].
+European expansion since the 15th century was closely tied to long-distance trade, exploration, and colonization, which often had profound impacts on Indigenous societies.
+The impacts also varied across regions depending on specific geographic and historical circumstances, resulting in different forms of European colonization and power relations [@Jordan2009].
+Understanding the diverse and dynamic colonial processes has become a key focus in archaeology, because it provides critical insights into the complex consequences of European expansion [@Gosden2004].
+In East Asia, European colonization was mostly driven by the pursuit of raw materials, natural resources, or luxury goods for commercial purposes [@Prakash1997].
+The new exchange systems and colonial institutions introduced by Europeans profoundly influenced local economies, social structures, or settlement patterns [@Christie1996; @Junker2004; @Lape2002].
+However, recent studies emphasize that Indigenous groups were not just passively accept European colonization; instead, they actively responded through resistance, negotiation, and cultural persistence or resilience, leading to complex and entangled colonial experiences [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998; @Noe2023].
 
@@
 
-# European Encounters in the Northern Taiwan
+# The European colonization period and regional trade network in Northern Taiwan
 <!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
-In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
+In 1626, the Spanish (Spanish East Indies) established the first colony of San Salvador at Heping Dao, and later Santo Domingo in Tamsui in northern Taiwan (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
 Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp.
 103-134; @Berrocal2018].
-The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81].
-The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort.
-Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171].
-This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement.
+The neighboring area of the fortress was inhabited by two Indigenous groups, the Taparris and the Quimaurris, who had lived at the fort location before being expelled by the Spanish [@Borao2001, SIT:165, p.81].
+The relations between the Spanish and Indigenous people might not involve extensive colonialism or intercultural domination, with limited territorial control only to the fort.
+Also, the Spanish relied on the exchange of local goods and food surpluses with Indigenous people, according to written records [@Borao2001, SIT:170-171].
+This corresponds with a higher proportion of Indigenous and Chinese items with little European materials found in San Salvador [@Berrocal2018], indicating cultural entanglement.
 
@@
 In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662.
-Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007].
-For example, the VOC governors used various colonial institutions to secure their authority, including the annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007].
-Among the political institutions, the tribute system operated from 1642 to 1647 and the village lease system introduced in 1644 had great impacts on the local indigenous economy and environment.
-Based on the tribute system, each household in allied villages was required to pay rice or high-quality deerskin annually.
-The amount of deerskins collected from indigenous people in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007].
-Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on indigenous lands during certain periods [@Heyns2001].
+Compared to the Spanish, the Dutch was more strategic in consolidating their power by establishing political and economic institutions, forming alliances with Indigenous communities, and conquering resistant communities [@Kang2016; @Andrade2007].
+For examples, the colonial institutions to Indigenous people include taxation, tribute system, village-leasehold system, census investigations, annual gatherings (Dutch: Landdag), and Indigenous representative from each aliened village [@Kang2016; @Andrade1997; @Andrade2007].
+Among those institutions, the tribute system operated from 1642 to 1647 and the village-leasehold system introduced in 1644 had great impacts on Indigenous economy and the environment they lived in.
+Based on the tribute system, each household in allied communities was required to pay rice or deer products annually, especially deerskin.
+The amount of deerskin collected from Indigenous communities in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007].
+Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on Indigenous lands during certain periods [@Heyns2001].
 From 1633 to 1661, the average number of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017].
-Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy.
+Those colonial activities might affect the economy of indigenous people since deer was an important resource in prehistoric Taiwan [@Chen2000animal].
 
 <!--variation of colonial powers across regions and indirect colonization at Kavalan-->
-Although the Dutch power expanded across nearly all of Taiwan in the middle of the 17th century, the extent of their colonial dominance and control differed greatly among different regions.
-Colonial rules in regions far from the European forts were usually difficult and unsuccessful.
-Yilan in northeastern Taiwan is an example of minimal European presence and indirect colonization, with fewer imposed colonial rules for most of the time [@Kang2012].
-Yilan is an alluvial plain, located about 70 kilometers from the European colony in Keelung, with mountains in between.
-The main inhabitants in Yilan were the local indigenous group, the Kavalan.
-Given its strategic location on the route to Torrobouan, where the Dutch believed gold mines existed, they were eager to bring this area under their control [@DZI2000; @Kang2016, p47].
-In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory and demanded tribute from the indigenous people [@Kang2016].
-Instead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the Basay people active in northern Taiwan to coordinate and collect tributes in Yilan [@Kang2016].
-This indirect approach allowed the maintenance of autonomy and flexibility for the Kavalan people, who were able to negotiate tribute requested by the VOC [@Kang2016].
+Although the Dutch power expanded across nearly whole Taiwan island in the mid-17th century, the extent of their colonial dominance differed greatly from region to region.
+Colonial power over regions far from the European fortress were usually difficult and unsuccessful.
+Yilan in northeastern Taiwan is an example of minimal European dominance or indirect colonization for most of the time [@Kang2012].
+As an plain located around 70 kilometers away from the European colonies with mountains in between, Yilan was hardly accessed by land but mostly from sea.
+The main Indigenous group in Yilan was the Kavalan.
+Given its strategic location on the route to Torrobouan, a place believed having gold mines, the Dutch was eager to integrate Yilan under their control [@DZI2000; @Kang2016, p47].
+In 1644, the Dutch defeated several Kavalan communities and incorporated them into their colonial territory [@Kang2016].
+Instead of engaging directly with the Kavalan people, the Dutch relied mostly on the assistance of the mediate, the Basay people active in northern Taiwan [@Kang2003], to coordinate and collect tributes from Yilan.
+This indirect approach allowed the maintenance of autonomy of the Kavalan people, and was even able to negotiate tributes requested by the VOC [@Kang2016].
 
-<!-- the pre-existing trade network in norhtern Taiwan-->
-Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the early 15th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005].
-Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the local Basay people [@Chen2005; @Andrade2007].
-This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003].
+<!-- the pre-existing trade network in norththern Taiwan-->
+Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the 14th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005].
+Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the Basay people [@Chen2005; @Andrade2007].
+This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].
+Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003].
 This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005].
-Yilan experienced frequent trading activities driven by the pursuit of foreign goods even before the Europeans arrival, and these activities continued during the European presence in northern Taiwan, likely with some occasional pressure from European powers.
-This unique historical and spatial context makes Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp.
+This trade network was then intensified during the European presence, and the participation of Kavalan in this network was driven by the pursuit of foreign goods, and also likely colonial pressure from European powers.
+This unique historical and spatial context makes Yilan an ideal location to study Indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp.
 31-78; @Trabert2017].
@@
 Kiwulan, located in the northern part of Yilan (@fig-KWL-map), is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
-As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed.
-A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ.
-It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004].
-The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004].
-According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938].
-In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].
+A large scale salvage excavation was conducted at Kiwulan that exposed a total of 262 excavation units (4 * 4 m), revealing abundant artifacts and numerous features, including middens, burials, ditches, and post-holes with wooden posts in situ.
+The excavtion suggests that Kiwulan was a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004; @Chiu2004].
+The main inhabitants of Kiwulan were the Kavalan people based on oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004].
+Kavalan people is the main Indigenous group in Yilan during the late Iron Age and the early modern period.
+According to a Dutch census conduced in the mid-17th century, there were about 40 Kavalan settlements with approximately 10,000 people in Yilan.
+Kiwulan, written as "Kibannoran" in Dutch records, was the largest settlement, with 160 households and 711 people [@Nakamura1938].
+In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which eventrually led to the relocation of Kavalan people to Hualien in eastern Taiwan [@Zhan1995].
 
 <!--archaeological materials-->
-The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005].
-In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials.
-Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through exchange, primarily with the Basay people active in northern Taiwan.
-This indicates that Kiwulan has continuously participated in regional trade networks closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@Wang2007].
-Moreover, the use of foreign goods appears to have cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].
+The most common artifact found at Kiwulan was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005].
+In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts and ornaments, were largely found in both the residential area and burials.
+Due to no evidence of metallurgy and local production of beads, these items are believed to have been acquired through exchange, primarily with the Basay people active in northern Taiwan.
+This indicates that Kiwulan has continuously participated in regional trade networks closely associated with northern Taiwan since the 14th century and influenced by the long-distance trade during the 17th century [@Wang2007].
+Moreover, the use of foreign goods appears to have cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021; @Cheng2008].
 
 <!--archaeological remains-->
-There were a wide range of animal bones and plant remains uncovered, reflecting the diversity of the local diet and the environment close to river and coastal area.
-The animals include deer, pigs, cows, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004].
-Rice farming seems to be common but limited to a household scale, according to historical records [@].
-Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments, which indicates a reliance on aquatic and marine resources.
-Those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities.
-According to those pieces of evidence, the subsistence practices at Kiwulan should be primarily reliant on rice farming, supplemented by hunting and fishing for wild resources.
-The resource use during the periods of external impacts still need further examination to better understands the changes over time.
+Abundant animal bones, shellfish, and plant remains were uncovered at Kiwulan, reflecting the diversity of natural resources in the plain and an environment close to estuary.
+The animal resource includes deer, suids, cattle, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004].
+Rice farming seems to be common but likely limited to a household scale rather than intensive agriculture according to historical records [@Chen1963].
+Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments.
+Based on those archaeological remains and environments, Kiwulan residents should have practiced a mixed subsistence that includes rice farming, hunting, and fishing for wild resources.
+Artifacts associated with those subsistence practices were also found at Kiwulan, such as iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities.
+However, the resource use across different time periods still need further investigations to better understands the subsistence adaptation relative to foreign presence and environmental changes.
 
@@
 # Materials and Methods
-To explore changes in dietary practices before and after the European arrival with associated long-distance trade, I examined faunal assemblages from 40 adjacent units at the residential area excavated during the first season (2001-2003) at Kiwulan [@Chen2007].
-The remains considered in this study are from dwelling spaces and trash pits that could reflect daily dietary practices, whereas those from burials that might relate to ritual activities are excluded.
-These 40 units were selected because of their location that were believed to be the center of the settlement with thick archaeological deposits and relatively intact archaeological contexts.
-Also, the spatial continuity can provides a more consistent framework for understanding food practices over time on a settlement scale.
-The same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences.
+The faunal assemblages from Kiwulan analyzed in this study were sampled from 40 adjacent units at the residential area excavated during the first season (2001-2003) [@Chen2007].
+These 40 units were selected because of their location that were believed to be the major residential area with thick archaeological deposits and relatively intact archaeological contexts without disruption from modern construction.
+The spatial continuity can provides a consistent framework for understanding food practices over time on a settlement scale.
+Also, the same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences.
+I mainly focus on vertebrate species due to their economic values in the trade market.
+The contexts of faunal remains include dwelling spaces and middens that were associated with daily dietary practices, whereas those from burials that might relate to ritual activities are excluded.
 
 To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six cultural layers (CL6-CL1) spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
-These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009].
-Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could offer a relatively robust framework for comparing archaeological records before and after the European influence.
-Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.
+The diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks relating to the 19th century [@Wang2011; @Hsieh2009].
+Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could still offer a relatively robust framework for comparing archaeological records before and after the foreign influence.
+Among these six temporal layers, the fourth cultural layer (CL4) can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second cultural layer (CL2) corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.
 
 Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences.
-Deer antlers without attaching to crania were excluded since they were less likely related to dietary practices.
-To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied.
-Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010].
+Deer antlers without attaching to cranium were excluded since they were less likely related to dietary practices.
+To obtain generalized pattern of animal use, measures of ubiquity and taxonomic abundance were applied.
+Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of consumtion of that taxa [@Mckechnie2016; @Vanderwarker2010].
 Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers.
 Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012].
-Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages.
-Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size [@Grayson1984; @Lyman2008].
-For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004].
+Different from MNI that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of remains by considering all fragmented bone assemblages.
+Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site scale with less impacts from sample size [@Grayson1984; @Lyman2008].
+For comparison across animal classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004].
 The normalized NISP is calculated by dividing NISP by the number of skeletal elements.
 
-Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies.
+Given that deerskin was a primary commodity in local regional exchange and in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies.
 This can shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network.
-Skeletal parts were quantified using NISP and classified into groups based on meat yield and portions: meaty (e.g., upperlimb), mid-meaty (e.g., vertebrae), and less meaty parts (e.g., skull and lower limb) [@Madrigal2004; @Munro2004].
-Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation and quantities compared to other parts.
+Skeletal parts were quantified using NISP and classified into groups based on meat yields and portions: meaty (e.g., upperlimb), mid-meaty (e.g., vertebrae), and less meaty parts (e.g., skull and lower limb) [@Madrigal2004; @Munro2004].
+Teeth without associated cranial elements were excluded to reduce overrepresentation due to their much higher preservation rate and quantities compared to other parts.
 While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002].
-Thus, less meaty elements were considered as secondary source of nutrition in the interpretation.
-Similarly, the degree of fragmentation could also imply the extraction of marrow or grease, which is calculated by dividing the Minimal Number of Elements (MNE) by NISP.
-Bone surface modifications relative to human butchering and cooking activities, such as cut marks, fracturing, and burning damage, were recorded, as well as natural marks such as weathering and root etching for understanding taphonomic effects.
-Cut marks on major limb joints, such as shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were further quantified to explore any changes in butchering patterns between temporal sequences.
+Thus, less meaty elements were considered as additional source of nutrition in the interpretation.
+Similarly, the degree of fragmentation could also imply the extraction of marrow or grease [@Lupo1997], which is calculated by dividing the Minimal Number of Elements (MNE) by NISP.
+
+Deer bone surface modifications relative to human butchering and cooking practices, such as cut marks, fracturing, and burning damage, were recorded, as well as natural marks such as weathering and root etching for understanding taphonomic effects.
+Cut marks on major limb joints, such as shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were quantified to explore butchering patterns over time.
 Moreover, cut marks on lower limbs relative to skinning processes were examined and compared.
-Those marks are charactersitic of circular and straight incisions left by the series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].
+Those marks are characteristic of circular and straight incisions left by the series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].
+
+The data and R code used for these analyses and visualizations are openly available at xxx for research reproducibility and transparency (@Marwick2017; @MarwickandWang2022).
 
@@
                                 str_detect(animal, "boar") ~ "suid", str_detect(animal, "fish") ~ "fish",
-                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
+                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "bovine",
                               animal == "aves" ~ "bird", TRUE ~ animal)) %>%
@@
   mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
-  mutate(category = fct_relevel(category, "deer","suid","muntjac","cattle","rat","bird","fish","turtle")) %>%
+  mutate(category = fct_relevel(category, "deer","suid","muntjac","bovine","rat","bird","fish","turtle")) %>%
   mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"), "yes", "no"),
@@
 unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "In")) %>% summarise(sum(NISP))
-deer_fami_NISP <- fauna_combined_taxa %>% 
-  filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "Ree")) %>% summarise(sum(NISP))
-deer_fami_NISP_per <- round(deer_fami_NISP/(total_NISP-unident), 3)*100
+deer_NISP <- fauna_combined_taxa %>% 
+  filter(str_detect(`Common name`, "eer")) %>% summarise(sum(NISP)) #|str_detect(Taxon, "Ree")
+suid_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Sus spp.")) %>% summarise(sum(NISP))
+buff_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Bub")) %>% summarise(sum(NISP))
+deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 2)*100
+suid_NISP_per <- round(suid_NISP/(total_NISP-unident), 2)*100
+buff_NISP_per <- round(buff_NISP/(total_NISP-unident), 2)*100
 ```
 
-The faunal assemblage analyszed in this paper from Kiwulan is composed of `r total_NISP` bone fragments based on NISP, which increased over time from CL6, peaked at CL4, and then declined dramatically after CL4 (@fig-NISP-proportion: A).
+The faunal assemblage analyzed in this study from Kiwulan is composed of `r total_NISP` bone fragments based on NISP (@tbl-counts).
+Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`) and water buffalo (`r buff_NISP_per`).
+To further compare general patterns of animal consumption in terms of their economic values over time, animals in the same family with similar body size were grouped together.
+Thus, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were group into the category, bovine.
+Similarly, domestic pig (n = 2, from CL4 and CL6) was integrated into the suid category.
+Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.
+In general, the faunal remains show the frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 ( (@tbl-counts, @fig-NISP-proportion: A).
 A similar trend was also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
-Among the faunal remains, the deer family is the majority accounting for about `r deer_fami_NISP_per`% of the bones identified to class level, followed by suid and cattle as the second and third taxon (@tbl-counts).
-Within the deer family, sika deer is prevalent, with a few bones identified as muntjac and a specimen indicating a possible affiliation with sambar deer.
-Although sika deer was the major animal source at the site, variations in their quantities and proportions could be observed between temporal sequences across 600 years.
-To further compare general patterns of dietary practices in terms of their economic values over time, the deer family were separated into two categories based on body size, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle were combined into one category, cattle.
-Similarly, domestic pig (n = 2, from CL4 and CL6) and wild boar were grouped into the category of suid.
-Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.
 
@@
 The reptile is represented by one piece of turtle shell found in CL4, which might not be a common food content at the site.
-Within those identifiable mammals, deer is the major source from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by cattle (@fig-NISP-proportion: D).
+Within those identifiable mammals, deer is the main animal from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by bovine (@fig-NISP-proportion: D).
 Suid was present from CL6 to CL3, and had a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5, and was continuously present until CL2.
@@
 Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model.
-The result shows no correlation between the diversity of animals and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), which rules out the potential sample size bias.
+The result shows no correlation between the diversity of animals and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), which suggests little bias from potential sample size.
 
@@
 
-@tbl-ubiquity shows deer had been the most frequent animal from CL6 to CL2, with the highest ubiquity in CL4 and CL5 (`r CL4_ubi_deer`), but disappeared in CL1.
-The second and third frequent animal are suid and muntjac.
-Although both animals reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for suid, and `r CL4_ubi_munt` for muntjac), their presence patterns are slightly different.
+@tbl-ubiquity shows deer had been the most widespread animal from CL6 to CL2, with the highest ubiquity in CL4 and CL5 (`r CL4_ubi_deer`), and lowest in CL2.
+The second and third animal regularly used are suid and muntjac.
+Although those two animals reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for suid, and `r CL4_ubi_munt` for muntjac), their presence patterns are slightly different.
 Suid was utilized as early as deer in CL6, declined in CL3, and disappeared in CL2, whereas muntjac showed a cultural lag, appearing one phase later than suid.
-Cattle had a low ubiquity in CL5 and high ubiquity in CL6.
+Bovine had a low ubiquity in CL5 and high ubiquity in CL6.
 Small animals (bird, fish, and turtle) had much lower ubiquity than large animals and they had a slightly increase in ubiquity in CL3.
-The result of animal ubiquity correspond with the result of their quantity (NISP), suggesting that deer is the most common resource at the site, with an increase utilization of suid and muntjac in CL4.
+The result of animal ubiquity correspond with their frequencies (NISP), suggesting that deer is the most common resource at the site, with an increase utilization of suid and muntjac in CL4.
 
-Since deer is the most frequent animal at Kiwulan, the frequency of skeletal parts was examined to explore any changes in processing decisions of meat portions between different phases.
-Deer bones were grouped into six body portions based on their nutritional values [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges).
+Although deer was predominant throughout all temporal sequences, variations in their skeletal parts could be identified.
+Deer bones were grouped into six body portions [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges).
 The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot.
-The consistency of higher numbers of upper limb and lower limb suggests that those elements should be transported together [e.g., @Stiner1991].
-The genereally low frequency of axial element, such as vertebrae and ribs, might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
+The consistency of higher numbers of upper limb and lower limb suggests that those parts should be transported together [e.g., @Stiner1991].
+The generally low frequency of axial elements might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
 A different pattern between layers is observed in the frequency of cranial bones, for which a few present in CL6, increased to the highest number in CL4, and then decreased thereafter.
 In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.
+Those differences might indicate changes in processing decisions of meat portions of deer.
 
-Considering the impacted of differential bone density on preservation of skeletal elements, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation.
+Considering the impacts of differential bone density on preservation of skeletal elements, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation.
 The values of bone survivorship were collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%.
-The result (@fig-deer-portion: B) shows that there is no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of deer bones.
-For the degree of bone fragmentation, CL3 shows higher fragmentation (index = `r frag_CL3_CL4[1]`), while bones in CL4 is less fragmented (index = `r frag_CL3_CL4[2]`).
+The result (@fig-deer-portion: B) shows no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of bones.
+For the degree of bone fragmentation, CL3 shows higher fragmentation (index = `r frag_CL3_CL4[1]`), while bones in CL4 are less fragmented (index = `r frag_CL3_CL4[2]`).
 However, the fragmentation rates are not significantly differed between layers.
@@
 Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching.
-@fig-deer-cutmark shows that at least 50 % of joints of each deer part in CL4 present cut marks, with a higher proportion of 75% at the hip.
+@fig-deer-cutmark shows that at least 50 % of joints in CL4 present cut marks, with a higher proportion of 75% at the hip.
 A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4.
-Not all types of joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult.
-Nevertheless, the overall high proprotion of joints with cut marks suggests intensive proceessing of deer throughout the site's occupation.
-In addition, cut marks on lower limbs can be associeted with skinning (@fig-cutmark-img: e) and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).
-Some joints displayed clean, smooth cut surfaces from chopping (@fig-cutmark-img: a, c), along with narrow V-shaped cut marks observed on most bones, suggesting they were likely made with metal tools [@Greenfield1999].
+Not all joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult.
+Nevertheless, the overall high proportion of joints with cut marks suggests intensive processing of deer throughout the site's occupation.
+In addition, cut marks on lower limbs should be associated with skinning (@fig-cutmark-img: e) and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).
+Some joints displayed clean, smooth cut surfaces from chopping (@fig-cutmark-img: a, c), along with narrow V-shaped cut marks observed on most bones, indicating the use of metal tools [@Greenfield1999].
 
@@
 CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5.
-Chewing marks were observed on a few bones that show a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023], suggesting carnivore gnawing.
-In addition, natural alterations to bones resulting from post-deposition impacts were also observed , such as weathering and root etching [@Gifford2018natural], were identified on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify).
-Those post-depositional impacts might have led to an underestimation of cut marks due to the difficulty of surface identification.
+Chewing marks were observed on a few bones represented by a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023], suggesting carnivore gnawing.
+In addition, natural alterations to bones resulting from post-deposition impacts were also observed, such as weathering and root etching [@Gifford2018natural], on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify).
+Those post-depositional impacts might have led to an underestimation of cut marks in general.
 
@@
 <!--results and chronology -->
-The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption from CL6 to CL2, and then shifted to cattle in CL1.
-Despite deer was a substantial part of dietary practices at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals show changes that could reflect the indigenous reactions to foreign presence.
+The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption from CL6 to CL2, and then shifted to bovine in CL1.
+Despite deer was a substantial part of subsistence at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals show changes that might relative to foreign impacts.
 Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences in relation to foreign presences for further discussion.
 The temporal sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1).
-It should be noted that the European period does not necessarily mean a solid European colonial control over Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial economic activities on indigenous people in Yilan, such as getting involved in the long-distance trade network, tribute paying, and suffered from occasional European attacks.
+It should be noted that the European period does not mean a solid European colonial control over Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial economic activities on indigenous people in Yilan, such as getting involved in the long-distance trade network, tribute paying, and suffered from occasional European attacks.
 The small sample size after separating them into different periods might limits a comprehensive interpretation.
-However, comparing the results in a continous sequences still provide evident patterns reflecting changes in animal use.
+However, comparing the results in a continuous sequences still provide evident patterns reflecting changes in animal use.
 
@@
 Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but in only half the quantity.
-This might indicate consistency in the frequency and distribution of deer at the settlement scale before and after European presence, but deer utilization was intensified during the European period.
-During the European period, in addition to the predominance of deer, an increased relative proportion and ubiquity of suid suggests a greater reliance on suid than in other periods.
+This might indicate consistency in the  regularity of deer consumption at the settlement scale before and after European presence, but deer utilization was intensified in the European period.
+During the European period, in addition to the predominance of deer, an increased relative proportion and ubiquity of suid suggests a more regular use of suid than in other periods.
 The dramatic decrease in deer bones after the European period might be related to both the decline of the Kiwulan population and a decrease in the wild deer population.
@@
 Despite this, deer remained a predominant resource for local consumption at Kiwulan during the post-European and early Chinese period.
-In the late Chinese period, the dietary preference shifts from deer to cattle, suggesting a Chinese influence and the relocation of the Kavalan people to Hualien due to population pressures from Chinese immigration [@Zhan1995].
+Bovine appeared in the pre-European period with just a few evidence (yellow cattle), but became the only animal (water buffalo) used at Kiwulan in the late Chinese period.
+Since bovine was not domestic aniamls that were beleived introduced from mainland China, this suggests a Chinese influence and the relocation of the Kavalan people to Hualien due to population pressures from Chinese immigration [@Zhan1995].
 
@@
 The presence of small chopped bones that could fit in local ceramics at Kiwulan indicates boiling as a possible practice since the European period.
-The shift in cooking practices suggests that the increased need for deer may have led people to adopt boiling due to its additional benefits, such as maximizing residual meat use, facilitating grease rendering, and improving quality of lean meat [@Lupo1997].
-This could be further confirmed through organic residue analysis in the future.
+This could be further confirmed through organic residue analysis on pottery in the future.
+Other meat preparation such as drying (or salting and smoking) cannot be ruled out [e.g., @Spyrou2019].
+Although deer vertebrae and ribs, portions potentially used for drying [@Friesen2001; @Binford1978], were in low frequencies at Kiwulan, dry meat consumption or exchange by the Kavalan people was reported in historical records [@Chen1963].
+The shift in cooking practices might result from increased need for deer, which could have led people to adopt boiling due to its additional benefits, such as maximizing residual meat use, facilitating grease rendering, and improving quality of lean meat [@Lupo1997], and drying for meat preserving and transporting.
 
@@
 One possible explanation is that femurs and tibiae were heavily fragmented for marrow extraction due to their high yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage.
-Marrow extraction can be supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation in general.
-Another likelihood is that some hindquarters might be used for trading, as European records mention deer legs and meats as commodities in addition to hides [@Blusse2006, pp.220-224], but without specifying which legs.
+Marrow extraction can be further supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation in general.
+Another likelihood is that some hindquarters might be used for trading, as Dutch records mention deer legs and meats as commodities in addition to hides [@Blusse2006, pp.220-224], but without specifying which legs.
+Deer legs imply bone-meat packages that would result in a low frequencies of bones of hindquarters or front quarters.
 Considering the overall high proportion of meaty portions at Kiwulan, deer were likely processed not only for local consumption but also for trade.
+The lower frequencies of hindquarters than front quarters might be attributed to both marrow extraction and export of deer products, rather than a sole contribution.
 
 <!--deer hides-->
-Deer hides trade had prevailed since 15th century in East Asia, and monopolized by the Europeans in the 17th century, when large hides were exported to Japan, especially from western Taiwan [@Cheng2017; @Hollmann1991; @Wang1981].
+Deerskin trade had prevailed since the 14th century in East Asia, and monopolized by the Europeans in the 17th century, when large hides were exported to Japan, especially from western Taiwan [@Cheng2017; @Hollmann1991; @Wang1981].
 The cut marks identified on deer lower limbs at Kiwulan confirm the removal of deer skins and likely being traded with external traders.
-Abundant foreign goods at Kiwulan, including ornaments, porcelains, and metal items, supports frequent exchange between residents and outsiders [@Wang2020ornament; @Chen2007].
-This indicates that not only western Taiwan but also northeastern Taiwan was intensively involved and impacted by the long-distance trade.
+Abundant foreign goods at Kiwulan, including ornaments, porcelains, and metal items, supports frequent exchanges between local residents and outsiders [@Wang2020ornament; @Chen2007].
+This suggests that not only western Taiwan but also northeastern Taiwan was intensively involved and impacted by the long-distance trade.
 Although lower limbs with skinning cuts did not show a substantial increase in the European period, the disproportionate increase in head parts might hint at a likelihood of brain tanning when preparing hides in addition to tissue consumption.
 Brain tanning is a traditional way for softening and preserving hides in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016].
-The present-day Bunun group in the central highlands of Taiwan has revived hide tanning based on oral history, primarily using peanuts or pig lard, with a few examples using animal brains [@Li1999, pp.
+The present-day Bunun group in the central highlands of Taiwan has revived hide tanning based on oral history, primarily using peanuts or pig lard as tannnig materials, with a few examples using animal brains [@Li1999, pp.
 48; @Xu1994].
@@
 However, the impacts from foreign presence can be still identified through deer butchery patterns, relative proportions of animals, and taxon ubiquity.
-First, lower limbs with skinning cuts were mostly in the European period, followed by the late pre-European phase, corresponding with the rise of regional trade since the 15th century that were later intensified in the 17th century.
+First, lower limbs with skinning cuts were mostly in the European period, followed by the late pre-European phase, corresponding with the rise of regional trade in the 14-15th century and was later intensified in the 17th century.
 Second, if the high frequencies of deer parts, especially heads for brains or tongues consumption, during the European period solely resulted from food stress due to population growth, a parallel rise in marrow extraction would be expected.
 However, long bones with fracture patterns were 10-30 % lower during the European period and the previous phase than in other periods.
-Furthrmore, population growth should have led to a more proportional increase in both large and small animals.
-Despite a high taxonomic diversity during the European period, suids present a disproportionate rise in their ubiquity, quantity, and proportion.
-This indicates that deer products trade may have driven indigenous people to rely more on suids as an alternative prey, likely because of their comparable body size and meat return rates to deer [@Broughton2011prey].
-Suids as a secondary resource after deer in indigenous cultures were observed in western Taiwan [@Chuang2021].
-This hints the role of suids as a supplement when deer was not sufficient.
-After the Dutch left, deer hide trade continued but in a small scale with around 10,000 hides exported annually, which is much less than 100,000 hides in the European period [@Nakamura1997; @Tsao2011].
-The small trade market and low population at Kiwulan might lead to a reduced reliance of suids, as shown by a low proportion of suids in the post-European period and eventually disappeared during the Chinese period.
+Third, population growth should have led to a more proportional increase in both large and small animals.
+Despite a high taxonomic diversity during the European period, suid presents a disproportionate rise in their ubiquity, quantity, and proportion.
+This indicates that deer products trade may have driven Indigenous people to rely more on suids as an alternative prey, likely because of their comparable body size and meat return rates to deer [@Broughton2011prey].
+Suids as a secondary resource after deer in Indigenous cultures were observed in western Taiwan [@Chuang2021].
+This hints the role of suid as a supplement when deer was not sufficient or available.
+After the Dutch left, deerskin trade continued but in a small scale with around 10,000 pieces exported annually, which is much less than 100,000 pieces in the European period [@Nakamura1997; @Tsao2011].
+The small trade market and low population at Kiwulan might lead to a reduced reliance of suid, as shown by a low proportion of suid in the post-European period and eventually disappeared during the Chinese period.
 
@@
 Deer is the first priority of food source and commodities at Kiwulan since the late pre-European period and their economic benefits were intensified during the European period.
-It should be recognized that the role of the Dutch is both colonist who demanded tribute from Yilan, and another trader who were incorporated into the pre-existing inter-insular trade involving Han Chinese merchants and other indigenous groups in Northern Taiwan [e.g., @Torrence2000].
-Local indigenous people might take a more active participation in the trade rather than resisted due to the benefits from exchanging foreign necessities (e.g.
+It should be recognized that the role of the Dutch is both colonist who demanded tributes from Yilan, and another trader who were incorporated into the pre-existing inter-insular trade involving Han Chinese merchants and other indigenous groups in Northern Taiwan [e.g., @Torrence2000].
+Local Indigenous people might take a more active participation in the trade rather than resisted due to the benefits of obtaining foreign necessities (e.g.
 metal knives and pan) and goods (e.g.
 ornaments and porcelains) that were highly valued in local cultures [@Wang2020ornament].
-Despite an indirect and weak Dutch colonial control in Yilan, the stress from colonial powers can not be totally neglected given the local experiences of the Dutch attacks occasionally.
+Despite an indirect and weak Dutch colonial control in Yilan, the stress from colonial powers cannot be totally neglected given the local experiences of the Dutch attacks occasionally.
 Everyday practices play an essential role for maintaining Indigenous traditions within a colonial setting or frequent cultural exchange [@Silliman2009; @Noe2023].
-The persistence of deer consumption with similar butchery patterns over time at Kiwulan reflect a maintained tradition even after other animals were introduced such as cattle, showing community identities.
+The persistence of deer consumption with similar butchery patterns over time at Kiwulan reflect a maintained tradition even after other animals were introduced, for example, bovine in the pre-European period.
+The increased reliance on suid during the European period could be viewed as a strategy to accommodate the high demands of deer products from trade network.
 
 # Conclusion
-By contextualizing zooarchaeological evidence into time periods relative to foreign presence, including the Europeans in the 17th century, and Chinese immigrants from the late 19th century, this study reveals the participation of indigenous groups in regional trade work.
-The frequencies, butchering patterns, and human modifications on deer bones confirms the active deer hides trade since the early 15th century and intensified during the European period.
-The results demonstrates even indigenous people in peripheral areas with relatively weak colonial control were also profoundly impacted than previously thought.
-In pericolonial setting, the continued prioritization of deer consumption at Kiwulan suggests a cultural persistence of food practices, while the strategic reliance on suids during the European period demonstrates adaptation to high demands for deer products, and later incorporating more muntjac in response to overhunting of sika deer.
-The This highlights how Kiwulan residents reacted to foreign presence and demonstrated cultural resilience.
-The long-term dynamics of animal use in northeastern Taiwan can enhance our understanding of Indigenous resilience in face of foreign impacts and environmental change.
-In addition, the role of indigenous people might be more active in the exchange network considering the large amount of foreign goods that were incorporated into local cultures and used as prestige items.
+By contextualizing zooarchaeological evidence into time periods relative to foreign presence, including the Europeans in the 17th century, and Chinese immigrants from the late 19th century, this study reveals the participation of Indigenous groups in regional trade work and adaptation strategies.
+The frequencies, butchering patterns, and human modifications on deer bones confirms the active deerskin trade since the 14th century and intensified during the European period.
+The results demonstrates even Indigenous people in the peripheral area with relatively weak colonial control were also profoundly impacted by colonial activities than previously thought.
+The continued prioritization of deer consumption at Kiwulan suggests a cultural persistence of food practices, while the strategic reliance on suids during the European period demonstrates adaptation to high demands for deer products, and later incorporating more muntjac in response to overhunting of sika deer.
+This highlights how Kiwulan residents reacted to foreign presence and environmental changes and demonstrated cultural resilience.
+The long-term dynamics of animal use in northeastern Taiwan can enhance our understanding of Indigenous reaction within a specific social-cultural context.
+The role of Indigenous people might be more active in the exchange network considering the large amount of foreign goods that were incorporated into local cultures and used as prestige items.
 A more comprehensive investigation across Indigenous settlements in northeastern Taiwan and including other dietary evidence, such as botanical remains and organic residue analysis in the future could provide a deeper insights into the regional dietary patterns and human-environment dynamics.
@@
 I also appreciate Jui-Yang Tsou for his valuable comments and assistance during the identification process.
-I would like to express my gratitude to Yin-Ru Chiang and my lab colleagues at the Biodiversity Research Center, Academia Sinica, for providing research space to finalize this work.
-I am also grateful to the reviewers of this manuscript.
+I would like to express my gratitude to Yin-Ru Chiang and his team at the Biodiversity Research Center, Academia Sinica, for providing me a lab space to work on data and finilizing this work.
+I am also grateful to the reviewers of this manuscript for their comments.
 

### 2025-03-09 analysis/paper/paper.qmd
draft editing

@@
 <!-- NE Taiwan as a case study -->
-While many studies focus on prominent colonies, places on the periphery of colonial bases were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017].
-Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion.
-Yilan, located in northeastern Taiwan, is an example of a remote region where European control was limited and indirect for most of the time but was intensively involved in regional trade activities with foreign groups.
-In the mid-17th century, the VOC (Dutch East India Company) monopolized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017].
-Despite many discussions on the operations and policies of the deerskin trade through Dutch and Chinese written records [@Lin2014deer], direct investigations on the colonial experiences of local indigenous people through archaeological evidence are rare.
-The interaction between Indigenous people and their environments in the colonial setting of economic demands requires more Indigenous perspectives from archaeological studies.
+While many studies focus on major colonial centers, peripheral regions also experienced varying degrees of European influence through colonial powers and associated trade networks, namely pericolonialism [@Acabado2017; @Trabert2017].
+Exploring these areas is important to for understanding the dirverse impacts of colonial expansion.
+Yilan, in northeastern Taiwan, is an example of a remote region where European control was limited and indirect but Indigenous communities were deeply involved in regional trade with foreign groups.
+In the mid-17th century, the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) monopolized the deerskin trade in Taiwan and exported deerskin on a large scale to Japan [@Heyns2001; @Cheng2017].
+While historical records by the Dutch and Chinese provide details about the operations and policies of the deerskin trade [@Lin2014deer], archaeological investigations into Indigenous reactions to these economic activities remain rare.
+To fully understand Indigenous responses to colonial economic demands, archaeological perspectives on their interactions with the environment are essential.
 
 <!-- research aim and questions -->
-In this paper, I will investigate the Indigenous subsistence in a colonial context through the case study of the Kiwulan site (AD 1350-1950), one of the largest Iron Age settlements in northeastern Taiwan [@Chen2007].
-Kiwulan encountered European colonial powers, including the Spanish and Dutch in the 17th century, followed by a large influx of Han Chinese migration in the 19th century, making it a suitable example to explore the reactions of Indigenous groups to the foreign presence over time.
-My research questions are: What was the resource use of Indigenous people before and after those foreign presences, especially during the Dutch period with large-scale trade?
-Can we identify the evidence of getting involved in the deer trade from faunal remains?
-If so, were there subsistence strategies in response to the impacts of economic demands by the Europeans and Chinese?
-To address these questions, I will analyze the faunal remains with a focus on vertebrate species from Kiwulan because zooarchaeology is one useful approach to studying local consumption and exchange systems [@Defrance2009; @Crabtree1990].
-Through zooarchaeological evidence, this study aims to stress the active role of Indigenous people at the periphery of colonial bases in regional trade networks.
+In this paper, I investigate Indigenous subsistence in a pericolonial context through the case study of the Kiwulan site (AD 1350-1850), one of the largest Iron Age settlements in northeastern Taiwan [@Chen2007].
+Kiwulan experienced European colonial powers, including the Spanish and Dutch in the 17th century, followed by a large influx of Han Chinese migrats in the 19th century, making it an ideal case to explore Indigenous responses to foreign presence over time.
+Key research questions include: How did Indigenous resource use change before and after these foreign encounters, particularly during the Dutch period of large-scale trade?
+Can we identify evidence of Indigenous participation in the deer trade from faunal remains?
+If so, any subsistence strategies emerged in response to economic demands from Europeans and Chinese traders?
+To address these questions, I analyze faunal remains with a focus on vertebrate species from Kiwulan, as zooarchaeology is an useful approach to studying local consumption and exchange systems [@Defrance2009; @Crabtree1990].
+Through this approach, this research highlights the active role of Indigenous communities at the periphery of colonial bases within regional trade networks.
 
@@
 <!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
-In 1626, the Spanish (Spanish East Indies) established the first colony of San Salvador at Heping Dao, and later Santo Domingo in Tamsui in northern Taiwan (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
-Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp.
+In 1626, the Spanish (Spanish East Indies) established their first colony in northern Taiwan, San Salvador at Heping Dao, follwed by Santo Domingo in Tamsui (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market.
+The VOC had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
+Although the Spanish intended to build a colonial city, San Salvador remained in its early stages of development throughout their colonization, consisting primarily of a fortress and a church [@Borao2009, pp.
 103-134; @Berrocal2018].
-The neighboring area of the fortress was inhabited by two Indigenous groups, the Taparris and the Quimaurris, who had lived at the fort location before being expelled by the Spanish [@Borao2001, SIT:165, p.81].
-The relations between the Spanish and Indigenous people might not involve extensive colonialism or intercultural domination, with limited territorial control only to the fort.
-Also, the Spanish relied on the exchange of local goods and food surpluses with Indigenous people, according to written records [@Borao2001, SIT:170-171].
-This corresponds with a higher proportion of Indigenous and Chinese items with little European materials found in San Salvador [@Berrocal2018], indicating cultural entanglement.
+The area surrounding the fortress was inhabited by Indigenous groups, the Taparris and the Quimaurris, who had lived at the site before being expelled by the Spanish [@Borao2001, SIT:165, p.81].
+The relations between the Spanish and Indigenous people may not have involved extensive colonial control or intercultural domination, as Spanish territorial control appears to have been limited to the fort.
+Instead, the Spanish relied on exchange of local goods and food surpluses with Indigenous communities based on historical documents [@Borao2001, SIT:170-171].
+This is consistent with the archaeological evidence from San Salvador, which shows a higher proportion of Indigenous and Chinese artifacts, with relatively few European materials [@Berrocal2018], indicating cultural entanglement.
 
 <!-- the Dutch domination, how and why colonial powers were able to exert control -->
-In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662.
-Compared to the Spanish, the Dutch was more strategic in consolidating their power by establishing political and economic institutions, forming alliances with Indigenous communities, and conquering resistant communities [@Kang2016; @Andrade2007].
-For examples, the colonial institutions to Indigenous people include taxation, tribute system, village-leasehold system, census investigations, annual gatherings (Dutch: Landdag), and Indigenous representative from each aliened village [@Kang2016; @Andrade1997; @Andrade2007].
-Among those institutions, the tribute system operated from 1642 to 1647 and the village-leasehold system introduced in 1644 had great impacts on Indigenous economy and the environment they lived in.
-Based on the tribute system, each household in allied communities was required to pay rice or deer products annually, especially deerskin.
-The amount of deerskin collected from Indigenous communities in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007].
-Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on Indigenous lands during certain periods [@Heyns2001].
-From 1633 to 1661, the average number of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017].
-Those colonial activities might affect the economy of indigenous people since deer was an important resource in prehistoric Taiwan [@Chen2000animal].
+In 1642, the Dutch defeated the Spanish and took control of San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662.
+Compared to the Spanish, the Dutch was more strategic in consolidating their power by establishing political and economic policies, forming alliances with Indigenous communities, and conquering resistant ones [@Kang2016; @Andrade2007].
+Examples of colonial institutions imposed on Indigenous people include taxation, a tribute system, a village-leasehold system, census investigations, annual gatherings (Dutch: Landdag), and Indigenous representative selection [@Kang2016; @Andrade1997; @Andrade2007].
+Among those colonial policies, the tribute system (1642-1647) and the village-leasehold system introduced in 1644 had profound impacts on Indigenous economies and the environment.
+Under the tribute system, each household in allied communities was required to pay deer products or rice to the Dutch annually, with deerskins being a primary item.
+In 1646, the amount of deerskin collected from Indigenous communities in northern Taiwan was 7,000 pieces [@Andrade2007].
+Moreover, to increase deerskin yields, the Dutch implemented a hunting license system (1636-1645), which permitted Han Chinese to hunt deer on Indigenous lands during specific periods [@Heyns2001].
+From 1633 to 1661, the average annual export of deerskins from Taiwan reached 67,000 [@Koo2011; @Cheng2017].
+These colonial activities likely impacted the economy of Indigenous peoples, because deer was an important resource in prehistoric Taiwan [@Chen2000animal].
 
 <!--variation of colonial powers across regions and indirect colonization at Kavalan-->
-Although the Dutch power expanded across nearly whole Taiwan island in the mid-17th century, the extent of their colonial dominance differed greatly from region to region.
-Colonial power over regions far from the European fortress were usually difficult and unsuccessful.
-Yilan in northeastern Taiwan is an example of minimal European dominance or indirect colonization for most of the time [@Kang2012].
-As an plain located around 70 kilometers away from the European colonies with mountains in between, Yilan was hardly accessed by land but mostly from sea.
-The main Indigenous group in Yilan was the Kavalan.
-Given its strategic location on the route to Torrobouan, a place believed having gold mines, the Dutch was eager to integrate Yilan under their control [@DZI2000; @Kang2016, p47].
+Although Dutch power expanded across nearly the entire Taiwan in the mid-17th century, the extent of their their colonial dominance varied greatly by region.
+Colonial control over areas far from the Dutch fortresses was often difficult and unsuccessful.
+Yilan in northeastern Taiwan is an example of weak colonial control or indirect colonization for most of the time [@Kang2012].
+Yilan is located around 70 kilometers from European colonies with mountains in between, which made it difficult to access by land and was primarily reached by sea.
+In Yilan, Kavalan people was the main Indigenous group since the Iron Age.
+Due to the strategic position of Yilan on the route to Torrobouan, a place believed to have gold mines, the Dutch were eager to bring Yilan under their control [@DZI2000; @Kang2016, p47].
 In 1644, the Dutch defeated several Kavalan communities and incorporated them into their colonial territory [@Kang2016].
-Instead of engaging directly with the Kavalan people, the Dutch relied mostly on the assistance of the mediate, the Basay people active in northern Taiwan [@Kang2003], to coordinate and collect tributes from Yilan.
-This indirect approach allowed the maintenance of autonomy of the Kavalan people, and was even able to negotiate tributes requested by the VOC [@Kang2016].
+However, instead of directly engaging with the Kavalan people, the Dutch relied on the assistance of intermediaries, the Basay people, who were active in Northern Taiwan, to collect tributes [@Kang2003].
+This approach of indirect control allowed the Kavalan to maintain a degree of autonomy and even negotiate tributes [@Kang2016].
 
 <!-- the pre-existing trade network in norththern Taiwan-->
-Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the 14th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005].
-Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the Basay people [@Chen2005; @Andrade2007].
-This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].
-Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003].
-This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005].
-This trade network was then intensified during the European presence, and the participation of Kavalan in this network was driven by the pursuit of foreign goods, and also likely colonial pressure from European powers.
-This unique historical and spatial context makes Yilan an ideal location to study Indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp.
+Before the arrival of Europeans, an inter-insular trade network had been established since the 15th century in northern Taiwan, involving Chinese merchants and Indigenous groups in Tamsui and Keelung [@Chen2005].
+Chinese traders transported goods such as textiles, porcelains, beads, and metal items in exchange for raw materials from Taiwan, including gold, sulfur, rattan, and animal hides and meats [@Chen2005; @Andrade2007].
+This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom around 15-16th centuries and later integrated into European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].
+Yilan was part of this pre-existing trade network, where local Kavalan people exchanged rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003].
+This trade network intensified during the period of European presence, which is consistent with the large quantities of imported foreign goods found at archaeological sites across Yilan, indicating frequent trade activity in the 17th century [@TaiwanHistory2005].
+This unique historical and spatial context makes Yilan an ideal location for exploring Indigenous responses to foreign presence and the dynamics of trade networks in a pericolonial context [@Acabado2017; @Kang2016, pp.
 31-78; @Trabert2017].
@@
 <!--introduction of Kiwulan and the population-->
-Kiwulan, located in the northern part of Yilan (@fig-KWL-map), is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
-A large scale salvage excavation was conducted at Kiwulan that exposed a total of 262 excavation units (4 * 4 m), revealing abundant artifacts and numerous features, including middens, burials, ditches, and post-holes with wooden posts in situ.
-The excavtion suggests that Kiwulan was a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004; @Chiu2004].
-The main inhabitants of Kiwulan were the Kavalan people based on oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004].
-Kavalan people is the main Indigenous group in Yilan during the late Iron Age and the early modern period.
-According to a Dutch census conduced in the mid-17th century, there were about 40 Kavalan settlements with approximately 10,000 people in Yilan.
-Kiwulan, written as "Kibannoran" in Dutch records, was the largest settlement, with 160 households and 711 people [@Nakamura1938].
-In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which eventrually led to the relocation of Kavalan people to Hualien in eastern Taiwan [@Zhan1995].
+Kiwulan, located in northern Yilan (@fig-KWL-map), is a multi-component site, with its upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
+A large-scale salvage excavation at Kiwulan uncovered a total of 262 excavation units (4 x 4 m), revealing abundant artifacts and numerous features, including middens, burials, ditches, and post-holes with wooden posts in situ.
+The excavation suggests that Kiwulan was a well-developed settlement with longhouses (approximately 4 x 8 m) arranged side by side, and burials located to the north of the residential area [@Chen2004].
+Based on oral history and the similarities between Kavalan material culture recorded in historical documents and the archaeological assemblage, Kiwulan is identified as a Kavalan settlement [@Chen2004].
+The Kavalan people were the main Indigenous group in Yilan during the late Iron Age and early modern period.
+According to a Dutch census from the mid-17th century, there were approximately 40 Kavalan settlements in Yilan, with a total population of about 10,000 people [@Nakamura1938].
+Among these Indigenous settlements, Kiwulan ("Kibannoran" in Dutch sources) was the largest one with 160 households and 711 inhabitants.
+In the early 19th century, increasing Han Chinese migration gradually led to the relocation of the Kavalan people to Hualien in eastern Taiwan [@Zhan1995].
+This is reflected in the dramatic decline of Kiwulan’s population to fewer than 100.
 
 <!--archaeological materials-->
-The most common artifact found at Kiwulan was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005].
-In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts and ornaments, were largely found in both the residential area and burials.
-Due to no evidence of metallurgy and local production of beads, these items are believed to have been acquired through exchange, primarily with the Basay people active in northern Taiwan.
-This indicates that Kiwulan has continuously participated in regional trade networks closely associated with northern Taiwan since the 14th century and influenced by the long-distance trade during the 17th century [@Wang2007].
-Moreover, the use of foreign goods appears to have cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021; @Cheng2008].
+The most common artifact found at Kiwulan was locally made pottery with geometric surface patterns, a typical type of ceramic in Yilan [@TaiwanHistory2005].
+In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts and ornaments, were largely found in both residential area and burials.
+Since there is no evidence of local metallurgy or bead production, these items were likely obtained through exchange.
+This suggests that Kiwulan maintained continuous participation in regional trade networks linked to northern Taiwan, which were further influenced by long-distance trade in the 17th century [@Wang2007].
+Moreover, foreign goods such as beads were considered prestige items or used for ritual purposes, and previous studies suggest that their introduction may have led to increasing social differentiation [@Wang2020ornament; @Wang2021; @Cheng2008].
 
@@
 Abundant animal bones, shellfish, and plant remains were uncovered at Kiwulan, reflecting the diversity of natural resources in the plain and an environment close to estuary.
-The animal resource includes deer, suids, cattle, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004].
-Rice farming seems to be common but likely limited to a household scale rather than intensive agriculture according to historical records [@Chen1963].
-Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments.
-Based on those archaeological remains and environments, Kiwulan residents should have practiced a mixed subsistence that includes rice farming, hunting, and fishing for wild resources.
-Artifacts associated with those subsistence practices were also found at Kiwulan, such as iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities.
-However, the resource use across different time periods still need further investigations to better understands the subsistence adaptation relative to foreign presence and environmental changes.
+The faunal remains include deer, suids, cattle, birds, fish, and shellfish, while plant remains primarily consist of rice, peach, pandanus, and cucurbits [@Chen2004].
+Rice farming appears to have been practiced, but likly limited a household scale rather than as intensive agriculture, according to archaeological evidence and historical records [@Chen1963].
+The shellfish remains include bivalves such as clams and oysters, and univalves from freshwater and brackish water environments.
+Based on the archaeological remains and local environment, it is likely that Kiwulan residents practiced a mixed subsistence strategy, including rice farming, hunting, and fishing.
+Artifacts associated with these subsistence practices were also found, such as iron arrowheads for hunting, iron sickles for harvesting, and wooden paddles for water-related activities.
+Further research is still needed to examine resource use across different time periods and to better understand how subsistence strategies responded to foreign influences and environmental changes.
 
@@
 
-To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six cultural layers (CL6-CL1) spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
+To explore changes in faunal remains over time, the original arbitrary levels of Kiwulan were assigned into six natural strata spanning from the 14th century to the 19th century, referred as cultural layers (CL1-Cl6) in this study.
+Each stratum was determined according to documentation of excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
 The diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks relating to the 19th century [@Wang2011; @Hsieh2009].
-Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could still offer a relatively robust framework for comparing archaeological records before and after the foreign influence.
+Although precisely identifying the stratum representing the specific time of European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could still offer a relatively robust framework for comparing archaeological records before and after the foreign influence.
 Among these six temporal layers, the fourth cultural layer (CL4) can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second cultural layer (CL2) corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.
@@
 <!--results and chronology -->
-The sampled faunal assemblage from the major residential area at Kiwulan suggests a continuation of sika deer consumption from CL6 to CL2, and then shifted to bovine in CL1.
-Despite deer was a substantial part of subsistence at Kiwulan, variations in their quantity, ubiquity, butchering patterns, and proportions relative to other animals show changes that might relative to foreign impacts.
-Based on radiocarbon dates and diagnostic artifacts indicating specific time [@Wang2020ornament; @Chen2007], the cultural layers were further grouped into temporal sequences in relation to foreign presences for further discussion.
-The temporal sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the arrival of Han-Chinese immigrants (CL2, CL1).
-It should be noted that the European period does not mean a solid European colonial control over Kiwulan, but a broad impact from European occupation in Northern Taiwan and associated colonial economic activities on indigenous people in Yilan, such as getting involved in the long-distance trade network, tribute paying, and suffered from occasional European attacks.
-The small sample size after separating them into different periods might limits a comprehensive interpretation.
-However, comparing the results in a continuous sequences still provide evident patterns reflecting changes in animal use.
+The faunal assemblage from the major residential area at Kiwulan suggests a continuous reliance on sika deer from CL6 to CL2, followed by a shift to bovine in CL1.
+Despite deer were the main animal resource during all times, variations in quantity, ubiquity, butchering patterns, and proportions relative to other animals suggest changes relative to foreign presence.
+The cultural layers were grouped into four temporal sequences in relation to foreign presences for the following interpretation.
+The sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the Chinese period with large Han Chinese settled down (CL2, CL1).
+The therm "European period" does not imply direct colonial control over Kiwulan but rather the broad impact of European colonial activities in northern Taiwan, such as long-distance trade, tribute systems, and occasional attacks affecting Indigenous communities in Yilan.
+Although the small sample size across periods limits comprehensive interpretation, examining results in a continuous sequence still reveals distinct patterns in animal use.
 
 <!--summary of overall observations across time periods -->
-The sampled area at Kiwulan shows a large quantity of faunal remains in the European period, followed by the late pre-European period.
-Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but in only half the quantity.
-This might indicate consistency in the  regularity of deer consumption at the settlement scale before and after European presence, but deer utilization was intensified in the European period.
-During the European period, in addition to the predominance of deer, an increased relative proportion and ubiquity of suid suggests a more regular use of suid than in other periods.
-The dramatic decrease in deer bones after the European period might be related to both the decline of the Kiwulan population and a decrease in the wild deer population.
-Sika deer populations in western Taiwan have significantly shrunk since the 18th century because of habitat loss and overhunting [@Chen2000animal].
-An increase in muntjac use may suggest they served as a substitute, replacing the role of suids in earlier periods.
-Despite this, deer remained a predominant resource for local consumption at Kiwulan during the post-European and early Chinese period.
-Bovine appeared in the pre-European period with just a few evidence (yellow cattle), but became the only animal (water buffalo) used at Kiwulan in the late Chinese period.
-Since bovine was not domestic aniamls that were beleived introduced from mainland China, this suggests a Chinese influence and the relocation of the Kavalan people to Hualien due to population pressures from Chinese immigration [@Zhan1995].
+The sampled area at Kiwulan contains the large quantity of faunal remains from the European period, followed by the late pre-European period.
+Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but only half the quantity.
+This might indicate a consistent pattern of regular deer consumption at the site both before and after European presence, with intensified deer use during the European period.
+In addition to the dominance of deer, the increased relative proportion and ubiquity of suids in the European period suggests a more regular use of suids compared to other periods.
+The dramatic decline in deer bones after the European period may be associated with both a reduction in the Kiwulan population and the decrease of wild deer.
+Sika deer populations in western Taiwan largely declined from the 18th century due to habitat loss and overhunting [@Chen2000animal].
+A slight increase in muntjac use suggests that they may have served as a substitute, replacing suids as a secondary animal resource from earlier periods.
+Despite these shifts, deer remained a primary resource for local consumption at Kiwulan during the post-European and early Chinese periods.
+Bovines were present in small numbers (yellow cattle) in the pre-European period, but became the sole large animal (water buffalo) used at Kiwulan in the late Chinese period.
+Since bovines were absent from prehistoric sites in Taiwan and are widely believed to have been introduced from mainland China in historical period [@Chen2000animal], their increasing presence suggests Chinese influence and aligns with the Indigenous relocation to eastern Taiwan since the 19th century due to Chinese population pressures [@Zhan1995].
 
 <!--skeletal part frequencies and butchery patterns of deer-->
-Although deer was the primary resource at Kiwulan, differences in skeletal completeness across periods suggests changes in butchery locations and cooking practices.
-There were more complete carcasses in the European and late pre-European period compared to the post-European and Chinese periods.
-Moreover, the substantial rise in fragmented head parts (cranium and mandible) during the European period indicates greater use of soft tissues, such as brains and tongues, or the transport of more intact deer to the settlement for processing.
-This aligns with the historical account in 1603, Record of Formosa (Dongfan ji) [@Chen1603], which describes deer tongue consumption and group hunting for deer in western Taiwan.
+Differences in deer skeletal completeness across periods suggests changes in butchery locations and cooking practices.
+There were more complete carcasses in the European and late pre-European periods compared to the post-European and Chinese periods.
+Moreover, the substantial rise in fragmented head parts (cranium and mandible) during the European period indicates increased use of soft tissues, such as brains and tongues, or the transport of more intact deer to the settlement for processing.
+This is consistent with the historical account in 1603, Record of Formosa (Dongfan ji) [@Chen1603], which describes the consumption of deer tongues and group hunting practices that could facilitate the transportation of deer.
 Changes in cooking practices are also evident.
-The decline in burned bones over time with a lowest percentage in the European period suggests a shift from roasting to other cooking methods, such as boiling.
-The presence of small chopped bones that could fit in local ceramics at Kiwulan indicates boiling as a possible practice since the European period.
-This could be further confirmed through organic residue analysis on pottery in the future.
-Other meat preparation such as drying (or salting and smoking) cannot be ruled out [e.g., @Spyrou2019].
-Although deer vertebrae and ribs, portions potentially used for drying [@Friesen2001; @Binford1978], were in low frequencies at Kiwulan, dry meat consumption or exchange by the Kavalan people was reported in historical records [@Chen1963].
-The shift in cooking practices might result from increased need for deer, which could have led people to adopt boiling due to its additional benefits, such as maximizing residual meat use, facilitating grease rendering, and improving quality of lean meat [@Lupo1997], and drying for meat preserving and transporting.
+The decline in burned bones over time, with the lowest percentage in the European period, suggests a shift from roasting to other cooking methods, such as boiling [@Lupo1997].
+The size of chopped bones, which could fit into local ceramics at Kiwulan, might be associated with boiling practices.
+This could be further confirmed through organic residue analysis on pottery.
+Other meat preparation methods, such as drying through salting and smoking, are also possible [e.g., @Spyrou2019].
+Although vertebrae and ribs, typically used for drying [@Friesen2001; @Binford1978], were present in low frequencies at Kiwulan, historical records mentioning the consumption or exchange of dried meat by the Kavalan people supports the possibility [@Chen1963].
+The shift in cooking practices might result from the increased need for deer, leading people to adopt boiling for its benefits, such as maximizing the use of residual meat, facilitating grease rendering, and improving quality of lean meat [@Lupo1997], while drying helped preserve and transport the meat.
 
-A consistent pattern in deer part frequencies was also identified throughout all periods.
-There were high frequencies of upper limb bones but low frequencies of vertebrae and ribs, suggesting selectively butchered and preferences of meaty portions.
-Moreover, among those meaty portions, front quarters (humerus and radius) are always more abundant than hindquarters (femur and tibia).
-One possible explanation is that femurs and tibiae were heavily fragmented for marrow extraction due to their high yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage.
-Marrow extraction can be further supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation in general.
-Another likelihood is that some hindquarters might be used for trading, as Dutch records mention deer legs and meats as commodities in addition to hides [@Blusse2006, pp.220-224], but without specifying which legs.
-Deer legs imply bone-meat packages that would result in a low frequencies of bones of hindquarters or front quarters.
-Considering the overall high proportion of meaty portions at Kiwulan, deer were likely processed not only for local consumption but also for trade.
-The lower frequencies of hindquarters than front quarters might be attributed to both marrow extraction and export of deer products, rather than a sole contribution.
+A consistent pattern of high frequencies of upper limb bones but low frequencies of vertebrae and ribs was identified across all periods.
+This might suggest selective butchering and a preference for meaty portions.
+Moreover, among the meaty portions, front quarters (humerus and radius) were consistently more abundant than hindquarters (femur and tibia).
+One possible explanation is that femurs and tibiae were heavily fragmented due to marrow extraction, as these bones have high marrow yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage.
+Evidence for marrow extraction is supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation overall.
+Another explanation is that some hindquarters were used for trade, as Dutch records mention that deer legs and meat were commodities in addition to hides [@Blusse2006, pp.220-224], but they did not specify which legs.
+If deer legs, particularly hindquarters, were traded as bone-meat packages, this could account for the lower frequencies of hindquarter bones compared to front quarters.
+While this hypothesis requires further evidence, the overall high proportion of meaty portions at Kiwulan suggests that deer meat might be not only used for local consumption but also processed for trade.
+Thus, both factors could contribute to this pattern, rather than a single factor alone.
 
 <!--deer hides-->
-Deerskin trade had prevailed since the 14th century in East Asia, and monopolized by the Europeans in the 17th century, when large hides were exported to Japan, especially from western Taiwan [@Cheng2017; @Hollmann1991; @Wang1981].
-The cut marks identified on deer lower limbs at Kiwulan confirm the removal of deer skins and likely being traded with external traders.
-Abundant foreign goods at Kiwulan, including ornaments, porcelains, and metal items, supports frequent exchanges between local residents and outsiders [@Wang2020ornament; @Chen2007].
-This suggests that not only western Taiwan but also northeastern Taiwan was intensively involved and impacted by the long-distance trade.
-Although lower limbs with skinning cuts did not show a substantial increase in the European period, the disproportionate increase in head parts might hint at a likelihood of brain tanning when preparing hides in addition to tissue consumption.
-Brain tanning is a traditional way for softening and preserving hides in many Indigenous cultures, such as North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016].
-The present-day Bunun group in the central highlands of Taiwan has revived hide tanning based on oral history, primarily using peanuts or pig lard as tannnig materials, with a few examples using animal brains [@Li1999, pp.
+Deerskin trade had been prevalent in East Asia since the 14th century and was monopolized by Europeans in the 17th century when a large number of hides, especially from western Taiwan, were exported to Japan [@Cheng2017; @Hollmann1991; @Wang1981].
+The cut marks identified on deer lower limbs at Kiwulan confirm the removal of deer skins, likely for trade with external traders.
+Moreover, the abundance of foreign goods at Kiwulan, including ornaments, porcelains, and metal items, further supports the frequent exchanges between local residents and outsiders [@Wang2020ornament; @Chen2007].
+This suggests that northeastern Taiwan, like western Taiwan, was actively involved in and impacted by long-distance trade.
+Although lower limbs with skinning cuts did not show a substantial increase during the European period, the disproportionate rise in head parts might hint at the likelihood of brain tanning when preparing hides.
+Brain tanning is a traditional method for softening and preserving hides in many Indigenous cultures, including those of North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016].
+In Taiwan, the present-day Bunun group in the central highlands has revived hide tanning based on oral history, primarily using peanuts or pig lard as tanning materials, with occasional use of animal brains [@Li1999, pp.
 48; @Xu1994].
-Despite the materials for tanning used by the Kavalan people being unknown due to the loss of traditions, brain tanning is a practicable method given the availability of deer heads from hunting.
+Despite the specific tanning materials used by the Kavalan people remain unknown due to the loss of traditions, brain tanning is a plausible method given the availability of deer heads from hunting.
 
 <!-- population dynamics and arguments-->
-The high population during the European period inferred by pottery could contribute to the increasing animal use overall [@Wang2020pot; @Hsieh2009].
-However, the impacts from foreign presence can be still identified through deer butchery patterns, relative proportions of animals, and taxon ubiquity.
-First, lower limbs with skinning cuts were mostly in the European period, followed by the late pre-European phase, corresponding with the rise of regional trade in the 14-15th century and was later intensified in the 17th century.
-Second, if the high frequencies of deer parts, especially heads for brains or tongues consumption, during the European period solely resulted from food stress due to population growth, a parallel rise in marrow extraction would be expected.
-However, long bones with fracture patterns were 10-30 % lower during the European period and the previous phase than in other periods.
-Third, population growth should have led to a more proportional increase in both large and small animals.
-Despite a high taxonomic diversity during the European period, suid presents a disproportionate rise in their ubiquity, quantity, and proportion.
-This indicates that deer products trade may have driven Indigenous people to rely more on suids as an alternative prey, likely because of their comparable body size and meat return rates to deer [@Broughton2011prey].
-Suids as a secondary resource after deer in Indigenous cultures were observed in western Taiwan [@Chuang2021].
-This hints the role of suid as a supplement when deer was not sufficient or available.
-After the Dutch left, deerskin trade continued but in a small scale with around 10,000 pieces exported annually, which is much less than 100,000 pieces in the European period [@Nakamura1997; @Tsao2011].
-The small trade market and low population at Kiwulan might lead to a reduced reliance of suid, as shown by a low proportion of suid in the post-European period and eventually disappeared during the Chinese period.
+The high population during the European period, as inferred by the frequencies of pottery, may have contributed to the overall increase in animal use [@Wang2020pot; @Hsieh2009].
+However, the impact of foreign presence can be still be identified in deer butchery patterns, the relative proportions of animals, and animal ubiquity.
+First, lower limbs with skinning cuts were mostly found in the European period, followed by the late pre-European phase, aligning with the rise of regional trade in the 14th–15th centuries and its intensification in the 17th century.
+Second, if the high frequency of deer heads during the European period was solely due to food stress from population growth, leading to increased brain or tongue consumption, a parallel rise in marrow extraction would be expected.
+However, long bones with fracture patterns showing 10–30% less frequent during the European and late pre-European periods than in other periods.
+Third, population growth would have likely led to a more proportional increase in both large and small animals.
+Despite high taxonomic diversity during the European period, suids showed a disproportionate rise in ubiquity, quantity, and proportion.
+This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and meat return rates to deer [@Broughton2011prey].
+The role of suids as a secondary resource after deer has been observed in Indigenous cultures in western Taiwan [@Chuang2021], suggesting they may have served as a supplement when deer were scarce or limited, possibly because some deer were reserved for trade.
+
+After the Dutch left, the deerskin trade continued but on a much smaller scale, with around 10,000 pieces exported annually in the 18th century, which is far fewer than the 100,000 pieces exported during the European period [@Nakamura1997; @Tsao2011].
+The shrinking trade market and declining population at Kiwulan may have contributed to the reduced reliance of suids, as shown in their lower proportion in the post-European period and eventual absence in the Chinese period.
+It should be noted that muntjac were consistently part of local subsistence, though in small numbers, until the early Chinese period.
+Their skins were used by many Indigenous groups such as the Tsou and Bunun for making clothing, similar to the use of deerskins [@Li1999], which may have also been practiced at Kiwulan.
+Although both suids and muntjacs served as substitutes for deer, their differing frequencies and ubiquity over time suggest that suids were the primary supplementary meat source, while muntjact were less frequently consumed but may have been valued for their hides in daily clothing.
 
 <!--indigenous reaction and resilience-->
-Compelling the zooarchaeological evidence and descriptions from historical records, the substantial increase in deer bones in the European period not just simply the result of population growth, but reflect intensified hunting driven by deer trade.
-Deer is the first priority of food source and commodities at Kiwulan since the late pre-European period and their economic benefits were intensified during the European period.
-It should be recognized that the role of the Dutch is both colonist who demanded tributes from Yilan, and another trader who were incorporated into the pre-existing inter-insular trade involving Han Chinese merchants and other indigenous groups in Northern Taiwan [e.g., @Torrence2000].
-Local Indigenous people might take a more active participation in the trade rather than resisted due to the benefits of obtaining foreign necessities (e.g.
-metal knives and pan) and goods (e.g.
-ornaments and porcelains) that were highly valued in local cultures [@Wang2020ornament].
-Despite an indirect and weak Dutch colonial control in Yilan, the stress from colonial powers cannot be totally neglected given the local experiences of the Dutch attacks occasionally.
-Everyday practices play an essential role for maintaining Indigenous traditions within a colonial setting or frequent cultural exchange [@Silliman2009; @Noe2023].
-The persistence of deer consumption with similar butchery patterns over time at Kiwulan reflect a maintained tradition even after other animals were introduced, for example, bovine in the pre-European period.
-The increased reliance on suid during the European period could be viewed as a strategy to accommodate the high demands of deer products from trade network.
+The Dutch played a dual role in Yilan as both colonizers demanding tribute, and traders integrated into the pre-existing inter-insular trade network involving Han Chinese merchants and different Indigenous groups [e.g., @Torrence2000].
+That means, on one hand, Indigenous group in Yilan faced pressures from colonial encounters, such as occasional Dutch attacks, despite the generally indirect and weak colonial for most of the time.
+On the other hand, rather than resisting, local Indigenous people might participate volentarily in trade to acquire foreign necessities (e.g.
+metal knives and wares) and goods (e.g.
+ornaments and porcelains) that were highly valued in local culture [@Wang2020ornament].
+The intensified hunting during the European period was not just a result of population growth but was driven by the deerskin trade.
+Zooarchaeological evidence from Kiwulan indicates that deer had been a primary food source and trade commodity since the late pre-European period, with its economic significance increasing during European presence.
+In addition, ethnographic studies suggest that deer hunting is closely tied to masculinity and coming-of-age ceremonies in Kavalan culture.
+That means, deer consumption at Kiwulan reflects not only daily practices but also the continuation of cultural traditions.
+Everyday practices play an essential role in maintaining Indigenous traditions within colonial settings and periods of frequent cultural exchange [@Silliman2009; @Noe2023].
+In the pericolonial context of northeastern Taiwan, the continued consumption of deer, even as other animals like bovines were introduced in the pre-European period, highlights the persistence of cultural practices, while the increased reliance on suids at Kiwulan likely represents an adaptive and temporary strategy to meet high trade demands for deer products.
 
 # Conclusion
-By contextualizing zooarchaeological evidence into time periods relative to foreign presence, including the Europeans in the 17th century, and Chinese immigrants from the late 19th century, this study reveals the participation of Indigenous groups in regional trade work and adaptation strategies.
-The frequencies, butchering patterns, and human modifications on deer bones confirms the active deerskin trade since the 14th century and intensified during the European period.
-The results demonstrates even Indigenous people in the peripheral area with relatively weak colonial control were also profoundly impacted by colonial activities than previously thought.
-The continued prioritization of deer consumption at Kiwulan suggests a cultural persistence of food practices, while the strategic reliance on suids during the European period demonstrates adaptation to high demands for deer products, and later incorporating more muntjac in response to overhunting of sika deer.
-This highlights how Kiwulan residents reacted to foreign presence and environmental changes and demonstrated cultural resilience.
-The long-term dynamics of animal use in northeastern Taiwan can enhance our understanding of Indigenous reaction within a specific social-cultural context.
-The role of Indigenous people might be more active in the exchange network considering the large amount of foreign goods that were incorporated into local cultures and used as prestige items.
-A more comprehensive investigation across Indigenous settlements in northeastern Taiwan and including other dietary evidence, such as botanical remains and organic residue analysis in the future could provide a deeper insights into the regional dietary patterns and human-environment dynamics.
+By situating zooarchaeological evidence within the context of foreign presence, the Europeans in the 17th century and Chinese immigrants in the 19th century, this study reveals Indigenous participation in regional trade networks and their adaptation strategies.
+The frequencies, butchery patterns, and modifications on deer bones from Kiwulan indicates an active deer processing that was intensified during the European period.
+Deer as a primary animal resource at Kiwulan suggests cultural continuity, while the strategic reliance on suids during the European period reflects adaptation to high trade demands for deer products.
+Additionally, the incorporation of muntjac into the diet during the Chinese period likely resulted from the overhunting of sika deer.
+These findings highlight how Kiwulan residents navigated foreign influence and environmental changes while maintaining cultural traditions.
+Through the long-term dynamics of animal use at Kiwulan, this research demonstrates that even Indigenous communities in peripheral areas with relatively weak colonial control were profoundly impacted by colonial activities than previously thought.
+Future research expanding to other Indigenous settlements in northeastern Taiwan, with additional dietary evidence such as botanical remains and organic residue analysis, could provide deeper insights into regional dietary patterns and human-environment interactions.
 
 # Acknowledgements
-
-I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the zooarchaeological remains used in this study and the archaeological team for the early documentation of faunal remains that provides a foundation for this study.
+I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the zooarchaeological remains used in the analyses and the archaeological team for the early documentation of faunal remains that provides a foundation for this research.
 Special thanks to Chien-Hsiang Lin at the Biodiversity Research Center, Academia Sinica, for his guidance and assistance in identifying fish bones.
 I also appreciate Jui-Yang Tsou for his valuable comments and assistance during the identification process.
-I would like to express my gratitude to Yin-Ru Chiang and his team at the Biodiversity Research Center, Academia Sinica, for providing me a lab space to work on data and finilizing this work.
+I would like to express my gratitude to Yin-Ru Chiang and his team at the Biodiversity Research Center, Academia Sinica, for providing me a lab space to work on data and finalizing this work.
 I am also grateful to the reviewers of this manuscript for their comments.

### 2025-03-09 analysis/paper/paper.qmd
draft editing

@@
 ---
-title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
+title: "Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
 author:
@@
       orcid: 0000-0001-9060-2125
-      institute: Academia Sinica
+      institute: Institute of History and Philology, Academia Sinica
 title-block-published: "Last updated"
@@
 highlights: |
-  Indigenous community in northeastern Taiwan engaged in regional deer trade were identified from faunal remains.
-  The high freqiencies of deer bones at Kiwulan reveals continously cultural persistence in deer consumption.
-  Changes in deer butchery patterns and cooking practices indicates an intensified hunting during the European colonization period.
-  Increased reliance on suids during the European period suggests adaptive strategies to sustain local subsistence.
+ Faunal analysis from Kiwulan suggests Indigenous participation in the European long-distance trade network.
+Evidence indicates the continued cultural persistence of deer consumption.
+Shifts in deer butchery patterns and cooking practices suggest intensified hunting during the European colonization period.
+The increased reliance on suids reflects adaptive strategies for sustaining local subsistence.
+
 ---
@@
 # Materials and Methods
-The faunal assemblages from Kiwulan analyzed in this study were sampled from 40 adjacent units at the residential area excavated during the first season (2001-2003) [@Chen2007].
-These 40 units were selected because of their location that were believed to be the major residential area with thick archaeological deposits and relatively intact archaeological contexts without disruption from modern construction.
-The spatial continuity can provides a consistent framework for understanding food practices over time on a settlement scale.
-Also, the same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences.
-I mainly focus on vertebrate species due to their economic values in the trade market.
-The contexts of faunal remains include dwelling spaces and middens that were associated with daily dietary practices, whereas those from burials that might relate to ritual activities are excluded.
+The faunal assemblages analyzed in this study were sampled from 40 adjacent units in the residential area excavated during the first season (2001-2003) [@Chen2007].
+These 40 units were selected because they represent the main residential area, with thick archaeological deposits and relatively intact contexts without modern disturbance.
+The spatial continuity of these units provides a consistent framework for understanding food practices over time at the settlement scale.
+Additionally, the use of the same excavation and recovery methods allows for comparison between assemblages from different temporal sequences.
+I primarily focus on vertebrate species due to their economic importance in local subsistence and trade.
+The faunal remains analyzed were from dwelling spaces and middens, as these are associated with daily subsistence practices, while remains from burials are excluded due to their potential ritual purposes rather than daily practices.
 
-To explore changes in faunal remains over time, the original arbitrary levels of Kiwulan were assigned into six natural strata spanning from the 14th century to the 19th century, referred as cultural layers (CL1-Cl6) in this study.
-Each stratum was determined according to documentation of excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
-The diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks relating to the 19th century [@Wang2011; @Hsieh2009].
-Although precisely identifying the stratum representing the specific time of European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could still offer a relatively robust framework for comparing archaeological records before and after the foreign influence.
-Among these six temporal layers, the fourth cultural layer (CL4) can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second cultural layer (CL2) corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.
+To explore changes in faunal remains over time, the original arbitrary excavation levels (10 cm) at Kiwulan were assigned into six natural strata or cultural layers (CL1-CL6), spanning from the 14th century to the 19th century.
+Each stratum was determined based on documentation of excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
+The diagnostic materials include pipes and stoneware commonly found at sites associated with European expansion in Southeast Asia, such as Anping jars and martavan  [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], and clay bricks associated with Chinese migrants [@Wang2011; @Hsieh2009].
+Although precisely identifying the stratum corresponding to the specific time of European arrival is challenging due to its brief duration, the strata separated based on the criteria above still provide a relatively robust framework for comparing archaeological records before and after foreign presence.
+Among these six temporal layers, the fourth cultural layer (CL4) is associated with the 17th century, when the Europeans occupied northern Taiwan, while the second cultural layer (CL2) corresponds to the 19th century, linked to the large influx of Chinese migrants in Yilan.
 
-Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences.
-Deer antlers without attaching to cranium were excluded since they were less likely related to dietary practices.
-To obtain generalized pattern of animal use, measures of ubiquity and taxonomic abundance were applied.
-Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of consumtion of that taxa [@Mckechnie2016; @Vanderwarker2010].
-Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers.
-Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012].
-Different from MNI that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of remains by considering all fragmented bone assemblages.
-Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site scale with less impacts from sample size [@Grayson1984; @Lyman2008].
-For comparison across animal classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004].
+Only faunal assemblages identified to at least the family level were considered in this study to ensure consistent comparison across time sequences.
+Deer antlers detached from the cranium were excluded because they were less likely to be related to dietary practices.
+To obtain a generalized pattern of animal consumption, measures of ubiquity and taxonomic abundance were applied.
+Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of consumption of that taxa [@Mckechnie2016; @Vanderwarker2010].
+Here ubiquity values are expressed as the percentage of units containing a given taxon to the total number of units per cultural layer.
+Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when estimating the expected NISP at the site [@Grayson2004; @Lyman2012].
+Unlike MNI, which determines the actual numbers of individual animals recovered from remains, NISP offers a broader overview by considering all fragmented bone assemblages.
+Despite criticism for potentially overestimating the number of animals, NISP better captures variation in species abundance with less impacts from sample size [@Grayson1984; @Lyman2008].
+For comparison across animal classes, the normalized NISP were used to avoid biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004].
 The normalized NISP is calculated by dividing NISP by the number of skeletal elements.
 
-Given that deerskin was a primary commodity in local regional exchange and in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies.
-This can shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network.
-Skeletal parts were quantified using NISP and classified into groups based on meat yields and portions: meaty (e.g., upperlimb), mid-meaty (e.g., vertebrae), and less meaty parts (e.g., skull and lower limb) [@Madrigal2004; @Munro2004].
-Teeth without associated cranial elements were excluded to reduce overrepresentation due to their much higher preservation rate and quantities compared to other parts.
-While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002].
-Thus, less meaty elements were considered as additional source of nutrition in the interpretation.
-Similarly, the degree of fragmentation could also imply the extraction of marrow or grease [@Lupo1997], which is calculated by dividing the Minimal Number of Elements (MNE) by NISP.
+Given that deerskin could be a primary commodity in local regional exchange [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies.
+This analysis can provide insights into how Indigenous people utilized deer in response to possible high economic demand when driven by their involvement in regional trade network.
+Skeletal parts were quantified using NISP and classified into groups based on meat yields and portions: meaty parts (e.g., upper limbs), mid-meaty parts (e.g., ribs), and less meaty parts (e.g., skulls and lower limbs) [@Madrigal2004; @Munro2004].
+Teeth without associated cranial elements were excluded to avoid overrepresentation, given their much higher preservation rate and quantities compared to other parts.
+While lower limbs and skull bones contain minimal meat, they are often valued in many cultures for marrow and soft tissues, such as the tongue and brain [@Madrigal2002].
+Thus, less meaty elements were considered as an additional source of nutrition in the later interpretation.
+Similarly, the degree of fragmentation could indicate the extraction of marrow or grease [@Lupo1997] and was calculated by dividing the Minimum Number of Elements (MNE) by NISP.
+In this approach, lower values represent a higher degree of fragmentation, and vice versa.
 
-Deer bone surface modifications relative to human butchering and cooking practices, such as cut marks, fracturing, and burning damage, were recorded, as well as natural marks such as weathering and root etching for understanding taphonomic effects.
-Cut marks on major limb joints, such as shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were quantified to explore butchering patterns over time.
-Moreover, cut marks on lower limbs relative to skinning processes were examined and compared.
-Those marks are characteristic of circular and straight incisions left by the series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].
+Deer bone surface modifications relative to human butchering and cooking practices, such as cut marks, fracturing, and burning damage, were recorded, along with natural marks like weathering and root etching to understand taphonomic effects.
+Cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were quantified to explore butchering patterns over time.
+Moreover, cut marks on lower limbs related to skinning processes, were examined and compared.
+Such skinning marks are characteristic of circular and straight incisions left by a series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].
 
-The data and R code used for these analyses and visualizations are openly available at xxx for research reproducibility and transparency (@Marwick2017; @MarwickandWang2022).
+The data and R code used for these analyses and visualizations are openly available at https://github.com/LiYingWang/kwlenv for research reproducibility and transparency [@Marwick2017; @MarwickandWang2022].
 
@@
 
-The faunal assemblage analyzed in this study from Kiwulan is composed of `r total_NISP` bone fragments based on NISP (@tbl-counts).
+The faunal assemblage analyzed in this study from Kiwulan consists of `r total_NISP` bone fragments based on NISP (@tbl-counts).
 Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`) and water buffalo (`r buff_NISP_per`).
-To further compare general patterns of animal consumption in terms of their economic values over time, animals in the same family with similar body size were grouped together.
-Thus, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were group into the category, bovine.
-Similarly, domestic pig (n = 2, from CL4 and CL6) was integrated into the suid category.
-Rats (Rattus sp.) were excluded from further analyses as they are normally considered commensal animals, and their relatively intact skeletons from one single archaeological context show no direct evidence of being part of the daily diet.
-In general, the faunal remains show the frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 ( (@tbl-counts, @fig-NISP-proportion: A).
-A similar trend was also observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
+To further compare general patterns of animal consumption in terms of their economic values over time, animals within the same family with of similar body size were grouped together.
+Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category.
+Similarly, domestic pig (n = 2, from CL4 and CL6) were integrated into the suid category.
+Rats (Rattus sp.) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes.
+In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 ( (@tbl-counts, @fig-NISP-proportion: A).
+A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
 
@@
 
-For animal classes, the relative abundance shows that mammals are the primary dietary source throughout all layers, with fish as the secondary source in CL4 and birds in CL3 and CL2 (@fig-NISP-proportion: C).
-Fish and birds are both identified from CL5 to CL3; however, their proportions (less than 10%) are much lower compared to mammals.
-The reptile is represented by one piece of turtle shell found in CL4, which might not be a common food content at the site.
-Within those identifiable mammals, deer is the main animal from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated only by bovine (@fig-NISP-proportion: D).
-Suid was present from CL6 to CL3, and had a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5, and was continuously present until CL2.
-In general, there is an increase in animal diversity in CL4 and CL5, especially in CL4, where suid and fish have slightly higher proportions compared to those in other layers.
-Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied from layer to layer, the relationship between the number of taxa and sample size (NISP) was examined by linear regression model.
-The result shows no correlation between the diversity of animals and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), which suggests little bias from potential sample size.
+For animal classes, the relative abundance shows that mammals are the primary dietary source across all layers, with fish as the secondary source in CL4 and birds in CL3 and CL2 (@fig-NISP-proportion: C).
+Fish and birds are both found from CL5 to CL3; however, their proportions (< 10%) are much lower compared to mammals.
+Reptiles are represented by a single piece of turtle shell found in CL4, suggesting that they might not be a common food source at the site.
+Among the identifiable mammals, deer is the main animal from CL6 to CL2 that accounts for at least 85% per layer, while CL1 is predominated by bovine (@fig-NISP-proportion: D).
+Suid remains were present from CL6 to CL3, with a higher proportion in CL4 (`r CL4_boar` %), while muntjac appeared in CL5 and was continuously present until CL2.
+There is an increase in animal diversity in CL4 and CL5.
+Given that the taxonomic richness, or animal diversity, could be impacted by sample size varied across layers, the relationship between the number of taxa and sample size (NISP) was examined by a linear regression model.
+The results show no correlation between animal diversity and sample size ($R^2$ = `r div_lm[1]`, P = `r div_lm[2]`), suggesting minimal bias from sample size variation.
 
@@
 
-@tbl-ubiquity shows deer had been the most widespread animal from CL6 to CL2, with the highest ubiquity in CL4 and CL5 (`r CL4_ubi_deer`), and lowest in CL2.
-The second and third animal regularly used are suid and muntjac.
-Although those two animals reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for suid, and `r CL4_ubi_munt` for muntjac), their presence patterns are slightly different.
-Suid was utilized as early as deer in CL6, declined in CL3, and disappeared in CL2, whereas muntjac showed a cultural lag, appearing one phase later than suid.
-Bovine had a low ubiquity in CL5 and high ubiquity in CL6.
-Small animals (bird, fish, and turtle) had much lower ubiquity than large animals and they had a slightly increase in ubiquity in CL3.
-The result of animal ubiquity correspond with their frequencies (NISP), suggesting that deer is the most common resource at the site, with an increase utilization of suid and muntjac in CL4.
+@tbl-ubiquity shows deer was the most widely used animal from CL6 to CL2, with the highest ubiquity in CL4 and CL5 (`r CL4_ubi_deer`), and the lowest in CL2.
+The second and third most commonly used animals are suid and muntjac.
+Although both reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for suid, ,and `r CL4_ubi_munt` for muntjac), their presence patterns differ slightly.
+Suid was utilized as early as CL6, declined in CL3, and absence in CL2, while muntjac showed a cultural lag, appearing one phase later than suid.
+Bovine had low ubiquity in CL5 and high ubiquity in CL1.
+Small animals (birds, fish, and turtle) had much lower ubiquity than large animals, with a slight increase in ubiquity in CL3.
+The result of ubiquity corresponds with the frequency data (NISP), suggesting that deer was the most common resource at the site, with an increased use of suid in CL4.
 
-Although deer was predominant throughout all temporal sequences, variations in their skeletal parts could be identified.
-Deer bones were grouped into six body portions [@Madrigal2004], including head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges).
-The result (@fig-deer-portion: A) shows that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by upper hindlimb and the less-meaty portions, lower limb and foot.
-The consistency of higher numbers of upper limb and lower limb suggests that those parts should be transported together [e.g., @Stiner1991].
-The generally low frequency of axial elements might be likely due to a high degree of fragmentation, butchery practices, or dietary preferences.
-A different pattern between layers is observed in the frequency of cranial bones, for which a few present in CL6, increased to the highest number in CL4, and then decreased thereafter.
-In addition, lower limb and foot bones appear to be proportionally less in CL2 and CL3 compared to earlier periods.
-Those differences might indicate changes in processing decisions of meat portions of deer.
+Although deer were predominant throughout all temporal sequences, variations in their skeletal parts can be identified.
+Deer bones were grouped into six body portions [@Madrigal2004], including the head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges).
+The results (@fig-deer-portion: A) show that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by the upper hindlimb and the less-meaty portions, lower limb and foot.
+The consistently higher frequencies of upper and lower limbs suggest that these parts were likely transported together [e.g., @Stiner1991].
+The generally low frequency of axial elements may be due to a high degree of fragmentation, butchery practices, or dietary preferences.
+A different pattern is observed in the frequency of cranial bones, which were scarce in CL6, increased to the highest number in CL4, and then decreased thereafter.
+This may reflect changes in deer processing decisions or transportation strategies over time.
 
-Considering the impacts of differential bone density on preservation of skeletal elements, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested by Pearson correlation.
-The values of bone survivorship were collected from @Lyman1994's research on deer, and calculated by dividing NISP by expected NISP, followed by multiplying 100%.
-The result (@fig-deer-portion: B) shows no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias has little impact on the taphonomic processes of bones.
-For the degree of bone fragmentation, CL3 shows higher fragmentation (index = `r frag_CL3_CL4[1]`), while bones in CL4 are less fragmented (index = `r frag_CL3_CL4[2]`).
-However, the fragmentation rates are not significantly differed between layers.
+Considering the potential impact of differential bone density on the preservation of skeletal elements, the relationship between the survivorship of deer skeletal elements and bone mineral density was tested using Pearson correlation.
+The values of bone survivorship were collected from @Lyman1994's research on deer and calculated by dividing NISP by the expected NISP, then multiplying by 100%.
+The results (@fig-deer-portion: B) show no significant positive correlation ($r$ = `r survivor_cor[1]`, P = `r survivor_cor[2]`)) between bone survivorship and bone density, suggesting that density-mediated bias had little impact on the taphonomic processes of the bones.
+As for bone fragmentation, CL3 shows a higher degree of fragmentation (`r frag_CL3_CL4[1]`), while bones in CL4 are less fragmented (`r frag_CL3_CL4[2]`).
 
@@
 
-Surface modifications, including cut marks, burning damage, and fractures were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify).
-Cut marks were present on 50% of bones in CL2, 30% of bones in earlier period (CL6 to CL4), and at a lower proportion of 23% in CL3.
-Most cut marks were located on major limb joints and long bones, suggesting butchering activities related to disarticulation and meat detaching.
-@fig-deer-cutmark shows that at least 50 % of joints in CL4 present cut marks, with a higher proportion of 75% at the hip.
+Surface modifications, including cut marks, burning damage, and fractures, were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify).
+Cut marks appeared on 50% of bones in CL2, 30% in earlier periods (CL6 to CL4), and at a lower proportion of 23% in CL3.
+Most cut marks were located on major limb joints and long bones, indicating butchering activities related to disarticulation and meat removal.
+At least 50% of joints in CL4 presents cut marks as shown in @fig-deer-cutmark.
 A similar pattern was observed in CL5, but with a lower proportion of cut marks on the ankle, knee, and hip compared to CL4.
-Not all joints were found in CL6, CL3 and CL2, making direct comparisons across all layers difficult.
-Nevertheless, the overall high proportion of joints with cut marks suggests intensive processing of deer throughout the site's occupation.
-In addition, cut marks on lower limbs should be associated with skinning (@fig-cutmark-img: e) and were primarily found on metapodials in CL6 (n = 1), CL5 (n = 4), CL4 (n = 6), CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).
-Some joints displayed clean, smooth cut surfaces from chopping (@fig-cutmark-img: a, c), along with narrow V-shaped cut marks observed on most bones, indicating the use of metal tools [@Greenfield1999].
+Direct comparisons across all layers are challenging since not all joints were found in CL6, CL3, and CL2.
+Nevertheless, the overall high frequency of joints with cut marks suggests intensive deer processing throughout the site's occupation.
+In addition, cut marks on lower limbs show traces associated with skinning (@fig-cutmark-img: e).
+Those cut marks were primarily found on metapodials in CL4 (n = 6), CL5 (n = 4), CL6 (n = 1), and CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).
+Some joints displayed clean, smooth cut surfaces from chopping (@fig-cutmark-img: a, c), accompanied by narrow V-shaped cut marks observed on most bones, indicating the use of metal tools [@Greenfield1999].
 
 Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4.
-Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percurssion scars [@Haynes1983; @Gifford2018; @Fisher1995], were observed frequently on long bones across different layers (@fig-fracture-img).
+Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percussion scars [@Haynes1983; @Gifford2018; @Fisher1995], were frequently observed on long bones across different layers (@fig-fracture-img).
 These fractures indicate intentional marrow extraction at Kiwulan.
 CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5.
-Chewing marks were observed on a few bones represented by a series of superficial elongated scores and circular depressions (@fig-fracture-img: e)[@Bello2023], suggesting carnivore gnawing.
-In addition, natural alterations to bones resulting from post-deposition impacts were also observed, such as weathering and root etching [@Gifford2018natural], on approximately one-third of the bones across all layers (@fig-fracture-img: c,d; @tbl-deer-modify).
-Those post-depositional impacts might have led to an underestimation of cut marks in general.
+Chewing marks, represented by a series of superficial elongated scores and circular depressions (@fig-fracture-img: e) [@Bello2023], were observed on a few bones, suggesting the evidence of carnivore gnawing.
+Additionally, natural post-depositional alterations, such as weathering and root etching [@Gifford2018natural], were observed on approximately one-third of the bones across all layers (@fig-fracture-img: c, d; @tbl-deer-modify).
+These post-depositional impacts may have contributed to an underestimation of cut marks.
 
@@
 This could be further confirmed through organic residue analysis on pottery.
-Other meat preparation methods, such as drying through salting and smoking, are also possible [e.g., @Spyrou2019].
+Other meat preparation methods, such as drying, are also possible [e.g., @Spyrou2019].
 Although vertebrae and ribs, typically used for drying [@Friesen2001; @Binford1978], were present in low frequencies at Kiwulan, historical records mentioning the consumption or exchange of dried meat by the Kavalan people supports the possibility [@Chen1963].
+Moreover, some features containing long, thin (~4 cm in diameter) or Y-shaped racks may have been associated with drying facilities for meat or hides [@Chen2007; @Friesen2001].
 The shift in cooking practices might result from the increased need for deer, leading people to adopt boiling for its benefits, such as maximizing the use of residual meat, facilitating grease rendering, and improving quality of lean meat [@Lupo1997], while drying helped preserve and transport the meat.

### 2025-03-10 analysis/code/kwl-faunal-layers.R
code update

@@ -9,6 +9,7 @@ kwl_fauna_HM <- readxl::read_excel(here::here("analysis","data","raw_data","KWL_
 kwl_fauna_broken <- readxl::read_excel(here::here("analysis","data","raw_data","broken_fauna.xlsx"), sheet = 1,
                                    col_types = "text")
 
+
 sample_p <- readr::read_csv(here::here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv"))
 
 kwl_chro_6 <- readxl::read_excel(here::here("analysis", "data", "raw_data", "KWL_chronology.xlsx"), sheet = 5)
@@ -99,7 +100,8 @@ fauna_combined_taxa <-
   select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
   bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>%
   mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`)) %>%
-  mutate(Taxon = ifelse(Taxon == "Sus scrofa", "Sus spp.", Taxon))
+  mutate(Taxon = ifelse(Taxon == "Reeves's muntjac", "Muntiacus reevesi", Taxon)) %>%
+  mutate(Taxon = ifelse(Taxon == "Cervus nippon", "Cervus nippon taiouanus", Taxon))
 
 # calculate MNI
 fauna_combined_MNI <-
@@ -288,13 +290,17 @@ fauna_deer_portion <-
     `部位/名稱` %in% c("上顎骨","上顎及齒","下顎及齒","下顎骨","枕骨","頭骨","顱骨", "角基部") ~ "head", #
     `部位/名稱` %in% c("第1趾骨","第2趾骨","第3趾骨","趾骨") ~ "foot"))  #"上顎齒","下顎齒","臼齒","齒"
 
+deer_axial <-
+  fauna_deer_portion %>%
+  filter(portion == "axial") %>%
+  select()
+
 deer_metapodials <-
   fauna_deer_only %>%
   mutate(metapodials = case_when(`部位/名稱` == "掌骨" ~ "metacarpals", `部位/名稱` == "蹠骨" ~ "metatarsals",
                                  `部位/名稱` == "掌骨或蹠骨" ~ "metapodials")) %>%
   filter(!is.na(metapodials)) %>%
   count(period, metapodials)
-
 
 deer_portion_plot <-
[…]

### 2025-03-10 analysis/paper/paper.qmd
finalizing the paper

@@
       orcid: 0000-0001-9060-2125
-      institute: Institute of History and Philology, Academia Sinica
+      institute: Academia Sinica
 title-block-published: "Last updated"
@@
   mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`)) %>%
-  mutate(Taxon = ifelse(Taxon == "Sus scrofa", "Sus spp.", Taxon))
+  mutate(Taxon = ifelse(Taxon == "Reeves's muntjac", "Muntiacus reevesi", Taxon)) %>%
+  mutate(Taxon = ifelse(Taxon == "Cervus nippon", "Cervus nippon taiouanus", Taxon))
 
@@
   filter(str_detect(`Common name`, "eer")) %>% summarise(sum(NISP)) #|str_detect(Taxon, "Ree")
-suid_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Sus spp.")) %>% summarise(sum(NISP))
+suid_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Sus")) %>% summarise(sum(NISP))
 buff_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Bub")) %>% summarise(sum(NISP))
@@
 suid_NISP_per <- round(suid_NISP/(total_NISP-unident), 2)*100
+
 buff_NISP_per <- round(buff_NISP/(total_NISP-unident), 2)*100
@@
 The faunal assemblage analyzed in this study from Kiwulan consists of `r total_NISP` bone fragments based on NISP (@tbl-counts).
-Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`) and water buffalo (`r buff_NISP_per`).
+Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`%) and water buffalo (`r buff_NISP_per`%).
 To further compare general patterns of animal consumption in terms of their economic values over time, animals within the same family with of similar body size were grouped together.
 Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category.
-Similarly, domestic pig (n = 2, from CL4 and CL6) were integrated into the suid category.
+Similarly, domestic pigs (CL4: n = 2, CL6: n = 1) were integrated into the suid category.
 Rats (Rattus sp.) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes.
@@
 Third, population growth would have likely led to a more proportional increase in both large and small animals.
-Despite high taxonomic diversity during the European period, suids showed a disproportionate rise in ubiquity, quantity, and proportion.
+Despite high taxonomic diversity during the European period, suid remains showed a disproportionate rise in ubiquity, quantity, and proportion.
 This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and meat return rates to deer [@Broughton2011prey].
-The role of suids as a secondary resource after deer has been observed in Indigenous cultures in western Taiwan [@Chuang2021], suggesting they may have served as a supplement when deer were scarce or limited, possibly because some deer were reserved for trade.
+
+The role of suids as a secondary resource after deer has been observed in the Indigenous cultures in southwestern Taiwan during the Neolithic period and Iron Age [@Chuang2021], suggesting that suids may have served as a supplementary resource while deer remained the prioritized prey.
+At Kiwulan, it is possible that portions of deer were reserved for trade, limiting local consumption and leading to an increased reliance on suids.
+Domestic pigs were identified at Kiwulan in both the European and pre-European periods based on their much smaller third molar size (length < 30 mm) and morphology, in comparison to archaeological data from Neolithic and Iron Age sites in Taiwan [@KtLi2015; @Chuang2021].
+However, a clear distinction between wild boars and domestic pigs in archaeological remains is challenging due to overlapping size ranges, regional variations, and the potential presence of hybrids [@Rowley2012; @Chuang2021].
+The third molar lengths of suids found at Kiwulan (28–36 mm) may reflect such hybridization patterns.
+While current evidence does not clearly indicate the emergence of intensified suid management, an increase in suid utilization, either through domestication or hunting of wild individuals, is evident during the European period.
 

### 2025-03-10 analysis/paper/paper.qmd
fixed typo

@@
   Colonial encounters between Europeans and Indigenous peoples often had profound impacts on Indigenous societies and cultures in many parts of the world.
-In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contact through regional trade networks in East Asia, which expanded further with the arrival of Europeans in the 17th century.
+In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foreign contact through regional trade networks in East Asia, which expanded further with the arrival of Europeans in the 17th century.
 This study focuses on local subsistence through faunal remains to explore how Indigenous communities responded to trade activities driven by Europeans and Han Chinese, particularly the high demand for natural resources such as deer.
-By analyzing faunal remains from the upper component of Kiwulan (AD 1350-1850), a late Iron Age site in northeastern Taiwan, this research dicusses the Indigenous participation in the deerskin trade and their adaptive strategies.
+By analyzing faunal remains from the upper component of Kiwulan (AD 1350-1850), a late Iron Age site in northeastern Taiwan, this research examines Indigenous participation in the deerskin trade and their adaptive subsistence strategies.
 The results suggest cultural persistence in deer consumption, along with increased reliance on suids during the European period as demand for deer products intensified.
@@
 While many studies focus on major colonial centers, peripheral regions also experienced varying degrees of European influence through colonial powers and associated trade networks, namely pericolonialism [@Acabado2017; @Trabert2017].
-Exploring these areas is important to for understanding the dirverse impacts of colonial expansion.
+Exploring these areas is important to understand the diverse impacts of colonial expansion.
 Yilan, in northeastern Taiwan, is an example of a remote region where European control was limited and indirect but Indigenous communities were deeply involved in regional trade with foreign groups.
-In the mid-17th century, the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) monopolized the deerskin trade in Taiwan and exported deerskin on a large scale to Japan [@Heyns2001; @Cheng2017].
+In the mid-17th century, the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) monopolized the regional trade in Taiwan and exported deerskin on a large scale to Japan [@Heyns2001; @Cheng2017].
 While historical records by the Dutch and Chinese provide details about the operations and policies of the deerskin trade [@Lin2014deer], archaeological investigations into Indigenous reactions to these economic activities remain rare.
-To fully understand Indigenous responses to colonial economic demands, archaeological perspectives on their interactions with the environment are essential.
+To fully understand Indigenous responses to colonial economic demands, it is necessary to examine their subsistence strategies and interactions with the environment from an archaeological perspective.
 
 <!-- research aim and questions -->
-In this paper, I investigate Indigenous subsistence in a pericolonial context through the case study of the Kiwulan site (AD 1350-1850), one of the largest Iron Age settlements in northeastern Taiwan [@Chen2007].
-Kiwulan experienced European colonial powers, including the Spanish and Dutch in the 17th century, followed by a large influx of Han Chinese migrats in the 19th century, making it an ideal case to explore Indigenous responses to foreign presence over time.
+In this paper, I investigate Indigenous subsistence in a pericolonial context through the case study of the Kiwulan site (AD 1350-1850), a large Iron Age settlement in northeastern Taiwan [@Chen2007].
+Kiwulan experienced European colonial powers, including the Spanish and Dutch in the 17th century, followed by a large influx of Han Chinese immigrants in the 19th century, making it an ideal case to explore Indigenous responses to foreign presence over time.
 Key research questions include: How did Indigenous resource use change before and after these foreign encounters, particularly during the Dutch period of large-scale trade?
@@
 <!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
-In 1626, the Spanish (Spanish East Indies) established their first colony in northern Taiwan, San Salvador at Heping Dao, follwed by Santo Domingo in Tamsui (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market.
+In 1626, the Spanish (Spanish East Indies) established their first colony in northern Taiwan, San Salvador at Heping Dao, followed by Santo Domingo in Tamsui (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market.
 The VOC had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007].
@@
 The area surrounding the fortress was inhabited by Indigenous groups, the Taparris and the Quimaurris, who had lived at the site before being expelled by the Spanish [@Borao2001, SIT:165, p.81].
-The relations between the Spanish and Indigenous people may not have involved extensive colonial control or intercultural domination, as Spanish territorial control appears to have been limited to the fort.
+The relations between the Spanish and Indigenous people may not have involved extensive colonial control or inter-cultural domination, as Spanish territorial control appears to have been limited to the fort.
 Instead, the Spanish relied on exchange of local goods and food surpluses with Indigenous communities based on historical documents [@Borao2001, SIT:170-171].
@@
 Moreover, to increase deerskin yields, the Dutch implemented a hunting license system (1636-1645), which permitted Han Chinese to hunt deer on Indigenous lands during specific periods [@Heyns2001].
-From 1633 to 1661, the average annual export of deerskins from Taiwan reached 67,000 [@Koo2011; @Cheng2017].
-These colonial activities likely impacted the economy of Indigenous peoples, because deer was an important resource in prehistoric Taiwan [@Chen2000animal].
+From 1633 to 1661, the average annual export of deerskins from Taiwan reached 67,000 pieces [@Koo2011; @Cheng2017].
+These colonial activities might impact the economy of Indigenous peoples, because deer was an important resource in prehistoric Taiwan [@Chen2000animal].
 
@@
 In 1644, the Dutch defeated several Kavalan communities and incorporated them into their colonial territory [@Kang2016].
-However, instead of directly engaging with the Kavalan people, the Dutch relied on the assistance of intermediaries, the Basay people, who were active in Northern Taiwan, to collect tributes [@Kang2003].
+However, instead of directly engaging with the Kavalan people, the Dutch relied on the assistance of intermediaries to collect tributes, especially the Basay people, who were active in Northern Taiwan [@Kang2003].
 This approach of indirect control allowed the Kavalan to maintain a degree of autonomy and even negotiate tributes [@Kang2016].
@@
 This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom around 15-16th centuries and later integrated into European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].
-Yilan was part of this pre-existing trade network, where local Kavalan people exchanged rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003].
-This trade network intensified during the period of European presence, which is consistent with the large quantities of imported foreign goods found at archaeological sites across Yilan, indicating frequent trade activity in the 17th century [@TaiwanHistory2005].
+Yilan was part of this pre-existing trade network, where the local Kavalan people exchanged rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003].
+This trade network intensified during the period of European presence, which is consistent with the large quantities of imported foreign goods found at archaeological sites across Yilan, indicating frequent trade activities in the 17th century [@TaiwanHistory2005].
 This unique historical and spatial context makes Yilan an ideal location for exploring Indigenous responses to foreign presence and the dynamics of trade networks in a pericolonial context [@Acabado2017; @Kang2016, pp.
@@
 <!--introduction of Kiwulan and the population-->
-Kiwulan, located in northern Yilan (@fig-KWL-map), is a multi-component site, with its upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007].
+Kiwulan, located in northern Yilan (@fig-KWL-map), is a multi-component site, with its upper component spanning from the late Iron Age to the historical period (AD 1350-1850) [@Chen2007].
 A large-scale salvage excavation at Kiwulan uncovered a total of 262 excavation units (4 x 4 m), revealing abundant artifacts and numerous features, including middens, burials, ditches, and post-holes with wooden posts in situ.
 The excavation suggests that Kiwulan was a well-developed settlement with longhouses (approximately 4 x 8 m) arranged side by side, and burials located to the north of the residential area [@Chen2004].
-Based on oral history and the similarities between Kavalan material culture recorded in historical documents and the archaeological assemblage, Kiwulan is identified as a Kavalan settlement [@Chen2004].
+Based on oral history and the similarities between Kavalan material culture and the archaeological assemblages, Kiwulan is believed as a Kavalan settlement [@Chen2004].
 The Kavalan people were the main Indigenous group in Yilan during the late Iron Age and early modern period.
-According to a Dutch census from the mid-17th century, there were approximately 40 Kavalan settlements in Yilan, with a total population of about 10,000 people [@Nakamura1938].
-Among these Indigenous settlements, Kiwulan ("Kibannoran" in Dutch sources) was the largest one with 160 households and 711 inhabitants.
-In the early 19th century, increasing Han Chinese migration gradually led to the relocation of the Kavalan people to Hualien in eastern Taiwan [@Zhan1995].
-This is reflected in the dramatic decline of Kiwulan’s population to fewer than 100.
+Dutch censuses from the mid-17th century recorded approximately 40 Kavalan settlements in Yilan, with a total population of about 10,000 people [@Nakamura1938].
+Among these settlements, Kiwulan ("Kibannoran" in Dutch sources) was the largest, with 160 households and 711 inhabitants.
+By the early 19th century, increasing Han Chinese migration gradually led to the relocation of the Kavalan people to Hualien in eastern Taiwan [@Zhan1995], resulting in a dramatic decline in the population of Kiwulan to fewer than 100.
 
 <!--archaeological materials-->
-The most common artifact found at Kiwulan was locally made pottery with geometric surface patterns, a typical type of ceramic in Yilan [@TaiwanHistory2005].
+The most common artifact found at Kiwulan was locally made pottery with geometric surface patterns, a typical type of earthenware in Yilan [@TaiwanHistory2005].
 In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts and ornaments, were largely found in both residential area and burials.
@@
 <!--archaeological remains-->
-Abundant animal bones, shellfish, and plant remains were uncovered at Kiwulan, reflecting the diversity of natural resources in the plain and an environment close to estuary.
-The faunal remains include deer, suids, cattle, birds, fish, and shellfish, while plant remains primarily consist of rice, peach, pandanus, and cucurbits [@Chen2004].
-Rice farming appears to have been practiced, but likly limited a household scale rather than as intensive agriculture, according to archaeological evidence and historical records [@Chen1963].
+Abundant animal bones, shellfish, and plant remains were uncovered at Kiwulan, reflecting the diversity of natural resources in the plain and an environment close to an estuary.
+The faunal remains include deer, suids, bovine, birds, fish, and shellfish, while plant remains primarily consist of rice, peach, pandanus, and cucurbits [@Chen2004].
+Rice farming seems to have been practiced, but was presumably limited to a small-scale or horticultural level rather than intensive agriculture, according to archaeological findings and historical records [@Chen1963].
 The shellfish remains include bivalves such as clams and oysters, and univalves from freshwater and brackish water environments.
@@
 #| echo: false
-#| fig-cap: "The geographic location of Kiwulan and European fortress in northern Taiwan (adapted from Wang and Marwick, 2020). Map data from naturalearthdata.com."
+#| fig-cap: "Geographic locations of Kiwulan and the European fortresses in northern Taiwan (adapted from Wang and Marwick, 2020). Map data from naturalearthdata.com."
 
@@
 The faunal assemblages analyzed in this study were sampled from 40 adjacent units in the residential area excavated during the first season (2001-2003) [@Chen2007].
-These 40 units were selected because they represent the main residential area, with thick archaeological deposits and relatively intact contexts without modern disturbance.
-The spatial continuity of these units provides a consistent framework for understanding food practices over time at the settlement scale.
-Additionally, the use of the same excavation and recovery methods allows for comparison between assemblages from different temporal sequences.
+These 40 units were selected because they represent the main residential area, with thick archaeological deposits and relatively intact contexts with minimal modern disturbance.
+The spatial continuity of these units provides a consistent framework for examining archaeological records at a settlement scale.
 I primarily focus on vertebrate species due to their economic importance in local subsistence and trade.
-The faunal remains analyzed were from dwelling spaces and middens, as these are associated with daily subsistence practices, while remains from burials are excluded due to their potential ritual purposes rather than daily practices.
+The faunal remains analyzed were collected from dwelling spaces and middens because they are closely linked to daily subsistence practices.
+Remains from burials are excluded due to their potential ritual purposes rather than daily practices.
 
@@
 Each stratum was determined based on documentation of excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators.
-The diagnostic materials include pipes and stoneware commonly found at sites associated with European expansion in Southeast Asia, such as Anping jars and martavan  [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], and clay bricks associated with Chinese migrants [@Wang2011; @Hsieh2009].
+The diagnostic materials include pipes and stoneware commonly found at sites associated with European expansion in Southeast Asia, such as Anping jars and martavan  [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], and clay bricks associated with Chinese immigrants [@Wang2011; @Hsieh2009].
 Although precisely identifying the stratum corresponding to the specific time of European arrival is challenging due to its brief duration, the strata separated based on the criteria above still provide a relatively robust framework for comparing archaeological records before and after foreign presence.
@@
 To obtain a generalized pattern of animal consumption, measures of ubiquity and taxonomic abundance were applied.
-Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of consumption of that taxa [@Mckechnie2016; @Vanderwarker2010].
-Here ubiquity values are expressed as the percentage of units containing a given taxon to the total number of units per cultural layer.
+Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of their consumption [@Mckechnie2016; @Vanderwarker2010].
+In this study, ubiquity values are expressed as the percentage of units containing a given taxon relative to the total number of units per cultural layer.
 Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when estimating the expected NISP at the site [@Grayson2004; @Lyman2012].
-Unlike MNI, which determines the actual numbers of individual animals recovered from remains, NISP offers a broader overview by considering all fragmented bone assemblages.
+MNI determines the actual numbers of individual animals recovered from remains, while NISP offers a broader overview by considering all fragmented bone assemblages.
 Despite criticism for potentially overestimating the number of animals, NISP better captures variation in species abundance with less impacts from sample size [@Grayson1984; @Lyman2008].
@@
 Teeth without associated cranial elements were excluded to avoid overrepresentation, given their much higher preservation rate and quantities compared to other parts.
-While lower limbs and skull bones contain minimal meat, they are often valued in many cultures for marrow and soft tissues, such as the tongue and brain [@Madrigal2002].
-Thus, less meaty elements were considered as an additional source of nutrition in the later interpretation.
-Similarly, the degree of fragmentation could indicate the extraction of marrow or grease [@Lupo1997] and was calculated by dividing the Minimum Number of Elements (MNE) by NISP.
-In this approach, lower values represent a higher degree of fragmentation, and vice versa.
+Despite having less meat, lower limbs and skulls are often valued in many cultures for their marrow and soft tissues, such as the tongue and brain [@Madrigal2002].
+Thus, less meaty elements were considered as an additional source of nutrition in later interpretations.
+Similarly, the degree of fragmentation could indicate marrow or grease extraction [@Lupo1997], and was calculated by dividing the Minimum Number of Elements (MNE) by NISP.
+In this approach, lower values represent a higher degree of fragmentation, and higher values indicate a lower degree of fragmentation.
 
@@
 Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`%) and water buffalo (`r buff_NISP_per`%).
-To further compare general patterns of animal consumption in terms of their economic values over time, animals within the same family with of similar body size were grouped together.
+To further compare general patterns of animal consumption based on economic values, animals from the same family and of similar body size were grouped together.
 Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category.
 Similarly, domestic pigs (CL4: n = 2, CL6: n = 1) were integrated into the suid category.
-Rats (Rattus sp.) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes.
-In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 ( (@tbl-counts, @fig-NISP-proportion: A).
+Rats (*Rattus sp.*) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes.
+In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 (@tbl-counts, @fig-NISP-proportion: A).
 A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
@@
 CL4 <- mammal_count %>% filter(period == "CL4") 
-CL4_boar <- round(CL4$n[3]/CL4$n[4], 3)*100
+CL4_boar <- round(CL4$n[2]/CL4$n[4], 2)*100
 
@@
 #| echo: false
-#| fig-cap: "A: The number of identified specimen per cultural layers.
-The NISP data is derived from Table 1 .
-CL6 represents the earliest layer and CL1 is the youngest.
-B: The weight of the unidentified small bone fragments for each cultural layer.
-C: Relative abundance of vertebrate classes by cultural layers based on normed NISP.
-D: Relative abundance of identified mammals by cultural layers based on NISP."
+#| fig-cap: "A: The number of identified specimens per cultural layer.
+NISP data is derived from Table 1 .
+CL6 represents the earliest layer, and CL1 is the youngest.
+B: The weight of unidentified small bone fragments per cultural layer.
+C: Relative abundance of vertebrate classes by cultural layer based on normed NISP.
+D: Relative abundance of identified mammals by cultural layer based on NISP."
 
@@
 The second and third most commonly used animals are suid and muntjac.
-Although both reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for suid, ,and `r CL4_ubi_munt` for muntjac), their presence patterns differ slightly.
+Although both reached their highest ubiquity in CL4 (`r CL4_ubi_boar` for suid, , and `r CL4_ubi_munt` for muntjac), their presence patterns differ slightly.
 Suid was utilized as early as CL6, declined in CL3, and absence in CL2, while muntjac showed a cultural lag, appearing one phase later than suid.
@@
 Small animals (birds, fish, and turtle) had much lower ubiquity than large animals, with a slight increase in ubiquity in CL3.
-The result of ubiquity corresponds with the frequency data (NISP), suggesting that deer was the most common resource at the site, with an increased use of suid in CL4.
+The result of ubiquity corresponds with the frequency data (NISP), suggesting that deer was the most common resource at the site, with an increased use of suid and muntjac in CL4.
 
 Although deer were predominant throughout all temporal sequences, variations in their skeletal parts can be identified.
-Deer bones were grouped into six body portions [@Madrigal2004], including the head (cranium, mandible), axial (vertebrae, sternum, ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges).
+Deer bones were grouped into six body portions [@Madrigal2004], including the head (cranium and mandible), axial (vertebrae, sternum, and ribs), upper forelimb (scapula, humerus, ulna, and radius), upper hindlimb (femur, tibia, innominate), lower limb (carpals, metacarpals, tarsals, metatarsals, astragalus, calcaneus), and foot (phalanges).
 The results (@fig-deer-portion: A) show that the meaty portion, especially the upper forelimb, is the most frequent part from CL6 to CL2, followed by the upper hindlimb and the less-meaty portions, lower limb and foot.
@@
 #| echo: false
-#| fig-cap: "A: The frequency of skeletal parts of deer using NISP across cultural layers at Kiwulan. B: Bone minerial density agaist %survivorship of deer remains."
+#| fig-cap: "A: The frequency of deer skeletal parts using NISP across cultural layers at Kiwulan. B: Bone minerial density agaist %survivorship of deer remains."
 
@@
 Surface modifications, including cut marks, burning damage, and fractures, were identified on deer bones from CL6 to CL2, with variations in their proportions (@tbl-deer-modify).
-Cut marks appeared on 50% of bones in CL2, 30% in earlier periods (CL6 to CL4), and at a lower proportion of 23% in CL3.
+Cut marks appeared on 54% of bones in CL2, around 30% in earlier periods (CL6 to CL4), and at a lower proportion of 23% in CL3.
 Most cut marks were located on major limb joints and long bones, indicating butchering activities related to disarticulation and meat removal.
@@
 Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4.
-Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percussion scars [@Haynes1983; @Gifford2018; @Fisher1995], were frequently observed on long bones across different layers (@fig-fracture-img).
+Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percussion scars [@Haynes1983; @Gifford2018; @Fisher1995], were frequently observed on long bones across different layers (@fig-fracture-img: a-d).
 These fractures indicate intentional marrow extraction at Kiwulan.
@@
 #| echo: false
-#| fig-cap: "The proportion of cut marks on major limb joints of deer across cultural layers. The method for clasifying and quantifiying limb joints is based on Lyman (2005)."
+#| fig-cap: "The proportion of cut marks on major limb joints of deer across cultural layers. The method for classifying and quantifying limb joints is based on Lyman (2005)."
 
@@
 Cut marks are indicated by red squares, and chop surfaces are indicated by black arrows.
-Bone fragments are (a) proximal radius with cut marks and a chop surface, (b) distal humerus with cut marks, (c) acetabulum with chop surfaces, (d) charred astragulus, (e) metatarsal with cut marks."
+Bone fragments include (a) proximal radius with cut marks and a chop surface, (b) distal humerus with cut marks, (c) acetabulum with chop surfaces, (d) charred astragulus, and (e) metatarsal with cut marks."
 
@@
 #| fig-cap: "Examples of deer bones with fractures, natural modifications, and chewing marks.
-Bone fragments are (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing spiral fracture, (c) distal radius showing tranverse fracture and root etching, (d) femur shaft showing longitudinal split, weathering, and a percurssion scar indicated by an arrow, (e) scapula with chewing marks."
+Bone fragments include (a) proximal metatarsal with a puncturing hole and longitudinal split, (b) distal tibia showing a spiral fracture, (c) distal radius showing a tranverse fracture and root etching, (d) femur shaft showing a longitudinal split, weathering, and a percurssion scar indicated by an arrow, and (e) scapula with chewing marks."
 
@@
 However, a clear distinction between wild boars and domestic pigs in archaeological remains is challenging due to overlapping size ranges, regional variations, and the potential presence of hybrids [@Rowley2012; @Chuang2021].
-The third molar lengths of suids found at Kiwulan (28–36 mm) may reflect such hybridization patterns.
+The third molar lengths of suids found at Kiwulan (28–36 mm) may also reflect such hybridization patterns.
 While current evidence does not clearly indicate the emergence of intensified suid management, an increase in suid utilization, either through domestication or hunting of wild individuals, is evident during the European period.
@@
 That means, on one hand, Indigenous group in Yilan faced pressures from colonial encounters, such as occasional Dutch attacks, despite the generally indirect and weak colonial for most of the time.
-On the other hand, rather than resisting, local Indigenous people might participate volentarily in trade to acquire foreign necessities (e.g.
-metal knives and wares) and goods (e.g.
-ornaments and porcelains) that were highly valued in local culture [@Wang2020ornament].
+On the other hand, rather than resisting, local Indigenous people might participate voluntarily in trade to acquire foreign necessities (e.g., metal knives and wares) and goods (e.g., ornaments and porcelains) that were highly valued in local culture [@Wang2020ornament].
 The intensified hunting during the European period was not just a result of population growth but was driven by the deerskin trade.
@@
 # Acknowledgements
-I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the zooarchaeological remains used in the analyses and the archaeological team for the early documentation of faunal remains that provides a foundation for this research.
+I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the zooarchaeological remains used in the analyses and the archaeological team for the early documentation of archaeological records that provides a foundation for this research.
 Special thanks to Chien-Hsiang Lin at the Biodiversity Research Center, Academia Sinica, for his guidance and assistance in identifying fish bones.
 I also appreciate Jui-Yang Tsou for his valuable comments and assistance during the identification process.
-I would like to express my gratitude to Yin-Ru Chiang and his team at the Biodiversity Research Center, Academia Sinica, for providing me a lab space to work on data and finalizing this work.
-I am also grateful to the reviewers of this manuscript for their comments.
+I would like to express my gratitude to Yin-Ru Chiang and his team at the Biodiversity Research Center, Academia Sinica, for providing lab space for data analysis.
+I am also grateful to the reviewers of this manuscript for their valuable comments.
 

### 2025-03-10 analysis/code/kwl-faunal-layers.R

@@ -184,7 +184,7 @@ verte_mammal_barplot <-
   ggplot(aes(x = period, fill = category))+
   geom_bar(position = "fill") +
   labs(y = "%NISP", x = NULL) +
-  scale_y_continuous(labels = scales::percent) +
+  scale_y_continuous(labels = scales::percent) +  #scales::percent
   theme_minimal() +
   scale_fill_viridis_d(option="magma", begin = 0.2, end = 0.9) +  #plasma
   guides(fill= guide_legend(title="mammal"))

### 2025-03-12 analysis/paper/paper.qmd
organized code

@@
 
-# get the bone mineral desity data
+# get the bone mineral density data
 deer_element_BMD <- readr::read_csv(here::here("analysis", "data", "derived_data", "deer_BMD.csv"))
@@
 #| echo: false
-#| fig-cap: "A: The frequency of deer skeletal parts using NISP across cultural layers at Kiwulan. B: Bone minerial density agaist %survivorship of deer remains."
+#| fig-cap: "A: The frequency of deer skeletal parts using NISP across cultural layers at Kiwulan. B: Bone mineral density against %survivorship of deer remains."
 

### 2025-03-12 analysis/code/kwl-faunal-layers.R

@@ -44,7 +44,7 @@ fauna_H_sam <-
   filter(componant == "上文化層") %>%
   left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area"))
 
-# combine and tidy data from midden and cultural layers
+# combine data from the contexts of midden and cultural layer and tidy
 fauna_combined_context_all <-
   rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
   filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
@@ -72,7 +72,7 @@ fauna_combined_context_all <-
          burning =  ifelse(str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒"), "yes", "no")) %>%
   filter(is.na(refitted)) # for NISP
 
-# taxa counts
+############### Tab 1: taxa counts ###############
 fauna_combined_taxa <-
   fauna_combined_context_all %>%
   count(taxa, `Weight (g)`) %>%
@@ -109,22 +109,11 @@ fauna_combined_MNI <-
   filter(!is.na(`部位/左右`)) %>%
   count(`部位/左右`, category, `部位/名稱`)
 
-fauna_combined_context <- fauna_combined_context_all %>% filter(!taxa == "Rattus sp.") # remove rat
+# remove rat
+fauna_combined_context <- fauna_combined_context_all %>% filter(!taxa == "Rattus sp.")
 
-################### NISP & Taxonomic abundance ###################
-# barplot I: NISP by period (NISP)
-NISP_barplot <-
-  fauna_combined_context %>%
-  count(period) %>%
-  drop_na() %>%
-  ggplot(aes(x = period, y = n)) +
-  geom_col(width = 0.6) + #geom_bar(stat = "identity", #  geom_col(width = 0.6) +  #
-           #position = position_dodge2(preserve = "single"), widtg = 0.6) +
-  labs(y = "NISP", x = NULL) +
-  theme_minimal() +
-  theme(legend.title=element_blank())
-
-# barplot: across taxon
+############### Fig 2: NISP & Taxonomic abundance ###############
+# explore: across taxon
 NISP_barplot_taxon <-
   fauna_combined_context %>%
   count(period, category) %>%
@@ -136,7 +125,19 @@ NISP_barplot_taxon <-
   theme_minimal() +
   theme(legend.title=element_blank())
 
-# barplot II: broken bones (all taxa)
[…]

### 2025-05-15 analysis/paper/paper.qmd
fixed the typo

@@
       correspondence: "yes"
-      email: liyingw15@gmail.com
+      email: liyingw15@gmail.com, liyingw@asihp.net
       orcid: 0000-0001-9060-2125
@@
 <!--variation of colonial powers across regions and indirect colonization at Kavalan-->
-Although Dutch power expanded across nearly the entire Taiwan in the mid-17th century, the extent of their their colonial dominance varied greatly by region.
+Although Dutch power expanded across nearly the entire Taiwan in the mid-17th century, the extent of their colonial dominance varied greatly by region.
 Colonial control over areas far from the Dutch fortresses was often difficult and unsuccessful.
@@
 The sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the Chinese period with large Han Chinese settled down (CL2, CL1).
-The therm "European period" does not imply direct colonial control over Kiwulan but rather the broad impact of European colonial activities in northern Taiwan, such as long-distance trade, tribute systems, and occasional attacks affecting Indigenous communities in Yilan.
+The term "European period" does not imply direct colonial control over Kiwulan but rather the broad impact of European colonial activities in northern Taiwan, such as long-distance trade, tribute systems, and occasional attacks affecting Indigenous communities in Yilan.
 Although the small sample size across periods limits comprehensive interpretation, examining results in a continuous sequence still reveals distinct patterns in animal use.
@@
 # Acknowledgements
-I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the zooarchaeological remains used in the analyses and the archaeological team for the early documentation of archaeological records that provides a foundation for this research.
-Special thanks to Chien-Hsiang Lin at the Biodiversity Research Center, Academia Sinica, for his guidance and assistance in identifying fish bones.
-I also appreciate Jui-Yang Tsou for his valuable comments and assistance during the identification process.
+I would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the zooarchaeological remains used in the analyses and the archaeological team led by Yu-Pei Chen for the early documentation of archaeological records that provides a foundation for this research.
+Special thanks to Chien-Hsiang Lin at the Biodiversity Research Center, Academia Sinica, for his assistance in identifying fish bones.
+I also appreciate the guidance and support of my academic advisor, Kuei-Chen Lin, at the Institute of History and Philology, Academia Sinica, and the valuable comments from Jui-Yang Tsou during the identification process.
 I would like to express my gratitude to Yin-Ru Chiang and his team at the Biodiversity Research Center, Academia Sinica, for providing lab space for data analysis.
-I am also grateful to the reviewers of this manuscript for their valuable comments.
+I also thank the reviewers of this manuscript for their valuable comments.
 

### 2025-05-25 analysis/paper/paper.qmd
revise the method section

@@
 
-Only faunal assemblages identified to at least the family level were considered in this study to ensure consistent comparison across time sequences.
-Deer antlers detached from the cranium were excluded because they were less likely to be related to dietary practices.
-To obtain a generalized pattern of animal consumption, measures of ubiquity and taxonomic abundance were applied.
-Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of their consumption [@Mckechnie2016; @Vanderwarker2010].
-In this study, ubiquity values are expressed as the percentage of units containing a given taxon relative to the total number of units per cultural layer.
 Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when estimating the expected NISP at the site [@Grayson2004; @Lyman2012].
 MNI determines the actual numbers of individual animals recovered from remains, while NISP offers a broader overview by considering all fragmented bone assemblages.
-Despite criticism for potentially overestimating the number of animals, NISP better captures variation in species abundance with less impacts from sample size [@Grayson1984; @Lyman2008].
-For comparison across animal classes, the normalized NISP were used to avoid biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004].
-The normalized NISP is calculated by dividing NISP by the number of skeletal elements.
+Despite criticism for potentially overestimating the number of animals, NISP can better capture variation in species abundance with less effect from sample size [@Grayson1984; @Lyman2008].
+Bone fragments identified to the class level (e.g., Mammalia, Aves) were included in the overall NISP count to reflect the full scope of vertebrate remains.
+However, to enable more detailed taxonomic comparisons, indeterminate Mammalia specimens were excluded from further analysis to focus on bones identified to at least the family level.
+Deer antlers detached from the cranium were not considered in the quantification because they were less likely to be associated with dietary practices.
+To reduce biases resulting from differential number and recovery of skeletal elements among animal classes, NISP values were normalized by dividing the NISP by the average number of skeletal elements in a complete individual of that class [@Grayson2004].
 
-Given that deerskin could be a primary commodity in local regional exchange [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies.
+To identify generalized patterns of animal consumption, measures of taxonomic abundance and ubiquity were applied.
+Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of their use [@Mckechnie2016; @Vanderwarker2010].
+In this study, ubiquity values are expressed as the percentage of units in which a given taxon is present relative to the total number of units per cultural layer.
+Given that mammals were the most commonly found animal resource, mammal bones identified to the family level were categorized individually for further discussion and compared in terms of their ubiquity with other animal classes.
+
+Since deerskin may have been a primary commodity in local regional exchange [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies.
 This analysis can provide insights into how Indigenous people utilized deer in response to possible high economic demand when driven by their involvement in regional trade network.

### 2025-05-27 analysis/paper/paper.qmd
revise code and text

@@
 Bone fragments identified to the class level (e.g., Mammalia, Aves) were included in the overall NISP count to reflect the full scope of vertebrate remains.
-However, to enable more detailed taxonomic comparisons, indeterminate Mammalia specimens were excluded from further analysis to focus on bones identified to at least the family level.
+However, to enable more meaningful taxonomic comparisons, abundance analysis focused primarily on specimens identified to at least the family level, excluding indeterminate Mammalia.
 Deer antlers detached from the cranium were not considered in the quantification because they were less likely to be associated with dietary practices.
@@
 
-To identify generalized patterns of animal consumption, measures of taxonomic abundance and ubiquity were applied.
+The measure of ubiquity was applied to identify generalized patterns of animal consumption.
 Ubiquity measures the frequency of taxa through their presence or absence in particular depositional contexts, suggesting the regularity of their use [@Mckechnie2016; @Vanderwarker2010].
@@
   count(taxa, `Weight (g)`) %>%
-  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
+  mutate(taxa = str_replace(taxa, "Large cervidae", "Cervidae large size")) %>%
+  mutate(taxa = str_replace(taxa, "Testudines", "Geoemydidae")) %>%
   mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
                           TRUE ~ taxa)) %>%
-  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~
+  mutate(`Common name` = case_when(str_detect(taxa, "Cervus") ~ "Sika deer", str_detect(taxa, "Bub") ~
                                      "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",
@@
                                      "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~
-                                     "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
-                                   str_detect(taxa, "cervid") ~ "Deer", taxa == "Sus scrofa domestica" ~
-                                     "Domestic pig", taxa == "Aves" ~ "Birds")) %>%
+                                     "Ray-finned fish", str_detect(taxa, "Geo") ~ "Turtle",
+                                   str_detect(taxa, "Cervidae") ~ "Deer", taxa == "Sus scrofa domestica" ~
+                                     "Domestic pig", taxa == "Aves" ~ "Bird")) %>%
   mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
-                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
-                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6,
-                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
+                           str_detect(taxa, "munt") ~ 3, str_detect(taxa, "Sus") ~ 4, 
+                           str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 5,  str_detect(taxa, "Rat") ~ 6,
+                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Geo") ~ 8,  
                            str_detect(taxa, "Ave") ~ 9,  str_detect(taxa, "large m") ~ 10, 
@@
   select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
-  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>%
-  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`)) %>%
+  bind_rows(summarise(head(., 13), across(where(is.numeric), sum), 
+                      across(where(is.character), ~"Total determinate"))) %>%
+  bind_rows(summarise(.[14:16, ], across(where(is.numeric), sum), 
+                      across(where(is.character), ~"Total indeterminate"))) %>%
+  bind_rows(summarise(head(., 16), across(where(is.numeric), sum), 
+                      across(where(is.character), ~"Total"))) %>%
+  #bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>%
+  mutate(`Weight (g)` = round(`Weight (g)`, 1)) %>% 
+  mutate(`Common name`= ifelse(is.na(`Common name`)|str_detect(`Common name`, "Total"), "-", `Common name`)) %>%
   mutate(Taxon = ifelse(Taxon == "Reeves's muntjac", "Muntiacus reevesi", Taxon)) %>%
@@
 suid_NISP_per <- round(suid_NISP/(total_NISP-unident), 2)*100
-
 buff_NISP_per <- round(buff_NISP/(total_NISP-unident), 2)*100
@@
 
-The faunal assemblage analyzed in this study from Kiwulan consists of `r total_NISP` bone fragments based on NISP (@tbl-counts).
-Among the identifiable bones, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`%) and water buffalo (`r buff_NISP_per`%).
+The faunal assemblage from the sampling area of Kiwulan consists of `r total_NISP` bone fragments, of which were identified to at least the family level (@tbl-counts).
+Among the mammal bones identified to the family level, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`%) and water buffalo (`r buff_NISP_per`%).
 To further compare general patterns of animal consumption based on economic values, animals from the same family and of similar body size were grouped together.

### 2025-05-28 analysis/paper/paper.qmd
editing

@@
   mutate(taxa = str_replace(taxa, "Testudines", "Geoemydidae")) %>%
-  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
-                          TRUE ~ taxa)) %>%
+  #mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Indeterminate", tolower(taxa), sep =" "),
+                          #TRUE ~ taxa)) %>%
   mutate(`Common name` = case_when(str_detect(taxa, "Cervus") ~ "Sika deer", str_detect(taxa, "Bub") ~
@@
                            str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 5,  str_detect(taxa, "Rat") ~ 6,
-                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Geo") ~ 8,  
-                           str_detect(taxa, "Ave") ~ 9,  str_detect(taxa, "large m") ~ 10, 
-                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
+                           str_detect(taxa, "Geo") ~ 7, str_detect(`Common name`, "fish") ~ 8,  
+                           str_detect(taxa, "Ave") ~ 9,  str_detect(taxa, "Large m") ~ 10, 
+                           str_detect(taxa, "Medium m") ~ 11, TRUE ~ 12)) %>%
   group_by(group) %>%
@@
   select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
-  bind_rows(summarise(head(., 13), across(where(is.numeric), sum), 
-                      across(where(is.character), ~"Total determinate"))) %>%
-  bind_rows(summarise(.[14:16, ], across(where(is.numeric), sum), 
-                      across(where(is.character), ~"Total indeterminate"))) %>%
   bind_rows(summarise(head(., 16), across(where(is.numeric), sum), 
@@
 total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
-unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "In")) %>% summarise(sum(NISP))
-deer_NISP <- fauna_combined_taxa %>% 
-  filter(str_detect(`Common name`, "eer")) %>% summarise(sum(NISP)) #|str_detect(Taxon, "Ree")
+unident <- fauna_combined_taxa %>% slice(12:16) %>% summarise(sum(NISP))
+family_ident <- total_NISP - unident
+cervid_NISP <- fauna_combined_taxa %>% filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "ree")) %>% summarise(sum(NISP))
 suid_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Sus")) %>% summarise(sum(NISP))
-buff_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Bub")) %>% summarise(sum(NISP))
-deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 2)*100
+buff_NISP <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Bub")|str_detect(Taxon, "Bos")) %>% summarise(sum(NISP))
+cer_NISP_per <- round(cervid_NISP/(total_NISP-unident), 2)*100
 suid_NISP_per <- round(suid_NISP/(total_NISP-unident), 2)*100
@@
 
-The faunal assemblage from the sampling area of Kiwulan consists of `r total_NISP` bone fragments, of which were identified to at least the family level (@tbl-counts).
-Among the mammal bones identified to the family level, large deer is the majority (`r deer_NISP_per`%), followed by suid (`r suid_NISP_per`%) and water buffalo (`r buff_NISP_per`%).
+The faunal assemblage from the sampling area of Kiwulan consists of `r total_NISP` bone fragments, of which `r family_ident` were identified to at least the family level (@tbl-counts).
+Among those bones identified to the family or more specific level, Cervidae is the majority (`r cer_NISP_per`%), followed by Suidae (`r suid_NISP_per`%) and Bovinae (`r buff_NISP_per`%).
 To further compare general patterns of animal consumption based on economic values, animals from the same family and of similar body size were grouped together.

### 2025-06-01 analysis/paper/paper.qmd
added a reference

@@
 Among those bones identified to the family or more specific level, Cervidae is the majority (`r cer_NISP_per`%), followed by Suidae (`r suid_NISP_per`%) and Bovinae (`r buff_NISP_per`%).
-To further compare general patterns of animal consumption based on economic values, animals from the same family and of similar body size were grouped together.
-Therefore, the deer category includes sika deer, sambar deer, and large deer, while water buffalo and yellow cattle were grouped into the bovine category.
-Similarly, domestic pigs (CL4: n = 2, CL6: n = 1) were integrated into the suid category.
+To further compare general patterns of animal consumption based on their economic values and meat return rates, animals of similar body size and from the same family were grouped together.
+For example, the "deer" category includes only medium and large deer species, while the smaller deer species, muntjac, which implies a lower meat yield, was excluded and treated as a separate category.
+Similarly, water buffalo and yellow cattle, which are comparable in size, were grouped into the "bovine" category, and domestic pigs (CL4: n = 2; CL6: n = 1) were included in the "suid" category.
 Rats (*Rattus sp.*) were excluded from further analyses as they are usually considered commensal animals, and the relatively intact skeletons recovered from a single archaeological context show no direct of being used for dietary purposes.

### 2025-06-02 analysis/paper/paper.qmd
editing

@@
 Among those bones identified to the family or more specific level, Cervidae is the majority (`r cer_NISP_per`%), followed by Suidae (`r suid_NISP_per`%) and Bovinae (`r buff_NISP_per`%).
-To further compare general patterns of animal consumption based on their economic values and meat return rates, animals of similar body size and from the same family were grouped together.
-For example, the "deer" category includes only medium and large deer species, while the smaller deer species, muntjac, which implies a lower meat yield, was excluded and treated as a separate category.
+Based on the assumption that differences in prey body size may be associated with distinct economic values [@Broughton2011prey; @Ugan2005], animals of similar body size and from the same family were grouped together to further compare general patterns of animal use.
+For example, the "deer" category includes only medium and large deer species, while the small deer species, muntjac, which implies a lower ranking resource, was excluded and treated as a separate category.
 Similarly, water buffalo and yellow cattle, which are comparable in size, were grouped into the "bovine" category, and domestic pigs (CL4: n = 2; CL6: n = 1) were included in the "suid" category.
@@
 In general, the faunal remains show that frequencies increased from CL6, peaked at CL4, and then declined dramatically after CL4 (@tbl-counts, @fig-NISP-proportion: A).
-A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B).
+A similar trend was observed in the weight of unidentified small bone fragments (@fig-NISP-proportion: B), which were too fragmented to be identified even to the class level.
 
@@
 CL6 represents the earliest layer, and CL1 is the youngest.
-B: The weight of unidentified small bone fragments per cultural layer.
+B: The weight of unidentified small bone fragments that could not be identified to the class level, shown per cultural layer.
 C: Relative abundance of vertebrate classes by cultural layer based on normed NISP.
@@
 Despite high taxonomic diversity during the European period, suid remains showed a disproportionate rise in ubiquity, quantity, and proportion.
-This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and meat return rates to deer [@Broughton2011prey].
+This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and similar return rates to deer [@Broughton2011prey].
 

### 2025-06-04 analysis/paper/paper.qmd
revised based on comments

@@
 Small animals (birds, fish, and turtle) had much lower ubiquity than large animals, with a slight increase in ubiquity in CL3.
-The result of ubiquity corresponds with the frequency data (NISP), suggesting that deer was the most common resource at the site, with an increased use of suid and muntjac in CL4.
+The results of the ubiquity analysis generally correspond with the frequency data (@fig-NISP-proportion: D), suggesting that deer was the most common resource at the site, with an increased use of suid in CL4.
+Although muntjac had relatively higher ubiquity in CL4 compared to other layers, their %NISP did not increase accordingly but instead decreased slightly, implying they may have been more widely used but only in small quantities.
 

### 2025-06-07 analysis/paper/paper.qmd
revise based on comments

@@
 Cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were quantified to explore butchering patterns over time.
-Moreover, cut marks on lower limbs related to skinning processes, were examined and compared.
-Such skinning marks are characteristic of circular and straight incisions left by a series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].
+Moreover, cut marks on lower limbs were particularly examined since they might be related to skinning processes.
+During skinning, circular and longitudinal incisions are required to obtain intact animal hides that often leave transverse, oblique, or longitudinal marks on lower limb [@Costamagno2019; @Isaakidou2007; @Val2011].
 

### 2025-06-09 analysis/paper/paper.qmd
revised the text based on comments

--- 2025-06-04 analysis/paper/paper.qmd
@@
 Cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were quantified to explore butchering patterns over time.
-Moreover, cut marks on lower limbs related to skinning processes, were examined and compared.
-Such skinning marks are characteristic of circular and straight incisions left by a series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].
+Moreover, cut marks on lower limbs were particularly examined since they might be related to skinning practices.
+Skinning typically involves circular and longitudinal incisions to obtain intact animal hides, which often leave transverse, oblique, or longitudinal cut marks on lower limb elements, such as metapodials [@Costamagno2019; @Isaakidou2007; @Val2011].
+These skinning marks are usually located on the medial or lateral sides of the metapodials and are characterized by shallow, elongated marks, which differ from the deeper, shorter marks left by tendon or meat removal [@Costamagno2019].
 

### 2025-06-14 analysis/paper/paper.qmd
revised based on comments

--- 2025-06-04 analysis/paper/paper.qmd
@@
 Cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were quantified to explore butchering patterns over time.
-Moreover, cut marks on lower limbs related to skinning processes, were examined and compared.
-Such skinning marks are characteristic of circular and straight incisions left by a series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].
+Moreover, cut marks on lower limbs were particularly examined since they might be related to skinning practices.
+Skinning typically involves circular and longitudinal incisions to remove intact animal hides, which often leave transverse, oblique, or longitudinal cut marks on lower limb elements, such as metapodials and phalanges [@Costamagno2019; @Isaakidou2007; @Val2011].
+On metapodials, these skinning marks are usually located on the medial or lateral sides and are characterized by shallow, elongated marks, which differ from the deeper, shorter marks associated with tendon or meat removal [@Costamagno2019, pp.
+212-220; @Soulier2017].
 
@@
 Nevertheless, the overall high frequency of joints with cut marks suggests intensive deer processing throughout the site's occupation.
-In addition, cut marks on lower limbs show traces associated with skinning (@fig-cutmark-img: e).
-Those cut marks were primarily found on metapodials in CL4 (n = 6), CL5 (n = 4), CL6 (n = 1), and CL3 (n = 1), followed by first phalanges in CL4 (n = 1) and CL5 (n = 1).
+In addition, cut marks on the lower limbs, particularly shallow, elongated traces on the medial and lateral sides of bones, are likely associated with skinning (@fig-cutmark-img: e).
+These marks were primarily observed on metapodials in CL4 (n = 6), CL5 (n = 4), CL6 (n = 1), and CL3 (n = 1), with additional occurrences on first phalanges in CL4 (n = 1) and CL5 (n = 1).
 Some joints displayed clean, smooth cut surfaces from chopping (@fig-cutmark-img: a, c), accompanied by narrow V-shaped cut marks observed on most bones, indicating the use of metal tools [@Greenfield1999].

### 2025-06-15 analysis/paper/paper.qmd
editing

@@
 Nevertheless, the overall high frequency of joints with cut marks suggests intensive deer processing throughout the site's occupation.
-In addition, cut marks on the lower limbs, particularly shallow, elongated traces on the medial and lateral sides of bones, are likely associated with skinning (@fig-cutmark-img: e).
-These marks were primarily observed on metapodials in CL4 (n = 6), CL5 (n = 4), CL6 (n = 1), and CL3 (n = 1), with additional occurrences on first phalanges in CL4 (n = 1) and CL5 (n = 1).
+In addition, cut marks on the lower limbs, particularly shallow, elongated traces on the medial and lateral sides of metapodials (@fig-cutmark-img: e), are consistent with marks left by skinning based on documented butchery experiments [@Costamagno2019; @Val2011].
+These skinning marks were identified on metapodials from CL4 (n = 6), CL5 (n = 4), CL6 (n = 1), and CL3 (n = 1), with additional occurrences on first phalanges from CL4 (n = 1) and CL5 (n = 1).
 Some joints displayed clean, smooth cut surfaces from chopping (@fig-cutmark-img: a, c), accompanied by narrow V-shaped cut marks observed on most bones, indicating the use of metal tools [@Greenfield1999].

### 2025-06-18 analysis/paper/paper.qmd
revise discussion section

@@
 Despite these shifts, deer remained a primary resource for local consumption at Kiwulan during the post-European and early Chinese periods.
-Bovines were present in small numbers (yellow cattle) in the pre-European period, but became the sole large animal (water buffalo) used at Kiwulan in the late Chinese period.
-Since bovines were absent from prehistoric sites in Taiwan and are widely believed to have been introduced from mainland China in historical period [@Chen2000animal], their increasing presence suggests Chinese influence and aligns with the Indigenous relocation to eastern Taiwan since the 19th century due to Chinese population pressures [@Zhan1995].
+
+Bovines were present in small numbers (yellow cattle) in the pre-European period but became the only animal (water buffalo) in the late Chinese period (CL1).
+One possible explanation for the shift in the animal taxonomic structure from deer to bovines is the dramatic decline of the Indigenous Kavalan population in Yilan during the 19th century, following a large influx of Han Chinese immigrants.
+Competition over land between the two groups eventually led to several waves of Kavalan relocation to eastern Taiwan, leaving only a small number of Kavalan in Yilan [@Zhan1995].
+In addition, bovines are widely believed to have been introduced from mainland China during the historical period due to the absence of bovine remains from prehistoric archaeological contexts in Taiwan [@Chen2000animal].
+The predominance of bovine remains in the final phase of Kiwulan may suggest both a decreased Indigenous population and a shift from traditional hunting practices to livestock use under Han influence.
+A similar cultural change is also observed in burial practices at Kiwulan, where coffin-like containers, a typical funerary practice of Han Chinese, were first adopted by local people during the Chinese period [@Chen2007; @Chiu2004].
 

### 2025-06-21 analysis/paper/paper.qmd
editing

--- 2025-06-15 analysis/paper/paper.qmd
@@
 Despite these shifts, deer remained a primary resource for local consumption at Kiwulan during the post-European and early Chinese periods.
-Bovines were present in small numbers (yellow cattle) in the pre-European period, but became the sole large animal (water buffalo) used at Kiwulan in the late Chinese period.
-Since bovines were absent from prehistoric sites in Taiwan and are widely believed to have been introduced from mainland China in historical period [@Chen2000animal], their increasing presence suggests Chinese influence and aligns with the Indigenous relocation to eastern Taiwan since the 19th century due to Chinese population pressures [@Zhan1995].
+
+Bovines were present in small numbers (yellow cattle) in the pre-European period but became the only animal (water buffalo) in the late Chinese period (CL1).
+One possible explanation for the shift in the animal taxonomic structure from deer to bovines is the dramatic decline of the Indigenous Kavalan population in Yilan during the 19th century, around the same time as a large influx of Han Chinese immigrants.
+Competition over land between the two groups eventually led to several waves of Kavalan relocation to eastern Taiwan, leaving only a small number of Kavalan in Yilan [@Zhan1995].
+Furthermore, bovines were absent in prehistoric Taiwan and are believed to have been introduced by Han populations during the historical period [@Chen2000animal].
+The reliance on bovines in CL1 may indicate both a decline in the Indigenous population and a shift from traditional hunting to livestock use introduced or intensified under Han influence.
+A similar cultural change is also observed in burial practices at Kiwulan, where coffin-like containers, typical funerary traditions of Han Chinese, were first adopted by Indigenous people in the late period (CL1 and CL2) [@Chen2007; @Chiu2004].
 

### 2025-07-08 analysis/paper/paper.qmd
revise the discussion section

--- 2025-06-15 analysis/paper/paper.qmd
@@
 The faunal assemblage from the major residential area at Kiwulan suggests a continuous reliance on sika deer from CL6 to CL2, followed by a shift to bovine in CL1.
-Despite deer were the main animal resource during all times, variations in quantity, ubiquity, butchering patterns, and proportions relative to other animals suggest changes relative to foreign presence.
+Despite deer being the main animal resource during all times, variations in quantity, ubiquity, butchering patterns, and proportions relative to other animals suggest changes relative to foreign presence.
 The cultural layers were grouped into four temporal sequences in relation to foreign presences for the following interpretation.
 The sequences include the pre-European period (CL6, CL5), the European period (CL4), the post-European period (CL3), and the Chinese period with large Han Chinese settled down (CL2, CL1).
-The term "European period" does not imply direct colonial control over Kiwulan but rather the broad impact of European colonial activities in northern Taiwan, such as long-distance trade, tribute systems, and occasional attacks affecting Indigenous communities in Yilan.
-Although the small sample size across periods limits comprehensive interpretation, examining results in a continuous sequence still reveals distinct patterns in animal use.
+The term "European period" here does not refer to direct colonial control over Kiwulan but the broad impact of colonial activities in northern Taiwan, including long-distance trade, tribute systems, and occasional attacks on Indigenous communities in Yilan.
+Although the small sample size across periods may limit the scope of interpretation, the consistent analysis of faunal remains from each period still provides valuable insights into changes in animal use.
 
 <!--summary of overall observations across time periods -->
-The sampled area at Kiwulan contains the large quantity of faunal remains from the European period, followed by the late pre-European period.
-Interestingly, deer bones from the late pre-European period presented the same ubiquity as in the European period but only half the quantity.
-This might indicate a consistent pattern of regular deer consumption at the site both before and after European presence, with intensified deer use during the European period.
-In addition to the dominance of deer, the increased relative proportion and ubiquity of suids in the European period suggests a more regular use of suids compared to other periods.
-The dramatic decline in deer bones after the European period may be associated with both a reduction in the Kiwulan population and the decrease of wild deer.
-Sika deer populations in western Taiwan largely declined from the 18th century due to habitat loss and overhunting [@Chen2000animal].
-A slight increase in muntjac use suggests that they may have served as a substitute, replacing suids as a secondary animal resource from earlier periods.
-Despite these shifts, deer remained a primary resource for local consumption at Kiwulan during the post-European and early Chinese periods.
-Bovines were present in small numbers (yellow cattle) in the pre-European period, but became the sole large animal (water buffalo) used at Kiwulan in the late Chinese period.
-Since bovines were absent from prehistoric sites in Taiwan and are widely believed to have been introduced from mainland China in historical period [@Chen2000animal], their increasing presence suggests Chinese influence and aligns with the Indigenous relocation to eastern Taiwan since the 19th century due to Chinese population pressures [@Zhan1995].
+Deer is the predominant animal, mostly from the European period, followed by the late pre-European period.
+Interestingly, deer bones from the European period show the same ubiquity as in the late pre-European period but appear in twice the quantity.
+This suggests a regular deer consumption at the site, but an intensified use during the European period.
+In addition to deer, suids were also used more frequently during the European period compared to other phases.
+The dramatic drop in deer bones after the European period may be associated with declines in both the wild deer population and the Kiwulan population.
+A similar pattern of shrunken sika deer populations has been recorded in western Taiwan since the 18th century due to habitat loss and overhunting [@Chen2000animal].
+A slight increase in muntjac remains in the post-European period suggests they may have served as a substitute, replacing suids as a secondary resource from earlier periods.
+Nevertheless, deer remained a primary source for local consumption at Kiwulan during the post-European and early Chinese periods.
 
 <!--skeletal part frequencies and butchery patterns of deer-->
-Differences in deer skeletal completeness across periods suggests changes in butchery locations and cooking practices.
-There were more complete carcasses in the European and late pre-European periods compared to the post-European and Chinese periods.
-Moreover, the substantial rise in fragmented head parts (cranium and mandible) during the European period indicates increased use of soft tissues, such as brains and tongues, or the transport of more intact deer to the settlement for processing.
-This is consistent with the historical account in 1603, Record of Formosa (Dongfan ji) [@Chen1603], which describes the consumption of deer tongues and group hunting practices that could facilitate the transportation of deer.
-Changes in cooking practices are also evident.
-The decline in burned bones over time, with the lowest percentage in the European period, suggests a shift from roasting to other cooking methods, such as boiling [@Lupo1997].
-The size of chopped bones, which could fit into local ceramics at Kiwulan, might be associated with boiling practices.
-This could be further confirmed through organic residue analysis on pottery.
-Other meat preparation methods, such as drying, are also possible [e.g., @Spyrou2019].
-Although vertebrae and ribs, typically used for drying [@Friesen2001; @Binford1978], were present in low frequencies at Kiwulan, historical records mentioning the consumption or exchange of dried meat by the Kavalan people supports the possibility [@Chen1963].
-Moreover, some features containing long, thin (~4 cm in diameter) or Y-shaped racks may have been associated with drying facilities for meat or hides [@Chen2007; @Friesen2001].
-The shift in cooking practices might result from the increased need for deer, leading people to adopt boiling for its benefits, such as maximizing the use of residual meat, facilitating grease rendering, and improving quality of lean meat [@Lupo1997], while drying helped preserve and transport the meat.
+Although deer remained a major resource prior to the late Chinese period, skeletal parts of deer indicate changes in butchery practices during the European period.
+More complete carcasses and a substantial increase in fragmented head parts (cranium and mandible) in the European period suggest that more intact deer were transported to and processed at the site, with greater use of soft tissues such as brains and tongues.
+These findings are consistent with the mentions about deer tongue consumption and Indigenous group hunting in *Record of Formosa (Dongfan ji)* in 1603 by Chinese travelers [@Chen1603].
+Group hunting may have become more frequent during the European period, enabling the transport of whole deer.
+Shifts in cooking practices are also evident.
+Burned bones were common in the pre-European period but decreased largely in the European period, suggesting a change from roasting to other cooking methods, such as boiling, an ideal practice for maximizing meat use and extracting grease [@Lupo1997].
+Many chopped bones at Kiwulan can fit well within the dimensions of local ceramics, implying the likelihood of boiling.
+This could be further confirmed through organic residue analysis on pottery in the future.
+Other meat preparation methods, such as drying, should also be considered [e.g., @Spyrou2019].
+Although skeletal elements, such as vertebrae and ribs [@Friesen2001; @Binford1978], typically associated with drying are present in low frequencies at the site, the consumption and exchange of dried meat by the Kavalan people were documented in Chinese historical records [@Chen1963].
+Moreover, wooden racks in long strips (~ 4 cm in diameter) and Y-shaped forms found in piles at Kiwulan might be associated with facilities for drying meat or hides [@Chen2007; @Friesen2001].
+Overall, the changes in butchering and cooking practices suggest an increased need for deer, with methods like boiling or drying being used to optimize animal use and extend meat preservation.
 
-A consistent pattern of high frequencies of upper limb bones but low frequencies of vertebrae and ribs was identified across all periods.
-This might suggest selective butchering and a preference for meaty portions.
-Moreover, among the meaty portions, front quarters (humerus and radius) were consistently more abundant than hindquarters (femur and tibia).
+For deer skeletal parts, the pattern of high frequencies of upper limb bones but low frequencies of vertebrae and ribs suggests selective butchering and a preference for meaty portions.
+It should be noted that among the meaty portions, front quarters (humerus and radius) were consistently more abundant than hindquarters (femur and tibia).
 One possible explanation is that femurs and tibiae were heavily fragmented due to marrow extraction, as these bones have high marrow yields [@Madrigal2002; @Lupo1998], making them less identifiable in the assemblage.
 Evidence for marrow extraction is supported by at least 25% of long bones showing fractures and a moderate degree of fragmentation overall.
-Another explanation is that some hindquarters were used for trade, as Dutch records mention that deer legs and meat were commodities in addition to hides [@Blusse2006, pp.220-224], but they did not specify which legs.
-If deer legs, particularly hindquarters, were traded as bone-meat packages, this could account for the lower frequencies of hindquarter bones compared to front quarters.
-While this hypothesis requires further evidence, the overall high proportion of meaty portions at Kiwulan suggests that deer meat might be not only used for local consumption but also processed for trade.
-Thus, both factors could contribute to this pattern, rather than a single factor alone.
+Another explanation is that some hindquarters could be used for trade, as Dutch records mention that deer legs and meat were commodities in addition to hides [@Blusse2006, pp.220-224], but they did not specify which legs.
+If deer legs, particularly hindquarters, were traded as bone-meat packages, this can account for the lower frequencies of hindquarter bones compared to front quarters.
+While this hypothesis requires further evidence, the overall high proportion of meaty portions at Kiwulan suggests that deer meat may have been not only consumed locally but also processed for trade.
 
 <!--deer hides-->
-Deerskin trade had been prevalent in East Asia since the 14th century and was monopolized by Europeans in the 17th century when a large number of hides, especially from western Taiwan, were exported to Japan [@Cheng2017; @Hollmann1991; @Wang1981].
-The cut marks identified on deer lower limbs at Kiwulan confirm the removal of deer skins, likely for trade with external traders.
-Moreover, the abundance of foreign goods at Kiwulan, including ornaments, porcelains, and metal items, further supports the frequent exchanges between local residents and outsiders [@Wang2020ornament; @Chen2007].
-This suggests that northeastern Taiwan, like western Taiwan, was actively involved in and impacted by long-distance trade.
-Although lower limbs with skinning cuts did not show a substantial increase during the European period, the disproportionate rise in head parts might hint at the likelihood of brain tanning when preparing hides.
-Brain tanning is a traditional method for softening and preserving hides in many Indigenous cultures, including those of North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016].
-In Taiwan, the present-day Bunun group in the central highlands has revived hide tanning based on oral history, primarily using peanuts or pig lard as tanning materials, with occasional use of animal brains [@Li1999, pp.
+According to documentary sources, deerskin trade flourished in East Asia from the 14th century and was monopolized by the Dutch in the 17th century, when large quantities of hides, particularly from western Taiwan, were exported to Japan [@Cheng2017; @Hollmann1991; @Wang1981].
+Archaeological evidence of hide preparation can be supported by cut marks found on the lower limbs of deer at Kiwulan.
+Moreover, foreign goods found at Kiwulan, such as glass beads, porcelains, and metal objects, might be the result of trade with outsiders [@Wang2020ornament; @Chen2007].
+These findings indicate that northeastern Taiwan was also actively engaged in the long-distance trade networks stimulated by European involvement.
+Although the frequency of lower limbs with skinning marks did not increase substantially during the European period, a disproportionate rise in head parts might imply the use of deer brains for hide tanning in addition to serving as a nutritional resource.
+Brain tanning is a method for softening and preserving hides used by many Indigenous cultures across North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016].
+In Taiwan, the use of peanuts or pig lard as tanning materials has been culturally revived by the present-day Bunun people in the central highlands based on oral history, with occasional use of animal brains [@Li1999, pp.
 48; @Xu1994].
-Despite the specific tanning materials used by the Kavalan people remain unknown due to the loss of traditions, brain tanning is a plausible method given the availability of deer heads from hunting.
+Although the tanning methods once practiced by the Kavalan people remain unclear, brain tanning is a plausible way, given the availability of deer heads from hunting and the large number of cranial fragments found at Kiwulan during the European period.
+However, further archaeological evidence is still needed to fully support this possibility.
 
 <!-- population dynamics and arguments-->
-The high population during the European period, as inferred by the frequencies of pottery, may have contributed to the overall increase in animal use [@Wang2020pot; @Hsieh2009].
-However, the impact of foreign presence can be still be identified in deer butchery patterns, the relative proportions of animals, and animal ubiquity.
-First, lower limbs with skinning cuts were mostly found in the European period, followed by the late pre-European phase, aligning with the rise of regional trade in the 14th–15th centuries and its intensification in the 17th century.
+Population growth inferred from the high frequencies of pottery during the European period may have contributed to the overall increase in animal use [@Wang2020pot; @Hsieh2009].
+Nevertheless, the impact of foreign presence is evident in deer butchery patterns, relative proportions of animals, and animal ubiquity.
+First, lower limbs with skinning cuts appear in the late pre-European phase and increase during the European period, which aligns with the development of the deerskin trade in East Asia in the 17th century.
 Second, if the high frequency of deer heads during the European period was solely due to food stress from population growth, leading to increased brain or tongue consumption, a parallel rise in marrow extraction would be expected.
-However, long bones with fracture patterns showing 10–30% less frequent during the European and late pre-European periods than in other periods.
-Third, population growth would have likely led to a more proportional increase in both large and small animals.
-Despite high taxonomic diversity during the European period, suid remains showed a disproportionate rise in ubiquity, quantity, and proportion.
-This indicates that the trade of deerskin and meat may have led Indigenous people to rely more on suids as an alternative prey due to their comparable body size and similar return rates to deer [@Broughton2011prey].
+In fact, fractured long bones are 10–30% less frequent during the European and the late pre-European periods compared to other periods.
+Third, population growth would normally lead to a proportional increase in both large and small animal remains.
+However, despite the high taxonomic diversity during the European period, only suid remains showed a substantial rise in ubiquity, quantity, and proportion.
+This indicates that the trade of deer products may have led Indigenous people to rely more heavily on suids as an alternative resource due to their comparable body size and similar return rates to deer [@Broughton2011prey].
 
-The role of suids as a secondary resource after deer has been observed in the Indigenous cultures in southwestern Taiwan during the Neolithic period and Iron Age [@Chuang2021], suggesting that suids may have served as a supplementary resource while deer remained the prioritized prey.
-At Kiwulan, it is possible that portions of deer were reserved for trade, limiting local consumption and leading to an increased reliance on suids.
-Domestic pigs were identified at Kiwulan in both the European and pre-European periods based on their much smaller third molar size (length < 30 mm) and morphology, in comparison to archaeological data from Neolithic and Iron Age sites in Taiwan [@KtLi2015; @Chuang2021].
-However, a clear distinction between wild boars and domestic pigs in archaeological remains is challenging due to overlapping size ranges, regional variations, and the potential presence of hybrids [@Rowley2012; @Chuang2021].
-The third molar lengths of suids found at Kiwulan (28–36 mm) may also reflect such hybridization patterns.
-While current evidence does not clearly indicate the emergence of intensified suid management, an increase in suid utilization, either through domestication or hunting of wild individuals, is evident during the European period.
+Suids may have been treated as a secondary resource, while deer remained the prioritized.
+This pattern is not only observed at Kiwulan, but also reported at some Neolithic and Iron Age sites in southwestern Taiwan [@Chuang2021].
+At Kiwulan, it is possible that particular deer parts were reserved for trade, which therefore limited local consumption and led to an increased reliance on suids during the European period.
+Among the suids, a few domestic pigs were identified in both the European and pre-European periods based on their small third molars (length < 30 mm) and dental morphology, when compared with reference data from other Neolithic and Iron Age sites in Taiwan [@KtLi2015; @Chuang2021].
+This suggests that small-scale pig domestication was practiced at Kiwulan.
+However, because domestic pigs and wild boars are difficult to distinguish due to similar body sizes, regional morphological variation, and the potential for hybridization, the number of domestic pigs may be underestimated [@Rowley2012; @Chuang2021].
+The third molar lengths of suids from Kiwulan mostly range from 28 mm to 36 mm, which may reflect such hybridization.
+While the current evidence does not clearly indicate intensified suid management, the increased utilization of suids, whether through domestication or the hunting of wild individuals, is evident during the European period.
 
-After the Dutch left, the deerskin trade continued but on a much smaller scale, with around 10,000 pieces exported annually in the 18th century, which is far fewer than the 100,000 pieces exported during the European period [@Nakamura1997; @Tsao2011].
+The deerskin trade continued after the Dutch left, but it was much smaller.
+In the 18th century, about 10,000 pieces were exported annually, far less than the 100,000 pieces exported during the European period [@Nakamura1997; @Tsao2011].
 The shrinking trade market and declining population at Kiwulan may have contributed to the reduced reliance of suids, as shown in their lower proportion in the post-European period and eventual absence in the Chinese period.
@@
 Although both suids and muntjacs served as substitutes for deer, their differing frequencies and ubiquity over time suggest that suids were the primary supplementary meat source, while muntjact were less frequently consumed but may have been valued for their hides in daily clothing.
+
+Bovines were present in small numbers (yellow cattle) in the pre-European period but became the only animal (water buffalo) in the late Chinese period (CL1).
+One possible explanation for the shift in the animal taxonomic structure from deer to bovines is the dramatic decline of the Indigenous Kavalan population in Yilan, which occurred around the same time as a large influx of Han Chinese immigrants in the 19th century.
+The occupation of land by Han settlers gradually led to Kavalan relocation to eastern Taiwan, leaving only a small Indigenous population in Yilan [@Zhan1995].
+Moreover, bovines are believed to have been introduced by Han populations because they are absent from prehistoric contexts in Taiwan [@Chen2000animal].
+Therefore, the predominance of bovines in CL1 may indicate both a decrease in the Indigenous population and a shift from hunting to livestock use introduced or intensified under Han influence.
+Similar cultural changes can also be seen in burial practices at Kiwulan during the Chinese period, when Indigenous people began using wooden containers for burials that resembled Chinese coffins [@Chen2007; @Chiu2004].
 
@@
 # Conclusion
-By situating zooarchaeological evidence within the context of foreign presence, the Europeans in the 17th century and Chinese immigrants in the 19th century, this study reveals Indigenous participation in regional trade networks and their adaptation strategies.
+The zooarchaeological evidence at Kiwulan reveals Indigenous participation in regional trade networks and their adaptation strategies by situating faunal remains within the context of foreign presence, including the Europeans in the 17th century and Chinese in the 19th century.
 The frequencies, butchery patterns, and modifications on deer bones from Kiwulan indicates an active deer processing that was intensified during the European period.
-Deer as a primary animal resource at Kiwulan suggests cultural continuity, while the strategic reliance on suids during the European period reflects adaptation to high trade demands for deer products.
-Additionally, the incorporation of muntjac into the diet during the Chinese period likely resulted from the overhunting of sika deer.
-These findings highlight how Kiwulan residents navigated foreign influence and environmental changes while maintaining cultural traditions.
-Through the long-term dynamics of animal use at Kiwulan, this research demonstrates that even Indigenous communities in peripheral areas with relatively weak colonial control were profoundly impacted by colonial activities than previously thought.
+Deer consumption at Kiwulan throughout most of time suggests cultural continuity, while the strategic reliance on suids during the European period reflects adaptation to high trade demands for deer products.
+On the other hand, the incorporation of muntjac into the diet during the Chinese period likely resulted from the overhunting of sika deer.
+These findings highlight the adaptation of Kiwulan residents to foreign impacts and environmental changes while maintaining cultural traditions.
+Through the long-term dynamics of animal use at Kiwulan, this research demonstrates that even Indigenous communities in peripheral areas with weak colonial control were profoundly impacted by colonial activities than previously thought.
 Future research expanding to other Indigenous settlements in northeastern Taiwan, with additional dietary evidence such as botanical remains and organic residue analysis, could provide deeper insights into regional dietary patterns and human-environment interactions.

### 2025-07-11 analysis/paper/paper.qmd
editing

@@
 ---
-title: "Preliminary Insights into Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
+title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
 author:
@@
 
-Burned bones (@fig-cutmark-img: d), characterized by charring and color changes, had the highest proportion in CL6 at 40% and gradually decreased to less than 10% after CL4.
-Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percussion scars [@Haynes1983; @Gifford2018; @Fisher1995], were frequently observed on long bones across different layers (@fig-fracture-img: a-d).
-These fractures indicate intentional marrow extraction at Kiwulan.
-CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5.
+Burned Bones, normally showing charring marks and color changes, peaked in CL6 at 40% and gradually decreased to less than 10% after CL4 (@fig-cutmark-img: d).
+Bone breakage marks, including spiral fractures, transverse fractures, longitudinal splits, and percussion scars [@Haynes1983; @Gifford2018; @Fisher1995], were commonly observed on long bones (@fig-fracture-img: a-d).
+These fractures indicate intentional marrow extraction.
+Specifically, CL2 had the highest proportion of fractured bones (> 50%), compared to 35% in CL3 and CL6, and 25% in CL4 and CL5.
 Chewing marks, represented by a series of superficial elongated scores and circular depressions (@fig-fracture-img: e) [@Bello2023], were observed on a few bones, suggesting the evidence of carnivore gnawing.
@@
 
+Bovines were present in small numbers (yellow cattle) in the pre-European period but became the only animal (water buffalo) in the late Chinese period (CL1).
+One possible explanation for the shift in the animal taxonomic structure from deer to bovines is the dramatic decline of the Indigenous Kavalan population in Yilan, which occurred around the same time as a large influx of Han Chinese immigrants in the 19th century.
+The occupation of land by Han settlers gradually led to Kavalan relocation to eastern Taiwan, leaving only a small Indigenous population in Yilan [@Zhan1995].
+Moreover, bovines are believed to have been introduced by Han populations because they are absent from prehistoric contexts in Taiwan [@Chen2000animal].
+Therefore, the predominance of bovines in CL1 may indicate both a decrease in the Indigenous population and a shift from hunting to livestock use introduced or intensified under Han influence.
+Similar cultural changes can also be seen in burial practices at Kiwulan during the Chinese period, when Indigenous people began using wooden containers for burials that resembled Chinese coffins [@Chen2007; @Chiu2004].
+
 <!--skeletal part frequencies and butchery patterns of deer-->
@@
 <!--deer hides-->
-According to documentary sources, deerskin trade flourished in East Asia from the 14th century and was monopolized by the Dutch in the 17th century, when large quantities of hides, particularly from western Taiwan, were exported to Japan [@Cheng2017; @Hollmann1991; @Wang1981].
-Archaeological evidence of hide preparation can be supported by cut marks found on the lower limbs of deer at Kiwulan.
-Moreover, foreign goods found at Kiwulan, such as glass beads, porcelains, and metal objects, might be the result of trade with outsiders [@Wang2020ornament; @Chen2007].
+The deerskin trade flourished in East Asia from the 14th century and was monopolized by the Dutch in the 17th century, when large quantities of hides, particularly from western Taiwan, were exported to Japan according to documentary sources [@Cheng2017; @Hollmann1991; @Wang1981].
+Skinning marks on deer lower limbs at Kiwulan provide additional archaeological support for the deerskin trade, which aligns with findings from Iron Age contexts in central Taiwan [@Yen2017].
+Moreover, foreign goods at Kiwulan, such as beads, porcelains, and metal objects, could be the result of trade with outsiders [@Wang2020ornament; @Chen2007].
 These findings indicate that northeastern Taiwan was also actively engaged in the long-distance trade networks stimulated by European involvement.
-Although the frequency of lower limbs with skinning marks did not increase substantially during the European period, a disproportionate rise in head parts might imply the use of deer brains for hide tanning in addition to serving as a nutritional resource.
+Although the frequency of skinning marks did not increase substantially during the European period, a disproportionate rise in head parts might imply the use of deer brains for hide tanning, in addition to serving as a nutritional resource.
 Brain tanning is a method for softening and preserving hides used by many Indigenous cultures across North America, Southern Africa, North Asia, and Japan [@Baillargeon2010; @Badenhorst2009; @Hatt1969; @Matsumoto2016].
@@
 
-The deerskin trade continued after the Dutch left, but it was much smaller.
-In the 18th century, about 10,000 pieces were exported annually, far less than the 100,000 pieces exported during the European period [@Nakamura1997; @Tsao2011].
-The shrinking trade market and declining population at Kiwulan may have contributed to the reduced reliance of suids, as shown in their lower proportion in the post-European period and eventual absence in the Chinese period.
-It should be noted that muntjac were consistently part of local subsistence, though in small numbers, until the early Chinese period.
-Their skins were used by many Indigenous groups such as the Tsou and Bunun for making clothing, similar to the use of deerskins [@Li1999], which may have also been practiced at Kiwulan.
-Although both suids and muntjacs served as substitutes for deer, their differing frequencies and ubiquity over time suggest that suids were the primary supplementary meat source, while muntjact were less frequently consumed but may have been valued for their hides in daily clothing.
-
-Bovines were present in small numbers (yellow cattle) in the pre-European period but became the only animal (water buffalo) in the late Chinese period (CL1).
-One possible explanation for the shift in the animal taxonomic structure from deer to bovines is the dramatic decline of the Indigenous Kavalan population in Yilan, which occurred around the same time as a large influx of Han Chinese immigrants in the 19th century.
-The occupation of land by Han settlers gradually led to Kavalan relocation to eastern Taiwan, leaving only a small Indigenous population in Yilan [@Zhan1995].
-Moreover, bovines are believed to have been introduced by Han populations because they are absent from prehistoric contexts in Taiwan [@Chen2000animal].
-Therefore, the predominance of bovines in CL1 may indicate both a decrease in the Indigenous population and a shift from hunting to livestock use introduced or intensified under Han influence.
-Similar cultural changes can also be seen in burial practices at Kiwulan during the Chinese period, when Indigenous people began using wooden containers for burials that resembled Chinese coffins [@Chen2007; @Chiu2004].
+The deerskin trade continued after the Dutch left, but on a much smaller scale.
+In the 18th century, about 10,000 hides were exported annually, far less than the 100,000 exported during the European period [@Nakamura1997; @Tsao2011].
+The shrinking trade market and declining population at Kiwulan may have led to reduced demand for exporting deer products, which subsequently decreased reliance on suids, as reflected in the lower frequencies of suid remains in the post-European period and their eventual absence in the Chinese period.
+Unlike suids, muntjacs remained a small part of local subsistence until the early Chinese period, possibly due to their value for hide.
+Many Indigenous groups in Taiwan, including the Tsou and Bunun, used muntjac skins to make clothing and shoes [@Li1999], which may have also been practiced at Kiwulan.
+Although it is possible that both suids and muntjacs were consumed as supplements for deer, their patterns of use suggest that suids were likely the main source of additional meat, whereas muntjacs might be valued more for their hides, especially as deer populations declined.
 
@@
 # Conclusion
-The zooarchaeological evidence at Kiwulan reveals Indigenous participation in regional trade networks and their adaptation strategies by situating faunal remains within the context of foreign presence, including the Europeans in the 17th century and Chinese in the 19th century.
-The frequencies, butchery patterns, and modifications on deer bones from Kiwulan indicates an active deer processing that was intensified during the European period.
+The zooarchaeological evidence at Kiwulan reveals Indigenous participation in large-scale trade networks and their adaptation strategies by situating faunal remains within the context of foreign presence, including the Europeans in the 17th century and Chinese in the 19th century.
+The frequencies, butchery patterns, and modifications on deer bones from Kiwulan indicate an active deer processing that was intensified during the European period.
 Deer consumption at Kiwulan throughout most of time suggests cultural continuity, while the strategic reliance on suids during the European period reflects adaptation to high trade demands for deer products.
 On the other hand, the incorporation of muntjac into the diet during the Chinese period likely resulted from the overhunting of sika deer.
-These findings highlight the adaptation of Kiwulan residents to foreign impacts and environmental changes while maintaining cultural traditions.
-Through the long-term dynamics of animal use at Kiwulan, this research demonstrates that even Indigenous communities in peripheral areas with weak colonial control were profoundly impacted by colonial activities than previously thought.
-Future research expanding to other Indigenous settlements in northeastern Taiwan, with additional dietary evidence such as botanical remains and organic residue analysis, could provide deeper insights into regional dietary patterns and human-environment interactions.
+Through the long-term dynamics of animal use at Kiwulan, this case study highlights the active responses of its residents to foreign impacts and environmental changes while maintaining cultural traditions.
+It also demonstrates that Indigenous communities in peripheral areas under limited colonial control were profoundly affected by colonial activities than previously thought.
+In the future, broader studies incorporating other Indigenous sites in northeastern Taiwan or additional evidence such as botanical remains and residue analysis could offer deeper insights into regional diets and human-environment interactions.
 
